### 2023-10-09 Analyzing perimetric data with the visualFields package/Analyzing perimetric data with the visualFields package.Rmd
Started with template.

---
documentclass: jss
author:
  - name: Cord Huchzermeyer
    orcid: 0000-0000-0000-0000
    affiliation: 'University Hospital Erlangen \AND Friedrich-Alexander-University Erlangen Nürnberg'
    # use this syntax to add text on several lines
    address: |
      | First line
      | Second line
    email: \email{huchzi@email.de}
    url: https://huchzi.github.io
  - name: Second Author
    orcid: 0000-0000-0000-0000
    affiliation: 'Affiliation \AND'
    # To add another line, use \AND at the end of the previous one as above
  - name: Third Author
    orcid: 0000-0000-0000-0000
    address: |
      | Department of Statistics and Mathematics,
      | Faculty of Biosciences,
      | Universitat Autònoma de Barcelona
    affiliation: |
      | Universitat Autònoma 
      | de Barcelona
    # use a different affiliation in adress field (differently formated here)
    affiliation2: Universitat Autònoma de Barcelona
title:
  formatted: "Analyzing Perimetric data with the \\pkg{visualFields} package"
  # If you use tex in the formatted title, also supply version without
  plain:     "Analyzing Perimetric data with the visualFields package"
  # For running headers, if needed
  short:     "\\pkg{visualFields}: Statistics for Perimetry"
abstract: >
  The abstract of the article.
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{R}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
---

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
```

# Introduction

# Psychophysical Thresholds

## Psychophysical Function

## Sensitivities, Defects, Probabilities

## Algorithms for Measuring Psychophysical Thresholds

# Mapping the Hill of Vision

## What is the Hill of Vision

## Global parameters

## Bebie-Curve

## Different patterns

## Plotting Visual Fields

# Technical details

## Humphrey

## Octopus

## Others

# Structure of perimetric data

# Importing perimetric data

# Setting up the Environment

# Statistical Analysis

# Plotting

# Conclusion

## Code formatting

In general, don't use Markdown, but use the more precise LaTeX commands instead:

* \proglang{Java}
* \pkg{plyr}

One exception is inline code, which can be written inside a pair of backticks (i.e., using the Markdown syntax).

If you want to use LaTeX commands in headers, you need to provide a `short-title` attribute. You can also provide a custom identifier if necessary. See the header of Section \ref{r-code} for example.

# \proglang{R} code {short-title="R code" #r-code}

Can be inserted in regular R markdown blocks.

```{r}
x <- 1:10
x
```

## Features specific to \pkg{rticles} {short-title="Features specific to rticles"}

* Adding short titles to section headers is a feature specific to \pkg{rticles} (implemented via a Pandoc Lua filter). This feature is currently not supported by Pandoc and we will update this template if [it is officially supported in the future](https://github.com/jgm/pandoc/issues/4409).
* Using the `\AND` syntax in the `author` field to add authors on a new line. This is a specific to the `rticles::jss_article` format.

### 2024-03-11 vignette.Rmd
Created RMD file as a draft.

---
title: "VF1 Vignette"
author: "Cord Huchzermeyer"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

The visualFields package facilitates statistical analysis and plotting of visual field data, focussing on static perimetry.

# What is visual field testing?
Visual field testing is a psychophysical test that measures a patient’s visual ability at different locations of the visual field. This ability is usually highest in the center of the visual field (i.e. the point that the observer looks at) and decreases toward the periphery (hill of vision).

Theoretically, different aspects of vision can be tested at different locations (ability to discriminate dots that are very close to each other, ability to discriminate similar colors, contrast sensitivity), but usually white-on-white perimetry is performed where a white spot is presented on a white background.

The most common type of visual field testing is *perimetry*, which is performed with in a dome-shaped cupule, so that the distance between the stimulus and the eye equal for all locations within the visual field. In contrast, *campimetry* is performed with a flat screen.

Psychophysical tests measure a subjective perception of a physical stimulus by systematically varying the stimulus and asking the observer to communicate his perception with a limited number of allowed responses. In visual field testing, the observer usually presses a button when a light spot is perceived. Usually, psychophysical testing results in a *threshold*, which denotes either the absolute stimulus strength that can just barely be perceived or the smallest difference that can be perceived as different. Real world scenarios where perception is close to threshold include going through a dark forest on a cloudy night or seeing the car in front through a dense fog. Therefore, visual field testing is often considered as tedious by patients and judges as a less reliable examination. The latter is not correct.

The visual fields package was created to analyze data from *static perimetry*, where a threshold is measured for each of a fixed number of locations in a grid pattern. In contrast, in *kinetic perimetry*, a light spot moves from the periphery toward the center along a number of meridians and the location there the spot is first perceivable for the subject is noted.

Different manufacturers have different conventions in perimetry.

# Starting with an example

The visualFields package can be installed from CRAN.

```{r}

library(visualFields)
# install.packages("visualFields", repos = "https://cloud.r-project.org")

```

## Loading data

The visualField package contains some example data.

```{r}

data(vfpwgSunyiu24d2)

str(vfpwgSunyiu24d2[, 1:6])

```
This dataset contains the results of 42 visual field tests of one patient (id: sample1). The patient has a chronic-neurodegenerative disease of the optic nerve called primary open-angle glaucoma (type: pwg).

```{r}

head(vfpwgSunyiu24d2, 4)

```

Each row corresponds to one visual field test. Visual field testing was performed monocularly (one eye at a time), so that two examinations exist for each date.

The abbreviations for the laterality is derived from latin (OD: right eye = oculus dexter, and OS: left eye = oculus sinister), when both eyes are examined, the abbreviation is OU.

For now, not that each column $l_{n}$ corresponds to one location in the visual field test.

# Visual field tests
For the typical visual field test, the subject or patient is comfortably seated in front of a perimeter and puts his head on a chin rest looking into a white Ganzfeld cupule (or hemisphere). The perimeter is located in a dimly lit, calm room and the subject has been explained what to do by an experienced technician before the examination. In his hand, the subject has a button in order to signal whether he saw a stimulus.

He now looks at a fixation light right in front of him in the center of the cupule and presses the button as soon as he sees a light appearing anywhere in the cupule. These lights are presented only for a short time span (typically 100-200ms), so that the subject in not tempted to look at the stimulus. Therefore, a press of the button will be linked to a presentation when it occurs within a given time window after the presentation. A computer will present stimuli several times at different locations in a grid pattern. For each location, the stimulus intensity (luminance) will be adjusted depending on the subjects responses, but the presentations will be randomly interleaved between different locations.

The grid, the stimulus size, presentation time and the algorithm responsible for varying intensities remain constant throughout the examination and are called a perimetry program. Not that examinations performed with different programs cannot be directly compared.

# Setting up the environment

A given visual field program has certain properties that remain unchanged during the test. In the visualFields packages these are saved in the *environment*. After setting this environment, you can only analyse data that is consistent with this environment.

The environment contains information on

  - Normative data (nv)
  - Grid (locmap)
  - Graphical parameters used for plotting the data (gpar)
  - Where to find visual field measurements in the tables (locini)

```{r see current environment variables}

getnv()$info

```
The standard environment assumes a static white-on-white perimetry with a Humphrey perimeter, using a 24-2 grid, a SITA standard algorithm, a stimulus size of Goldman Size III. The stimulus duration and the $L_{max}$ are not explicitly stated in the info.

## Grid

The spatial position of the points $l_1 $to $ln$ in the visual field is found in the locmap and depends on the pattern used by the perimeter. When analysing data, you need to specify the locmap.

By default, the 24-2 pattern which uses locations that are equally spaced across the field is used.

```{r locmaps}

default_locmap <- getlocmap()
default_locmap$name
default_locmap$desc

```

Beside a name and a description, a locmap dataset contains a coordinate table with the x- and y- coordinates of the locations measured in degrees, and the id of the points that cover the physiological blind spot (the id is the number x in $l_x$ and corresponds to the row in the coord table).

```{r}

plot(x = default_locmap$coord$x, 
     y = default_locmap$coord$y, 
     type = "point", 
     asp = 1,
     xlim = c(-24, 24),
     ylim = c(-24, 24))
points(x = default_locmap$coord[default_locmap$bs, "x"],
       y = default_locmap$coord[default_locmap$bs, "y"],
       col = "red")

```

Locations of the locmap that cover the blind spot are reported in the locmaps lists under bs. If the blind spot is outside the pattern (for example in the 10° visual fields) or is spared by the grid as in the Octopus G1 pattern, the list item bs is set to numeric(0).

A new locmap can be specified manually or selected from the locmaps dataset included in the package.

```{r}

new_locmap <- locmaps$p10d2
setlocmap(new_locmap)

```

Note that some grids spare the blind spot, either on purpose or because they cover only locations central to the blind spot. In this case, locmap$bs is numeric(0).

```{r}

blind_spot <- getlocmap()$bs

plot(x = new_locmap$coord$x, 
     y = new_locmap$coord$y, 
     type = "point", 
     asp = 1,
     xlim = c(-24, 24),
     ylim = c(-24, 24))
points(x = default_locmap$coord[default_locmap$bs, "x"],
       y = default_locmap$coord[default_locmap$bs, "y"],
       col = "red")

```

## Resetting the enviroment

```{r}

setdefaults()

```

## Using environment variables

Note that the visualFields package will always override the local environment from which the call was made.

```{r}

# new_env <- new.env(parent = get)

```

# Pathophysiology of the visual field
Interindividual variability and age
Photoreceptors / Retina
Ganglion cells
Other factors

# Measurements
Although usually not seen in visual field data, the most elementary unit of a static visual field test is a subject's response. It consists of the constant stimulus parameters (size, presentation time, background luminance), the stimulus location, the stimulus intensity and whether the subject pressed the button within a predefined time window to indicate that he saw the stimulus.

Below a certain *threshold* luminance, a stimulus is usually not perceivable, above the threshold, the stimulus is usually perceived. However, close to threshold, a stimulus is not always reliably perceived. By definition, the threshold is the luminance of the stimulus where the subject will see the stimulus with a probability of 50%.

The most straightforward to measure such a threshold would be to present stimuli several time at a number different, fixed stimulus strengths. This method is called the method of *constant stimuli* and results in the *psychometric function*, which is usually modeled with a cumulative distribution function (CDF), for example a Gaussian CDF. However, this is not the most effective way to estimate a threshold, because stimulus strength far from the threshold will have probabilities to be seen that are close to either 0% or 100%. Testing these strengths several times does not yield much information.

Therefore, advanced algorithms were developed for reducing the number of stimulus presentations with limited trade-off concerning the accuracy of the measurements. A staircase algorithm where the intensity is decreased by a certain step size when the stimulus was seen, increased when not seen, and the step size is halved upon each revearsal (seen to not seen or vice versa) is a more efficient algorithms that is frequently used. More efficient algorithms usually make a trade-off between precision of the threshold estimate and test time, and often incorporate information on responses in neighboring locations into the algorithm. Examples include the Swedish Interactive Threshold Algorithm (SITA) of the Humphery Perimeters. Note that long test times negatively affect attention and therefore both accuracy and precision of the threshold estimates.

# Differential light sensitivity and other mathematical derivatives of threshold
However, the data analyzed with the visual fields package usually consists of the subjects *differential light sensitivity* for a stimulus at a given location in decibel (dB) as the most elementary unit. This sensitivity is calculated from the psychophysical threshold, which is derived from the aforementioned responses at this location.

The light sensitivity [dB] is defined as $10*log_{10}{\frac{L_{max}}{L}}$. Where $L$ is the threshold luminance and $L_{max}$ is the maximal stimulus intensity the perimeter uses.

The light sensitivity fluctuates considerably between measurements in the same observer, between observers, but it also decreases with age.

## False negatives / positives
Residual deviation from the psychometric function may occur due to inattention of the observer but also due to short-term fluctuation of the threshold. For example, it has been shown that the threshold underlies short-term fluctuates in the areas where light sensitivity was negatively affected by glaucoma, a disease of the optic nerve.

The cooperation of the subject is also tested in some perimeters by presenting deliberately presenting a stimulus at the location of the *physiological blind spot*. If the subjects sees that stimulus, it must be assumed that he did not fixate correctly (fixation loss) and that the reliability of the field is decreased.

## Defect / Deviation

Clinically it is often more interesting to map the deviation from a theoretical "normal" sensitivity than the absolute sensitivity, because this facilitates the identification of disease. Furthermore, a decline of DLS with age occurs also in normal subjects. Therefore, age-related normative date must be collected which results in a function the describes the relationship between DLS and age for each location in the visual field. This function is usually linear.

The resulting difference is also measured in dB and is called either *deviaton* (loss of function has a negative sign; used in Humphrey perimeters) or *defect* (loss of function has a positive sign; used in Octopus perimeters). In this package, *deviation* is used, and it is called *total deviation*. Make sure that you get the right sign if you import data.

In eye diseases like glaucoma, general defects that affect the whole visual field are often less relevant than focal scotomas. Therefore, deviation is sometimes corrected for the average sensitivity of the visual field. This is called *pattern deviation*.

### Sensitivity

```{r sensitivity}

vfpwgSunyiu24d2[1:4, "1:12"]

```

### Total deviation

```{r total deviation}

gettd(vfpwgSunyiu24d2)[1:4, 1:12]

```

### Pattern deviation

```{r pattern deviation}

getpd(vfpwgSunyiu24d2)[1:4, 1:12]

```

## Probabilities

Psychophysical values vary considerably between subjects, but they also fluctuate between measurements. Therefore, deviation will be different from zero most of the time, and the clinician has to know, if a given deviation is likely to be caused by disease, or whether it may just reflect normal variability.

This is calculated based on measurements in the reference population that is also used to calculate the normative data.

### Total deviation

```{r total deviation probabilities}

gettdp(vfpwgSunyiu24d2)[1:4, 1:12]

```

### Pattern deviation

```{r pattern deviation probabilities}

getpdp(vfpwgSunyiu24d2)[1:4, 1:12]

```

# Plotting the visual field

There are different ways to visualize perimetric results. These visualizations have to be interpreted like maps in that they represent a simplified version of the hill of vision.

The most straightforward plots show the sensitivity/total deviation/pattern deviation/probabilitites at the corresponding locations. These representations are very exact, but may be difficult to interpret.

```{r}

setdefaults()

example1 <- vffilter(vfpwgSunyiu24d2, 
                     id == "sample1" & 
                       eye == "OD" & 
                       date == min(vfpwgSunyiu24d2$date))
vfisvalid(example1)

example1_dataframe <- data.frame(x = getlocmap()$coord$x, y = getlocmap()$coord$y, td = as.vector(unlist(example1[1, 11:64])))
ggplot(example1_dataframe, aes(x = x, y = y)) +
  geom_text(aes(label = td)) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  theme_void()

```

The standard already contains some additional information with a border presenting a color-coded visualisation of the deviation and a graphical representation of the blind spot.

```{r}

vfplot(example1, type = "td")

vfplot(example1, type = "s")

```

Note that you have to specify what type of data is contained in the visual field tables. The default is set to *total deviation*.

# Global parameters

Global parameters summarize the visual field in a few parameters. These global parameters can be calculated with the visualField package.

The mean sensitivity (MS) is the arithmetic mean of all sensitivity values, the mean deviation (MD) is the arithmetic mean of all total deviation values in the field. The pattern standard deviation (PSD) is the standard deviation of the total deviation values.

```{r}

getgl(example1)[, 11:18]

```
Again, the probability of the observed values given that the visual field is normal can be calculated.

```{r}

getglp(getgl(example1))

```

# Structure function correlations in glaucoma patients

In glaucoma, the damage occurs at the levels of the nerve fiber bundles in the optic nerve. Therefore, visual field defect frequently follow this anatomy and correspond to structural changes in the retinal nerve fiber layer that can be imaged along a ring around the optic nerve. Locations in the visual field correspond to a location where the nerves that transmit signals from the photoreceptors that detect a signal via the retinal ganglion cells cross this ring.

```{r}

example1_dataframe

loc2psi(locmaps$p24d2$coord[1, ])

```

# Longitudinal analyses

Monitoring function over time in glaucoma patients is the most common clinical application of visual fields in ophthalmology. For the next example, we will select a series of 27 visual fields that were obtained over several years.

```{r}

example2 <- vffilter(vfpwgSunyiu24d2, 
                     id == "sample1" & 
                       eye == "OS")
```

```{r}

vflegoplot(example2)
vfsparklines(example2)

```

Statistical analysis can be performed with linear regression of global parameters, pointwise linear regression (one regression line for each location), which results in a number of p-values, or with Permutation analysis of Pointwise Linear Regression (PoPLR).

```{r}

glr(getgl(example2))

```
```{r}

res <- plr(example2)

vfplotplr(example2)

```

```{r}

poplr(example2)

```

Progression reports can be created as pdfs or as interactive reports with shiny.

```{r}

vfspa(example2, file = "test.pdf")
# vfspashiny(example2, file = "test.pdf")

```

# DICOM standard of visual fields

DICOM (Digital Imaging and Communications in Medicine) standards for ophthalmic static perimetry are specified in Supplement 146. Currently, the data representation in the VisualFields package differs from these standards. However, there are two functions for loading data generated in the DICOMM format created by the Humphrey Field Analyzer (HFA) by Zeiss. Theoretically, this function should also load data from other perimeters.

# Using a different perimeter

The VisualField package provided normative data for the Humphrey Field Analyzer but currently not for other perimeters. The loadoctopus function also imports some normative data from the Octopus perimeter by Haag-Streit, but this does not allow full use of the functionality of the visualFields package, currently.

### 2024-03-12 vignette.Rmd
Commit before restructuring.

@@
 # What is visual field testing?
-Visual field testing is a psychophysical test that measures a patient’s visual ability at different locations of the visual field.
+
+Visual field testing is a psychophysical test that measures a patient's visual ability at different locations of the visual field.
 This ability is usually highest in the center of the visual field (i.e.
@@
 ```
+
 This dataset contains the results of 42 visual field tests of one patient (id: sample1).
@@
 
+## Plotting the visual field
+
+The most straightforward plots show the sensitivity/total deviation/pattern deviation/probabilitites at the corresponding locations.
+These representations are very exact, but may be difficult to interpret.
+
+```{r}
+
+setdefaults()
+
+example1 <- vffilter(vfpwgSunyiu24d2, 
+                     id == "sample1" & 
+                       eye == "OD" & 
+                       date == min(vfpwgSunyiu24d2$date))
+vfisvalid(example1)
+
+example1_dataframe <- data.frame(x = getlocmap()$coord$x, y = getlocmap()$coord$y, td = as.vector(unlist(example1[1, 11:64])))
+ggplot(example1_dataframe, aes(x = x, y = y)) +
+  geom_text(aes(label = td)) +
+  geom_hline(yintercept = 0) +
+  geom_vline(xintercept = 0) +
+  theme_void()
+
+```
+
 # Visual field tests
+
 For the typical visual field test, the subject or patient is comfortably seated in front of a perimeter and puts his head on a chin rest looking into a white Ganzfeld cupule (or hemisphere).
@@
 ```
+
 The standard environment assumes a static white-on-white perimetry with a Humphrey perimeter, using a 24-2 grid, a SITA standard algorithm, a stimulus size of Goldman Size III.
@@
 
-The spatial position of the points $l_1 $to $ln$ in the visual field is found in the locmap and depends on the pattern used by the perimeter. When analysing data, you need to specify the locmap.
+The spatial position of the points \$l_1 \$to $ln$ in the visual field is found in the locmap and depends on the pattern used by the perimeter. When analysing data, you need to specify the locmap.
 
@@
 
-Note that some grids spare the blind spot, either on purpose or because they cover only locations central to the blind spot. In this case, locmap$bs is numeric(0).
+Note that some grids spare the blind spot, either on purpose or because they cover only locations central to the blind spot. In this case, locmap\$bs is numeric(0).
 
@@
 
-# Pathophysiology of the visual field
-Interindividual variability and age
-Photoreceptors / Retina
-Ganglion cells
-Other factors
+# Measurements
 
-# Measurements
 Although usually not seen in visual field data, the most elementary unit of a static visual field test is a subject's response.
@@
 # Differential light sensitivity and other mathematical derivatives of threshold
+
 However, the data analyzed with the visual fields package usually consists of the subjects *differential light sensitivity* for a stimulus at a given location in decibel (dB) as the most elementary unit.
@@
 ## False negatives / positives
+
 Residual deviation from the psychometric function may occur due to inattention of the observer but also due to short-term fluctuation of the threshold.
@@
 
-The resulting difference is also measured in dB and is called either *deviaton* (loss of function has a negative sign; used in Humphrey perimeters) or *defect* (loss of function has a positive sign; used in Octopus perimeters).
+The resulting difference is also measured in dB and is called either *deviation* (loss of function has a negative sign; used in Humphrey perimeters) or *defect* (loss of function has a positive sign; used in Octopus perimeters).
 In this package, *deviation* is used, and it is called *total deviation*.
@@
 
-### Sensitivity
+### Dataset
+
+We use the dataset included in the package for demonstrating calculation of global parameters.
 
@@
 
-vfpwgSunyiu24d2[1:4, "1:12"]
+vfpwgSunyiu24d2[1:4, 1:12]
 
@@
 
-The most straightforward plots show the sensitivity/total deviation/pattern deviation/probabilitites at the corresponding locations.
-These representations are very exact, but may be difficult to interpret.
+The standard already contains some additional information with a border presenting a color-coded visualization of the deviation and a graphical representation of the blind spot.
+Note that the vfplot with the option "td" needs information on differential light-sensitivities in a normal reference population and on the probability of given deviations from these normal values.
+These informations are included for the typical Humphrey Field Analyzer fields, but you need to collect this information if you use other parameters.
 
 ```{r}
-
-setdefaults()
-
-example1 <- vffilter(vfpwgSunyiu24d2, 
-                     id == "sample1" & 
-                       eye == "OD" & 
-                       date == min(vfpwgSunyiu24d2$date))
-vfisvalid(example1)
-
-example1_dataframe <- data.frame(x = getlocmap()$coord$x, y = getlocmap()$coord$y, td = as.vector(unlist(example1[1, 11:64])))
-ggplot(example1_dataframe, aes(x = x, y = y)) +
-  geom_text(aes(label = td)) +
-  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = 0) +
-  theme_void()
-
-```
-
-The standard already contains some additional information with a border presenting a color-coded visualisation of the deviation and a graphical representation of the blind spot.
-
-```{r}
-
-vfplot(example1, type = "td")
 
@@
 Note that you have to specify what type of data is contained in the visual field tables. The default is set to *total deviation*.
+
+```{r}
+
+vfplot(example1, type = "td")
+
+```
+
+```{r}
+
+vfplot(example1, type = "tds")
+
+```
+
+```{r}
+
+vfplot(example1, type = "pd")
+
+```
+
+```{r}
+
+vfplot(example1, type = "pds")
+
+```
 
@@
 ```
+
 Again, the probability of the observed values given that the visual field is normal can be calculated.
@@
 Monitoring function over time in glaucoma patients is the most common clinical application of visual fields in ophthalmology.
+
+For several reasons, measurements will fluctuate over time.
+Fluctuation between measurements will be higher in eyes with glaucoma, especially in areas of glaucoma-related visual field defects.
+Therefore, it is not possible to compare two field tests and claim that the disease has progressed only because the second test yields poorer sensitivities.
+
+Thus, statistical methods and a minimum number of tests are necessary in order to demonstrate disease progression in the visual fields.
+Statistical methods that can be subdivided into *event-based* and *trend-based* methods.
+Furthermore, methods can be subdivided into those that require datasets which describe the "normal fluctuation" in a given population, those that rely on a mathematical model of visual field progression, and those that do not rely on either.
+
+Event-based methods are not implemented in the visualFields pacakge.
+Briefly, they look for progression between the most recent field and a baseline test.
+Progression is characterized by a decrease in sensitivity with a magnitude that is unlikely (< 5%) in the population under examination (population based data is necessary) and needs to be confirmed in further tests (usually at least two).
+
+Trend-based methods assume that progression takes place more slowly, so that several field tests are carried out during this time span.
+This concept is unsettling for the clinician who wants to stop progression the first time that he sees the patient, but regularly takes place even in the case of optimal treatment.
+The advantage is that all measurements are taken into account, not only the baseline test and the most recent tests.
+
+## Dataset
+
 For the next example, we will select a series of 27 visual fields that were obtained over several years.
@@
 
+## Plotting progression
+
+The sparkline plot only shows the raw data.
+
 ```{r}
 
-vflegoplot(example2)
 vfsparklines(example2)
@@
 
+The lego plot shows probabilities for progression which relies on population based data.
+
+```{r}
+
+vflegoplot(example2)
+
+```
+
+## Statistical analysis
+
 Statistical analysis can be performed with linear regression of global parameters, pointwise linear regression (one regression line for each location), which results in a number of p-values, or with Permutation analysis of Pointwise Linear Regression (PoPLR).
+
+The *glr* function performs linear regression on global parameters.
 
@@
 ```
+
+The *plr* function performs pointwise linear regression.
+
 ```{r}
 
-res <- plr(example2)
+prog1 <- plr(example2)
+str(prog1, max.level = 1)
+
+print(sum(prog1$pval < 0.05))
 
@@
 
+The *poplr* function performs pointwise linear regression, and then calculates an S-score on the visual field series and on random permutations of the series, in order to control the Type I error (diagnosing progression in a non-progressing series) without relying on population data or a mathematical model of progression.
+It is computationally intensive.
+
+The *poplr* function yields a large list with many intermediate computations.
+The main result is the S-score and the probability of getting this S-score in the absence of progression (Type I error).
+Note that there is a left and a right probability (deterioration and improvement) in the current implementation.
+
 ```{r}
 
-poplr(example2)
+prog2 <- poplr(example2)
+str(prog2, max.level = 1)
 
@@
 
-Progression reports can be created as pdfs or as interactive reports with shiny.
+An improvement is highly unlikely in this series.
 
@@
 
+prog2$csr
+prog2$csrp
+
+```
+
+In contrast, a deterioration is very likely (p < 0.0001).
+
+```{r}
+
+prog2$csl
+prog2$cslp
+
+```
+
+## Creating pretty reports
+
+Progression reports can be created as pdfs or as interactive reports with shiny.
+
+```{r eval=FALSE, include=TRUE}
+
 vfspa(example2, file = "test.pdf")
-# vfspashiny(example2, file = "test.pdf")
+
+vfspashiny(example2, file = "test.pdf")
 
@@
 DICOM (Digital Imaging and Communications in Medicine) standards for ophthalmic static perimetry are specified in Supplement 146.
-Currently, the data representation in the VisualFields package differs from these standards.
 However, there are two functions for loading data generated in the DICOMM format created by the Humphrey Field Analyzer (HFA) by Zeiss.
-Theoretically, this function should also load data from other perimeters.
+
+Currently, the data representation in the VisualFields package differs from these standards and data cannot be exported from the package to DICOMM.
 
@@
 
+Humphrey Field Analyzer (Carl-Zeiss-Meditec)
+
 The VisualField package provided normative data for the Humphrey Field Analyzer but currently not for other perimeters.
+
+Octopus 900 Perimeter (Haag-Streit)
+
 The loadoctopus function also imports some normative data from the Octopus perimeter by Haag-Streit, but this does not allow full use of the functionality of the visualFields package, currently.
+
+For making full use of the visusalField package with other parameters, you ideally need a database of visual fields from a normal reference population, but the creation of normal values with the visualFields package goes beyond the scope of this vignette.

### 2024-03-12 vignette.Rmd
Re-structured text.

@@
 library(ggplot2)
+library(visualFields)
+
+example1 <- vffilter(vfpwgSunyiu24d2, 
+                     id == "sample1" & 
+                       eye == "OD" & 
+                       date == min(vfpwgSunyiu24d2$date))
+
+# example1_dataframe <- data.frame(x = getlocmap()$coord$x, y = getlocmap()$coord$y, td = as.vector(unlist(example1[1, 11:64])))
+
 ```
@@
 
-Visual field testing is a psychophysical test that measures a patient's visual ability at different locations of the visual field.
+Visual field testing measures a patient's visual ability at different locations of the visual field.
 This ability is usually highest in the center of the visual field (i.e.
-the point that the observer looks at) and decreases toward the periphery (hill of vision).
-
-Theoretically, different aspects of vision can be tested at different locations (ability to discriminate dots that are very close to each other, ability to discriminate similar colors, contrast sensitivity), but usually white-on-white perimetry is performed where a white spot is presented on a white background.
-
+the point that the observer looks at) and decreases toward the periphery (*hill of vision*).
 The most common type of visual field testing is *perimetry*, which is performed with in a dome-shaped cupule, so that the distance between the stimulus and the eye equal for all locations within the visual field.
@@
 
-Psychophysical tests measure a subjective perception of a physical stimulus by systematically varying the stimulus and asking the observer to communicate his perception with a limited number of allowed responses.
-In visual field testing, the observer usually presses a button when a light spot is perceived.
-Usually, psychophysical testing results in a *threshold*, which denotes either the absolute stimulus strength that can just barely be perceived or the smallest difference that can be perceived as different.
-Real world scenarios where perception is close to threshold include going through a dark forest on a cloudy night or seeing the car in front through a dense fog.
-Therefore, visual field testing is often considered as tedious by patients and judges as a less reliable examination.
-The latter is not correct.
-
+Usually, perimetry tests the ability to perceive differences in luminance with an achromatic (not colored) stimulus (white-on-white perimetry).
+However, theoretically, other aspects of vision can be tested at different locations (ability to discriminate dots that are very close to each other, ability to discriminate similar colors, contrast sensitivity).
 The visual fields package was created to analyze data from *static perimetry*, where a threshold is measured for each of a fixed number of locations in a grid pattern.
@@
 
-Different manufacturers have different conventions in perimetry.
+# Performing visual field tests
 
-# Starting with an example
+For the typical visual field test, the subject or patient is comfortably seated in front of a perimeter and puts his head on a chin rest looking into a white Ganzfeld cupule (or hemisphere).
+The perimeter is located in a dimly lit, calm room and the subject has been explained what to do by an experienced technician before the examination.
+In his hand, the subject has a button in order to signal whether he saw a stimulus.
 
-The visualFields package can be installed from CRAN.
+He now looks at a fixation light right in front of him in the center of the cupule and presses the button as soon as he sees a light appearing anywhere in the cupule.
+These lights are presented only for a short time span (typically 100-200ms), so that the subject in not tempted to look at the stimulus.
+Therefore, a press of the button will be linked to a presentation when it occurs within a given time window after the presentation.
+A computer will present stimuli several times at different locations in a grid pattern.
+For each location, the stimulus intensity (luminance) will be adjusted depending on the subjects responses, but the presentations will be randomly interleaved between different locations.
 
-```{r}
+## Grid pattern
 
-library(visualFields)
-# install.packages("visualFields", repos = "https://cloud.r-project.org")
+The spatial position of the points $l_1$ to $l_n$ in the visual field is found in the locmap and depends on the pattern used by the perimeter. When analysing data, you need to specify the locmap.
+
+By default, the 24-2 pattern which uses locations that are equally spaced across the field is used.
+
+Beside a name and a description, a locmap dataset contains a coordinate table with the x- and y- coordinates of the locations measured in degrees, and the id of the points that cover the physiological blind spot (the id is the number x in $l_x$ and corresponds to the row in the coord table).
+
+```{r echo=FALSE}
+
+default_locmap <- getlocmap()
+plot(x = default_locmap$coord$x, 
+     y = default_locmap$coord$y, 
+     type = "point", 
+     asp = 1,
+     xlim = c(-24, 24),
+     ylim = c(-24, 24))
+points(x = default_locmap$coord[default_locmap$bs, "x"],
+       y = default_locmap$coord[default_locmap$bs, "y"],
+       col = "red")
 
 ```
+
+Locations of the locmap that cover the blind spot are reported in the locmaps lists under bs.
+If the blind spot is outside the pattern (for example in the 10° visual fields) or is spared by the grid as in the Octopus G1 pattern, the list item bs is set to numeric(0).
+
+The grid, the stimulus size, presentation time and the algorithm responsible for varying intensities remain constant throughout the examination and are called a perimetry program.
+Not that examinations performed with different programs cannot be directly compared.
+
+# Clinical applications of visual field tests
+
+In clinical practice, visual field testing is used identify decreases in visual function caused by disease.
+Such *visual field defects* or *scotomas* can be caused by diseases that affect tissue anywhere along the optic pathway.
+This ranges from the optic media (cornea / lens / vitreous) which transmit light, over the retina which detects lights and preprocesses visual information, the optic nerve which transmits information, up to the brain.
+
+Often, clinicians can draw reliable conclusions about the localisation of the disease by the pattern of the visual field defects.
+For example, an infarction of the right visual cortex leads complete blindness in the left hemispheres of both eyes with a very distinct border along the vertical meridian.
+Therefore, the patient does not perceive anything to the right of the point they look at.
+On the other side of vertical meridian the visual field is normal.
+In contrast, patients with glaucoma start with areas of slightly perceived light sensitivity often in the upper or lower hemifield with a bow-shaped pattern cause by the localisation of damage in the optic nerve head.
+
+Visual field tests can be used to diagnose disease, to appraise the functional consequences of disease (for example: can the patient still drive a vehicle?), and to monitor progression of disease.
+
+The most advanced statistical methods have been created for monitoring glaucoma progression.
+The open-angle glaucomas are a group of slowly progressing, chronic neurodegenerative diseases of the optic nerve, often caused by increased eye pressure.
+Patients need consequent monitoring over many years and slight changes in the visual field must be detected reliably.
+Therefore, many of the functions of the visualField package focus on glaucoma research.
+
+# Getting started with the package
 
@@
 The visualField package contains some example data.
+This dataset contains the results of 42 visual field tests of one patient (id: sample1).
+The patient has a chronic-neurodegenerative disease of the optic nerve called primary open-angle glaucoma (type: pwg).
 
 ```{r}
+
+# install.packages("visualFields")
+library(visualFields)
 
@@
 ```
-
-This dataset contains the results of 42 visual field tests of one patient (id: sample1).
-The patient has a chronic-neurodegenerative disease of the optic nerve called primary open-angle glaucoma (type: pwg).
 
@@
 
-# Visual field tests
+# Psychophysical testing
 
-For the typical visual field test, the subject or patient is comfortably seated in front of a perimeter and puts his head on a chin rest looking into a white Ganzfeld cupule (or hemisphere).
-The perimeter is located in a dimly lit, calm room and the subject has been explained what to do by an experienced technician before the examination.
-In his hand, the subject has a button in order to signal whether he saw a stimulus.
-
-He now looks at a fixation light right in front of him in the center of the cupule and presses the button as soon as he sees a light appearing anywhere in the cupule.
-These lights are presented only for a short time span (typically 100-200ms), so that the subject in not tempted to look at the stimulus.
-Therefore, a press of the button will be linked to a presentation when it occurs within a given time window after the presentation.
-A computer will present stimuli several times at different locations in a grid pattern.
-For each location, the stimulus intensity (luminance) will be adjusted depending on the subjects responses, but the presentations will be randomly interleaved between different locations.
-
-The grid, the stimulus size, presentation time and the algorithm responsible for varying intensities remain constant throughout the examination and are called a perimetry program.
-Not that examinations performed with different programs cannot be directly compared.
-
-# Setting up the environment
-
-A given visual field program has certain properties that remain unchanged during the test.
-In the visualFields packages these are saved in the *environment*.
-After setting this environment, you can only analyse data that is consistent with this environment.
-
-The environment contains information on
-
--   Normative data (nv)
--   Grid (locmap)
--   Graphical parameters used for plotting the data (gpar)
--   Where to find visual field measurements in the tables (locini)
-
-```{r see current environment variables}
-
-getnv()$info
-
-```
-
-The standard environment assumes a static white-on-white perimetry with a Humphrey perimeter, using a 24-2 grid, a SITA standard algorithm, a stimulus size of Goldman Size III.
-The stimulus duration and the $L_{max}$ are not explicitly stated in the info.
-
-## Grid
-
-The spatial position of the points \$l_1 \$to $ln$ in the visual field is found in the locmap and depends on the pattern used by the perimeter. When analysing data, you need to specify the locmap.
-
-By default, the 24-2 pattern which uses locations that are equally spaced across the field is used.
-
-```{r locmaps}
-
-default_locmap <- getlocmap()
-default_locmap$name
-default_locmap$desc
-
-```
-
-Beside a name and a description, a locmap dataset contains a coordinate table with the x- and y- coordinates of the locations measured in degrees, and the id of the points that cover the physiological blind spot (the id is the number x in $l_x$ and corresponds to the row in the coord table).
-
-```{r}
-
-plot(x = default_locmap$coord$x, 
-     y = default_locmap$coord$y, 
-     type = "point", 
-     asp = 1,
-     xlim = c(-24, 24),
-     ylim = c(-24, 24))
-points(x = default_locmap$coord[default_locmap$bs, "x"],
-       y = default_locmap$coord[default_locmap$bs, "y"],
-       col = "red")
-
-```
-
-Locations of the locmap that cover the blind spot are reported in the locmaps lists under bs.
-If the blind spot is outside the pattern (for example in the 10° visual fields) or is spared by the grid as in the Octopus G1 pattern, the list item bs is set to numeric(0).
-
-A new locmap can be specified manually or selected from the locmaps dataset included in the package.
-
-```{r}
-
-new_locmap <- locmaps$p10d2
-setlocmap(new_locmap)
-
-```
-
-Note that some grids spare the blind spot, either on purpose or because they cover only locations central to the blind spot. In this case, locmap\$bs is numeric(0).
-
-```{r}
-
-blind_spot <- getlocmap()$bs
-
-plot(x = new_locmap$coord$x, 
-     y = new_locmap$coord$y, 
-     type = "point", 
-     asp = 1,
-     xlim = c(-24, 24),
-     ylim = c(-24, 24))
-points(x = default_locmap$coord[default_locmap$bs, "x"],
-       y = default_locmap$coord[default_locmap$bs, "y"],
-       col = "red")
-
-```
-
-## Resetting the enviroment
-
-```{r}
-
-setdefaults()
-
-```
-
-## Using environment variables
-
-Note that the visualFields package will always override the local environment from which the call was made.
-
-```{r}
-
-# new_env <- new.env(parent = get)
-
-```
+Psychophysical tests systematically measure the relationship between physical stimulus strength and the subjective perception.
+This is performed by varying the stimulus and asking the observer to communicate his perception with a limited number of allowed responses.
+In visual field testing, the observer usually presses a button when a light spot is perceived.
+The relationship between stimulus and perception is analyzed using mathematical model.
+Commonly, these models result in an estimate of a *threshold*, which denotes either the absolute stimulus strength that can just barely be perceived or the smallest difference that can be perceived as different.
+Real world scenarios where perception is close to threshold include going through a dark forest on a cloudy night or seeing the car in front through a dense fog.
+Stimuli are not clearly seen when close to threshold.
+For this reason, patients sometimes feel that lengthy visual field testing is tedious and that it is as an unreliable examination.
+The latter is not clearly correct when fluctuations are taken into account and data are analyzed with appropriate statistical methods.
 
@@
 
-# Differential light sensitivity and other mathematical derivatives of threshold
+## False negatives / positives
+
+Residual deviation from the psychometric function may occur due to inattention of the observer but also due to short-term fluctuation of the threshold.
+For example, it has been shown that the threshold underlies short-term fluctuates in the areas where light sensitivity was negatively affected by glaucoma, a disease of the optic nerve.
+
+The cooperation of the subject is also tested in some perimeters by presenting deliberately presenting a stimulus at the location of the *physiological blind spot*.
+If the subjects sees that stimulus, it must be assumed that he did not fixate correctly (fixation loss) and that the reliability of the field is decreased.
+
+# Mathematical derivatives of threshold
+
+## Sensitivity and other mathematical derivatives of threshold
 
@@
 The light sensitivity fluctuates considerably between measurements in the same observer, between observers, but it also decreases with age.
-
-## False negatives / positives
-
-Residual deviation from the psychometric function may occur due to inattention of the observer but also due to short-term fluctuation of the threshold.
-For example, it has been shown that the threshold underlies short-term fluctuates in the areas where light sensitivity was negatively affected by glaucoma, a disease of the optic nerve.
-
-The cooperation of the subject is also tested in some perimeters by presenting deliberately presenting a stimulus at the location of the *physiological blind spot*.
-If the subjects sees that stimulus, it must be assumed that he did not fixate correctly (fixation loss) and that the reliability of the field is decreased.
 
@@
 
+Different manufacturers have different conventions in perimetry.
+
 Humphrey Field Analyzer (Carl-Zeiss-Meditec)
@@
 For making full use of the visusalField package with other parameters, you ideally need a database of visual fields from a normal reference population, but the creation of normal values with the visualFields package goes beyond the scope of this vignette.
+
+```{r}
+
+```
+
+# Advanced topics
+
+# Setting up the environment
+
+A given visual field program has certain properties that remain unchanged during the test.
+In the visualFields packages these are saved in the *environment*.
+After setting this environment, you can only analyse data that is consistent with this environment.
+
+The environment contains information on
+
+-   Normative data (nv)
+-   Grid (locmap)
+-   Graphical parameters used for plotting the data (gpar)
+-   Where to find visual field measurements in the tables (locini)
+
+```{r see current environment variables}
+
+getnv()$info
+
+```
+
+The standard environment assumes a static white-on-white perimetry with a Humphrey perimeter, using a 24-2 grid, a SITA standard algorithm, a stimulus size of Goldman Size III.
+The stimulus duration and the $L_{max}$ are not explicitly stated in the info.
+
+## Locini and Number of locations / vfisvalid
+
+## Locmaps
+
+```{r locmaps}
+
+default_locmap <- getlocmap()
+default_locmap$name
+default_locmap$desc
+
+```
+
+A new locmap can be specified manually or selected from the locmaps dataset included in the package.
+
+```{r}
+
+new_locmap <- locmaps$p10d2
+setlocmap(new_locmap)
+
+```
+
+Note that some grids spare the blind spot, either on purpose or because they cover only locations central to the blind spot. In this case, locmap\$bs is numeric(0).
+
+```{r}
+
+blind_spot <- getlocmap()$bs
+
+plot(x = new_locmap$coord$x, 
+     y = new_locmap$coord$y, 
+     type = "point", 
+     asp = 1,
+     xlim = c(-24, 24),
+     ylim = c(-24, 24))
+points(x = default_locmap$coord[default_locmap$bs, "x"],
+       y = default_locmap$coord[default_locmap$bs, "y"],
+       col = "red")
+
+```
+
+## Resetting the enviroment
+
+```{r}
+
+setdefaults()
+
+```
+
+## Using environment variables
+
+Note that the visualFields package will always override the local environment from which the call was made.
+
+```{r}
+
+# new_env <- new.env(parent = get)
+
+```

### 2024-03-13 vignette.Rmd
First draft ready. Asking for feedback.

@@
 However, theoretically, other aspects of vision can be tested at different locations (ability to discriminate dots that are very close to each other, ability to discriminate similar colors, contrast sensitivity).
+
+```{r echo=FALSE}
+
+default_locmap <- getlocmap()
+
+plot(x = default_locmap$coord$x, 
+     y = default_locmap$coord$y, 
+     type = "point", 
+     asp = 1,
+     xlim = c(-32, 26),
+     ylim = c(-29, 26),
+     yaxt = 'n',
+     xaxt = 'n',
+     bty = 'n',
+     xlab = "",
+     ylab = "")
+text(x = 0, y = -28, "horizontal eccentricity [degree]")
+text(x = -34, y = -0, "vertical eccentricity [degree]", srt = 90)
+points(x = default_locmap$coord[default_locmap$bs, "x"],
+       y = default_locmap$coord[default_locmap$bs, "y"],
+       col = "red")
+lines(x = c(-2.5, 2.5), y = c(0, 0))
+lines(x = c(0, 0), y = c(-2.5, 2.5))
+
+```
+
 The visual fields package was created to analyze data from *static perimetry*, where a threshold is measured for each of a fixed number of locations in a grid pattern.
@@
 The perimeter is located in a dimly lit, calm room and the subject has been explained what to do by an experienced technician before the examination.
-In his hand, the subject has a button in order to signal whether he saw a stimulus.
+In his hand, the subject has a button.
 
 He now looks at a fixation light right in front of him in the center of the cupule and presses the button as soon as he sees a light appearing anywhere in the cupule.
-These lights are presented only for a short time span (typically 100-200ms), so that the subject in not tempted to look at the stimulus.
-Therefore, a press of the button will be linked to a presentation when it occurs within a given time window after the presentation.
+These lights are presented only for a short time span (typically 100-200ms), so that the subject has no time to shift his gaze to the stimulus.
+A press of the button will be linked to a presentation when it occurs within a given time window after the presentation.
 A computer will present stimuli several times at different locations in a grid pattern.
@@
 
-The spatial position of the points $l_1$ to $l_n$ in the visual field is found in the locmap and depends on the pattern used by the perimeter. When analysing data, you need to specify the locmap.
+The spatial position of the points $l_1$ to $l_n$ in the visual field is found in the location map (locmap) and depends on the pattern used by the perimeter.
+When analyzing data, you need to specify the locmap.
 
-By default, the 24-2 pattern which uses locations that are equally spaced across the field is used.
+By default, the 24-2 pattern (shown above) is used, which consists of 54 locations that are equally spaced across the field.
+
+```{r}
+
+# install.packages("visualFields")
+library(visualFields)
+
+default_locmap <- getlocmap()
+
+str(default_locmap)
+
+```
 
@@
 
-```{r echo=FALSE}
+Locations of the locmap that cover the blind spot are reported in the locmaps lists under the variable bs.
+If the blind spot is outside the pattern (for example in the 10° visual fields) or is spared by the grid as in the Octopus G1 pattern, the list item bs is set to numeric(0).
 
-default_locmap <- getlocmap()
-plot(x = default_locmap$coord$x, 
-     y = default_locmap$coord$y, 
-     type = "point", 
-     asp = 1,
-     xlim = c(-24, 24),
-     ylim = c(-24, 24))
-points(x = default_locmap$coord[default_locmap$bs, "x"],
-       y = default_locmap$coord[default_locmap$bs, "y"],
-       col = "red")
+```{r}
+
+ten_two_locmap <- locmaps$p10d2
+
+str(ten_two_locmap)
 
 ```
-
-Locations of the locmap that cover the blind spot are reported in the locmaps lists under bs.
-If the blind spot is outside the pattern (for example in the 10° visual fields) or is spared by the grid as in the Octopus G1 pattern, the list item bs is set to numeric(0).
-
 The grid, the stimulus size, presentation time and the algorithm responsible for varying intensities remain constant throughout the examination and are called a perimetry program.
@@
 
-The visualField package contains some example data.
-This dataset contains the results of 42 visual field tests of one patient (id: sample1).
-The patient has a chronic-neurodegenerative disease of the optic nerve called primary open-angle glaucoma (type: pwg).
+The visualField package contains some example data. This dataset contains the results of 42 visual field tests of one patient (id: sample1). The patient has primary open-angle glaucoma (type: pwg).
 
@@
 
-# install.packages("visualFields")
-library(visualFields)
-
 data(vfpwgSunyiu24d2)
 
-str(vfpwgSunyiu24d2[, 1:6])
+str(vfpwgSunyiu24d2[, 1:10])
 
@@
 
+The first columns characterize the patient, the time and duration of the examination, the diagnosis, and the reliability parameters 'false positive responses', 'false negative responses' and 'fixation losses'.
+
+Each row corresponds to one visual field test.
+Visual field testing is performed monocularly (one eye at a time), so that two examinations exist for each date.
+The abbreviations for the laterality ('eye')is derived from latin (OD: right eye = oculus dexter, and OS: left eye = oculus sinister), when both eyes are examined, the abbreviation is OU.
+
+The environment variable locini contains the number of the first column that contains the actual visual field date.
+Each column $l_{n}$ corresponds to one location in the visual field test.
+A concise way to select the columns $l_1$ to $l_{n}$ is to use the *getvfcols* function.
+
 ```{r}
 
-head(vfpwgSunyiu24d2, 4)
+getlocini()
+getvfcols()
+
+vfpwgSunyiu24d2[1, getlocini():ncol(vfpwgSunyiu24d2)]
 
@@
 
-Each row corresponds to one visual field test. Visual field testing was performed monocularly (one eye at a time), so that two examinations exist for each date.
+There mustn't be any columns after these columns.
+Use the *vffilter* function to select specific fields and teh *vfisvalid* function to check whether a table fulfills the criteria.
+The latter will also tell you what the problem is when it doesn't.
 
-The abbreviations for the laterality is derived from latin (OD: right eye = oculus dexter, and OS: left eye = oculus sinister), when both eyes are examined, the abbreviation is OU.
+```{r}
 
-For now, not that each column $l_{n}$ corresponds to one location in the visual field test.
+example1 <- vffilter(vfpwgSunyiu24d2, 
+                     id == "sample1" & 
+                       eye == "OD")[1, ]
+
+example1$newVariable <- "not valid"
+
+vfisvalid(example1)
+
+```
 
@@
 
-The most straightforward plots show the sensitivity/total deviation/pattern deviation/probabilitites at the corresponding locations.
+The most straightforward plots show the sensitivities at the corresponding locations.
 These representations are very exact, but may be difficult to interpret.
+For learning purposes, here, we manually plot some visual field data.
 
@@
 
-example1 <- vffilter(vfpwgSunyiu24d2, 
+example2 <- vffilter(vfpwgSunyiu24d2, 
                      id == "sample1" & 
-                       eye == "OD" & 
-                       date == min(vfpwgSunyiu24d2$date))
-vfisvalid(example1)
+                       eye == "OD")[1, ]
 
-example1_dataframe <- data.frame(x = getlocmap()$coord$x, y = getlocmap()$coord$y, td = as.vector(unlist(example1[1, 11:64])))
-ggplot(example1_dataframe, aes(x = x, y = y)) +
+vfisvalid(example2)
+
+example2_dataframe <- data.frame(x = getlocmap()$coord$x, 
+                                 y = getlocmap()$coord$y, 
+                                 td = as.vector(unlist(example2[, getvfcols()])))
+
+ggplot(example2_dataframe, aes(x = x, y = y)) +
   geom_text(aes(label = td)) +
@@
 ```
+Note that left eyes have to be treated differently if you plot the points manually, because the location labels are mirrored.
+
+By convention, when both eyes are presented side by side, the left eye is shown on the left and the right eye on the right.
+This corresponds to the visual field of the observer, so that the physiological blindspots are directed temporally (toward the temple, away from the nose) for both eyes.
+This is in contrast to presentation of structural data, which is usually shown like it would look in a patient sitting in front of you, where his right eye is on the observers left side.
+
+```{r}
+
+OD_dataframe <- data.frame(x = getlocmap()$coord$x, 
+                             y = getlocmap()$coord$y, 
+                             loc_label = paste0("l", 1:54),
+                             is_blindspot = FALSE,
+                             eye = "OD")
+
+# Note the minus in the x variable for the left eye
+OS_dataframe <- data.frame(x = -getlocmap()$coord$x, 
+                             y = getlocmap()$coord$y, 
+                             loc_label = paste0("l", 1:54),
+                             is_blindspot = FALSE,
+                             eye = "OS")
+
+OD_dataframe[getlocmap()$bs, "is_blindspot"] <- TRUE
+OS_dataframe[getlocmap()$bs, "is_blindspot"] <- TRUE
+
+ODOS_dataframe <- rbind(OS_dataframe, OD_dataframe)
+
+# Use factor in order to show left eye on the left side
+ODOS_dataframe$eye <- factor(ODOS_dataframe$eye, 
+                             levels = c("OS", "OD"), 
+                             ordered = TRUE)
+
+ggplot(ODOS_dataframe, aes(x = x, y = y)) +
+  geom_label(aes(label = loc_label, color = is_blindspot)) +
+  scale_color_manual(values = c("black", "red")) +
+  facet_wrap(~ eye)
+
+```
+
+The visualField package provides functions for more sophisticated plotting, which will be explained later.
+*These functions automatically take care that right eyes and left eyes are plotted correctly.*
 
@@
 In visual field testing, the observer usually presses a button when a light spot is perceived.
+
 The relationship between stimulus and perception is analyzed using mathematical model.
 Commonly, these models result in an estimate of a *threshold*, which denotes either the absolute stimulus strength that can just barely be perceived or the smallest difference that can be perceived as different.
+
 Real world scenarios where perception is close to threshold include going through a dark forest on a cloudy night or seeing the car in front through a dense fog.
@@
 
-The cooperation of the subject is also tested in some perimeters by presenting deliberately presenting a stimulus at the location of the *physiological blind spot*.
+When the observer does not see a stimulus that is clearly above a level that has been seen before, this is called a false negative response (fnr).
+In contrast, when the subject presses the button although no stimulus is presented, this is a false positive response (fpr).
+
+The cooperation of the subject is also tested in some perimeters by presenting deliberately presenting a stimulus at the location of the blind spot.
 If the subjects sees that stimulus, it must be assumed that he did not fixate correctly (fixation loss) and that the reliability of the field is decreased.
@@
 
-### Dataset
+### Sensitivity
 
-We use the dataset included in the package for demonstrating calculation of global parameters.
+In the visualField datasets, $l_1$ to $l_n$ represents sensitivities.
 
-```{r sensitivity}
+```{r}
 
-vfpwgSunyiu24d2[1:4, 1:12]
+example2[, getlocini()+1:4]
 
@@
 
+Total deviation is calculated from the sensitivities and from age-related normal values. These are saved in the environment variable *nv* (normal values).
+
 ```{r total deviation}
 
-gettd(vfpwgSunyiu24d2)[1:4, 1:12]
+age <- example2[1, "age"]
+
+print(age)
+
+normal_values <- getnv()$agem$model(age)[1:4]
+
+print(normal_values)
+
+total_defects <- example2[, getlocini()+1:4] - normal_values
+
+print(total_defects)
+
+```
+
+These values can be automatically calculated with a dedicated function.
+
+```{r}
+
+gettd(example2)[, getlocini()+1:4]
 
@@
 
+The pattern deviation corrects for the mean deviation, i.e.
+the average sensitivity loss.
+It highlights local defects but does not show global sensitivity loss, for example due to opacification of the optic media.
+
 ```{r pattern deviation}
 
-getpd(vfpwgSunyiu24d2)[1:4, 1:12]
+getpd(example2)[, getlocini()+1:4]
 
@@
 
-gettdp(vfpwgSunyiu24d2)[1:4, 1:12]
+gettdp(example2)[, getlocini()+1:4]
 
@@
 Note that the vfplot with the option "td" needs information on differential light-sensitivities in a normal reference population and on the probability of given deviations from these normal values.
-These informations are included for the typical Humphrey Field Analyzer fields, but you need to collect this information if you use other parameters.
+These data are included for the typical Humphrey Field Analyzer fields, but you need to collect this information if you use other perimeters
 
@@
 
-vfplot(example1, type = "s")
+vfplot(example2, type = "s")
 
@@
 
-vfplot(example1, type = "td")
+vfplot(example2, type = "td")
 
@@
 
-vfplot(example1, type = "tds")
+vfplot(example2, type = "tds")
 
@@
 
-vfplot(example1, type = "pd")
+vfplot(example2, type = "pd")
 
@@
 
-vfplot(example1, type = "pds")
+vfplot(example2, type = "pds")
 
@@
 
-Global parameters summarize the visual field in a few parameters. These global parameters can be calculated with the visualField package.
+Global parameters summarize the visual field in a few parameters.
+These global parameters also can be calculated with the visualField package.
+The resulting table contains the visual field header data (patient id, eye, etc.), but contains 8 columns (msens, ssens, tmd, tsd, pmd, psd, gh, and vfi) instead.
 
-The mean sensitivity (MS) is the arithmetic mean of all sensitivity values, the mean deviation (MD) is the arithmetic mean of all total deviation values in the field.
+The mean sensitivity (MS, msens) is the arithmetic mean of all sensitivity values, the mean deviation (MD, tmd) is the arithmetic mean of all total deviation values in the field.
 The pattern standard deviation (PSD) is the standard deviation of the total deviation values.
@@
 
-getgl(example1)[, 11:18]
+getgl(example2)[, -(1:getlocini()-1)]
 
 ```
-
 Again, the probability of the observed values given that the visual field is normal can be calculated.
@@
 
-getglp(getgl(example1))
+getglp(getgl(example2))[, -(1:getlocini()-1)]
 
@@
 
-In glaucoma, the damage occurs at the levels of the nerve fiber bundles in the optic nerve.
-Therefore, visual field defect frequently follow this anatomy and correspond to structural changes in the retinal nerve fiber layer that can be imaged along a ring around the optic nerve.
-Locations in the visual field correspond to a location where the nerves that transmit signals from the photoreceptors that detect a signal via the retinal ganglion cells cross this ring.
+In the last years, techniques detailed imaging of the retina and the optic nerve have made incredible advances.
+Therefore, correlation between structure and function is becoming even more important than ever.
+
+When photoreceptors are lost in an specific retinal area, the scotoma always occurs in the directions in space that is perceived by that retinal area. This makes structure-function correlations easy.
+
+In glaucoma, the damage occurs in the optic nerve at the levels of the nerve fiber bundles, which consist of the axons of the retinal ganglion cells.
+In the retina, these cells form two layers: the ganglion cell layer that contains the cell bodies, and the retinal nerve fiber layer.
+However, the cell bodies of the retinal ganglion cells are displaced from the area where the photoreceptors are located that they receive input from.
+This displacement is relevant mostly in the center of the visual field.
+
+Therefore, the visualFields package provides function to calculate where the ganglion cells that subserve a given location in the visual field are located and to calculate at which angle the axons of these ganglion cells enter the optic nerve head (optic disc), where they pass through the sclera (wall of the eye) to form the optic nerve.
 
@@
 
-example1_dataframe
+locmaps$p24d2$coord[23, ]
 
-loc2psi(locmaps$p24d2$coord[1, ])
+loc2psi(locmaps$p24d2$coord[23, ])
+
+vf2gc(locmaps$p24d2$coord[23, ])
 
 ```
+
+From the angle of incidence, the nerve fiber bundle path can be calculated.
+Note that the function for calculating the trajectory will extrapolate the trajectory beyond the location of the visual field (usually to the raphe).
+
+```{r}
+
+# use loc2psi to calculate the incidence of the nerve fiber corresponding to a given location at the ONH
+angle_of_incidence_l19 <-loc2psi(locmaps$p24d2$coord[19, ])
+
+# use cart2jpolar for calculating the position of the visual field location in the polar coordinate system
+locations <- rbind(cart2jpolar(locmaps$p24d2$coord[19, ]),
+                   cart2jpolar(locmaps$p24d2$coord[23, ]),
+                   cart2jpolar(locmaps$p24d2$coord[47, ]),
+                   cart2jpolar(data.frame(x = 0, y = 0)))
+locations$labels <- c("l19", "l23", "l47", "fixation")
+
+# use bundePath to create a function for calculating the bundle path
+path_function_l19 <- bundlePath(angle_of_incidence_l19)
+path_function_l23 <- bundlePath(loc2psi(locmaps$p24d2$coord[23, ]))
+path_function_l47 <- bundlePath(loc2psi(locmaps$p24d2$coord[47, ]))
+
+bundle_path <- data.frame(y = 0:90,
+                          x_l19 = path_function_l19(0:90),
+                          x_l23 = path_function_l23(0:90),
+                          x_l47 = path_function_l47(0:90))
+
+ggplot(bundle_path, aes(y = y)) +
+  geom_line(aes(x = x_l19)) +
+  geom_point(data = locations, aes(x = -psi, y = r, color = labels, shape = labels == "fixation")) +
+  geom_line(aes(x = x_l23), color = "red") +
+  geom_line(aes(x = x_l47), color = "blue") +
+  coord_polar(start = pi / 2, direction = -1) +
+  scale_x_continuous(breaks = c(-180, -90, 0, 90, 180), limits = c(-180, 180)) +
+  scale_y_continuous(limits = c(0, 60)) +
+  scale_color_manual(values = c("l19" = "black", "l23" = "red", "l47" = "blue", "fixation" = "black")) +
+  guides(color = FALSE, shape = FALSE)
+
+```
+
+In this plot, the origin corresponds to the center of the ONH, the black triangle corresponds to the fovea, and the 180degree angle corrsponds to the line between the ONH center and the fovea, which is, in reality, slightly tilted downward in most patients.
 
@@
 
-## Dataset
-
 For the next example, we will select a series of 27 visual fields that were obtained over several years.
@@
 
-example2 <- vffilter(vfpwgSunyiu24d2, 
+example3 <- vffilter(vfpwgSunyiu24d2, 
                      id == "sample1" & 
@@
 
-vfsparklines(example2)
+vfsparklines(example3)
 
@@
 
-vflegoplot(example2)
+vflegoplot(example3)
 
@@
 
-The *glr* function performs linear regression on global parameters.
+The *glr* function performs linear regression on global parameters. In the following example, the patient loses 0.24dB each year and this change is statistically significant.
 
@@
 
-glr(getgl(example2))
+md_regression <- data.frame(time = glr(getgl(example3), type = "md")$years,
+           md = glr(getgl(example3), type = "md")$data)
+
+slope <- glr(getgl(example3), type = "md")$sl
+intercept <- glr(getgl(example3), type = "md")$int
+
+ggplot(md_regression, aes(x = time, y =md)) +
+  geom_point() +
+  geom_abline(slope = slope, intercept = intercept)
+
+glr(getgl(example3), type = "md")$pval
 
@@
 
-The *plr* function performs pointwise linear regression.
+The *plr* function performs pointwise linear regression (i.e.
+linear regression for each location in the field separately).
+Specific parameters have to be chosen, in order to detect disease progression.
+For example, progression might be defined as at least 3 locations having a slope less than 0dB/year and a p-value less than 0.05.
 
@@
 
-prog1 <- plr(example2)
+prog1 <- plr(example3)
 str(prog1, max.level = 1)
@@
 
-vfplotplr(example2)
+vfplotplr(example3)
 
@@
 
-prog2 <- poplr(example2)
+prog2 <- poplr(example3)
 str(prog2, max.level = 1)
@@
 
-In contrast, a deterioration is very likely (p < 0.0001).
+In contrast, a deterioration is very likely (p < 0.0001). In the histogram, you can see that a S-score of 176 was only very rarely found in the random permutations of the visual field series.
 
@@
 prog2$cslp
+
+hist(prog2$cstats$cslall)
+abline(v = prog2$csl)
 
@@
 
-vfspa(example2, file = "test.pdf")
+# vfspa(example3, file = "test.pdf")
 
-vfspashiny(example2, file = "test.pdf")
+# vfspashiny(example3, file = "test.pdf")
 
@@
 Different manufacturers have different conventions in perimetry.
-
-Humphrey Field Analyzer (Carl-Zeiss-Meditec)
+For example, clinicians working with Humphrey Field Analyzers (Carl-Zeiss-Meditec) are often used to report loss of function as negativ values (MD: mean *deviation* of -10dB), while those working with Octopus perimeters (Haag-Streit) often report loss as a positive value (MD: mean *defect* of 10dB).
+Furthermore, clinicians working with Octopus perimeters often use different grids, even though the 24-2 and 10-2 grids are theoretically available.
+Usually, fields cannot be directly compared because other setting ($L_max$) are also different and because different algorithms are used.
 
@@
 
-Octopus 900 Perimeter (Haag-Streit)
-
-The loadoctopus function also imports some normative data from the Octopus perimeter by Haag-Streit, but this does not allow full use of the functionality of the visualFields package, currently.
+The loadoctopus function also imports some normative data from the Octopus perimeter by Haag-Streit, but, as of now, this does not allow full use of the functionality of the visualFields package
 
@@
 
-```{r}
-
-```
-
 # Advanced topics
 
-# Setting up the environment
+## Setting up the environment
 
@@
 
+The nv environment variable also contains functions for calculating defects, probabilities and global parameters that can be accessed via the functions stated above.
+
 ```{r see current environment variables}
 
-getnv()$info
+getnv()$info        # current normal values
+str(getnv(), 1)
+
+str(getlocmap(), 1)   # current locmap
+str(getgpar(), 1)     # current parameters for plotting
+getlocini()   # position of the column l1 in the datafram
 
@@
 The stimulus duration and the $L_{max}$ are not explicitly stated in the info.
-
-## Locini and Number of locations / vfisvalid
 
@@
 
+Note that for switching to another perimetry program, you need to set all environment variables.
+
+```{r}
+
+setnv(normvals$iowa_PC26_pw)
+
+setlocmap(locmaps$pPC26)
+
+setgpar(gpars$pPC26)
+
+```
+
 ## Resetting the enviroment
+
+If you reset the environment, you return to the 24-2 pattern measured with the SITA algorithm on a Humphrey Field Analyser.
 
@@
 ```
-
-## Using environment variables
-
-Note that the visualFields package will always override the local environment from which the call was made.
-
-```{r}
-
-# new_env <- new.env(parent = get)
-
-```

### 2024-08-17 vignette.Rmd
IMF revision. Did some general editing, added references, a classic Hill of Vision figure, and an example with Octopus G1 grid and made up data

@@
 ---
-title: "VF1 Vignette"
-author: "Cord Huchzermeyer"
+title: visualFields 1.x
+author:
+  - Cord Huchzermeyer, INSTITUTION
+  - Iván Marín-Franch, Computational Optometry
 date: "`r Sys.Date()`"
 output: html_document
+bibliography: references.bib
+biblio-style: "apalike"
+link-citations: true
 ---
@@
 knitr::opts_chunk$set(echo = TRUE)
-library(ggplot2)
-library(visualFields)
-
-example1 <- vffilter(vfpwgSunyiu24d2, 
-                     id == "sample1" & 
-                       eye == "OD" & 
-                       date == min(vfpwgSunyiu24d2$date))
-
-# example1_dataframe <- data.frame(x = getlocmap()$coord$x, y = getlocmap()$coord$y, td = as.vector(unlist(example1[1, 11:64])))
-
 ```
 
-The visualFields package facilitates statistical analysis and plotting of visual field data, focussing on static perimetry.
+# The Open Perimetry Initiative (OPI)
+The OPI [@IMF:Tur2012] [@IMF:Marin2022a] is a standard for interfacing with visual field testing machines (perimeters).
+It specifies basic functions that allow many visual field tests to be constructed.
+As of October 2017 it is fully implemented on the Octopus 900 and partially on the Heidelberg Edge Perimeter, the Kowa AP 7000, the CrewT IMO and the Centervue Compass.
+
+The visualFields package, part of the OPI, is a collection of tools for analyzing the field of vision.
+It provides a framework for the development and use of innovative methods for visualization, statistical analysis, and clinical interpretation of visual-field loss and its change over time.
+
+To install the released version of the package, run 
+```{r eval=FALSE}
+install.packages("visualFields")
+```
+
+To install the newestversion of the package, run:
+```{r eval=FALSE}
+install.packages("devtools")
+library(devtools)
+install_github("imarinfr/vf1/source")
+```
+
+To load the package run
+```{r}
+library(visualFields)
+```
+
+```{r echo=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
+library(ggplot2)
+library(plotrix)
+```
+
+# What is the visual field?
+The visual field of an eye is the extent of the world that can be seen.
+It refers to both the central and the peripheral vision.
+It is measured in degrees of visual angle, with the center of vision being $0^\circ$, points on the temporal and superior regions of the visual field being positive angles and those in the inferior and nasal being negative.
+On healthy eyes, the sensitivity to light contrast changes as the magnitude of the visual angle increases (or as it is commonly said in perimetry, eccentricity).
+The difference in light sensitivity over the visual field as a function of eccentricity is known as the Traquair's *hill of vision* [@IMF:Traquair1938].
+The figure below shows the hill of vision as rendered by Aulhorn & Harms [@IMF:Aul1967].
+Notice in the discontinuity in the hill of vision in the lower right panel.
+Sensitivity in this region, centered at approximately $(15^\circ, -3^\circ)$ in the visual field is is completely blind, as it corresponds to the anatomical region of the retina known as the *blind spot*, where the optical fiber exits from the eye towards the visual cortex.
+
+<figure>
+  <center><img src="./AulhornHillOfVision.png" alt="Hill of Vision" width="75%"></center>
+  <center><figcaption>Aulhorn and Harms' rendering of Traquair's hill of vision.</figcaption></center>
+</figure>
 
 # What is visual field testing?
+Visual field testing measures a patient's visual sensitivity at different locations of the visual field.
+Its purpose is to reconstruct as best as possible within the confines and limitations of a clinical setting the *hill of vision*.
+The most common type of visual field testing is *static automated perimetry (SAP)*, which is performed with in a dome-shaped device, so that the distance between the stimulus and the eye equal for all locations within the visual field.
+[I WOULD DELETE THE FOLLOWING AS IT MAY CONFUSE RATHER THAN CLARIFY GIVEN THAT NOBODY USES CAMPIMETRY ANYMORE, AT LEAST IN THE UK OR US] In contrast, *campimetry* is performed with a flat screen.
 
-Visual field testing measures a patient's visual ability at different locations of the visual field.
-This ability is usually highest in the center of the visual field (i.e.
-the point that the observer looks at) and decreases toward the periphery (*hill of vision*).
-The most common type of visual field testing is *perimetry*, which is performed with in a dome-shaped cupule, so that the distance between the stimulus and the eye equal for all locations within the visual field.
-In contrast, *campimetry* is performed with a flat screen.
-
-Usually, perimetry tests the ability to perceive differences in luminance with an achromatic (not colored) stimulus (white-on-white perimetry).
+Usually, perimetry tests the ability to detect increments in stimulus luminance with respect to the background with an achromatic (not colored) stimulus.
+This is a form of perimetry known as *white-on-white perimetry*.
 However, theoretically, other aspects of vision can be tested at different locations (ability to discriminate dots that are very close to each other, ability to discriminate similar colors, contrast sensitivity).
+The figure below shows the most common grid of locations tested during *SAP*.
+The gray ellipse corresponds to the blind spot of the hill of vision and, therefore, the red dots are expected to have no sensitivity.
 
@@
      y = default_locmap$coord$y, 
-     type = "point", 
+     type = "n", 
      asp = 1,
-     xlim = c(-32, 26),
-     ylim = c(-29, 26),
+     xlim = c(-30, 30),
+     ylim = c(-34, 34),
      yaxt = 'n',
@@
      xlab = "",
-     ylab = "")
-text(x = 0, y = -28, "horizontal eccentricity [degree]")
-text(x = -34, y = -0, "vertical eccentricity [degree]", srt = 90)
-points(x = default_locmap$coord[default_locmap$bs, "x"],
-       y = default_locmap$coord[default_locmap$bs, "y"],
-       col = "red")
-lines(x = c(-2.5, 2.5), y = c(0, 0))
-lines(x = c(0, 0), y = c(-2.5, 2.5))
+     ylab = "",
+     main = "24-2 grid of test locations")
+lines(-c(30, 2.5), c(0, 0), lwd = 2, col = "lightgray")
+lines(c(2.5, 30), c(0, 0), lwd = 2, col = "lightgray")
+lines(c(0, 0), -c(30, 2.5), lwd = 2, col = "lightgray")
+lines(c(0, 0), c(2.5, 30), lwd = 2, col = "lightgray")
+lines(x = c(-2.5, 2.5), y = c(0, 0), lwd = 1.5)
+lines(x = c(0, 0), y = c(-2.5, 2.5), lwd = 1.5)
+for(r in c(10, 20, 30))
+  draw.circle(0, 0, r, lwd = 2, col = NA, border = "lightgray")
+draw.ellipse(15, -1.5, 2.75, 3.75, col = "gray", border = NA)
+text(x = 0, y = -34, "horizontal eccentricity [degrees]")
+text(x = -34, y = -0, "vertical eccentricity [degrees]", srt = 90)
+text(x = 31, y = 0, "30\u00b0", adj = c(0, 0.5))
+text(x = 0, y = 31, "30\u00b0", adj = c(0.5, 0))
+points(x = default_locmap$coord$x[-default_locmap$bs],
+       y = default_locmap$coord$y[-default_locmap$bs])
+points(x = default_locmap$coord$x[default_locmap$bs],
+       y = default_locmap$coord$y[default_locmap$bs],
+       col = "red", pch = 19)
 
@@
 
-The visual fields package was created to analyze data from *static perimetry*, where a threshold is measured for each of a fixed number of locations in a grid pattern.
-In contrast, in *kinetic perimetry*, a light spot moves from the periphery toward the center along a number of meridians and the location there the spot is first perceivable for the subject is noted.
+The visual fields package was created to analyze data from *SAP*.
+In SAP, the goal is to quantify sensitivity by increasing and decreasing the light differential (or contrast) between stimulus and background and asking the observer to press the button when they detect the presence of the visual stimulus.
+The contrast threshold is determined at each location of the visual field as the one for which the observer is able to detect the presence of the stimulus 50% of the times.
+Such thresholds are estimated at a fixed number of locations forming a regular grid as the one in the figure above.
+There is another common form of perimetry, *kinetic perimetry*, where a light spot is moved from the periphery towards the center along a number of meridians.
+In this form of perimetry the location where the spot is first detected by the subject is noted and those points at different meridians are connected to find isopters of equal sensitivity on the retina.
+The statistical analysis for this form of perimetry is very different than for static perimetry and beyond the scope of this visualFields package.
 
 # Performing visual field tests
+For the typical visual field test, an observer (usually a patient in an ophthalmology or optimetry clinic) is comfortably seated in front of a perimeter and puts his head on a chin rest looking into a white Ganzfeld dome (or hemisphere).
+The perimeter is located in a dimly lit, calm room and the observer has been explained what to do by an experienced technician before the examination.
+In his hand, the observer has access to a clicker.
 
-For the typical visual field test, the subject or patient is comfortably seated in front of a perimeter and puts his head on a chin rest looking into a white Ganzfeld cupule (or hemisphere).
-The perimeter is located in a dimly lit, calm room and the subject has been explained what to do by an experienced technician before the examination.
-In his hand, the subject has a button.
+Observers now look at a fixation target (a dot, a cross, a diamond, or other) right in front of them in the center of the dome and are presented visual stimuli over their visual field.
+Observer are to press the clicker any time a stimulus appears anywhere in the dome These lights are presented only for a short time span (typically 100-200ms), so that the observer has no time to shift his gaze to the stimulus.
+A press of the clicker will be linked to a presentation when it occurs within a given time window after the presentation.
+A computer will present stimuli several times at different locations in a grid pattern and in a spatially random order.
+For each location, the stimulus contrast will be decreased if the observer responded to the previous response and increased if not.
 
-He now looks at a fixation light right in front of him in the center of the cupule and presses the button as soon as he sees a light appearing anywhere in the cupule.
-These lights are presented only for a short time span (typically 100-200ms), so that the subject has no time to shift his gaze to the stimulus.
-A press of the button will be linked to a presentation when it occurs within a given time window after the presentation.
-A computer will present stimuli several times at different locations in a grid pattern.
-For each location, the stimulus intensity (luminance) will be adjusted depending on the subjects responses, but the presentations will be randomly interleaved between different locations.
+At the end of the test, the inverse of the contrast threshold, $\Delta L_{\rm threshold}$ at each location is recorded in a decibel scale (dB).
+This dB scale is normally defined as attenuation from the maximum stimulus possible $\Delta L_{\rm max}$ as:
+\begin{equation}
+dB = -10\ \rm log \left (\frac{\Delta L_{\rm thr}}{\Delta L_{\rm max}} \right)\ .
+\end{equation}
+For the most common devices, The Humprhery Field Analyzer and the Octopus 900, the maximum luminance differential $\Delta L_{\rm max}$ is $10000\ (\rm apostilb) / \pi = 3183.1\ cd\ m^{-2}$.
 
-## Grid pattern
+These are the data expected by the visualFields package for statistical analysis and generation of visualization devices.
 
-The spatial position of the points $l_1$ to $l_n$ in the visual field is found in the location map (locmap) and depends on the pattern used by the perimeter.
+## Grid patterns
+The spatial position of the points $l_1$ to $l_k$ [CAN WE USE K FOR LOCATIONS, TO PROTECT N FOR NUMBER OF VISUAL FIELDS OR SAMPLE SIZE?] in the visual field is found in the location map (locmap) and depends on the pattern used by the perimeter.
 When analyzing data, you need to specify the locmap.
 
-By default, the 24-2 pattern (shown above) is used, which consists of 54 locations that are equally spaced across the field.
+By default, the 24-2 pattern (shown in the previous figure) is used, which consists of 54 locations that are equally spaced across the field.
 
 ```{r}
-
-# install.packages("visualFields")
-library(visualFields)
-
-default_locmap <- getlocmap()
-
 str(default_locmap)
-
 ```
 
-Beside a name and a description, a locmap dataset contains a coordinate table with the x- and y- coordinates of the locations measured in degrees, and the id of the points that cover the physiological blind spot (the id is the number x in $l_x$ and corresponds to the row in the coord table).
+Beside a name and a description, a locmap dataset contains a coordinate table with the x- and y-coordinates of the locations measured in degrees, and the id of the points that cover the physiological blind spot (the id is the number x in $l_x$ and corresponds to the row in the coord table).
+
+Other grid patterns are also defined in the package. For instance, the 10-2 grid:
+
+```{r}
+ten_two_locmap <- locmaps$p10d2
+str(ten_two_locmap)
+```
+
+And many others:
+
+```{r}
+names(locmaps)
+```
 
@@
 
-```{r}
-
-ten_two_locmap <- locmaps$p10d2
-
-str(ten_two_locmap)
-
-```
 The grid, the stimulus size, presentation time and the algorithm responsible for varying intensities remain constant throughout the examination and are called a perimetry program.
@@
 
-# Clinical applications of visual field tests
-
+# visualFields as an aid for clinical assessment of the visual field
 In clinical practice, visual field testing is used identify decreases in visual function caused by disease.
 Such *visual field defects* or *scotomas* can be caused by diseases that affect tissue anywhere along the optic pathway.
-This ranges from the optic media (cornea / lens / vitreous) which transmit light, over the retina which detects lights and preprocesses visual information, the optic nerve which transmits information, up to the brain.
+This ranges from the optic media (cornea / lens / vitreous) which transmit light, to the retina, which detects lights and preprocesses visual information, to the optic nerve, which transmits information, and up to the visual cortex.
 
@@
 Therefore, the patient does not perceive anything to the right of the point they look at.
-On the other side of vertical meridian the visual field is normal.
-In contrast, patients with glaucoma start with areas of slightly perceived light sensitivity often in the upper or lower hemifield with a bow-shaped pattern cause by the localisation of damage in the optic nerve head.
+On the other side of the vertical meridian the visual field is normal.
+In contrast, patients with glaucoma start with areas of slightly decreased light sensitivity often in the upper or lower hemifield with a bow-shaped (arcuate) pattern cause by the localisation of damage in the optic nerve head.
 
-Visual field tests can be used to diagnose disease, to appraise the functional consequences of disease (for example: can the patient still drive a vehicle?), and to monitor progression of disease.
+Visual field tests can be used to diagnose disease, to appraise the functional consequences of disease (for example: can the patient still drive a vehicle?), and to monitor progression.
 
+[THIS SEEMS TO BE A BIT OUF OF PLACE, BUT I PREFER NOT TO EDIT THIS PART MYSELF]
 The most advanced statistical methods have been created for monitoring glaucoma progression.
@@
 # Getting started with the package
-
 ## Loading data
-
 The visualField package contains some example data. This dataset contains the results of 42 visual field tests of one patient (id: sample1). The patient has primary open-angle glaucoma (type: pwg).
@@
 ```{r}
-
-data(vfpwgSunyiu24d2)
 
@@
 Visual field testing is performed monocularly (one eye at a time), so that two examinations exist for each date.
-The abbreviations for the laterality ('eye')is derived from latin (OD: right eye = oculus dexter, and OS: left eye = oculus sinister), when both eyes are examined, the abbreviation is OU.
+The abbreviations for the laterality ('eye') is derived from latin (OD: right eye = oculus dexter, and OS: left eye = oculus sinister), when both eyes are examined, the abbreviation is OU.
 
@@
 
-vfpwgSunyiu24d2[1, getlocini():ncol(vfpwgSunyiu24d2)]
+vfpwgSunyiu24d2[1, getvfcols()]
 
@@
 ## Plotting the visual field
-
 The most straightforward plots show the sensitivities at the corresponding locations.
@@
 setdefaults()
+###[I HAVE CHANGED THE EXAMPLE TO ONE FROM THE RETEST DATASET BECAUSE THE ONE FROM SUNY/IU IS TOO EXTREME]
 
-example2 <- vffilter(vfpwgSunyiu24d2, 
-                     id == "sample1" & 
-                       eye == "OD")[1, ]
+example2 <- vffilter(vfpwgRetest24d2, 
+                     id == "1" & 
+                       eye == "OD")[1,]
 
@@
                                  y = getlocmap()$coord$y, 
-                                 td = as.vector(unlist(example2[, getvfcols()])))
+                                 td = as.vector(unlist(example2[,getvfcols()])))
 
@@
   geom_vline(xintercept = 0) +
+  coord_fixed() +
   theme_void()
@@
 ```
-Note that left eyes have to be treated differently if you plot the points manually, because the location labels are mirrored.
+
+Note that left eyes have to be treated differently if you plot the points manually, because the location labels are mirrored along the y axis.
+In other words, to plot the data manually, for a left eye x should be `-getlocmap()$coord$x`.
 
@@
 ggplot(ODOS_dataframe, aes(x = x, y = y)) +
-  geom_label(aes(label = loc_label, color = is_blindspot)) +
+  coord_fixed() +
+  xlim(-30, 30) +
+  ylim(-24, 24) +
+  geom_label(aes(label = loc_label, color = is_blindspot), size = 2.75) +
   scale_color_manual(values = c("black", "red")) +
@@
 # Psychophysical testing
-
+[THIS SECTION COULD BE MERGED WITH "Performing visual field tests"]
 Psychophysical tests systematically measure the relationship between physical stimulus strength and the subjective perception.
@@
 # Measurements
-
+[THIS SECTION COULD BE AFTER "Performing visual field tests"]
 Although usually not seen in visual field data, the most elementary unit of a static visual field test is a subject's response.
@@
 
-## False negatives / positives
+## False negatives and false positives
+[THIS SECTION COULD BE AFTER "Performing visual field tests" TOO?]
 
@@
 
-# Mathematical derivatives of threshold
+# Mathematical derivatives of contrast thresholds for statisitcal analysis
+## Identifying defective locations on the visual field
+[A PART OF THIS HAS BEEN EXPLAINED IN MY EDITED VERSION OF "Performing visual field tests". I HAVE REWORKED THIS SECTION. SEE IF MY EDITS MAKE SENSE]
+As mentioned earlier, the sensitivity in dB of attenuation from the maximum stimulus at each location is computed from the contrast thresholds.
 
-## Sensitivity and other mathematical derivatives of threshold
+The sensitivity to light fluctuates considerably between measurements in the same observer (test-retest variability ---and more as the disease progresses) and between observers (inter-individual differences).
+They also decrease at different rates depending on the eccentricity with age.
 
-However, the data analyzed with the visual fields package usually consists of the subjects *differential light sensitivity* for a stimulus at a given location in decibel (dB) as the most elementary unit.
-This sensitivity is calculated from the psychophysical threshold, which is derived from the aforementioned responses at this location.
+With these predictable and unpredictable differences in the pointwise sensitivity measurements, how can we provide tools and reports to aid the clinical assessment of the visual field?
+Or more to the point, how can we discriminate between healthy locations and locations that are damaged or suspected of being damaged?
+Clinically it is more useful to compute and report the deviations from an age-corrected mean normal visual field, since it becomes easier the identification of spatial patterns of damage thus facilitating the clinical assessment of the visual field.
 
-The light sensitivity [dB] is defined as $10*log_{10}{\frac{L_{max}}{L}}$. Where $L$ is the threshold luminance and $L_{max}$ is the maximal stimulus intensity the perimeter uses.
+But this analysis requires the use of a *normative dataset* from which to obtain models describing the age effect on sensitivity at each location.
+These functions are usually linear and those *age linear models* are used to obtain the *age-corrected mean normal visual field*.
 
-The light sensitivity fluctuates considerably between measurements in the same observer, between observers, but it also decreases with age.
-
-## Defect / Deviation
-
-Clinically it is often more interesting to map the deviation from a theoretical "normal" sensitivity than the absolute sensitivity, because this facilitates the identification of disease.
-Furthermore, a decline of DLS with age occurs also in normal subjects.
-Therefore, age-related normative date must be collected which results in a function the describes the relationship between DLS and age for each location in the visual field.
-This function is usually linear.
-
-The resulting difference is also measured in dB and is called either *deviation* (loss of function has a negative sign; used in Humphrey perimeters) or *defect* (loss of function has a positive sign; used in Octopus perimeters).
-In this package, *deviation* is used, and it is called *total deviation*.
-Make sure that you get the right sign if you import data.
+The resulting *deviations* from age-corrected mean normal visual field are also measured in dB and they are interpreted as follows.
+In the Humphrey Field Analyzer, a negative deviation value means that the sensitivity at that location is below mean normal.
+By contrast, on Octopus perimeters, negative is above and positive below mean normal.
+In this package, those differences are called *total deviation* and their interpretation are as for the Humphrey perimeter.
+In case you use visualFields with data from a different perimeter it is important to know what is the definition of *deviation* and change the sign if necessary.
 
 In eye diseases like glaucoma, general defects that affect the whole visual field are often less relevant than focal scotomas.
-Therefore, deviation is sometimes corrected for the average sensitivity of the visual field.
-This is called *pattern deviation*.
+And often ocular opacities caused by cataract can mask such focal defects.
+For this reason, a statistical post-hoc correction to *total deviation* values is further applied to obtain *pattern deviation* values.
+The correction factor is often called the *general height* and for the 24-2 Humphrey is obtained by selecting the 7th greatest total deviation value.
+The *pattern deviation* values equal the *total deviation* values minus the *general height*.
 
-### Sensitivity
+## Normative dataset and normative references
+The visualFields package contains three datasets of healthy subjects that can be used to obtain normative age linear models and references.
 
-In the visualField datasets, $l_1$ to $l_n$ represents sensitivities.
+The SUNY-IU dataset [@IMF:Mar2013b] for the 24-2 SITA Standard and the Goldmann Size III:
+```{r SUNYIU datasets}
+str(vfctrSunyiu24d2[,1:10])
+```
 
-```{r}
+The Iowa dataset [@IMF:Marin2018c] for a bimodal ZEST algorithm and the Goldmann Size V:
+```{r Iowa datasets}
+str(vfctrIowaPC26[,1:10])
+str(vfctrIowaPeri[,1:10])
+```
 
-example2[, getlocini()+1:4]
+From these datasets, or any other dataset of healthy eyes, normative references can be computed as follows:
+
+[CORD: I ADDED THIS TO SHOW THAT VISUALFIELDS IS NOT JUST TO ANALYZE THE DATA IN A CERTAIN WAY, YOU GENERATE YOUR OWN NORMATIVE VALUES AND WORK WITH THEM.
+SO FROM HERE ALL EXAMPLES ARE FROM nv INSTEAD OF getnv().
+AT THE END OF THE DOCUMENT, WE CAN EXPLAIN setnv(), getnv(), setlocini(), setdefaults(), ETC.
+OK?]
+
+```{r nv}
+nv <- nvgenerate(vfctrSunyiu24d2)
+```
+
+For future use, we set `nv` as the normative values to be used in visualFields for the calculation of total and pattern deviation values and other statistical analyses.
+```{r setnv}
+setnv(nv)
+```
+
+### Age linear models and mean normal visual fields
+The normative values and references stored in `nv` contains the coefficients of the age linear models:
+
+```{r agelm}
+print(nv$agem$coeff)
+```
+
+And a function from where the age-corrected mean normal visual fields can be computed:
+```{r agemeannormal}
+age <- 50
+print(nv$agem$model(age))
+```
+
+Notice that the locations corresponding to the anatomical region of the blind spot all have `NA` as values, as they are excluded from the statistical analysis.
+
+## Total deviation
+Total deviation is calculated from the sensitivities and from age-related normal values. These are saved in the environment variable *nv* (normal values).
+
+```{r sensitivities}
+
+print(example2[, getvfcols()[1:10]])
 
@@
 
-### Total deviation
-
-Total deviation is calculated from the sensitivities and from age-related normal values. These are saved in the environment variable *nv* (normal values).
-
 ```{r total deviation}
 
-age <- example2[1, "age"]
+age <- example2$age[1]
 
@@
 
-normal_values <- getnv()$agem$model(age)[1:4]
+normal_values <- getnv()$agem$model(age)[1:10]
 
@@
 
-total_defects <- example2[, getlocini()+1:4] - normal_values
+total_defects <- example2[, getvfcols()[1:10]] - normal_values
 
@@
 
-gettd(example2)[, getlocini()+1:4]
+td <- gettd(example2)
+print(td[, getvfcols()[1:10]])
 
@@
 
-### Pattern deviation
+## Pattern deviation
+Pattern deviation can be calculated after obtaining the general height.
+```{r gh}
+gh <- getgh(td)
+print(gh)
+pattern_defects <- td[, getvfcols()[1:10]] - gh
+print(pattern_defects)
+```
 
-The pattern deviation corrects for the mean deviation, i.e.
-the average sensitivity loss.
-It highlights local defects but does not show global sensitivity loss, for example due to opacification of the optic media.
+These values can be automatically calculated with a dedicated function.
 
@@
 
-getpd(example2)[, getlocini()+1:4]
+pd <- getpd(td)
+print(pd[, getvfcols()[1:10]])
 
@@
 
-## Probabilities
+## Total and Pattern deviation probabilities
+As mentioned psychophysical sensitivity values vary considerably from test to test and between subjects Therefore, even in perfectly average eyes deviations will be different from zero most of the time.
+Given a visual field from a patient, the clinician has to know if a given deviation is likely to be caused by disease as opposed to normal retest and inter-individual variability.
 
-Psychophysical values vary considerably between subjects, but they also fluctuate between measurements.
-Therefore, deviation will be different from zero most of the time, and the clinician has to know, if a given deviation is likely to be caused by disease, or whether it may just reflect normal variability.
+The function `nvgenerate` computes for the normative dataset, the total deviation and pattern deviation values for healthy eyes and records pointwise empirical probability distribution functions [@IMF:Hei1987b] that can be used to compute the probability that a specific total deviation (or pattern deviation value) value is from a healthy location [@IMF:Hei1987a] [@IMF:Heijl1991].
+Thus, a low probability implies a greater likelihood of the location being damaged.
+Typically (and in our example), the probabilities are categorized as
+```{r probability categories}
+print(getnv()$lut$probs)
+```
 
-This is calculated based on measurements in the reference population that is also used to calculate the normative data.
+where `0.005` means probability $< 0.005$, where `0.01` means probability $< 0.01$, etc. A probability greater than 0.05 is considered within the normal range.
 
-### Total deviation
+The probability values for the previous values are as follows.
 
-```{r total deviation probabilities}
+```{r probability maps}
+print(gettdp(td)[, getvfcols()[1:10]])
+print(getpdp(pd)[, getvfcols()[1:10]])
+```
 
-gettdp(example2)[, getlocini()+1:4]
+# Plotting the visual field
+There are different ways to visualize the perimetric results. These visualizations have to be interpreted like maps in that they represent a simplified version of the hill of vision.
+
+[WE PROBABLY NEED SOME INTRO TO VFPLOT BEFORE TALKING MAKING NOTES, ETC?]
+The standard already contains some additional information with a border presenting a color-coded visualization of the deviation and a graphical representation of the blind spot.
+Note that the vfplot with the option "td" needs information on differential light-sensitivities in a normal reference population and on the probability of given deviations from these normal values.
+These data are included for the typical Humphrey Field Analyzer fields, but you need to collect this information if you use other perimeters
+
+Note that you have to specify what type of data is contained in the visual field tables. The default is set to *total deviation*, `td`.
+
+```{r}
+
+vfplot(example2)
 
@@
 
-### Pattern deviation
+In this graph, yellow represents probabilities < 0.05, orange probabilities < 0.02, red < 0.01, and dark red < 0.005.
 
-```{r pattern deviation probabilities}
-
-getpdp(vfpwgSunyiu24d2)[1:4, 1:12]
-
-```
-
-# Plotting the visual field
-
-There are different ways to visualize perimetric results. These visualizations have to be interpreted like maps in that they represent a simplified version of the hill of vision.
-
-The standard already contains some additional information with a border presenting a color-coded visualization of the deviation and a graphical representation of the blind spot.
-Note that the vfplot with the option "td" needs information on differential light-sensitivities in a normal reference population and on the probability of given deviations from these normal values.
-These data are included for the typical Humphrey Field Analyzer fields, but you need to collect this information if you use other perimeters
+It is possible to print sensitivities (with darker representing locations with lower sensitivities)
 
@@
 
-Note that you have to specify what type of data is contained in the visual field tables. The default is set to *total deviation*.
+Pattern deviation values
 
@@
 
-vfplot(example2, type = "td")
+vfplot(example2, type = "pd")
 
 ```
+
+And an hybrid plot showing total deviation values and probabilities along with a grayscale representing the sensitivities.
 
@@
 
-```{r}
-
-vfplot(example2, type = "pd")
-
-```
+And a similar hybrid plotbut with pattern deviation values and probabilities.
 
@@
 
-# Global parameters
-
-Global parameters summarize the visual field in a few parameters.
-These global parameters also can be calculated with the visualField package.
-The resulting table contains the visual field header data (patient id, eye, etc.), but contains 8 columns (msens, ssens, tmd, tsd, pmd, psd, gh, and vfi) instead.
-
-The mean sensitivity (MS, msens) is the arithmetic mean of all sensitivity values, the mean deviation (MD, tmd) is the arithmetic mean of all total deviation values in the field.
-The pattern standard deviation (PSD) is the standard deviation of the total deviation values.
+# Global indices
+Global indices summarize the visual field in a few parameters and their values and probability level can be calculated with the visualField package.
+The resulting table contains the visual field header data (patient id, eye, etc.) and 8 additional columns (msens, ssens, tmd, tsd, pmd, psd, gh, and vfi).
 
@@
 
-getgl(example2)[, -(1:getlocini()-1)]
+print(getgl(example2)[,-(1:getlocini()-1)])
 
 ```
-Again, the probability of the observed values given that the visual field is normal can be calculated.
+
+The column `msens` is the mean sensitivity (MS); the arithmetic mean of all sensitivity values. The column `ssens` is the standard deviation of sensitivity values.
+
+The column `tmd` is the mean deviation (MD); the arithmetic mean of all total deviation values. The column `tsd` is the standard deviation of total deviation values.
+
+The column `pmd` is the mean of pattern deviation values. The column `psd` is the standard pattern deviation (PSD) or the standard deviation of the pattern deviation values.
+
+The last column, `vfi`, is the visual field index (VFI) [@IMF:Ben2008], and the second to last is the general height (GH).
+
+The probability of the observed values given that the visual field is normal can also be calculated.
 
@@
 
-getglp(getgl(example2))[, -(1:getlocini()-1)]
+print(getglp(getgl(example2))[, -(1:getlocini()-1)])
 
@@
 
+# Putting everything together.
+Single field analysis reports can be created as pdfs or as interactive reports with Shiny. To explore them, type in the R Console the following:
+
+```{r eval = FALSE, include = TRUE}
+vfsfa(example2, file = "test.pdf")
+vfsfashiny(example2)
+```
+
 # Structure function correlations in glaucoma patients
-
-In the last years, techniques detailed imaging of the retina and the optic nerve have made incredible advances.
+In the last years, techniques that provide detailed imaging of the retina and the optic nerve have made incredible advances.
 Therefore, correlation between structure and function is becoming even more important than ever.
 
-When photoreceptors are lost in an specific retinal area, the scotoma always occurs in the directions in space that is perceived by that retinal area. This makes structure-function correlations easy.
+When [SINCE THIS IS MORE FOCUSED TO GLAUCOMA, I REMOVED PHOTORECEPTORS AND PUT INSTEAD GANGLION CELLS, AS PHOTORECEPTORS REMAIN INTACT IN GLAUCOMA] ganglion cells are lost in characteristic patterns at the retinal, [I DO NOT UNDERSTAND THIS SENTENCE] the scotoma always occurs in the directions in space that is perceived by that retinal area.
+
+To aid comparisons between structural insult and functional defect, it is useful to understand and model the anatomy of a typical healthy eye, even if such modeling through is imperfect given the large inter-individual differences in anatomical features [@IMF:Jan2009] [@IMF:Jan2012] [@IMF:Drasdo2007] [@IMF:Montesano2020]
 
@@
 However, the cell bodies of the retinal ganglion cells are displaced from the area where the photoreceptors are located that they receive input from.
-This displacement is relevant mostly in the center of the visual field.
+This displacement is relevant mostly in the center of the visual field, so it would greatly affect the structure-function comparisons on grids such as the 10-2.
+The following figure shows the visual field locations and the corresponding ganglion cell soma being stimulated.
+
+```{r echo=FALSE}
+
+locmap <- locmaps$p10d2
+gcmap <- vf2gc(locmap$coord)
+par(mfrow = c(1, 2))
+plot(x = locmap$coord$x, 
+     y = locmap$coord$y, 
+     type = "n", 
+     asp = 1,
+     xlim = c(-10, 12),
+     ylim = c(-10, 10),
+     yaxt = 'n',
+     xaxt = 'n',
+     bty = 'n',
+     xlab = "",
+     ylab = "",
+     main = "10-2 grid of test locations")
+lines(-c(10, 0.5), c(0, 0), lwd = 2, col = "lightgray")
+lines(c(0.5, 10), c(0, 0), lwd = 2, col = "lightgray")
+lines(c(0, 0), -c(10, 0.5), lwd = 2, col = "lightgray")
+lines(c(0, 0), c(0.5, 10), lwd = 2, col = "lightgray")
+lines(x = c(-0.5, 0.5), y = c(0, 0), lwd = 1.5)
+lines(x = c(0, 0), y = c(-0.5, 0.5), lwd = 1.5)
+for(r in c(2.5, 5, 7.5, 10))
+  draw.circle(0, 0, r, lwd = 2, col = NA, border = "lightgray")
+text(x = 11, y = 0, "10\u00b0", adj = c(0, 0.5))
+text(x = 0, y = 11, "10\u00b0", adj = c(0.5, 0))
+points(x = locmap$coord$x, y = locmap$coord$y, bg = "white")
+plot(x = locmap$coord$x, 
+     y = locmap$coord$y, 
+     type = "n", 
+     asp = 1,
+     xlim = c(-10, 12),
+     ylim = c(-10, 10),
+     yaxt = 'n',
+     xaxt = 'n',
+     bty = 'n',
+     xlab = "",
+     ylab = "",
+     main = "Corresponding ganglion cells")
+lines(-c(10, 0.5), c(0, 0), lwd = 2, col = "lightgray")
+lines(c(0.5, 10), c(0, 0), lwd = 2, col = "lightgray")
+lines(c(0, 0), -c(10, 0.5), lwd = 2, col = "lightgray")
+lines(c(0, 0), c(0.5, 10), lwd = 2, col = "lightgray")
+lines(x = c(-0.5, 0.5), y = c(0, 0), lwd = 1.5)
+lines(x = c(0, 0), y = c(-0.5, 0.5), lwd = 1.5)
+for(r in c(2.5, 5, 7.5, 10))
+  draw.circle(0, 0, r, lwd = 2, col = NA, border = "lightgray")
+text(x = 11, y = 0, "10\u00b0", adj = c(0, 0.5))
+text(x = 0, y = 11, "10\u00b0", adj = c(0.5, 0))
+points(x = gcmap$x, y = gcmap$y, pch = 21, bg = "white")
+```
 
 Therefore, the visualFields package provides function to calculate where the ganglion cells that subserve a given location in the visual field are located and to calculate at which angle the axons of these ganglion cells enter the optic nerve head (optic disc), where they pass through the sclera (wall of the eye) to form the optic nerve.
+This function is based on Montesano et al.
+[@IMF:Montesano2020] which revisited the original work by Drasdo et al.
+[@IMF:Drasdo2007].
 
@@
 
-locmaps$p24d2$coord[23, ]
-
-loc2psi(locmaps$p24d2$coord[23, ])
-
-vf2gc(locmaps$p24d2$coord[23, ])
+print(locmaps$p24d2$coord[23,])
+print(vf2gc(locmaps$p24d2$coord[23,]))
 
@@
 
-From the angle of incidence, the nerve fiber bundle path can be calculated.
+The Montesano Ganglion-cell displacement function can be in conjunction with the Jansonius map describing average ganglion cell bundles paths reaching different regions of the retina [@IMF:Jan2009] [@IMF:Jan2012].
+First, we can obtain the angle of incidence on the optic never head (ONH)
+```{r}
+print(loc2psi(vf2gc(locmaps$p24d2$coord[23,])))
+```
+
+And from the angle of incidence, the nerve fiber bundle path can be calculated.
 Note that the function for calculating the trajectory will extrapolate the trajectory beyond the location of the visual field (usually to the raphe).
@@
 
-bundle_path <- data.frame(y = 0:90,
-                          x_l19 = path_function_l19(0:90),
-                          x_l23 = path_function_l23(0:90),
-                          x_l47 = path_function_l47(0:90))
-
+bundle_path <- data.frame(y = 4:90,
+                          x_l19 = path_function_l19(4:90),
+                          x_l23 = path_function_l23(4:90),
+                          x_l47 = path_function_l47(4:90))
 ggplot(bundle_path, aes(y = y)) +
@@
 ```
+THERE ARE SOME WARNINGS IN THIS CODE THAT I DO NOT KNOW HOW TO TACKLE
 
-In this plot, the origin corresponds to the center of the ONH, the black triangle corresponds to the fovea, and the 180degree angle corrsponds to the line between the ONH center and the fovea, which is, in reality, slightly tilted downward in most patients.
+In this plot, the origin corresponds to the center of the ONH, the black triangle corresponds to the fovea, and the $180^\circ$ angle corresponds to the line between the ONH center and the fovea, which is, in reality, slightly tilted downward in most patients.
+I DO NOT SEE THIS, BUT I MAY BE MISSING SOMETHING.
+ALSO THE FOVEA (TRIANGLE IF I GOT IT RIGHT?) SHOULD BE A BIT BELOW THE HORIZONTAL MERIDIAN.
+THE POSITION OF THE CENTER OF THE BLIND SPOT IS (15,+2) IN THE MODEL (PLUS AND NOT MINUS BECAUSE THE MODEL IS SPECIFIED FOR THE ANATOMICAL RETINA, AND IT NEEDS TO BE TURNED UPSIDE DOWN TO MATCH THE VISUAL FIELD)
+
+For convenience, visualFields also contains a function to map the angle of incidence of bundles on the ONH to the corresponding angle of the circumpapillary retinal nerve fiber layer (cpRNFL) scan.
+This allows to map a visual field location to its corresponding ganglion cell location, to its angle of incidence on the ONH and finally to the angle in the cpRNFL scan.
+
+```{r}
+
+l23 <- locmaps$p24d2$coord[23,]
+
+# location on the visual field to corresponding ganglion cell
+gc23 <- vf2gc(l23)
+print(gc23)
+
+# ganglion cell location to angle of incidence on the ONH
+psi23 <- loc2psi(gc23)
+print(psi23)
+
+# angle of incidence on the ONH to OCT cpRNFL scan angle
+ang23 <- psi2oct(psi23)
+print(ang23 - 360)
+
+```
 
 # Longitudinal analyses
-
-Monitoring function over time in glaucoma patients is the most common clinical application of visual fields in ophthalmology.
+Monitoring visual function over time in glaucoma patients is an essential clinical application of visual fields in ophthalmology and optometry practices.
 
 For several reasons, measurements will fluctuate over time.
-Fluctuation between measurements will be higher in eyes with glaucoma, especially in areas of glaucoma-related visual field defects.
-Therefore, it is not possible to compare two field tests and claim that the disease has progressed only because the second test yields poorer sensitivities.
+Fluctuation between measurements will be greater in eyes with glaucoma, especially in areas of glaucoma-related visual field defects.
+Therefore, it is not possible to compare two field tests and claim that the disease has progressed only because the second test yielded lower sensitivities.
 
-Thus, statistical methods and a minimum number of tests are necessary in order to demonstrate disease progression in the visual fields.
-Statistical methods that can be subdivided into *event-based* and *trend-based* methods.
+Statistical methods and a series with a minimum number of tests are necessary in order to demonstrate disease progression in the visual fields.
+Statistical methods for progression in glaucoma that can be subdivided into *event-based* and *trend-based* methods.
 Furthermore, methods can be subdivided into those that require datasets which describe the "normal fluctuation" in a given population, those that rely on a mathematical model of visual field progression, and those that do not rely on either.
 
-Event-based methods are not implemented in the visualFields pacakge.
-Briefly, they look for progression between the most recent field and a baseline test.
+Event-based methods are not implemented in the visualFields package.
+Briefly, they look for progression between the most recent field and a baseline.
 Progression is characterized by a decrease in sensitivity with a magnitude that is unlikely (< 5%) in the population under examination (population based data is necessary) and needs to be confirmed in further tests (usually at least two).
@@
 
-For the next example, we will select a series of 27 visual fields that were obtained over several years.
+There are several tools to analyze progression with visualFields. To review them, we will select a series of 27 visual fields that were obtained over several years.
 
 ```{r}
-
 example3 <- vffilter(vfpwgSunyiu24d2, 
@@
 ## Plotting progression
-
-The sparkline plot only shows the raw data.
+The sparkline plot only shows the raw data: the series of visual field data (by default total deviation values) over time at each location of the visual field.
+The lines are plotted in red if they are very "noisy", i.e., if the median absolut deviation of the residuals from simple linear regression are greater than a certain threshold `thr` (by default 2 dB).
+Here is an example with the threshold set at 4 dB.
 
@@
 
-vfsparklines(example3)
+vfsparklines(example3, thr = 4)
 
@@
 
-The lego plot shows probabilities for progression which relies on population based data.
+The legoplot is a visualization device designed to compare the first visual fields against the last ones on a series in a sort of "wait-and-see" fashion.
+By default it compares the first 3 visual fields against the last 3 of a series.
+And by default it compares total deviation values (and probabilities of damage, which relies on population based statistics).
+This example, however, shows sensitivities.
 
@@
 
-vflegoplot(example3)
+vflegoplot(example3, type = "s")
 
@@
 
+The outer polygon shows the grayscale for the average sensitivity over the first 3 visual fields, whereas the inner circles show the grayscale for the average of the last 3.
+The numbers inside the inner circles are the difference in sensitivity between the last 3 and first 3 visual fields.
+
 ## Statistical analysis
+Statistical analysis can be performed with linear regression of global indices or with pointwise linear regression (one regression line for each location), which results in as many slopes and $p$-values as there are visual field locations (52 for the 24-2 grid after removing the two that fall on the anatomical region of the blind spot).
 
-Statistical analysis can be performed with linear regression of global parameters, pointwise linear regression (one regression line for each location), which results in a number of p-values, or with Permutation analysis of Pointwise Linear Regression (PoPLR).
-
-The *glr* function performs linear regression on global parameters. In the following example, the patient loses 0.24dB each year and this change is statistically significant.
+The *glr* function performs linear regression on global parameters.
+In the following example, the patient loses `r round(abs(glr(getgl(example3), type = "md")$sl), 2)` dB each year and this change is statistically significant, with a $p$-value < `r format(ceiling(1000 * glr(getgl(example3), type = "md")$pval) / 1000, nsmall = 3)`.
 
@@
 
-slope <- glr(getgl(example3), type = "md")$sl
-intercept <- glr(getgl(example3), type = "md")$int
+globallm <- glr(getgl(example3), type = "md")
+slope <- globallm$sl
+intercept <- globallm$int
+pval <- globallm$pval
 
@@
 
-glr(getgl(example3), type = "md")$pval
+print(paste("intercept:", intercept))
+print(paste("slope:", slope))
+print(paste("p-value:", pval))
 
@@
 
-The *plr* function performs pointwise linear regression (i.e.
+The `plr` function performs pointwise linear regression (i.e.
 linear regression for each location in the field separately).
 Specific parameters have to be chosen, in order to detect disease progression.
-For example, progression might be defined as at least 3 locations having a slope less than 0dB/year and a p-value less than 0.05.
+For example, progression might be defined as at least 3 locations having a slope less than 0 dB / year and a $p$-value < 0.05.
 
@@
 
-print(sum(prog1$pval < 0.05))
-
-vfplotplr(example3)
+# show the total number of siginficantly progressions locations
+# (out of the 52 pointwise linear regressions)
+print(sum(prog1$pval[-getlocmap()$bs] < 0.05))
 
@@
 
-The *poplr* function performs pointwise linear regression, and then calculates an S-score on the visual field series and on random permutations of the series, in order to control the Type I error (diagnosing progression in a non-progressing series) without relying on population data or a mathematical model of progression.
-It is computationally intensive.
+The results of `plr` can be displayed using the following:
+
+```{r}
+vfplotplr(example3)
+```
+In this graphical display, yellow means that the $p$-value for this location is < 0.05, orange that it is < 0.02, red < 0.01, and dark red < 0.005.
+
+Criteria such as the one suggested earlier based on clusters of locations significantly progressing are useful clinically but do not control for specificity.
+An alternative method that ensure that specificity of progression is mantained at 95% or any other desired level is Permutation of Pointwise Linear Regression or PoPLR [@IMF:Ole2012] [@IMF:Marin2021].
+For this method, all 52 $p$-values are combined into a single $S$-statistic, originally introduced by Fisher [@IMF:Fisher1925].
+In order to avoid reliance on population statistics or parametric statistics.
+(Fisher's original work showed that such statistic followed a chi-square distribution but only if the 52 significance tests were independent of each other and the data was Gaussian, two assumptions that are far from reality.) Thus, random permutations of the series of visual fields are generated and, for each, the $S$-statistic computed.
+If enough random permutations are performed, the resulting set of $S$-values represent an empirical distribution for the null hypothesis that there is no progression.
+A combined $p$-value can then be obtained to test this hypothesis against deterioration of the visual field.
+The interpretation of the $p$-value is as with any significance test.
 
 The *poplr* function yields a large list with many intermediate computations.
-The main result is the S-score and the probability of getting this S-score in the absence of progression (Type I error).
-Note that there is a left and a right probability (deterioration and improvement) in the current implementation.
+The main result is the $S$-value and its $p$-value.
+Note that there is a left tailed (deterioration), a right tailed (improvement) and a two-tailed (visual field not stable) versions in the current implementation.
 
@@
 
-prog2$csr
-prog2$csrp
+print(prog2$csr)
+print(prog2$csrp)
 
@@
 
-In contrast, a deterioration is very likely (p < 0.0001). In the histogram, you can see that a S-score of 176 was only very rarely found in the random permutations of the visual field series.
+In contrast, a deterioration is very likely (p < 0.001).
 
@@
 
-hist(prog2$cstats$cslall)
-abline(v = prog2$csl)
+```
+
+In the histogram representing the null distribution, you can see that a $S$-value of `r round(prog2$csl)` was only very rarely found in the random permutations of the visual field series.
+
+```{r}
+
+hist(prog2$cstats$cslall, xlab = "S-values", xlim = c(0, 200))
+abline(v = prog2$csl, col = "red", lwd = 2)
 
@@
 ## Creating pretty reports
+Progression reports can be created as pdfs or as interactive reports with shiny. To explore them, type in the R Console the following:
 
-Progression reports can be created as pdfs or as interactive reports with shiny.
+```{r eval = FALSE, include = TRUE}
 
-```{r eval=FALSE, include=TRUE}
-
-# vfspa(example3, file = "test.pdf")
-
-# vfspashiny(example3, file = "test.pdf")
+vfspa(example3, file = "test.pdf")
+vfspashiny(example3)
 
@@
 # DICOM standard of visual fields
-
-DICOM (Digital Imaging and Communications in Medicine) standards for ophthalmic static perimetry are specified in Supplement 146.
+DICOM (Digital Imaging and Communications in Medicine) standards for ophthalmic static perimetry are specified in Supplement 146 [CAN YOU PROVIDE A REFERENCE OR SPECIFY SUPPLEMENT IN WHAT?].
 However, there are two functions for loading data generated in the DICOMM format created by the Humphrey Field Analyzer (HFA) by Zeiss.
 
-Currently, the data representation in the VisualFields package differs from these standards and data cannot be exported from the package to DICOMM.
+Currently, the data representation in the visualFields package differs from these standards and data cannot be exported from the package to DICOMM.
 
 # Using a different perimeter
-
 Different manufacturers have different conventions in perimetry.
-For example, clinicians working with Humphrey Field Analyzers (Carl-Zeiss-Meditec) are often used to report loss of function as negativ values (MD: mean *deviation* of -10dB), while those working with Octopus perimeters (Haag-Streit) often report loss as a positive value (MD: mean *defect* of 10dB).
-Furthermore, clinicians working with Octopus perimeters often use different grids, even though the 24-2 and 10-2 grids are theoretically available.
+For example, clinicians working with Humphrey Field Analyzers (Carl-Zeiss-Meditec) are often used to report loss of function as negative values (e.g., MD: mean *deviation* of $-10$ dB), while those working with Octopus perimeters (Haag-Streit) often report loss as a positive value (e.g., MD: mean *defect* of 10 dB).
+Furthermore, clinicians working with Octopus perimeters often use different grids, even though the 24-2 and 10-2 grids are also available.
 Usually, fields cannot be directly compared because other setting ($L_max$) are also different and because different algorithms are used.
 
-The VisualField package provided normative data for the Humphrey Field Analyzer but currently not for other perimeters.
-
-The loadoctopus function also imports some normative data from the Octopus perimeter by Haag-Streit, but, as of now, this does not allow full use of the functionality of the visualFields package
-
-For making full use of the visusalField package with other parameters, you ideally need a database of visual fields from a normal reference population, but the creation of normal values with the visualFields package goes beyond the scope of this vignette.
+The visualField package provided normative data for the Humphrey Field Analyzer 24-2 and Goldmann size III and for Iowa's full field perimetry [@IMF:Marin2018c], which comprises two different grids, Central $26^\circ$ and Periphery (from $30^\circ$ to more than $60^\circ$), with the Goldmann Size V stimulus on the Octopus 900.
+The `loadoctopus` function also imports some normative data from the Octopus perimeter by Haag-Streit, but, as of now, this does not allow full use of the functionality of the visualFields package.
+Currently, there are no other normative references for any other perimeters or grids.
+Research labs can, however, use their suitable normative datasets to generate normative values for analysis with visual fields.
 
 # Advanced topics
-
 ## Setting up the environment
-
 A given visual field program has certain properties that remain unchanged during the test.
-In the visualFields packages these are saved in the *environment*.
+In the visualFields packages these are saved in an *environment*.
 After setting this environment, you can only analyse data that is consistent with this environment.
 
-The environment contains information on
+The core elements of the environment that need to be setup are
 
--   Normative data (nv)
--   Grid (locmap)
--   Graphical parameters used for plotting the data (gpar)
--   Where to find visual field measurements in the tables (locini)
+-   The normative reference values (`nv`)
+-   The test grid (`locmap`)
+-   Graphical parameters used for the different visualization devices (`gpar`)
+-   The column where the visual field measurements start in the tables (`locini`)
 
-The nv environment variable also contains functions for calculating defects, probabilities and global parameters that can be accessed via the functions stated above.
+The `nv` environment variable also contains functions for calculating defects, probabilities and global parameters that can be accessed via the functions stated above.
 
-```{r see current environment variables}
+visualFields provides predefined test grids `locmaps`, normative values `normvals`, and parameters for graphical display (`gpars`)
 
-getnv()$info        # current normal values
-str(getnv(), 1)
+```{r}
 
-str(getlocmap(), 1)   # current locmap
-str(getgpar(), 1)     # current parameters for plotting
-getlocini()   # position of the column l1 in the datafram
+print(names(locmaps))
+print(names(normvals))
+print(names(gpars))
 
@@
 
-The standard environment assumes a static white-on-white perimetry with a Humphrey perimeter, using a 24-2 grid, a SITA standard algorithm, a stimulus size of Goldman Size III.
-The stimulus duration and the $L_{max}$ are not explicitly stated in the info.
+They should be consistent.
+For instance, `locmaps$p24d2` can only be used with one of the following `normvals`: `sunyiu_24d2_pw`, `sunyiu_24d2`, `sunyiu_24d2_pw_cps` or `sunyiu_24d2_cps`, but not with, e.g., `iowa_PC26_pw`.
 
-## Locmaps
+Likewise, `locmaps$p24d2` and `normvals$sunyiu_24d2_pw` or `normvals$sunyiu_24d2` can only be used with `gpars$p24d2`, but not the "_csp" version `gpars$p24d2_cps`
 
-```{r locmaps}
+```{r getters}
 
-default_locmap <- getlocmap()
-default_locmap$name
-default_locmap$desc
+# current normal values
+print(getnv()$info)
+str(getnv(), 1)
+
+# current locmap
+str(getlocmap(), 1)
+# current parameters for plotting
+str(getgpar(), 1)
+# position of the column l1 in the dataframe
+print(getlocini())
 
@@
 
-A new locmap can be specified manually or selected from the locmaps dataset included in the package.
+The default environment assumes a static white-on-white perimetry with a Humphrey perimeter, using a 24-2 grid, a SITA standard algorithm, a stimulus size of Goldman Size III.
+The stimulus duration and the $\Delta L_{max}$ are not explicitly stated in the info, because they are generally known and it does not impact the statistical analysis of the visual field.
+
+Note that for switching to another perimetry program, you need to set all environment variables. visualFields has setter functions for this purpose.
 
@@
 
-new_locmap <- locmaps$p10d2
-setlocmap(new_locmap)
-
-```
-
-Note that some grids spare the blind spot, either on purpose or because they cover only locations central to the blind spot. In this case, locmap\$bs is numeric(0).
-
-```{r}
-
-blind_spot <- getlocmap()$bs
-
-plot(x = new_locmap$coord$x, 
-     y = new_locmap$coord$y, 
-     type = "point", 
-     asp = 1,
-     xlim = c(-24, 24),
-     ylim = c(-24, 24))
-points(x = default_locmap$coord[default_locmap$bs, "x"],
-       y = default_locmap$coord[default_locmap$bs, "y"],
-       col = "red")
-
-```
-
-Note that for switching to another perimetry program, you need to set all environment variables.
-
-```{r}
-
+# Set locations for Iowa Central 26 grid
+setlocmap(locmaps$pPC26)
+print(getlocmap()$name)
+# Set normative values for Iowa Central 26 grid
 setnv(normvals$iowa_PC26_pw)
-
-setlocmap(locmaps$pPC26)
-
+print(getnv()$info$name)
+# Set graphical parameters for Iowa Central 26 grid
 setgpar(gpars$pPC26)
@@
 
-## Resetting the enviroment
+If the structure of the dataframe containing visual field data has changed, and the sensitivity locations start at a differen column, this can also be changed with another setter function.
 
-If you reset the environment, you return to the 24-2 pattern measured with the SITA algorithm on a Humphrey Field Analyser.
+```{r}
+
+# Set first column with sensitivity values
+setlocini(24)
+print(getlocini())
+
+```
+
+Here is the figure with the test locations of the Iowa central $26^\circ$. Notice that this map does not have any locations falling on the anatomical blind spot.
+
+```{r}
+
+new_locmap <- getlocmap()
+
+plot(x = new_locmap$coord$x, 
+     y = new_locmap$coord$y, 
+     type = "n", 
+     asp = 1,
+     xlim = c(-30, 30),
+     ylim = c(-34, 34),
+     yaxt = 'n',
+     xaxt = 'n',
+     bty = 'n',
+     xlab = "",
+     ylab = "",
+     main = "Iowa Central 26 grid of test locations")
+lines(-c(30, 2.5), c(0, 0), lwd = 2, col = "lightgray")
+lines(c(2.5, 30), c(0, 0), lwd = 2, col = "lightgray")
+lines(c(0, 0), -c(30, 2.5), lwd = 2, col = "lightgray")
+lines(c(0, 0), c(2.5, 30), lwd = 2, col = "lightgray")
+lines(x = c(-2.5, 2.5), y = c(0, 0), lwd = 1.5)
+lines(x = c(0, 0), y = c(-2.5, 2.5), lwd = 1.5)
+for(r in c(10, 20, 30))
+  draw.circle(0, 0, r, lwd = 2, col = NA, border = "lightgray")
+draw.ellipse(15, -1.5, 2.75, 3.75, col = "gray", border = NA)
+text(x = 0, y = -34, "horizontal eccentricity [degrees]")
+text(x = -34, y = -0, "vertical eccentricity [degrees]", srt = 90)
+text(x = 31, y = 0, "30\u00b0", adj = c(0, 0.5))
+text(x = 0, y = 31, "30\u00b0", adj = c(0.5, 0))
+points(x = new_locmap$coord$x, y = new_locmap$coord$y)
+
+```
+
+### Defaults
+The environment can be reset to its default values, the 24-2 pattern measured with the SITA algorithm on a Humphrey Field Analyzer, and with a data frame structure where the sensitivity values start at column 11.
 
@@
 
+print(getlocini())
+print(getlocmap()$name)
+print(getnv()$info$name)
+print(str(getnv(), 1))
+
 ```
+
+## Customizing visualFields
+To make full use of the visualField package with other perimeters, grids, or testing algorithms (staircase, full threshold, different implementations of ZEST, etc.), it is necessary to configure visualFields to the custom test.
+
+The first decision to make is the structure of visual field result data.
+In a data frame (and a loadable csv file) the first 4 columns must have the uniquely identifiable key (`id`, `eye`, `date`, `time`).
+To compute *deviations* from age-corrected mean normal, the age must also be included.
+And it is recommended to add the information about the reliability indices of the exam, such as false positive rate (`fpr`), false negative rate (`fnr`), and fixation losses (`fl`), although the last two have recently been found not to be "reliable".
+To design your structure, looking at the datasets included in the visualFields package may be useful.
+
+The first 10 columns of `vfpwgRetest24d2` contains the key, age, and other information.
+
+```{r}
+str(vfpwgRetest24d2[,1:10])
+```
+
+After the information regarding the test, etc, must come the columns representing the visual field locations, each row containing the sensitivities for a specific exam of the eye of a patient.
+
+```{r}
+str(vfpwgRetest24d2[,11:21])
+```
+
+The name of the columns with sensitivity values of an exam must be prefixed by "l" and followed by the corresponding conecutive location number.
+So for a patient using the 24-2 grid, the full record is:
+
+```{r}
+print(vfpwgRetest24d2[1,])
+```
+
+Depending on the structure the visualField environment variable `locini` must be adjusted so that its value is the column number of the first sensitivity value, `l1`.
+This may be done with the setter `setlocini()`.
+
+For a custom grid of location, it is possible to create and set the location map and its graphical parameters.
+visualFields has a function `vfgpar()` to generate graphical parameters for a given grid.
+Here is an example with the Octopus "G1" map, that uses `vfread` to read visualField's ready visual field data.
+```{r}
+# read the Octopus G1 coordinates and create the locmap
+g1 <- list(coord = read.csv("./g1.csv"), name = "G1", desc = "The G1 location map used by Octopus perimeters", bs = numeric())
+str(g1)
+# set the G1 locmap
+setlocmap(g1)
+# create and set the graphical parameters
+setgpar(vfgpar(g1$coord))
+# read visual fields and plot the sensitivities
+vfplot(vfread("./vfg1.csv"), type = "s")
+```
+
+The function `vfgpar()` that generates the graphical parameters has many input argument to provide as much flexibility as possible regarding the color schemes to use for probability scales, grayscale, etc.
+
+We cannot compute or visualize total deviations, pattern deviations, or their corresponding probability maps, because we do not have normative reference values for this particular test.
+If we had a visualField's ready csv file with a suitable normative dataset, we could read it with `vfread()`, use the previously used function `nvgenerate()`, and set the generated normative values with `setnv()`.
+
+Finally, if we want to go to the default settings, we ran again:
+
+```{r}
+setdefaults()
+```
+
+# Acknowledgments
+The visualField's package would not have existed if it wasn't for the important contribution of many contributors who partially funded the development of the package, and/or contributed datasets, code, and above all insight on visual fields analysis and about perimetry and visual sciences at large.
+Huge thanks to:
+  - William H Swanson
+  - Michael Wall
+  - Andrew Turpin
+  - Paul H Artes
+  - Giovanni Montesano
+  - Mitchell W Dul
+
+# References

### 2024-08-21 vignette.Rmd
Reaction to modifications by imarinfr.

@@
 
-To install the newestversion of the package, run:
+To install the newest version of the package, run:
 ```{r eval=FALSE}
@@
 It is measured in degrees of visual angle, with the center of vision being $0^\circ$, points on the temporal and superior regions of the visual field being positive angles and those in the inferior and nasal being negative.
-On healthy eyes, the sensitivity to light contrast changes as the magnitude of the visual angle increases (or as it is commonly said in perimetry, eccentricity).
+In healthy eyes, the sensitivity to light contrast decreases as the visual angle increases (or as it is commonly said in perimetry, eccentricity).
 The difference in light sensitivity over the visual field as a function of eccentricity is known as the Traquair's *hill of vision* [@IMF:Traquair1938].
 The figure below shows the hill of vision as rendered by Aulhorn & Harms [@IMF:Aul1967].
-Notice in the discontinuity in the hill of vision in the lower right panel.
-Sensitivity in this region, centered at approximately $(15^\circ, -3^\circ)$ in the visual field is is completely blind, as it corresponds to the anatomical region of the retina known as the *blind spot*, where the optical fiber exits from the eye towards the visual cortex.
+Notice the discontinuity in the hill of vision in the lower right panel.
+Sensitivity in this region, the *blind spot*, centered at approximately $(15^\circ, -3^\circ)$ in the visual field corresponds to the optic nerve head [THE BLIND SPOT IS NOT AN ANATOMICAL TERM, BUT A PERIMETRIC ONE], where the optical fiber exits from the eye towards the visual cortex.
 
@@
 The most common type of visual field testing is *static automated perimetry (SAP)*, which is performed with in a dome-shaped device, so that the distance between the stimulus and the eye equal for all locations within the visual field.
-[I WOULD DELETE THE FOLLOWING AS IT MAY CONFUSE RATHER THAN CLARIFY GIVEN THAT NOBODY USES CAMPIMETRY ANYMORE, AT LEAST IN THE UK OR US] In contrast, *campimetry* is performed with a flat screen.
+In contrast, *campimetry* is performed with a flat screen.
+[CAMPIMETRY MAY BE USED MORE OFTEN IN THE FUTURE BECAUSE COMPUTER SCREENS/ VR DEVICES MAY BE USED FOR TELEMEDICINE; LEAVE IT?]
 
@@
 Such thresholds are estimated at a fixed number of locations forming a regular grid as the one in the figure above.
+
 There is another common form of perimetry, *kinetic perimetry*, where a light spot is moved from the periphery towards the center along a number of meridians.
@@
 # Performing visual field tests
-For the typical visual field test, an observer (usually a patient in an ophthalmology or optimetry clinic) is comfortably seated in front of a perimeter and puts his head on a chin rest looking into a white Ganzfeld dome (or hemisphere).
+For the typical visual field test, an observer (usually a patient in an ophthalmology or optometry clinic) is comfortably seated in front of a perimeter and puts his head on a chin rest looking into a white Ganzfeld dome (or hemisphere).
 The perimeter is located in a dimly lit, calm room and the observer has been explained what to do by an experienced technician before the examination.
@@
 \end{equation}
-For the most common devices, The Humprhery Field Analyzer and the Octopus 900, the maximum luminance differential $\Delta L_{\rm max}$ is $10000\ (\rm apostilb) / \pi = 3183.1\ cd\ m^{-2}$.
+For the most common devices, the Humprhery Field Analyzer and the Octopus 900, the maximum luminance differential $\Delta L_{\rm max}$ is $10000\ (\rm apostilb) / \pi = 3183.1\ cd\ m^{-2}$.
 
@@
 
+## Psychophysical testing
+Psychophysical tests systematically measure the relationship between physical stimulus strength and the subjective perception.
+This is performed by varying the stimulus and asking the observer to communicate his perception with a limited number of allowed responses.
+In visual field testing, the observer usually presses a button when a light spot is perceived.
+
+The relationship between stimulus and perception is analyzed using mathematical model.
+Commonly, these models result in an estimate of a *threshold*, which denotes either the absolute stimulus strength that can just barely be perceived or the smallest difference that can be perceived as different.
+
+Real world scenarios where perception is close to threshold include going through a dark forest on a cloudy night or seeing the car in front through a dense fog.
+Stimuli are not clearly seen when close to threshold.
+For this reason, patients sometimes feel that lengthy visual field testing is tedious and that it is as an unreliable examination.
+The latter is not clearly correct when fluctuations are taken into account and data are analyzed with appropriate statistical methods.
+
 ## Grid patterns
-The spatial position of the points $l_1$ to $l_k$ [CAN WE USE K FOR LOCATIONS, TO PROTECT N FOR NUMBER OF VISUAL FIELDS OR SAMPLE SIZE?] in the visual field is found in the location map (locmap) and depends on the pattern used by the perimeter.
+The spatial position of the points $l_1$ to $l_k$ in the visual field is found in the location map (locmap) and depends on the pattern used by the perimeter.
 When analyzing data, you need to specify the locmap.
@@
 Locations of the locmap that cover the blind spot are reported in the locmaps lists under the variable bs.
-If the blind spot is outside the pattern (for example in the 10° visual fields) or is spared by the grid as in the Octopus G1 pattern, the list item bs is set to numeric(0).
+If the blind spot is outside the pattern (for example in the 10° visual fields) or is spared by the grid as in the Octopus G1 pattern, the list item bs is set to `r numeric(0)`.
 
@@
 
-# visualFields as an aid for clinical assessment of the visual field
-In clinical practice, visual field testing is used identify decreases in visual function caused by disease.
+## Measurements
+Although usually not seen in visual field data, the most elementary unit of a static visual field test is a subject's response.
+It consists of the constant stimulus parameters (size, presentation time, background luminance), the stimulus location, the stimulus intensity and whether the subject pressed the button within a predefined time window to indicate that he saw the stimulus.
+
+Below a certain *threshold* luminance, a stimulus is usually not perceivable, above the threshold, the stimulus is usually perceived.
+However, close to threshold, a stimulus is not always reliably perceived.
+By definition, the threshold is the luminance of the stimulus where the subject will see the stimulus with a probability of 50%.
+
+The most straightforward to measure such a threshold would be to present stimuli several time at a number different, fixed stimulus strengths.
+This method is called the method of *constant stimuli* and results in the *psychometric function*, which is usually modeled with a cumulative distribution function (CDF), for example a Gaussian CDF.
+However, this is not the most effective way to estimate a threshold, because stimulus strength far from the threshold will have probabilities to be seen that are close to either 0% or 100%.
+Testing these strengths several times does not yield much information.
+
+Therefore, advanced algorithms were developed for reducing the number of stimulus presentations with limited trade-off concerning the accuracy of the measurements.
+A staircase algorithm where the intensity is decreased by a certain step size when the stimulus was seen, increased when not seen, and the step size is halved upon each revearsal (seen to not seen or vice versa) is a more efficient algorithms that is frequently used.
+More efficient algorithms usually make a trade-off between precision of the threshold estimate and test time, and often incorporate information on responses in neighboring locations into the algorithm.
+Examples include the Swedish Interactive Threshold Algorithm (SITA) of the Humphery Perimeters.
+Note that long test times negatively affect attention and therefore both accuracy and precision of the threshold estimates.
+
+## False negatives and false positives
+
+Residual deviation from the psychometric function may occur due to inattention of the observer but also due to short-term fluctuation of the threshold.
+For example, it has been shown that the threshold underlies short-term fluctuates in the areas where light sensitivity was negatively affected by glaucoma, a disease of the optic nerve.
+
+When the observer does not see a stimulus that is clearly above a level that has been seen before, this is called a false negative response (fnr).
+In contrast, when the subject presses the button although no stimulus is presented, this is a false positive response (fpr).
+
+The cooperation of the subject is also tested in some perimeters by presenting deliberately presenting a stimulus at the location of the blind spot.
+If the subjects sees that stimulus, it must be assumed that he did not fixate correctly (fixation loss) and that the reliability of the field is decreased.
+
+# Clinical applications of perimetry
+In clinical practice, visual field testing is used to identify decreases in visual function caused by disease.
 Such *visual field defects* or *scotomas* can be caused by diseases that affect tissue anywhere along the optic pathway.
@@
 For example, an infarction of the right visual cortex leads complete blindness in the left hemispheres of both eyes with a very distinct border along the vertical meridian.
-Therefore, the patient does not perceive anything to the right of the point they look at.
+Therefore, the patient does not perceive anything to the right of the point he looks at.
 On the other side of the vertical meridian the visual field is normal.
@@
 The environment variable locini contains the number of the first column that contains the actual visual field date.
-Each column $l_{n}$ corresponds to one location in the visual field test.
-A concise way to select the columns $l_1$ to $l_{n}$ is to use the *getvfcols* function.
+Each column $l_{x}$ corresponds to one of {k} locations in the visual field test.
+A concise way to select the columns $l_1$ to $l_{k}$ is to use the *getvfcols* function.
 
@@
 There mustn't be any columns after these columns.
-Use the *vffilter* function to select specific fields and teh *vfisvalid* function to check whether a table fulfills the criteria.
+Use the *vffilter* function to select specific fields and the *vfisvalid* function to check whether a table fulfills the criteria.
 The latter will also tell you what the problem is when it doesn't.
@@
 setdefaults()
-###[I HAVE CHANGED THE EXAMPLE TO ONE FROM THE RETEST DATASET BECAUSE THE ONE FROM SUNY/IU IS TOO EXTREME]
 
@@
 
-# Psychophysical testing
-[THIS SECTION COULD BE MERGED WITH "Performing visual field tests"]
-Psychophysical tests systematically measure the relationship between physical stimulus strength and the subjective perception.
-This is performed by varying the stimulus and asking the observer to communicate his perception with a limited number of allowed responses.
-In visual field testing, the observer usually presses a button when a light spot is perceived.
-
-The relationship between stimulus and perception is analyzed using mathematical model.
-Commonly, these models result in an estimate of a *threshold*, which denotes either the absolute stimulus strength that can just barely be perceived or the smallest difference that can be perceived as different.
-
-Real world scenarios where perception is close to threshold include going through a dark forest on a cloudy night or seeing the car in front through a dense fog.
-Stimuli are not clearly seen when close to threshold.
-For this reason, patients sometimes feel that lengthy visual field testing is tedious and that it is as an unreliable examination.
-The latter is not clearly correct when fluctuations are taken into account and data are analyzed with appropriate statistical methods.
-
-# Measurements
-[THIS SECTION COULD BE AFTER "Performing visual field tests"]
-Although usually not seen in visual field data, the most elementary unit of a static visual field test is a subject's response.
-It consists of the constant stimulus parameters (size, presentation time, background luminance), the stimulus location, the stimulus intensity and whether the subject pressed the button within a predefined time window to indicate that he saw the stimulus.
-
-Below a certain *threshold* luminance, a stimulus is usually not perceivable, above the threshold, the stimulus is usually perceived.
-However, close to threshold, a stimulus is not always reliably perceived.
-By definition, the threshold is the luminance of the stimulus where the subject will see the stimulus with a probability of 50%.
-
-The most straightforward to measure such a threshold would be to present stimuli several time at a number different, fixed stimulus strengths.
-This method is called the method of *constant stimuli* and results in the *psychometric function*, which is usually modeled with a cumulative distribution function (CDF), for example a Gaussian CDF.
-However, this is not the most effective way to estimate a threshold, because stimulus strength far from the threshold will have probabilities to be seen that are close to either 0% or 100%.
-Testing these strengths several times does not yield much information.
-
-Therefore, advanced algorithms were developed for reducing the number of stimulus presentations with limited trade-off concerning the accuracy of the measurements.
-A staircase algorithm where the intensity is decreased by a certain step size when the stimulus was seen, increased when not seen, and the step size is halved upon each revearsal (seen to not seen or vice versa) is a more efficient algorithms that is frequently used.
-More efficient algorithms usually make a trade-off between precision of the threshold estimate and test time, and often incorporate information on responses in neighboring locations into the algorithm.
-Examples include the Swedish Interactive Threshold Algorithm (SITA) of the Humphery Perimeters.
-Note that long test times negatively affect attention and therefore both accuracy and precision of the threshold estimates.
-
-## False negatives and false positives
-[THIS SECTION COULD BE AFTER "Performing visual field tests" TOO?]
-
-Residual deviation from the psychometric function may occur due to inattention of the observer but also due to short-term fluctuation of the threshold.
-For example, it has been shown that the threshold underlies short-term fluctuates in the areas where light sensitivity was negatively affected by glaucoma, a disease of the optic nerve.
-
-When the observer does not see a stimulus that is clearly above a level that has been seen before, this is called a false negative response (fnr).
-In contrast, when the subject presses the button although no stimulus is presented, this is a false positive response (fpr).
-
-The cooperation of the subject is also tested in some perimeters by presenting deliberately presenting a stimulus at the location of the blind spot.
-If the subjects sees that stimulus, it must be assumed that he did not fixate correctly (fixation loss) and that the reliability of the field is decreased.
-
 # Mathematical derivatives of contrast thresholds for statisitcal analysis
+
 ## Identifying defective locations on the visual field
-[A PART OF THIS HAS BEEN EXPLAINED IN MY EDITED VERSION OF "Performing visual field tests". I HAVE REWORKED THIS SECTION. SEE IF MY EDITS MAKE SENSE]
-As mentioned earlier, the sensitivity in dB of attenuation from the maximum stimulus at each location is computed from the contrast thresholds.
+
+As mentioned earlier, the sensitivity in dB at each location is calculated from the contrast threshold relative to the maximal contrast.
 
@@
 Or more to the point, how can we discriminate between healthy locations and locations that are damaged or suspected of being damaged?
-Clinically it is more useful to compute and report the deviations from an age-corrected mean normal visual field, since it becomes easier the identification of spatial patterns of damage thus facilitating the clinical assessment of the visual field.
+Clinically, it is more useful to compute and report the deviations from an age-corrected mean normal visual field, since the identification of spatial patterns of damage becomes easier, thus facilitating the clinical assessment of the visual field.
 
@@
 The resulting *deviations* from age-corrected mean normal visual field are also measured in dB and they are interpreted as follows.
-In the Humphrey Field Analyzer, a negative deviation value means that the sensitivity at that location is below mean normal.
-By contrast, on Octopus perimeters, negative is above and positive below mean normal.
+By convention, in the Humphrey Field Analyzer, a negative deviation value means that the sensitivity at that location is below mean normal.
+In contrast, on Octopus perimeters, negative is above and positive below mean normal (therefore, values are called *defects*).
 In this package, those differences are called *total deviation* and their interpretation are as for the Humphrey perimeter.
@@
 From these datasets, or any other dataset of healthy eyes, normative references can be computed as follows:
-
-[CORD: I ADDED THIS TO SHOW THAT VISUALFIELDS IS NOT JUST TO ANALYZE THE DATA IN A CERTAIN WAY, YOU GENERATE YOUR OWN NORMATIVE VALUES AND WORK WITH THEM.
-SO FROM HERE ALL EXAMPLES ARE FROM nv INSTEAD OF getnv().
-AT THE END OF THE DOCUMENT, WE CAN EXPLAIN setnv(), getnv(), setlocini(), setdefaults(), ETC.
-OK?]
 
@@
 In the last years, techniques that provide detailed imaging of the retina and the optic nerve have made incredible advances.
-Therefore, correlation between structure and function is becoming even more important than ever.
-
-When [SINCE THIS IS MORE FOCUSED TO GLAUCOMA, I REMOVED PHOTORECEPTORS AND PUT INSTEAD GANGLION CELLS, AS PHOTORECEPTORS REMAIN INTACT IN GLAUCOMA] ganglion cells are lost in characteristic patterns at the retinal, [I DO NOT UNDERSTAND THIS SENTENCE] the scotoma always occurs in the directions in space that is perceived by that retinal area.
+Therefore, correlation between structure and function is becoming more important than ever.
 
@@
 
-```{r}
+```{r, warning=FALSE}
 
@@
   scale_color_manual(values = c("l19" = "black", "l23" = "red", "l47" = "blue", "fixation" = "black")) +
-  guides(color = FALSE, shape = FALSE)
-
-```
-THERE ARE SOME WARNINGS IN THIS CODE THAT I DO NOT KNOW HOW TO TACKLE
+  guides(color = "none", shape = "none")
+
+```
 
@@
 THE POSITION OF THE CENTER OF THE BLIND SPOT IS (15,+2) IN THE MODEL (PLUS AND NOT MINUS BECAUSE THE MODEL IS SPECIFIED FOR THE ANATOMICAL RETINA, AND IT NEEDS TO BE TURNED UPSIDE DOWN TO MATCH THE VISUAL FIELD)
+
+[IVAN: I THOUGHT THAT PSI WAS THE ANGLE OF INCIDENCE IN THE OPTIC NERVE HEAD.
+THEREFORE, I CHOSE A POLAR PLOT AND NOT AN X-Y-PLOT.
+ALSO, I THOUGHT THAT THE 0° ANGLE OF INCIDENCE WAS DEFINED AS THE LINE BETWEEN FOVEA AND ONH.
+I DID NOT TURN THE POLAR COORDINATED ACCORDINGLY EVEN THOUGHT I MENTIONED THAT THE FOVEA WAS "TILTED DOWNWARD".
+BUT I DO NOT HAVE MUCH EXPERIENCE WITH THE MODEL, SO PLEASE CORRECT ME IF I'M WRONG.
+HOW WOULD I CALCULATE XY-COORDINATES?]
 
@@
 Furthermore, clinicians working with Octopus perimeters often use different grids, even though the 24-2 and 10-2 grids are also available.
-Usually, fields cannot be directly compared because other setting ($L_max$) are also different and because different algorithms are used.
+Usually, fields cannot be directly compared because other setting ($L_{max}$) are also different and because different algorithms are used.
 
@@
 Here is an example with the Octopus "G1" map, that uses `vfread` to read visualField's ready visual field data.
+
+[I WANTED TO SHOW SOME OF THE FUNCTIONALITY OF THE LOADOCTOPUS FUNCTION, BUT I COULD DOCUMENT THIS ELSEWHERE]
+
+[THE PROBLEM HERE IS THAT NOW PEOPLE WORKING WITH THE OCTOPUS HAVE TO FIND THE COORDINATES ELSEWHERE, AGAIN, BECAUSE THEY DO NOT HAVE ACCESS TO g1.csv. COULD IT BE PROVIDED IN THE PACKAGE?]
+
 ```{r}

### 2024-08-22 vignette.Rmd
deleted comment

@@
 In this plot, the origin corresponds to the center of the ONH, the black triangle corresponds to the fovea, and the $180^\circ$ angle corresponds to the line between the ONH center and the fovea, which is, in reality, slightly tilted downward in most patients.
-I DO NOT SEE THIS, BUT I MAY BE MISSING SOMETHING.
-ALSO THE FOVEA (TRIANGLE IF I GOT IT RIGHT?) SHOULD BE A BIT BELOW THE HORIZONTAL MERIDIAN.
-THE POSITION OF THE CENTER OF THE BLIND SPOT IS (15,+2) IN THE MODEL (PLUS AND NOT MINUS BECAUSE THE MODEL IS SPECIFIED FOR THE ANATOMICAL RETINA, AND IT NEEDS TO BE TURNED UPSIDE DOWN TO MATCH THE VISUAL FIELD)
-
-[IVAN: I THOUGHT THAT PSI WAS THE ANGLE OF INCIDENCE IN THE OPTIC NERVE HEAD.
-THEREFORE, I CHOSE A POLAR PLOT AND NOT AN X-Y-PLOT.
-ALSO, I THOUGHT THAT THE 0° ANGLE OF INCIDENCE WAS DEFINED AS THE LINE BETWEEN FOVEA AND ONH.
-I DID NOT TURN THE POLAR COORDINATED ACCORDINGLY EVEN THOUGHT I MENTIONED THAT THE FOVEA WAS "TILTED DOWNWARD".
-BUT I DO NOT HAVE MUCH EXPERIENCE WITH THE MODEL, SO PLEASE CORRECT ME IF I'M WRONG.
-HOW WOULD I CALCULATE XY-COORDINATES?]
 

### 2025-05-09 jss-article-rnw/article.Rnw
Start des jstatsoft-Artikels.

\documentclass[article]{jss}

%% -- LaTeX packages and custom commands ---------------------------------------

%% recommended packages
\usepackage{orcidlink,thumbpdf,lmodern}

%% another package (only for this demo article)
\usepackage{framed}

%% new custom commands
\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}

%% For Sweave-based articles about R packages:
%% need no \usepackage{Sweave}
\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE}
<<preliminaries, echo=FALSE, results=hide>>=
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
library("MASS")
@

%% -- Article metainformation (author, title, ...) -----------------------------

\author{Cord Huchzermeyer\\INSTITUTION
   \And Iván Marín-Franch\\Computational Optometry}
\Plainauthor{Cord Huchzermeyer, Iván Marín-Franch}

\title{visualFields 1.x: Tools for Analyzing the Field of Vision in \proglang{R}}
\Plaintitle{visualFields 1.x: Tools for Analyzing the Field of Vision in R}
\Shorttitle{visualFields 1.x in \proglang{R}}

\Abstract{
  The \pkg{visualFields} package is part of the Open Perimetry Initiative (OPI), 
  a standard for interfacing with visual field testing machines. This article 
  introduces the package, which provides tools for analyzing the field of vision, 
  including visualization, statistical analysis, and clinical interpretation of 
  visual-field loss and its change over time.
}

\Keywords{visualFields, Open Perimetry Initiative, visual field analysis, \proglang{R}}
\Plainkeywords{visualFields, Open Perimetry Initiative, visual field analysis, R}

\Address{
  Cord Huchzermeyer\\
  INSTITUTION\\
  E-mail: \email{cord@example.com}\\
  \\
  Iván Marín-Franch\\
  Computational Optometry\\
  E-mail: \email{ivan@example.com}\\
  URL: \url{https://www.computationaloptometry.com/}
}

\begin{document}

%% -- Introduction -------------------------------------------------------------

\section[Introduction: Visual Fields in R]{Introduction: Visual Fields in \proglang{R}} \label{sec:intro}

The Open Perimetry Initiative (OPI) \citep{IMF:Tur2012, IMF:Marin2022a} is a standard for interfacing with visual field testing machines (perimeters). It specifies basic functions that allow many visual field tests to be constructed. As of October 2017, it is fully implemented on the Octopus 900 and partially on the Heidelberg Edge Perimeter, the Kowa AP 7000, the CrewT IMO, and the Centervue Compass.

The \pkg{visualFields} package, part of the OPI, is a collection of tools for analyzing the field of vision. It provides a framework for the development and use of innovative methods for visualization, statistical analysis, and clinical interpretation of visual-field loss and its change over time.

\subsection{Installation}

To install the released version of the package, run:
\begin{CodeInput}
install.packages("visualFields")
\end{CodeInput}

To install the newest version of the package, run:
\begin{CodeInput}
install.packages("devtools")
library(devtools)
install_github("imarinfr/vf1/source")
\end{CodeInput}

To load the package, use:
\begin{CodeInput}
library(visualFields)
\end{CodeInput}

\subsection{What is the visual field?}

The visual field of an eye is the extent of the world that can be seen. It refers to both the central and the peripheral vision. It is measured in degrees of visual angle, with the center of vision being $0^\circ$, points on the temporal and superior regions of the visual field being positive angles, and those in the inferior and nasal being negative. In healthy eyes, the sensitivity to light contrast decreases as the visual angle increases (or as it is commonly said in perimetry, eccentricity). The difference in light sensitivity over the visual field as a function of eccentricity is known as Traquair's \emph{hill of vision} \citep{IMF:Traquair1938}. The figure below shows the hill of vision as rendered by Aulhorn \& Harms \citep{IMF:Aul1967}.

\begin{figure}[t!]
\centering
\includegraphics[width=0.75\textwidth]{AulhornHillOfVision.png}
\caption{\label{fig:hill} Hill of vision as rendered by Aulhorn \& Harms.}
\end{figure}

%% -- Methods ------------------------------------------------------------------

\section{Methods} \label{sec:methods}

\begin{leftbar}
This section can include details about the methods implemented in the \pkg{visualFields} package, such as statistical models, algorithms, or visualization techniques. For now, this section is left as a placeholder.
\end{leftbar}

%% -- Results ------------------------------------------------------------------

\section{Results} \label{sec:results}

\begin{leftbar}
This section can include results obtained using the \pkg{visualFields} package, such as example analyses or visualizations. For now, this section is left as a placeholder.
\end{leftbar}

%% -- Summary and discussion ---------------------------------------------------

\section{Summary and discussion} \label{sec:summary}

The \pkg{visualFields} package provides a comprehensive set of tools for analyzing visual field data. It is part of the Open Perimetry Initiative and supports a wide range of analyses, from basic visualizations to advanced statistical modeling. Future work could include extending the package to support additional perimeters and integrating new methods for visual field analysis.

%% -- Bibliography -------------------------------------------------------------

\bibliography{references}

%% -- Appendix (if any) --------------------------------------------------------
%% - After the bibliography with page break.
%% - With proper section titles and _not_ just "Appendix".

\newpage

\begin{appendix}

\section{More technical details} \label{app:technical}

\begin{leftbar}
Appendices can be included after the bibliography (with a page break). Each
section within the appendix should have a proper section title (rather than
just \emph{Appendix}).

For more technical style details, please check out JSS's style FAQ at
\url{https://www.jstatsoft.org/pages/view/style#frequently-asked-questions}
which includes the following topics:
\begin{itemize}
  \item Title vs.\ sentence case.
  \item Graphics formatting.
  \item Naming conventions.
  \item Turning JSS manuscripts into \proglang{R} package vignettes.
  \item Trouble shooting.
  \item Many other potentially helpful details\dots
\end{itemize}
\end{leftbar}

\section[Using BibTeX]{Using \textsc{Bib}{\TeX}} \label{app:bibtex}

\begin{leftbar}
References need to be provided in a \textsc{Bib}{\TeX} file (\code{.bib}). All
references should be made with \verb|\cite|, \verb|\citet|, \verb|\citep|,
\verb|\citealp| etc.\ (and never hard-coded). This commands yield different
formats of author-year citations and allow to include additional details (e.g.,
pages, chapters, \dots) in brackets. In case you are not familiar with these
commands see the JSS style FAQ for details.

Cleaning up \textsc{Bib}{\TeX} files is a somewhat tedious task -- especially
when acquiring the entries automatically from mixed online sources. However,
it is important that informations are complete and presented in a consistent
style to avoid confusions. JSS requires the following format.
\begin{itemize}
  \item JSS-specific markup (\verb|\proglang|, \verb|\pkg|, \verb|\code|) should
    be used in the references.
  \item Titles should be in title case.
  \item Journal titles should not be abbreviated and in title case.
  \item DOIs should be included where available.
  \item Software should be properly cited as well. For \proglang{R} packages
    \code{citation("pkgname")} typically provides a good starting point.
\end{itemize}
\end{leftbar}

\end{appendix}

%% -----------------------------------------------------------------------------

\end{document}

### 2025-05-09 vignette.Rmd
Start des jstatsoft-Artikels.

--- 2024-08-21 vignette.Rmd
@@
 To install the released version of the package, run 
-```{r eval=FALSE}
+```{r install from CRAN, eval=FALSE}
 install.packages("visualFields")
@@
 To install the newest version of the package, run:
-```{r eval=FALSE}
+```{r install from github, eval=FALSE}
 install.packages("devtools")
@@
 To load the package run
-```{r}
+```{r load package}
 library(visualFields)

### 2025-05-09 vignette.Rmd
Commented code in Vignette because of incompatibility with glr function.

@@
 The *glr* function performs linear regression on global parameters.
-In the following example, the patient loses `r round(abs(glr(getgl(example3), type = "md")$sl), 2)` dB each year and this change is statistically significant, with a $p$-value < `r format(ceiling(1000 * glr(getgl(example3), type = "md")$pval) / 1000, nsmall = 3)`.
-
-```{r}
-
-md_regression <- data.frame(time = glr(getgl(example3), type = "md")$years,
-           md = glr(getgl(example3), type = "md")$data)
-
-globallm <- glr(getgl(example3), type = "md")
-slope <- globallm$sl
-intercept <- globallm$int
-pval <- globallm$pval
-
-ggplot(md_regression, aes(x = time, y =md)) +
-  geom_point() +
-  geom_abline(slope = slope, intercept = intercept)
-
-print(paste("intercept:", intercept))
-print(paste("slope:", slope))
-print(paste("p-value:", pval))
+In the following example, the patient loses `r # round(abs(glr(getgl(example3))$sl), 2)` dB each year and this change is statistically significant, with a $p$-value < `r # format(ceiling(1000 * glr(getgl(example3))$pval) / 1000, nsmall = 3)`.
+
+```{r}
+
+# md_regression <- data.frame(time = glr(getgl(example3))$years,
+#            md = glr(getgl(example3))$data)
+# 
+# globallm <- glr(getgl(example3))
+# slope <- globallm$sl
+# intercept <- globallm$int
+# pval <- globallm$pval
+# 
+# ggplot(md_regression, aes(x = time, y = md)) +
+#   geom_point() +
+#   geom_abline(slope = slope, intercept = intercept)
+# 
+# print(paste("intercept:", intercept))
+# print(paste("slope:", slope))
+# print(paste("p-value:", pval))
 

### 2025-05-09 jss-article-rnw/article.Rnw
Successful weave of JSTAT article.
Started including whole vignette.

@@
   Cord Huchzermeyer\\
-  INSTITUTION\\
-  E-mail: \email{cord@example.com}\\
+  Department of Ophthalmology, University Hospital Erlangen \\
+  E-mail: \email{c.huchzermeyer@posteo.de}\\
   \\
@@
 \begin{document}
+\SweaveOpts{concordance=TRUE}
 
@@
 To install the released version of the package, run:
+\begin{CodeChunk}
 \begin{CodeInput}
@@
 \end{CodeInput}
+\end{CodeChunk}
 
 To install the newest version of the package, run:
+\begin{CodeChunk}
 \begin{CodeInput}
@@
 \end{CodeInput}
+\end{CodeChunk}
 
 To load the package, use:
-\begin{CodeInput}
+<<>>=
 library(visualFields)
-\end{CodeInput}
+@
 
@@
 \end{figure}
+
+\subsection{What is visual field testing?}
+
+Visual field testing measures a patient's visual sensitivity at different locations of the visual field.
+Its purpose is to reconstruct as best as possible within the confines and limitations of a clinical setting the \textbf{hill of vision}.
+The most common type of visual field testing is \textbf{static automated perimetry (SAP)}, which is performed with in a dome-shaped device, so that the distance between the stimulus and the eye equal for all locations within the visual field.
+In contrast, \textbf{campimetry} is performed with a flat screen.
+[CAMPIMETRY MAY BE USED MORE OFTEN IN THE FUTURE BECAUSE COMPUTER SCREENS/ VR DEVICES MAY BE USED FOR TELEMEDICINE; LEAVE IT?]
+
+Usually, perimetry tests the ability to detect increments in stimulus luminance with respect to the background with an achromatic (not colored) stimulus.
+This is a form of perimetry known as \textbf{white-on-white perimetry}.
+However, theoretically, other aspects of vision can be tested at different locations (ability to discriminate dots that are very close to each other, ability to discriminate similar colors, contrast sensitivity).
+The figure below shows the most common grid of locations tested during {SAP}.
+The gray ellipse corresponds to the blind spot of the hill of vision and, therefore, the red dots are expected to have no sensitivity.
+
+<<echo=FALSE, fig=TRUE>>=
+
+library(plotrix)
+
+default_locmap <- getlocmap()
+
+plot(x = default_locmap$coord$x, 
+     y = default_locmap$coord$y, 
+     type = "n", 
+     asp = 1,
+     xlim = c(-30, 30),
+     ylim = c(-34, 34),
+     yaxt = 'n',
+     xaxt = 'n',
+     bty = 'n',
+     xlab = "",
+     ylab = "",
+     main = "24-2 grid of test locations")
+lines(-c(30, 2.5), c(0, 0), lwd = 2, col = "lightgray")
+lines(c(2.5, 30), c(0, 0), lwd = 2, col = "lightgray")
+lines(c(0, 0), -c(30, 2.5), lwd = 2, col = "lightgray")
+lines(c(0, 0), c(2.5, 30), lwd = 2, col = "lightgray")
+lines(x = c(-2.5, 2.5), y = c(0, 0), lwd = 1.5)
+lines(x = c(0, 0), y = c(-2.5, 2.5), lwd = 1.5)
+for(r in c(10, 20, 30))
+  draw.circle(0, 0, r, lwd = 2, col = NA, border = "lightgray")
+draw.ellipse(15, -1.5, 2.75, 3.75, col = "gray", border = NA)
+text(x = 0, y = -34, "horizontal eccentricity [degrees]")
+text(x = -34, y = -0, "vertical eccentricity [degrees]", srt = 90)
+text(x = 31, y = 0, "30\u00b0", adj = c(0, 0.5))
+text(x = 0, y = 31, "30\u00b0", adj = c(0.5, 0))
+points(x = default_locmap$coord$x[-default_locmap$bs],
+       y = default_locmap$coord$y[-default_locmap$bs])
+points(x = default_locmap$coord$x[default_locmap$bs],
+       y = default_locmap$coord$y[default_locmap$bs],
+       col = "red", pch = 19)
+
+@
+
+The visual fields package was created to analyze data from \textbf{SAP}.
+In SAP, the goal is to quantify sensitivity by increasing and decreasing the light differential (or contrast) between stimulus and background and asking the observer to press the button when they detect the presence of the visual stimulus.
+The contrast threshold is determined at each location of the visual field as the one for which the observer is able to detect the presence of the stimulus 50\% of the times.
+Such thresholds are estimated at a fixed number of locations forming a regular grid as the one in the figure above.
+
+There is another common form of perimetry, \textbf{kinetic perimetry}, where a light spot is moved from the periphery towards the center along a number of meridians.
+In this form of perimetry the location where the spot is first detected by the subject is noted and those points at different meridians are connected to find isopters of equal sensitivity on the retina.
+The statistical analysis for this form of perimetry is very different than for static perimetry and beyond the scope of this visualFields package.
+
+\subsection{Performing visual field tests}
+For the typical visual field test, an observer (usually a patient in an ophthalmology or optometry clinic) is comfortably seated in front of a perimeter and puts his head on a chin rest looking into a white Ganzfeld dome (or hemisphere).
+The perimeter is located in a dimly lit, calm room and the observer has been explained what to do by an experienced technician before the examination.
+In his hand, the observer has access to a clicker.
+
+Observers now look at a fixation target (a dot, a cross, a diamond, or other) right in front of them in the center of the dome and are presented visual stimuli over their visual field.
+Observer are to press the clicker any time a stimulus appears anywhere in the dome These lights are presented only for a short time span (typically 100-200ms), so that the observer has no time to shift his gaze to the stimulus.
+A press of the clicker will be linked to a presentation when it occurs within a given time window after the presentation.
+A computer will present stimuli several times at different locations in a grid pattern and in a spatially random order.
+For each location, the stimulus contrast will be decreased if the observer responded to the previous response and increased if not.
+
+At the end of the test, the inverse of the contrast threshold, $\Delta L_{\rm threshold}$ at each location is recorded in a decibel scale (dB).
+This dB scale is normally defined as attenuation from the maximum stimulus possible $\Delta L_{\rm max}$ as:
+\begin{equation}
+dB = -10\ \rm log \left (\frac{\Delta L_{\rm thr}}{\Delta L_{\rm max}} \right)\ .
+\end{equation}
+For the most common devices, the Humprhery Field Analyzer and the Octopus 900, the maximum luminance differential $\Delta L_{\rm max}$ is $10000\ (\rm apostilb) / \pi = 3183.1\ cd\ m^{-2}$.
+
+These are the data expected by the visualFields package for statistical analysis and generation of visualization devices.
+
+\subsection{Psychophysical testing}
+Psychophysical tests systematically measure the relationship between physical stimulus strength and the subjective perception.
+This is performed by varying the stimulus and asking the observer to communicate his perception with a limited number of allowed responses.
+In visual field testing, the observer usually presses a button when a light spot is perceived.
+
+The relationship between stimulus and perception is analyzed using mathematical model.
+Commonly, these models result in an estimate of a *threshold*, which denotes either the absolute stimulus strength that can just barely be perceived or the smallest difference that can be perceived as different.
+
+Real world scenarios where perception is close to threshold include going through a dark forest on a cloudy night or seeing the car in front through a dense fog.
+Stimuli are not clearly seen when close to threshold.
+For this reason, patients sometimes feel that lengthy visual field testing is tedious and that it is as an unreliable examination.
+The latter is not clearly correct when fluctuations are taken into account and data are analyzed with appropriate statistical methods.
+
+\subsection{Grid patterns}
+The spatial position of the points $l_1$ to $l_k$ in the visual field is found in the location map (locmap) and depends on the pattern used by the perimeter.
+When analyzing data, you need to specify the locmap.
+
+By default, the 24-2 pattern (shown in the previous figure) is used, which consists of 54 locations that are equally spaced across the field.
+
+<<>>=
+str(default_locmap)
+@
+
+Beside a name and a description, a locmap dataset contains a coordinate table with the x- and y-coordinates of the locations measured in degrees, and the id of the points that cover the physiological blind spot (the id is the number x in $l_x$ and corresponds to the row in the coord table).
+
+Other grid patterns are also defined in the package. For instance, the 10-2 grid:
+
+<<>>=
+ten_two_locmap <- locmaps$p10d2
+str(ten_two_locmap)
+@
+
+And many others:
+
+<<>>=
+names(locmaps)
+@
+
+Locations of the locmap that cover the blind spot are reported in the locmaps lists under the variable bs.
+If the blind spot is outside the pattern (for example in the 10° visual fields) or is spared by the grid as in the Octopus G1 pattern, the list item bs is set to `r numeric(0)`.
+
+The grid, the stimulus size, presentation time and the algorithm responsible for varying intensities remain constant throughout the examination and are called a perimetry program.
+Not that examinations performed with different programs cannot be directly compared.
+
+\subsection{Measurements
+Although usually not seen in visual field data, the most elementary unit of a static visual field test is a subject's response.
+It consists of the constant stimulus parameters (size, presentation time, background luminance), the stimulus location, the stimulus intensity and whether the subject pressed the button within a predefined time window to indicate that he saw the stimulus.
+
+Below a certain *threshold* luminance, a stimulus is usually not perceivable, above the threshold, the stimulus is usually perceived.
+However, close to threshold, a stimulus is not always reliably perceived.
+By definition, the threshold is the luminance of the stimulus where the subject will see the stimulus with a probability of 50%.
+
+The most straightforward to measure such a threshold would be to present stimuli several time at a number different, fixed stimulus strengths.
+This method is called the method of *constant stimuli* and results in the *psychometric function*, which is usually modeled with a cumulative distribution function (CDF), for example a Gaussian CDF.
+However, this is not the most effective way to estimate a threshold, because stimulus strength far from the threshold will have probabilities to be seen that are close to either 0% or 100%.
+Testing these strengths several times does not yield much information.
+
+Therefore, advanced algorithms were developed for reducing the number of stimulus presentations with limited trade-off concerning the accuracy of the measurements.
+A staircase algorithm where the intensity is decreased by a certain step size when the stimulus was seen, increased when not seen, and the step size is halved upon each revearsal (seen to not seen or vice versa) is a more efficient algorithms that is frequently used.
+More efficient algorithms usually make a trade-off between precision of the threshold estimate and test time, and often incorporate information on responses in neighboring locations into the algorithm.
+Examples include the Swedish Interactive Threshold Algorithm (SITA) of the Humphery Perimeters.
+Note that long test times negatively affect attention and therefore both accuracy and precision of the threshold estimates.
+
+\subsection{False negatives and false positives
+
+Residual deviation from the psychometric function may occur due to inattention of the observer but also due to short-term fluctuation of the threshold.
+For example, it has been shown that the threshold underlies short-term fluctuates in the areas where light sensitivity was negatively affected by glaucoma, a disease of the optic nerve.
+
+When the observer does not see a stimulus that is clearly above a level that has been seen before, this is called a false negative response (fnr).
+In contrast, when the subject presses the button although no stimulus is presented, this is a false positive response (fpr).
+
+The cooperation of the subject is also tested in some perimeters by presenting deliberately presenting a stimulus at the location of the blind spot.
+If the subjects sees that stimulus, it must be assumed that he did not fixate correctly (fixation loss) and that the reliability of the field is decreased.
+
+\subsection{Clinical applications of perimetry}
+In clinical practice, visual field testing is used to identify decreases in visual function caused by disease.
+Such *visual field defects* or *scotomas* can be caused by diseases that affect tissue anywhere along the optic pathway.
+This ranges from the optic media (cornea / lens / vitreous) which transmit light, to the retina, which detects lights and preprocesses visual information, to the optic nerve, which transmits information, and up to the visual cortex.
+
+Often, clinicians can draw reliable conclusions about the localisation of the disease by the pattern of the visual field defects.
+For example, an infarction of the right visual cortex leads complete blindness in the left hemispheres of both eyes with a very distinct border along the vertical meridian.
+Therefore, the patient does not perceive anything to the right of the point he looks at.
+On the other side of the vertical meridian the visual field is normal.
+In contrast, patients with glaucoma start with areas of slightly decreased light sensitivity often in the upper or lower hemifield with a bow-shaped (arcuate) pattern cause by the localisation of damage in the optic nerve head.
+
+Visual field tests can be used to diagnose disease, to appraise the functional consequences of disease (for example: can the patient still drive a vehicle?), and to monitor progression.
+
+[THIS SEEMS TO BE A BIT OUF OF PLACE, BUT I PREFER NOT TO EDIT THIS PART MYSELF]
+The most advanced statistical methods have been created for monitoring glaucoma progression.
+The open-angle glaucomas are a group of slowly progressing, chronic neurodegenerative diseases of the optic nerve, often caused by increased eye pressure.
+Patients need consequent monitoring over many years and slight changes in the visual field must be detected reliably.
+Therefore, many of the functions of the visualField package focus on glaucoma research.
+
+\subsection{Getting started with the package}
+\subsection{Loading data}
+The visualField package contains some example data. This dataset contains the results of 42 visual field tests of one patient (id: sample1). The patient has primary open-angle glaucoma (type: pwg).
+
+<<>>=
+
+str(vfpwgSunyiu24d2[, 1:10])
+
+@
+
+The first columns characterize the patient, the time and duration of the examination, the diagnosis, and the reliability parameters 'false positive responses', 'false negative responses' and 'fixation losses'.
+
+Each row corresponds to one visual field test.
+Visual field testing is performed monocularly (one eye at a time), so that two examinations exist for each date.
+The abbreviations for the laterality ('eye') is derived from latin (OD: right eye = oculus dexter, and OS: left eye = oculus sinister), when both eyes are examined, the abbreviation is OU.
+
+The environment variable locini contains the number of the first column that contains the actual visual field date.
+Each column $l_{x}$ corresponds to one of {k} locations in the visual field test.
+A concise way to select the columns $l_1$ to $l_{k}$ is to use the *getvfcols* function.
+
+<<>>=
+
+getlocini()
+getvfcols()
+
+vfpwgSunyiu24d2[1, getvfcols()]
+
+@
+
+There mustn't be any columns after these columns.
+Use the *vffilter* function to select specific fields and the *vfisvalid* function to check whether a table fulfills the criteria.
+The latter will also tell you what the problem is when it doesn't.
+
+<<>>=
+
+example1 <- vffilter(vfpwgSunyiu24d2, 
+                     id == "sample1" & 
+                       eye == "OD")[1, ]
+
+example1$newVariable <- "not valid"
+
+vfisvalid(example1)
+
+@
+
+\subsection{Plotting the visual field}
+The most straightforward plots show the sensitivities at the corresponding locations.
+These representations are very exact, but may be difficult to interpret.
+For learning purposes, here, we manually plot some visual field data.
+
+<<>>=
+
+setdefaults()
+
+example2 <- vffilter(vfpwgRetest24d2, 
+                     id == "1" & 
+                       eye == "OD")[1,]
+
+vfisvalid(example2)
+
+example2_dataframe <- data.frame(x = getlocmap()$coord$x, 
+                                 y = getlocmap()$coord$y, 
+                                 td = as.vector(unlist(example2[,getvfcols()])))
+
+ggplot(example2_dataframe, aes(x = x, y = y)) +
+  geom_text(aes(label = td)) +
+  geom_hline(yintercept = 0) +
+  geom_vline(xintercept = 0) +
+  coord_fixed() +
+  theme_void()
+
+@
+
+Note that left eyes have to be treated differently if you plot the points manually, because the location labels are mirrored along the y axis.
+In other words, to plot the data manually, for a left eye x should be `-getlocmap()$coord$x`.
+
+By convention, when both eyes are presented side by side, the left eye is shown on the left and the right eye on the right.
+This corresponds to the visual field of the observer, so that the physiological blindspots are directed temporally (toward the temple, away from the nose) for both eyes.
+This is in contrast to presentation of structural data, which is usually shown like it would look in a patient sitting in front of you, where his right eye is on the observers left side.
+
+<<>>=
+
+OD_dataframe <- data.frame(x = getlocmap()$coord$x, 
+                             y = getlocmap()$coord$y, 
+                             loc_label = paste0("l", 1:54),
+                             is_blindspot = FALSE,
+                             eye = "OD")
+
+# Note the minus in the x variable for the left eye
+OS_dataframe <- data.frame(x = -getlocmap()$coord$x, 
+                             y = getlocmap()$coord$y, 
+                             loc_label = paste0("l", 1:54),
+                             is_blindspot = FALSE,
+                             eye = "OS")
+
+OD_dataframe[getlocmap()$bs, "is_blindspot"] <- TRUE
+OS_dataframe[getlocmap()$bs, "is_blindspot"] <- TRUE
+
+ODOS_dataframe <- rbind(OS_dataframe, OD_dataframe)
+
+# Use factor in order to show left eye on the left side
+ODOS_dataframe$eye <- factor(ODOS_dataframe$eye, 
+                             levels = c("OS", "OD"), 
+                             ordered = TRUE)
+
+ggplot(ODOS_dataframe, aes(x = x, y = y)) +
+  coord_fixed() +
+  xlim(-30, 30) +
+  ylim(-24, 24) +
+  geom_label(aes(label = loc_label, color = is_blindspot), size = 2.75) +
+  scale_color_manual(values = c("black", "red")) +
+  facet_wrap(~ eye)
+
+@
+
+The visualField package provides functions for more sophisticated plotting, which will be explained later.
+*These functions automatically take care that right eyes and left eyes are plotted correctly.*
+
+\subsection{Mathematical derivatives of contrast thresholds for statisitcal analysis}
+
+\subsection{Identifying defective locations on the visual field}
+
+As mentioned earlier, the sensitivity in dB at each location is calculated from the contrast threshold relative to the maximal contrast.
+
+The sensitivity to light fluctuates considerably between measurements in the same observer (test-retest variability ---and more as the disease progresses) and between observers (inter-individual differences).
+They also decrease at different rates depending on the eccentricity with age.
+
+With these predictable and unpredictable differences in the pointwise sensitivity measurements, how can we provide tools and reports to aid the clinical assessment of the visual field?
+Or more to the point, how can we discriminate between healthy locations and locations that are damaged or suspected of being damaged?
+Clinically, it is more useful to compute and report the deviations from an age-corrected mean normal visual field, since the identification of spatial patterns of damage becomes easier, thus facilitating the clinical assessment of the visual field.
+
+But this analysis requires the use of a *normative dataset* from which to obtain models describing the age effect on sensitivity at each location.
+These functions are usually linear and those *age linear models* are used to obtain the *age-corrected mean normal visual field*.
+
+The resulting *deviations* from age-corrected mean normal visual field are also measured in dB and they are interpreted as follows.
+By convention, in the Humphrey Field Analyzer, a negative deviation value means that the sensitivity at that location is below mean normal.
+In contrast, on Octopus perimeters, negative is above and positive below mean normal (therefore, values are called *defects*).
+In this package, those differences are called *total deviation* and their interpretation are as for the Humphrey perimeter.
+In case you use visualFields with data from a different perimeter it is important to know what is the definition of *deviation* and change the sign if necessary.
+
+In eye diseases like glaucoma, general defects that affect the whole visual field are often less relevant than focal scotomas.
+And often ocular opacities caused by cataract can mask such focal defects.
+For this reason, a statistical post-hoc correction to *total deviation* values is further applied to obtain *pattern deviation* values.
+The correction factor is often called the *general height* and for the 24-2 Humphrey is obtained by selecting the 7th greatest total deviation value.
+The *pattern deviation* values equal the *total deviation* values minus the *general height*.
+
+\subsection{Normative dataset and normative references}
+The visualFields package contains three datasets of healthy subjects that can be used to obtain normative age linear models and references.
+
+The SUNY-IU dataset [@IMF:Mar2013b] for the 24-2 SITA Standard and the Goldmann Size III:
+<<SUNYIU datasets>>=
+str(vfctrSunyiu24d2[,1:10])
+@
+
+The Iowa dataset [@IMF:Marin2018c] for a bimodal ZEST algorithm and the Goldmann Size V:
+<<Iowa datasets>>=
+str(vfctrIowaPC26[,1:10])
+str(vfctrIowaPeri[,1:10])
+@
+
+From these datasets, or any other dataset of healthy eyes, normative references can be computed as follows:
+
+<<nv>>=
+nv <- nvgenerate(vfctrSunyiu24d2)
+@
+
+For future use, we set `nv` as the normative values to be used in visualFields for the calculation of total and pattern deviation values and other statistical analyses.
+<<setnv>>=
+setnv(nv)
+@
+
+\subsubsection{Age linear models and mean normal visual fields}
+The normative values and references stored in `nv` contains the coefficients of the age linear models:
+
+<<agelm>>=
+print(nv$agem$coeff)
+@
+
+And a function from where the age-corrected mean normal visual fields can be computed:
+<<agemeannormal>>=
+age <- 50
+print(nv$agem$model(age))
+@
+
+Notice that the locations corresponding to the anatomical region of the blind spot all have `NA` as values, as they are excluded from the statistical analysis.
+
+\subsection{Total deviation}
+Total deviation is calculated from the sensitivities and from age-related normal values. These are saved in the environment variable *nv* (normal values).
+
+<<sensitivities>>=
+
+print(example2[, getvfcols()[1:10]])
+
+@
+
+<<total deviation>>=
+
+age <- example2$age[1]
+
+print(age)
+
+normal_values <- getnv()$agem$model(age)[1:10]
+
+print(normal_values)
+
+total_defects <- example2[, getvfcols()[1:10]] - normal_values
+
+print(total_defects)
+
+@
+
+These values can be automatically calculated with a dedicated function.
+
+<<>>=
+
+td <- gettd(example2)
+print(td[, getvfcols()[1:10]])
+
+@
+
+\subsection{Pattern deviation}
+Pattern deviation can be calculated after obtaining the general height.
+<<gh>>=
+gh <- getgh(td)
+print(gh)
+pattern_defects <- td[, getvfcols()[1:10]] - gh
+print(pattern_defects)
+@
+
+These values can be automatically calculated with a dedicated function.
+
+<<pattern deviation>>=
+
+pd <- getpd(td)
+print(pd[, getvfcols()[1:10]])
+
+@
+
+\subsection{Total and Pattern deviation probabilities}
+As mentioned psychophysical sensitivity values vary considerably from test to test and between subjects Therefore, even in perfectly average eyes deviations will be different from zero most of the time.
+Given a visual field from a patient, the clinician has to know if a given deviation is likely to be caused by disease as opposed to normal retest and inter-individual variability.
+
+The function `nvgenerate` computes for the normative dataset, the total deviation and pattern deviation values for healthy eyes and records pointwise empirical probability distribution functions [@IMF:Hei1987b] that can be used to compute the probability that a specific total deviation (or pattern deviation value) value is from a healthy location [@IMF:Hei1987a] [@IMF:Heijl1991].
+Thus, a low probability implies a greater likelihood of the location being damaged.
+Typically (and in our example), the probabilities are categorized as
+<<probability categories>>=
+print(getnv()$lut$probs)
+@
+
+where `0.005` means probability $< 0.005$, where `0.01` means probability $< 0.01$, etc. A probability greater than 0.05 is considered within the normal range.
+
+The probability values for the previous values are as follows.
+
+<<>>= probability maps}
+print(gettdp(td)[, getvfcols()[1:10]])
+print(getpdp(pd)[, getvfcols()[1:10]])
+@
+
+\subsection{Plotting the visual field}
+There are different ways to visualize the perimetric results. These visualizations have to be interpreted like maps in that they represent a simplified version of the hill of vision.
+
+[WE PROBABLY NEED SOME INTRO TO VFPLOT BEFORE TALKING MAKING NOTES, ETC?]
+The standard already contains some additional information with a border presenting a color-coded visualization of the deviation and a graphical representation of the blind spot.
+Note that the vfplot with the option "td" needs information on differential light-sensitivities in a normal reference population and on the probability of given deviations from these normal values.
+These data are included for the typical Humphrey Field Analyzer fields, but you need to collect this information if you use other perimeters
+
+Note that you have to specify what type of data is contained in the visual field tables. The default is set to *total deviation*, `td`.
+
+<<>>=
+
+vfplot(example2)
+
+@
+
+In this graph, yellow represents probabilities < 0.05, orange probabilities < 0.02, red < 0.01, and dark red < 0.005.
+
+It is possible to print sensitivities (with darker representing locations with lower sensitivities)
+
+<<>>=
+
+vfplot(example2, type = "s")
+
+@
+
+Pattern deviation values
+
+<<>>=
+
+vfplot(example2, type = "pd")
+
+@
+
+And an hybrid plot showing total deviation values and probabilities along with a grayscale representing the sensitivities.
+
+<<>>=
+
+vfplot(example2, type = "tds")
+
+@
+
+And a similar hybrid plotbut with pattern deviation values and probabilities.
+
+<<>>=
+
+vfplot(example2, type = "pds")
+
+@
+
+\subsection{Global indices}
+Global indices summarize the visual field in a few parameters and their values and probability level can be calculated with the visualField package.
+The resulting table contains the visual field header data (patient id, eye, etc.) and 8 additional columns (msens, ssens, tmd, tsd, pmd, psd, gh, and vfi).
+
+<<>>=
+
+print(getgl(example2)[,-(1:getlocini()-1)])
+
+@
+
+The column `msens` is the mean sensitivity (MS); the arithmetic mean of all sensitivity values. The column `ssens` is the standard deviation of sensitivity values.
+
+The column `tmd` is the mean deviation (MD); the arithmetic mean of all total deviation values. The column `tsd` is the standard deviation of total deviation values.
+
+The column `pmd` is the mean of pattern deviation values. The column `psd` is the standard pattern deviation (PSD) or the standard deviation of the pattern deviation values.
+
+The last column, `vfi`, is the visual field index (VFI) [@IMF:Ben2008], and the second to last is the general height (GH).
+
+The probability of the observed values given that the visual field is normal can also be calculated.
+
+<<>>=
+
+print(getglp(getgl(example2))[, -(1:getlocini()-1)])
+
+@
+
+\subsection{Putting everything together}
+Single field analysis reports can be created as pdfs or as interactive reports with Shiny. To explore them, type in the R Console the following:
+
+<<>>= eval = FALSE, include = TRUE}
+vfsfa(example2, file = "test.pdf")
+vfsfashiny(example2)
+@
+
+\subsection{Structure function correlations in glaucoma patients}
+In the last years, techniques that provide detailed imaging of the retina and the optic nerve have made incredible advances.
+Therefore, correlation between structure and function is becoming more important than ever.
+
+To aid comparisons between structural insult and functional defect, it is useful to understand and model the anatomy of a typical healthy eye, even if such modeling through is imperfect given the large inter-individual differences in anatomical features [@IMF:Jan2009] [@IMF:Jan2012] [@IMF:Drasdo2007] [@IMF:Montesano2020]
+
+In glaucoma, the damage occurs in the optic nerve at the levels of the nerve fiber bundles, which consist of the axons of the retinal ganglion cells.
+In the retina, these cells form two layers: the ganglion cell layer that contains the cell bodies, and the retinal nerve fiber layer.
+However, the cell bodies of the retinal ganglion cells are displaced from the area where the photoreceptors are located that they receive input from.
+This displacement is relevant mostly in the center of the visual field, so it would greatly affect the structure-function comparisons on grids such as the 10-2.
+The following figure shows the visual field locations and the corresponding ganglion cell soma being stimulated.
+
+<<>>= echo=FALSE}
+
+locmap <- locmaps$p10d2
+gcmap <- vf2gc(locmap$coord)
+par(mfrow = c(1, 2))
+plot(x = locmap$coord$x, 
+     y = locmap$coord$y, 
+     type = "n", 
+     asp = 1,
+     xlim = c(-10, 12),
+     ylim = c(-10, 10),
+     yaxt = 'n',
+     xaxt = 'n',
+     bty = 'n',
+     xlab = "",
+     ylab = "",
+     main = "10-2 grid of test locations")
+lines(-c(10, 0.5), c(0, 0), lwd = 2, col = "lightgray")
+lines(c(0.5, 10), c(0, 0), lwd = 2, col = "lightgray")
+lines(c(0, 0), -c(10, 0.5), lwd = 2, col = "lightgray")
+lines(c(0, 0), c(0.5, 10), lwd = 2, col = "lightgray")
+lines(x = c(-0.5, 0.5), y = c(0, 0), lwd = 1.5)
+lines(x = c(0, 0), y = c(-0.5, 0.5), lwd = 1.5)
+for(r in c(2.5, 5, 7.5, 10))
+  draw.circle(0, 0, r, lwd = 2, col = NA, border = "lightgray")
+text(x = 11, y = 0, "10\u00b0", adj = c(0, 0.5))
+text(x = 0, y = 11, "10\u00b0", adj = c(0.5, 0))
+points(x = locmap$coord$x, y = locmap$coord$y, bg = "white")
+plot(x = locmap$coord$x, 
+     y = locmap$coord$y, 
+     type = "n", 
+     asp = 1,
+     xlim = c(-10, 12),
+     ylim = c(-10, 10),
+     yaxt = 'n',
+     xaxt = 'n',
+     bty = 'n',
+     xlab = "",
+     ylab = "",
+     main = "Corresponding ganglion cells")
+lines(-c(10, 0.5), c(0, 0), lwd = 2, col = "lightgray")
+lines(c(0.5, 10), c(0, 0), lwd = 2, col = "lightgray")
+lines(c(0, 0), -c(10, 0.5), lwd = 2, col = "lightgray")
+lines(c(0, 0), c(0.5, 10), lwd = 2, col = "lightgray")
+lines(x = c(-0.5, 0.5), y = c(0, 0), lwd = 1.5)
+lines(x = c(0, 0), y = c(-0.5, 0.5), lwd = 1.5)
+for(r in c(2.5, 5, 7.5, 10))
+  draw.circle(0, 0, r, lwd = 2, col = NA, border = "lightgray")
+text(x = 11, y = 0, "10\u00b0", adj = c(0, 0.5))
+text(x = 0, y = 11, "10\u00b0", adj = c(0.5, 0))
+points(x = gcmap$x, y = gcmap$y, pch = 21, bg = "white")
+@
+
+Therefore, the visualFields package provides function to calculate where the ganglion cells that subserve a given location in the visual field are located and to calculate at which angle the axons of these ganglion cells enter the optic nerve head (optic disc), where they pass through the sclera (wall of the eye) to form the optic nerve.
+This function is based on Montesano et al.
+[@IMF:Montesano2020] which revisited the original work by Drasdo et al.
+[@IMF:Drasdo2007].
+
+<<>>=
+
+print(locmaps$p24d2$coord[23,])
+print(vf2gc(locmaps$p24d2$coord[23,]))
+
+@
+
+The Montesano Ganglion-cell displacement function can be in conjunction with the Jansonius map describing average ganglion cell bundles paths reaching different regions of the retina [@IMF:Jan2009] [@IMF:Jan2012].
+First, we can obtain the angle of incidence on the optic never head (ONH)
+<<>>=
+print(loc2psi(vf2gc(locmaps$p24d2$coord[23,])))
+@
+
+And from the angle of incidence, the nerve fiber bundle path can be calculated.
+Note that the function for calculating the trajectory will extrapolate the trajectory beyond the location of the visual field (usually to the raphe).
+
+<<>>=, warning=FALSE}
+
+# use loc2psi to calculate the incidence of the nerve fiber corresponding to a given location at the ONH
+angle_of_incidence_l19 <-loc2psi(locmaps$p24d2$coord[19, ])
+
+# use cart2jpolar for calculating the position of the visual field location in the polar coordinate system
+locations <- rbind(cart2jpolar(locmaps$p24d2$coord[19, ]),
+                   cart2jpolar(locmaps$p24d2$coord[23, ]),
+                   cart2jpolar(locmaps$p24d2$coord[47, ]),
+                   cart2jpolar(data.frame(x = 0, y = 0)))
+locations$labels <- c("l19", "l23", "l47", "fixation")
+
+# use bundePath to create a function for calculating the bundle path
+path_function_l19 <- bundlePath(angle_of_incidence_l19)
+path_function_l23 <- bundlePath(loc2psi(locmaps$p24d2$coord[23, ]))
+path_function_l47 <- bundlePath(loc2psi(locmaps$p24d2$coord[47, ]))
+
+bundle_path <- data.frame(y = 4:90,
+                          x_l19 = path_function_l19(4:90),
+                          x_l23 = path_function_l23(4:90),
+                          x_l47 = path_function_l47(4:90))
+ggplot(bundle_path, aes(y = y)) +
+  geom_line(aes(x = x_l19)) +
+  geom_point(data = locations, aes(x = -psi, y = r, color = labels, shape = labels == "fixation")) +
+  geom_line(aes(x = x_l23), color = "red") +
+  geom_line(aes(x = x_l47), color = "blue") +
+  coord_polar(start = pi / 2, direction = -1) +
+  scale_x_continuous(breaks = c(-180, -90, 0, 90, 180), limits = c(-180, 180)) +
+  scale_y_continuous(limits = c(0, 60)) +
+  scale_color_manual(values = c("l19" = "black", "l23" = "red", "l47" = "blue", "fixation" = "black")) +
+  guides(color = "none", shape = "none")
+
+@
+
+In this plot, the origin corresponds to the center of the ONH, the black triangle corresponds to the fovea, and the $180^\circ$ angle corresponds to the line between the ONH center and the fovea, which is, in reality, slightly tilted downward in most patients.
+I DO NOT SEE THIS, BUT I MAY BE MISSING SOMETHING.
+ALSO THE FOVEA (TRIANGLE IF I GOT IT RIGHT?) SHOULD BE A BIT BELOW THE HORIZONTAL MERIDIAN.
+THE POSITION OF THE CENTER OF THE BLIND SPOT IS (15,+2) IN THE MODEL (PLUS AND NOT MINUS BECAUSE THE MODEL IS SPECIFIED FOR THE ANATOMICAL RETINA, AND IT NEEDS TO BE TURNED UPSIDE DOWN TO MATCH THE VISUAL FIELD)
+
+[IVAN: I THOUGHT THAT PSI WAS THE ANGLE OF INCIDENCE IN THE OPTIC NERVE HEAD.
+THEREFORE, I CHOSE A POLAR PLOT AND NOT AN X-Y-PLOT.
+ALSO, I THOUGHT THAT THE 0° ANGLE OF INCIDENCE WAS DEFINED AS THE LINE BETWEEN FOVEA AND ONH.
+I DID NOT TURN THE POLAR COORDINATED ACCORDINGLY EVEN THOUGHT I MENTIONED THAT THE FOVEA WAS "TILTED DOWNWARD".
+BUT I DO NOT HAVE MUCH EXPERIENCE WITH THE MODEL, SO PLEASE CORRECT ME IF I'M WRONG.
+HOW WOULD I CALCULATE XY-COORDINATES?]
+
+For convenience, visualFields also contains a function to map the angle of incidence of bundles on the ONH to the corresponding angle of the circumpapillary retinal nerve fiber layer (cpRNFL) scan.
+This allows to map a visual field location to its corresponding ganglion cell location, to its angle of incidence on the ONH and finally to the angle in the cpRNFL scan.
+
+<<>>=
+
+l23 <- locmaps$p24d2$coord[23,]
+
+# location on the visual field to corresponding ganglion cell
+gc23 <- vf2gc(l23)
+print(gc23)
+
+# ganglion cell location to angle of incidence on the ONH
+psi23 <- loc2psi(gc23)
+print(psi23)
+
+# angle of incidence on the ONH to OCT cpRNFL scan angle
+ang23 <- psi2oct(psi23)
+print(ang23 - 360)
+
+@
+
+\subsection{Longitudinal analyses}
+Monitoring visual function over time in glaucoma patients is an essential clinical application of visual fields in ophthalmology and optometry practices.
+
+For several reasons, measurements will fluctuate over time.
+Fluctuation between measurements will be greater in eyes with glaucoma, especially in areas of glaucoma-related visual field defects.
+Therefore, it is not possible to compare two field tests and claim that the disease has progressed only because the second test yielded lower sensitivities.
+
+Statistical methods and a series with a minimum number of tests are necessary in order to demonstrate disease progression in the visual fields.
+Statistical methods for progression in glaucoma that can be subdivided into *event-based* and *trend-based* methods.
+Furthermore, methods can be subdivided into those that require datasets which describe the "normal fluctuation" in a given population, those that rely on a mathematical model of visual field progression, and those that do not rely on either.
+
+Event-based methods are not implemented in the visualFields package.
+Briefly, they look for progression between the most recent field and a baseline.
+Progression is characterized by a decrease in sensitivity with a magnitude that is unlikely (< 5%) in the population under examination (population based data is necessary) and needs to be confirmed in further tests (usually at least two).
+
+Trend-based methods assume that progression takes place more slowly, so that several field tests are carried out during this time span.
+This concept is unsettling for the clinician who wants to stop progression the first time that he sees the patient, but regularly takes place even in the case of optimal treatment.
+The advantage is that all measurements are taken into account, not only the baseline test and the most recent tests.
+
+There are several tools to analyze progression with visualFields. To review them, we will select a series of 27 visual fields that were obtained over several years.
+
+<<>>=
+example3 <- vffilter(vfpwgSunyiu24d2, 
+                     id == "sample1" & 
+                       eye == "OS")
+@
+
+\subsection{Plotting progression}
+The sparkline plot only shows the raw data: the series of visual field data (by default total deviation values) over time at each location of the visual field.
+The lines are plotted in red if they are very "noisy", i.e., if the median absolut deviation of the residuals from simple linear regression are greater than a certain threshold `thr` (by default 2 dB).
+Here is an example with the threshold set at 4 dB.
+
+<<>>=
+
+vfsparklines(example3, thr = 4)
+
+@
+
+The legoplot is a visualization device designed to compare the first visual fields against the last ones on a series in a sort of "wait-and-see" fashion.
+By default it compares the first 3 visual fields against the last 3 of a series.
+And by default it compares total deviation values (and probabilities of damage, which relies on population based statistics).
+This example, however, shows sensitivities.
+
+<<>>=
+
+vflegoplot(example3, type = "s")
+
+@
+
+The outer polygon shows the grayscale for the average sensitivity over the first 3 visual fields, whereas the inner circles show the grayscale for the average of the last 3.
+The numbers inside the inner circles are the difference in sensitivity between the last 3 and first 3 visual fields.
+
+\subsection{Statistical analysis}
+Statistical analysis can be performed with linear regression of global indices or with pointwise linear regression (one regression line for each location), which results in as many slopes and $p$-values as there are visual field locations (52 for the 24-2 grid after removing the two that fall on the anatomical region of the blind spot).
+
+The *glr* function performs linear regression on global parameters.
+In the following example, the patient loses `r # round(abs(glr(getgl(example3))$sl), 2)` dB each year and this change is statistically significant, with a $p$-value < `r # format(ceiling(1000 * glr(getgl(example3))$pval) / 1000, nsmall = 3)`.
+
+<<>>=
+
+# md_regression <- data.frame(time = glr(getgl(example3))$years,
+#            md = glr(getgl(example3))$data)
+# 
+# globallm <- glr(getgl(example3))
+# slope <- globallm$sl
+# intercept <- globallm$int
+# pval <- globallm$pval
+# 
+# ggplot(md_regression, aes(x = time, y = md)) +
+#   geom_point() +
+#   geom_abline(slope = slope, intercept = intercept)
+# 
+# print(paste("intercept:", intercept))
+# print(paste("slope:", slope))
+# print(paste("p-value:", pval))
+
+@
+
+The `plr` function performs pointwise linear regression (i.e.
+linear regression for each location in the field separately).
+Specific parameters have to be chosen, in order to detect disease progression.
+For example, progression might be defined as at least 3 locations having a slope less than 0 dB / year and a $p$-value < 0.05.
+
+<<>>=
+
+prog1 <- plr(example3)
+str(prog1, max.level = 1)
+
+# show the total number of siginficantly progressions locations
+# (out of the 52 pointwise linear regressions)
+print(sum(prog1$pval[-getlocmap()$bs] < 0.05))
+
+@
+
+The results of `plr` can be displayed using the following:
+
+<<>>=
+vfplotplr(example3)
+@
+In this graphical display, yellow means that the $p$-value for this location is < 0.05, orange that it is < 0.02, red < 0.01, and dark red < 0.005.
+
+Criteria such as the one suggested earlier based on clusters of locations significantly progressing are useful clinically but do not control for specificity.
+An alternative method that ensure that specificity of progression is mantained at 95% or any other desired level is Permutation of Pointwise Linear Regression or PoPLR [@IMF:Ole2012] [@IMF:Marin2021].
+For this method, all 52 $p$-values are combined into a single $S$-statistic, originally introduced by Fisher [@IMF:Fisher1925].
+In order to avoid reliance on population statistics or parametric statistics.
+(Fisher's original work showed that such statistic followed a chi-square distribution but only if the 52 significance tests were independent of each other and the data was Gaussian, two assumptions that are far from reality.) Thus, random permutations of the series of visual fields are generated and, for each, the $S$-statistic computed.
+If enough random permutations are performed, the resulting set of $S$-values represent an empirical distribution for the null hypothesis that there is no progression.
+A combined $p$-value can then be obtained to test this hypothesis against deterioration of the visual field.
+The interpretation of the $p$-value is as with any significance test.
+
+The *poplr* function yields a large list with many intermediate computations.
+The main result is the $S$-value and its $p$-value.
+Note that there is a left tailed (deterioration), a right tailed (improvement) and a two-tailed (visual field not stable) versions in the current implementation.
+
+<<>>=
+
+prog2 <- poplr(example3)
+str(prog2, max.level = 1)
+
+@
+
+An improvement is highly unlikely in this series.
+
+<<>>=
+
+print(prog2$csr)
+print(prog2$csrp)
+
+@
+
+In contrast, a deterioration is very likely (p < 0.001).
+
+<<>>=
+
+prog2$csl
+prog2$cslp
+
+@
+
+In the histogram representing the null distribution, you can see that a $S$-value of `r round(prog2$csl)` was only very rarely found in the random permutations of the visual field series.
+
+<<>>=
+
+hist(prog2$cstats$cslall, xlab = "S-values", xlim = c(0, 200))
+abline(v = prog2$csl, col = "red", lwd = 2)
+
+@
+
+\subsection{Creating pretty reports}
+Progression reports can be created as pdfs or as interactive reports with shiny. To explore them, type in the R Console the following:
+
+<<>>= eval = FALSE, include = TRUE}
+
+vfspa(example3, file = "test.pdf")
+vfspashiny(example3)
+
+@
+
+\subsection{DICOM standard of visual fields}
+DICOM (Digital Imaging and Communications in Medicine) standards for ophthalmic static perimetry are specified in Supplement 146 [CAN YOU PROVIDE A REFERENCE OR SPECIFY SUPPLEMENT IN WHAT?].
+However, there are two functions for loading data generated in the DICOMM format created by the Humphrey Field Analyzer (HFA) by Zeiss.
+
+Currently, the data representation in the visualFields package differs from these standards and data cannot be exported from the package to DICOMM.
+
+\subsection{Using a different perimeter}
+Different manufacturers have different conventions in perimetry.
+For example, clinicians working with Humphrey Field Analyzers (Carl-Zeiss-Meditec) are often used to report loss of function as negative values (e.g., MD: mean *deviation* of $-10$ dB), while those working with Octopus perimeters (Haag-Streit) often report loss as a positive value (e.g., MD: mean *defect* of 10 dB).
+Furthermore, clinicians working with Octopus perimeters often use different grids, even though the 24-2 and 10-2 grids are also available.
+Usually, fields cannot be directly compared because other setting ($L_{max}$) are also different and because different algorithms are used.
+
+The visualField package provided normative data for the Humphrey Field Analyzer 24-2 and Goldmann size III and for Iowa's full field perimetry [@IMF:Marin2018c], which comprises two different grids, Central $26^\circ$ and Periphery (from $30^\circ$ to more than $60^\circ$), with the Goldmann Size V stimulus on the Octopus 900.
+The `loadoctopus` function also imports some normative data from the Octopus perimeter by Haag-Streit, but, as of now, this does not allow full use of the functionality of the visualFields package.
+Currently, there are no other normative references for any other perimeters or grids.
+Research labs can, however, use their suitable normative datasets to generate normative values for analysis with visual fields.
+
+\subsection{Advanced topics}
+\subsection{Setting up the environment}
+A given visual field program has certain properties that remain unchanged during the test.
+In the visualFields packages these are saved in an *environment*.
+After setting this environment, you can only analyse data that is consistent with this environment.
+
+The core elements of the environment that need to be setup are
+
+-   The normative reference values (`nv`)
+-   The test grid (`locmap`)
+-   Graphical parameters used for the different visualization devices (`gpar`)
+-   The column where the visual field measurements start in the tables (`locini`)
+
+The `nv` environment variable also contains functions for calculating defects, probabilities and global parameters that can be accessed via the functions stated above.
+
+visualFields provides predefined test grids `locmaps`, normative values `normvals`, and parameters for graphical display (`gpars`)
+
+<<>>=
+
+print(names(locmaps))
+print(names(normvals))
+print(names(gpars))
+
+@
+
+They should be consistent.
+For instance, `locmaps$p24d2` can only be used with one of the following `normvals`: `sunyiu_24d2_pw`, `sunyiu_24d2`, `sunyiu_24d2_pw_cps` or `sunyiu_24d2_cps`, but not with, e.g., `iowa_PC26_pw`.
+
+Likewise, `locmaps$p24d2` and `normvals$sunyiu_24d2_pw` or `normvals$sunyiu_24d2` can only be used with `gpars$p24d2`, but not the "_csp" version `gpars$p24d2_cps`
+
+<<>>= getters}
+
+# current normal values
+print(getnv()$info)
+str(getnv(), 1)
+
+# current locmap
+str(getlocmap(), 1)
+# current parameters for plotting
+str(getgpar(), 1)
+# position of the column l1 in the dataframe
+print(getlocini())
+
+@
+
+The default environment assumes a static white-on-white perimetry with a Humphrey perimeter, using a 24-2 grid, a SITA standard algorithm, a stimulus size of Goldman Size III.
+The stimulus duration and the $\Delta L_{max}$ are not explicitly stated in the info, because they are generally known and it does not impact the statistical analysis of the visual field.
+
+Note that for switching to another perimetry program, you need to set all environment variables. visualFields has setter functions for this purpose.
+
+<<>>=
+
+# Set locations for Iowa Central 26 grid
+setlocmap(locmaps$pPC26)
+print(getlocmap()$name)
+# Set normative values for Iowa Central 26 grid
+setnv(normvals$iowa_PC26_pw)
+print(getnv()$info$name)
+# Set graphical parameters for Iowa Central 26 grid
+setgpar(gpars$pPC26)
+
+@
+
+If the structure of the dataframe containing visual field data has changed, and the sensitivity locations start at a differen column, this can also be changed with another setter function.
+
+<<>>=
+
+# Set first column with sensitivity values
+setlocini(24)
+print(getlocini())
+
+@
+
+Here is the figure with the test locations of the Iowa central $26^\circ$. Notice that this map does not have any locations falling on the anatomical blind spot.
+
+<<>>=
+
+new_locmap <- getlocmap()
+
+plot(x = new_locmap$coord$x, 
+     y = new_locmap$coord$y, 
+     type = "n", 
+     asp = 1,
+     xlim = c(-30, 30),
+     ylim = c(-34, 34),
+     yaxt = 'n',
+     xaxt = 'n',
+     bty = 'n',
+     xlab = "",
+     ylab = "",
+     main = "Iowa Central 26 grid of test locations")
+lines(-c(30, 2.5), c(0, 0), lwd = 2, col = "lightgray")
+lines(c(2.5, 30), c(0, 0), lwd = 2, col = "lightgray")
+lines(c(0, 0), -c(30, 2.5), lwd = 2, col = "lightgray")
+lines(c(0, 0), c(2.5, 30), lwd = 2, col = "lightgray")
+lines(x = c(-2.5, 2.5), y = c(0, 0), lwd = 1.5)
+lines(x = c(0, 0), y = c(-2.5, 2.5), lwd = 1.5)
+for(r in c(10, 20, 30))
+  draw.circle(0, 0, r, lwd = 2, col = NA, border = "lightgray")
+draw.ellipse(15, -1.5, 2.75, 3.75, col = "gray", border = NA)
+text(x = 0, y = -34, "horizontal eccentricity [degrees]")
+text(x = -34, y = -0, "vertical eccentricity [degrees]", srt = 90)
+text(x = 31, y = 0, "30\u00b0", adj = c(0, 0.5))
+text(x = 0, y = 31, "30\u00b0", adj = c(0.5, 0))
+points(x = new_locmap$coord$x, y = new_locmap$coord$y)
+
+@
+
+\subsubsection{Defaults}
+The environment can be reset to its default values, the 24-2 pattern measured with the SITA algorithm on a Humphrey Field Analyzer, and with a data frame structure where the sensitivity values start at column 11.
+
+<<>>=
+
+setdefaults()
+
+print(getlocini())
+print(getlocmap()$name)
+print(getnv()$info$name)
+print(str(getnv(), 1))
+
+@
+
+\subsection{Customizing visualFields}
+To make full use of the visualField package with other perimeters, grids, or testing algorithms (staircase, full threshold, different implementations of ZEST, etc.), it is necessary to configure visualFields to the custom test.
+
+The first decision to make is the structure of visual field result data.
+In a data frame (and a loadable csv file) the first 4 columns must have the uniquely identifiable key (`id`, `eye`, `date`, `time`).
+To compute *deviations* from age-corrected mean normal, the age must also be included.
+And it is recommended to add the information about the reliability indices of the exam, such as false positive rate (`fpr`), false negative rate (`fnr`), and fixation losses (`fl`), although the last two have recently been found not to be "reliable".
+To design your structure, looking at the datasets included in the visualFields package may be useful.
+
+The first 10 columns of `vfpwgRetest24d2` contains the key, age, and other information.
+
+<<>>=
+str(vfpwgRetest24d2[,1:10])
+@
+
+After the information regarding the test, etc, must come the columns representing the visual field locations, each row containing the sensitivities for a specific exam of the eye of a patient.
+
+<<>>=
+str(vfpwgRetest24d2[,11:21])
+@
+
+The name of the columns with sensitivity values of an exam must be prefixed by "l" and followed by the corresponding conecutive location number.
+So for a patient using the 24-2 grid, the full record is:
+
+<<>>=
+print(vfpwgRetest24d2[1,])
+@
+
+Depending on the structure the visualField environment variable `locini` must be adjusted so that its value is the column number of the first sensitivity value, `l1`.
+This may be done with the setter `setlocini()`.
+
+For a custom grid of location, it is possible to create and set the location map and its graphical parameters.
+visualFields has a function `vfgpar()` to generate graphical parameters for a given grid.
+Here is an example with the Octopus "G1" map, that uses `vfread` to read visualField's ready visual field data.
+
+[I WANTED TO SHOW SOME OF THE FUNCTIONALITY OF THE LOADOCTOPUS FUNCTION, BUT I COULD DOCUMENT THIS ELSEWHERE]
+
+[THE PROBLEM HERE IS THAT NOW PEOPLE WORKING WITH THE OCTOPUS HAVE TO FIND THE COORDINATES ELSEWHERE, AGAIN, BECAUSE THEY DO NOT HAVE ACCESS TO g1.csv. COULD IT BE PROVIDED IN THE PACKAGE?]
+
+<<>>=
+# read the Octopus G1 coordinates and create the locmap
+g1 <- list(coord = read.csv("./g1.csv"), name = "G1", desc = "The G1 location map used by Octopus perimeters", bs = numeric())
+str(g1)
+# set the G1 locmap
+setlocmap(g1)
+# create and set the graphical parameters
+setgpar(vfgpar(g1$coord))
+# read visual fields and plot the sensitivities
+vfplot(vfread("./vfg1.csv"), type = "s")
+@
+
+The function `vfgpar()` that generates the graphical parameters has many input argument to provide as much flexibility as possible regarding the color schemes to use for probability scales, grayscale, etc.
+
+We cannot compute or visualize total deviations, pattern deviations, or their corresponding probability maps, because we do not have normative reference values for this particular test.
+If we had a visualField's ready csv file with a suitable normative dataset, we could read it with `vfread()`, use the previously used function `nvgenerate()`, and set the generated normative values with `setnv()`.
+
+Finally, if we want to go to the default settings, we ran again:
+
+<<>>=
+setdefaults()
+@
+
+\subsection{Acknowledgments}
+The visualField's package would not have existed if it wasn't for the important contribution of many contributors who partially funded the development of the package, and/or contributed datasets, code, and above all insight on visual fields analysis and about perimetry and visual sciences at large.
+Huge thanks to:
+  - William H Swanson
+  - Michael Wall
+  - Andrew Turpin
+  - Paul H Artes
+  - Giovanni Montesano
+  - Mitchell W Dul
 

### 2025-05-09 gallery/gallery.Rmd
Started a gallery with visualField plots.

---
title: visualFields 1.x
author:
  - Cord Huchzermeyer, Department of Ophthalmology, University Hospital Erlangen
  - Iván Marín-Franch, Computational Optometry
date: "`r Sys.Date()`"
output:   
  distill::distill_article:
    theme: styles.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(visualFields)

example1 <- vffilter(vfpwgSunyiu24d2, 
                     id == "sample1" & 
                       eye == "OD")[1, ]

example2 <- vffilter(vfpwgRetest24d2, 
                     id == "1" & 
                       eye == "OD")[1,]

example3 <- vffilter(vfpwgSunyiu24d2, 
                     id == "sample1" & 
                       eye == "OS")

```

# Plotting single fields

[![sensitivity-plot](thumbnails/sensitivity.png)](#sensitivity-plot)
[![Projekt 1](thumbnails/total_deviation.png)](#total-deviation-plot)
[![Projekt 1](thumbnails/pattern_deviation.png)](#pattern-deviation-plot)
[![Projekt 1](thumbnails/td_sens.png)](#hybrid-td-and-sensitivity-plot)
[![Projekt 1](thumbnails/pd_sens.png)](#hybrid-pd-and-sensitivity-plot)

# Plotting progression

[![Projekt 1](thumbnails/spark.png)](#sparkline-plot)
[![Projekt 1](thumbnails/lego.png)](#lego-plot)
[![Projekt 1](thumbnails/plr.png)](#pointwise-linear-regression-plot)

# Sensitivity plot

The sensitivity plot shows the perimetric sensitivities [dB] at the different locations.

These values are plotted directly and encoded in grayscale.

```{r}

example2 <- vffilter(vfpwgRetest24d2, 
                     id == "1" & eye == "OD")[1,]

vfplot(example2, type = "s")

```

This plot works without normal values.

[Back](#plotting-single-fields)

# Total deviation plot

The total deviation plot shows the total deviation, that is the difference between the subjects sensitivity and an age-corrected normal sensitivity. Negative values signify loss of function. The probability that a given deviation occurs by chance in a normal observers is encoded by the color of the borders (yellow: < 0.05, orange: < 0.02, red: < 0.01, and dark red: < 0.005).

```{r}

example2 <- vffilter(vfpwgRetest24d2, 
                     id == "1" & eye == "OD")[1,]

vfplot(example2)

```

*Prerequisite:* A function for age-related normal values and probability tables.

[Back](#plotting-single-fields)

# Pattern deviation plot

The pattern deviation plot shows the difference between the the subject's total deviation and the mean deviation, therefore correcting for global loss of function. Negative values signify loss of function. The probability that a given deviation occurs by chance in a normal observers is encoded by the color of the borders (yellow: < 0.05, orange: < 0.02, red: < 0.01, and dark red: < 0.005).

```{r}

example2 <- vffilter(vfpwgRetest24d2, 
                     id == "1" & eye == "OD")[1,]

example2 <- vffilter(vfpwgRetest24d2, 
                     id == "1" & eye == "OD")[1,]

vfplot(example2, type = "pd")

```

*Prerequisite:* A function for age-related normal values and probability tables.

[Back](#plotting-single-fields)

# Hybrid TD and Sensitivity plot

This plot is a hybrid plot between a TD plot (values and borders) and a sensitivity plot (grayscale).

```{r}

example2 <- vffilter(vfpwgRetest24d2, 
                     id == "1" & eye == "OD")[1,]

vfplot(example2, type = "tds")

```

*Prerequisite:* Same as for TD plot -> function for age-related normal values and probability tables.

[Back](#plotting-single-fields)

# Hybrid PD and Sensitivity plot

This plot is a hybrid plot between a PD plot (values and borders) and a sensitivity plot (grayscale).

```{r}

example2 <- vffilter(vfpwgRetest24d2, 
                     id == "1" & eye == "OD")[1,]

vfplot(example2, type = "pds")

```

*Prerequisite:* Same as for PD plot -> function for age-related normal values and probability tables.

[Back](#plotting-single-fields)

# Sparkline plot

The sparkline plot only shows the raw data: the series of visual field data (by default total deviation values) over time at each location of the visual field. The lines are plotted in red if they are very "noisy", i.e., if the median absolut deviation of the residuals from simple linear regression are greater than a certain threshold `thr` (by default 2 dB). Here is an example with the threshold set at 4 dB.

```{r}

example3 <- vffilter(vfpwgSunyiu24d2, 
                     id == "sample1" & 
                       eye == "OS")

vfsparklines(example3, thr = 4)

```

*Prerequisite:* Same as for TD plot -> function for age-related normal values and probability tables.

[Back](#plotting-single-fields)

# Lego plot

The legoplot is a visualization device designed to compare the first visual fields against the last ones on a series in a sort of "wait-and-see" fashion. By default it compares the first 3 visual fields against the last 3 of a series. And by default it compares total deviation values (and probabilities of damage, which relies on population based statistics). This example, however, shows sensitivities.

The outer polygon shows the grayscale for the average sensitivity over the first 3 visual fields, whereas the inner circles show the grayscale for the average of the last 3. The numbers inside the inner circles are the difference in sensitivity between the last 3 and first 3 visual fields.

```{r}

example3 <- vffilter(vfpwgSunyiu24d2, 
                     id == "sample1" & 
                       eye == "OS")

vflegoplot(example3, type = "s")

```

[Back](#plotting-single-fields)

# Pointwise linear regression plot

```{r}

example3 <- vffilter(vfpwgSunyiu24d2, 
                     id == "sample1" & 
                       eye == "OS")

vfplotplr(example3)

```

[Back](#plotting-single-fields)

### 2025-05-10 jss-article-rnw/article.Rnw
Started creating gallery for different plots.

@@
 
-\author{Cord Huchzermeyer\\INSTITUTION
+\author{Cord Huchzermeyer~\orcidlink{0000-0002-0408-8981}\\Department of Ophthalmology, University Hospital Erlangen
    \And Iván Marín-Franch\\Computational Optometry}
@@
 
+<<echo=FALSE>>=
+
+library(ggplot2)
+library(plotrix)
+
+@
+
 \subsection{What is the visual field?}
@@
 <<echo=FALSE, fig=TRUE>>=
-
-library(plotrix)
 
@@
 \end{equation}
-For the most common devices, the Humprhery Field Analyzer and the Octopus 900, the maximum luminance differential $\Delta L_{\rm max}$ is $10000\ (\rm apostilb) / \pi = 3183.1\ cd\ m^{-2}$.
+For the most common devices, the Humphrey Field Analyzer and the Octopus 900, the maximum luminance differential $\Delta L_{\rm max}$ is $10000\ (\rm apostilb) / \pi = 3183.1\ cd\ m^{-2}$.
 
@@
 The relationship between stimulus and perception is analyzed using mathematical model.
-Commonly, these models result in an estimate of a *threshold*, which denotes either the absolute stimulus strength that can just barely be perceived or the smallest difference that can be perceived as different.
+Commonly, these models result in an estimate of a \textbf{threshold}, which denotes either the absolute stimulus strength that can just barely be perceived or the smallest difference that can be perceived as different.
 
@@
 
-\subsection{Measurements
+\subsection{Measurements}
 Although usually not seen in visual field data, the most elementary unit of a static visual field test is a subject's response.
@@
 
-Below a certain *threshold* luminance, a stimulus is usually not perceivable, above the threshold, the stimulus is usually perceived.
+Below a certain \textbf{threshold} luminance, a stimulus is usually not perceivable, above the threshold, the stimulus is usually perceived.
 However, close to threshold, a stimulus is not always reliably perceived.
-By definition, the threshold is the luminance of the stimulus where the subject will see the stimulus with a probability of 50%.
+By definition, the threshold is the luminance of the stimulus where the subject will see the stimulus with a probability of 50\%.
 
 The most straightforward to measure such a threshold would be to present stimuli several time at a number different, fixed stimulus strengths.
-This method is called the method of *constant stimuli* and results in the *psychometric function*, which is usually modeled with a cumulative distribution function (CDF), for example a Gaussian CDF.
-However, this is not the most effective way to estimate a threshold, because stimulus strength far from the threshold will have probabilities to be seen that are close to either 0% or 100%.
+This method is called the method of \textbf{constant stimuli} and results in the \textbf{psychometric function}, which is usually modeled with a cumulative distribution function (CDF), for example a Gaussian CDF.
+However, this is not the most effective way to estimate a threshold, because stimulus strength far from the threshold will have probabilities to be seen that are close to either 0\% or 100\%.
 Testing these strengths several times does not yield much information.
@@
 
-\subsection{False negatives and false positives
+\subsection{False negatives and false positives}
 
@@
 In clinical practice, visual field testing is used to identify decreases in visual function caused by disease.
-Such *visual field defects* or *scotomas* can be caused by diseases that affect tissue anywhere along the optic pathway.
+Such \textbf{visual field defects} or \textbf{scotomas} can be caused by diseases that affect tissue anywhere along the optic pathway.
 This ranges from the optic media (cornea / lens / vitreous) which transmit light, to the retina, which detects lights and preprocesses visual information, to the optic nerve, which transmits information, and up to the visual cortex.
@@
 Each column $l_{x}$ corresponds to one of {k} locations in the visual field test.
-A concise way to select the columns $l_1$ to $l_{k}$ is to use the *getvfcols* function.
+A concise way to select the columns $l_1$ to $l_{k}$ is to use the \textbf{getvfcols} function.
 
@@
 There mustn't be any columns after these columns.
-Use the *vffilter* function to select specific fields and the *vfisvalid* function to check whether a table fulfills the criteria.
+Use the \textbf{vffilter} function to select specific fields and the \textbf{vfisvalid} function to check whether a table fulfills the criteria.
 The latter will also tell you what the problem is when it doesn't.
@@
 
-<<>>=
+<<fig=TRUE>>=
 
@@
 The visualField package provides functions for more sophisticated plotting, which will be explained later.
-*These functions automatically take care that right eyes and left eyes are plotted correctly.*
+\textbf{These functions automatically take care that right eyes and left eyes are plotted correctly.}
 
@@
 
-But this analysis requires the use of a *normative dataset* from which to obtain models describing the age effect on sensitivity at each location.
-These functions are usually linear and those *age linear models* are used to obtain the *age-corrected mean normal visual field*.
-
-The resulting *deviations* from age-corrected mean normal visual field are also measured in dB and they are interpreted as follows.
+But this analysis requires the use of a \textbf{normative dataset} from which to obtain models describing the age effect on sensitivity at each location.
+These functions are usually linear and those \textbf{age linear models} are used to obtain the \textbf{age-corrected mean normal visual field}.
+
+The resulting \textbf{deviations} from age-corrected mean normal visual field are also measured in dB and they are interpreted as follows.
 By convention, in the Humphrey Field Analyzer, a negative deviation value means that the sensitivity at that location is below mean normal.
-In contrast, on Octopus perimeters, negative is above and positive below mean normal (therefore, values are called *defects*).
-In this package, those differences are called *total deviation* and their interpretation are as for the Humphrey perimeter.
-In case you use visualFields with data from a different perimeter it is important to know what is the definition of *deviation* and change the sign if necessary.
+In contrast, on Octopus perimeters, negative is above and positive below mean normal (therefore, values are called \textbf{defects}).
+In this package, those differences are called \textbf{total deviation} and their interpretation are as for the Humphrey perimeter.
+In case you use visualFields with data from a different perimeter it is important to know what is the definition of \textbf{deviation} and change the sign if necessary.
 
@@
 And often ocular opacities caused by cataract can mask such focal defects.
-For this reason, a statistical post-hoc correction to *total deviation* values is further applied to obtain *pattern deviation* values.
-The correction factor is often called the *general height* and for the 24-2 Humphrey is obtained by selecting the 7th greatest total deviation value.
-The *pattern deviation* values equal the *total deviation* values minus the *general height*.
+For this reason, a statistical post-hoc correction to \textbf{total deviation} values is further applied to obtain \textbf{pattern deviation} values.
+The correction factor is often called the \textbf{general height} and for the 24-2 Humphrey is obtained by selecting the 7th greatest total deviation value.
+The \textbf{pattern deviation} values equal the \textbf{total deviation} values minus the \textbf{general height}.
 
@@
 \subsection{Total deviation}
-Total deviation is calculated from the sensitivities and from age-related normal values. These are saved in the environment variable *nv* (normal values).
+Total deviation is calculated from the sensitivities and from age-related normal values. These are saved in the environment variable \textbf{nv} (normal values).
 
@@
 
-<<>>= probability maps}
+<<probability maps>>=
 print(gettdp(td)[, getvfcols()[1:10]])
@@
 
-Note that you have to specify what type of data is contained in the visual field tables. The default is set to *total deviation*, `td`.
-
-<<>>=
+Note that you have to specify what type of data is contained in the visual field tables. The default is set to \textbf{total deviation}, `td`.
+
+<<fig=TRUE>>=
 
@@
 
-And a similar hybrid plotbut with pattern deviation values and probabilities.
-
-<<>>=
+And a similar hybrid plot, but with pattern deviation values and probabilities.
+
+<<fig=TRUE>>=
 
@@
 
-<<>>= eval = FALSE, include = TRUE}
+<<eval = FALSE>>=
 vfsfa(example2, file = "test.pdf")
@@
 
-<<>>= echo=FALSE}
+<<echo=FALSE, fig=TRUE>>=
 
@@
 
-<<>>=, warning=FALSE}
+<<warning=FALSE, fig=TRUE>>=
 
@@
 Statistical methods and a series with a minimum number of tests are necessary in order to demonstrate disease progression in the visual fields.
-Statistical methods for progression in glaucoma that can be subdivided into *event-based* and *trend-based* methods.
+Statistical methods for progression in glaucoma that can be subdivided into \textbf{event-based} and \textbf{trend-based} methods.
 Furthermore, methods can be subdivided into those that require datasets which describe the "normal fluctuation" in a given population, those that rely on a mathematical model of visual field progression, and those that do not rely on either.
@@
 Briefly, they look for progression between the most recent field and a baseline.
-Progression is characterized by a decrease in sensitivity with a magnitude that is unlikely (< 5%) in the population under examination (population based data is necessary) and needs to be confirmed in further tests (usually at least two).
+Progression is characterized by a decrease in sensitivity with a magnitude that is unlikely (< 5\%) in the population under examination (population based data is necessary) and needs to be confirmed in further tests (usually at least two).
 
@@
 
-The *glr* function performs linear regression on global parameters.
-In the following example, the patient loses `r # round(abs(glr(getgl(example3))$sl), 2)` dB each year and this change is statistically significant, with a $p$-value < `r # format(ceiling(1000 * glr(getgl(example3))$pval) / 1000, nsmall = 3)`.
+The \textbf{glr} function performs linear regression on global parameters.
+In the following example, the patient loses `r round(abs(glr(getgl(example3))$sl), 2)` dB each year and this change is statistically significant, with a $p$-value < `r format(ceiling(1000 * glr(getgl(example3))$pval) / 1000, nsmall = 3)`.
 
@@
 
-<<>>=
+<<fig=TRUE>>=
+
 vfplotplr(example3)
+
 @
@@
 Criteria such as the one suggested earlier based on clusters of locations significantly progressing are useful clinically but do not control for specificity.
-An alternative method that ensure that specificity of progression is mantained at 95% or any other desired level is Permutation of Pointwise Linear Regression or PoPLR [@IMF:Ole2012] [@IMF:Marin2021].
+An alternative method that ensure that specificity of progression is mantained at 95\% or any other desired level is Permutation of Pointwise Linear Regression or PoPLR [@IMF:Ole2012] [@IMF:Marin2021].
 For this method, all 52 $p$-values are combined into a single $S$-statistic, originally introduced by Fisher [@IMF:Fisher1925].
@@
 
-The *poplr* function yields a large list with many intermediate computations.
+The \textbf{poplr} function yields a large list with many intermediate computations.
 The main result is the $S$-value and its $p$-value.
@@
 
-<<>>= eval = FALSE, include = TRUE}
+<<eval = FALSE>>=
 
@@
 Different manufacturers have different conventions in perimetry.
-For example, clinicians working with Humphrey Field Analyzers (Carl-Zeiss-Meditec) are often used to report loss of function as negative values (e.g., MD: mean *deviation* of $-10$ dB), while those working with Octopus perimeters (Haag-Streit) often report loss as a positive value (e.g., MD: mean *defect* of 10 dB).
+For example, clinicians working with Humphrey Field Analyzers (Carl-Zeiss-Meditec) are often used to report loss of function as negative values (e.g., MD: mean \textbf{deviation} of $-10$ dB), while those working with Octopus perimeters (Haag-Streit) often report loss as a positive value (e.g., MD: mean \textbf{defect} of 10 dB).
 Furthermore, clinicians working with Octopus perimeters often use different grids, even though the 24-2 and 10-2 grids are also available.
@@
 A given visual field program has certain properties that remain unchanged during the test.
-In the visualFields packages these are saved in an *environment*.
+In the visualFields packages these are saved in an \textbf{environment}.
 After setting this environment, you can only analyse data that is consistent with this environment.
@@
 
-<<>>= getters}
+<<getters>>=
 
@@
 In a data frame (and a loadable csv file) the first 4 columns must have the uniquely identifiable key (`id`, `eye`, `date`, `time`).
-To compute *deviations* from age-corrected mean normal, the age must also be included.
+To compute \textbf{deviations} from age-corrected mean normal, the age must also be included.
 And it is recommended to add the information about the reliability indices of the exam, such as false positive rate (`fpr`), false negative rate (`fnr`), and fixation losses (`fl`), although the last two have recently been found not to be "reliable".
@@
 
-<<>>=
+<<eval=FALSE>>=
 # read the Octopus G1 coordinates and create the locmap

### 2025-05-10 jss-article-rnw/article.Rnw
Started restructuring article.

@@
 \SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE}
+
 <<preliminaries, echo=FALSE, results=hide>>=
@@
 
-\title{visualFields 1.x: Tools for Analyzing the Field of Vision in \proglang{R}}
-\Plaintitle{visualFields 1.x: Tools for Analyzing the Field of Vision in R}
-\Shorttitle{visualFields 1.x in \proglang{R}}
+\title{visualFields 1.0.7: Tools for Analyzing the Field of Vision in \proglang{R}}
+\Plaintitle{visualFields 1.0.7: Tools for Analyzing the Field of Vision in R}
+\Shorttitle{visualFields 1.0.7 in \proglang{R}}
 
@@
 \begin{document}
-\SweaveOpts{concordance=TRUE}
+%%\SweaveOpts{concordance=TRUE}
 
@@
 It provides a framework for the development and use of innovative methods for visualization, statistical analysis, and clinical interpretation of visual-field loss and its change over time.
-
-\subsection{Installation}
-
-To install the released version of the package, run:
-\begin{CodeChunk}
-\begin{CodeInput}
-install.packages("visualFields")
-\end{CodeInput}
-\end{CodeChunk}
-
-To install the newest version of the package, run:
-\begin{CodeChunk}
-\begin{CodeInput}
-install.packages("devtools")
-library(devtools)
-install_github("imarinfr/vf1/source")
-\end{CodeInput}
-\end{CodeChunk}
-
-To load the package, use:
-<<>>=
-library(visualFields)
-@
-
-<<echo=FALSE>>=
-
-library(ggplot2)
-library(plotrix)
-
-@
 
@@
 <<echo=FALSE, fig=TRUE>>=
+
+library(visualFields)
+library(plotrix)
 
@@
 
-\subsection{Getting started with the package}
+%% -- Methods ------------------------------------------------------------------
+
+\section{Methods} \label{sec:methods}
+
+\subsection{Installation}
+
+The package is available on CRAN under the Apache 2.0 license and can be installed with the following command:
+
+<<eval=FALSE>>=
+
+install.packages("visualFields")
+
+@
+
+To install the newest version of the package, run:
+
+<<eval=FALSE>>=
+
+library(devtools)
+install_github("imarinfr/vf1/source")
+
+@
+
+<<echo=FALSE>>=
+
+library(ggplot2)
+library(plotrix)
+
+@
+
 \subsection{Loading data}
@@
 
-\subsection{Plotting the visual field}
-The most straightforward plots show the sensitivities at the corresponding locations.
-These representations are very exact, but may be difficult to interpret.
-For learning purposes, here, we manually plot some visual field data.
-
-<<fig=TRUE>>=
-
-setdefaults()
-
-example2 <- vffilter(vfpwgRetest24d2, 
-                     id == "1" & 
-                       eye == "OD")[1,]
-
-vfisvalid(example2)
-
-example2_dataframe <- data.frame(x = getlocmap()$coord$x, 
-                                 y = getlocmap()$coord$y, 
-                                 td = as.vector(unlist(example2[,getvfcols()])))
-
-ggplot(example2_dataframe, aes(x = x, y = y)) +
-  geom_text(aes(label = td)) +
-  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = 0) +
-  coord_fixed() +
-  theme_void()
-
-@
-
-Note that left eyes have to be treated differently if you plot the points manually, because the location labels are mirrored along the y axis.
-In other words, to plot the data manually, for a left eye x should be `-getlocmap()$coord$x`.
-
-By convention, when both eyes are presented side by side, the left eye is shown on the left and the right eye on the right.
-This corresponds to the visual field of the observer, so that the physiological blindspots are directed temporally (toward the temple, away from the nose) for both eyes.
-This is in contrast to presentation of structural data, which is usually shown like it would look in a patient sitting in front of you, where his right eye is on the observers left side.
-
-<<fig=TRUE>>=
-
-OD_dataframe <- data.frame(x = getlocmap()$coord$x, 
-                             y = getlocmap()$coord$y, 
-                             loc_label = paste0("l", 1:54),
-                             is_blindspot = FALSE,
-                             eye = "OD")
-
-# Note the minus in the x variable for the left eye
-OS_dataframe <- data.frame(x = -getlocmap()$coord$x, 
-                             y = getlocmap()$coord$y, 
-                             loc_label = paste0("l", 1:54),
-                             is_blindspot = FALSE,
-                             eye = "OS")
-
-OD_dataframe[getlocmap()$bs, "is_blindspot"] <- TRUE
-OS_dataframe[getlocmap()$bs, "is_blindspot"] <- TRUE
-
-ODOS_dataframe <- rbind(OS_dataframe, OD_dataframe)
-
-# Use factor in order to show left eye on the left side
-ODOS_dataframe$eye <- factor(ODOS_dataframe$eye, 
-                             levels = c("OS", "OD"), 
-                             ordered = TRUE)
-
-ggplot(ODOS_dataframe, aes(x = x, y = y)) +
-  coord_fixed() +
-  xlim(-30, 30) +
-  ylim(-24, 24) +
-  geom_label(aes(label = loc_label, color = is_blindspot), size = 2.75) +
-  scale_color_manual(values = c("black", "red")) +
-  facet_wrap(~ eye)
-
-@
-
-The visualField package provides functions for more sophisticated plotting, which will be explained later.
-\textbf{These functions automatically take care that right eyes and left eyes are plotted correctly.}
-
 \subsection{Mathematical derivatives of contrast thresholds for statisitcal analysis}
 
-\subsection{Identifying defective locations on the visual field}
+\subsubsection{Identifying defective locations on the visual field}
 
@@
 
-\subsection{Normative dataset and normative references}
+\subsubsection{Normative dataset and normative references}
 The visualFields package contains three datasets of healthy subjects that can be used to obtain normative age linear models and references.
@@
 
-\subsubsection{Age linear models and mean normal visual fields}
+\subsubsubsection{Age linear models and mean normal visual fields}
 The normative values and references stored in `nv` contains the coefficients of the age linear models:
@@
 
-\subsection{Total deviation}
+\subsubsection{Total deviation}
 Total deviation is calculated from the sensitivities and from age-related normal values. These are saved in the environment variable \textbf{nv} (normal values).
@@
 <<sensitivities>>=
+
+example2 <- vffilter(vfpwgRetest24d2, 
+                     id == "1" & 
+                       eye == "OD")[1,]
 
@@
 
-\subsection{Pattern deviation}
+\subsubsection{Pattern deviation}
 Pattern deviation can be calculated after obtaining the general height.
@@
 
-\subsection{Total and Pattern deviation probabilities}
+\subsubsection{Total and Pattern deviation probabilities}
 As mentioned psychophysical sensitivity values vary considerably from test to test and between subjects Therefore, even in perfectly average eyes deviations will be different from zero most of the time.
@@
 
-\subsection{Plotting the visual field}
-There are different ways to visualize the perimetric results. These visualizations have to be interpreted like maps in that they represent a simplified version of the hill of vision.
-
-[WE PROBABLY NEED SOME INTRO TO VFPLOT BEFORE TALKING MAKING NOTES, ETC?]
-The standard already contains some additional information with a border presenting a color-coded visualization of the deviation and a graphical representation of the blind spot.
-Note that the vfplot with the option "td" needs information on differential light-sensitivities in a normal reference population and on the probability of given deviations from these normal values.
-These data are included for the typical Humphrey Field Analyzer fields, but you need to collect this information if you use other perimeters
-
-Note that you have to specify what type of data is contained in the visual field tables. The default is set to \textbf{total deviation}, `td`.
-
-<<fig=TRUE>>=
-
-vfplot(example2)
-
-@
-
-In this graph, yellow represents probabilities < 0.05, orange probabilities < 0.02, red < 0.01, and dark red < 0.005.
-
-It is possible to print sensitivities (with darker representing locations with lower sensitivities)
-
-<<fig=TRUE>>=
-
-vfplot(example2, type = "s")
-
-@
-
-Pattern deviation values
-
-<<fig=TRUE>>=
-
-vfplot(example2, type = "pd")
-
-@
-
-And an hybrid plot showing total deviation values and probabilities along with a grayscale representing the sensitivities.
-
-<<fig=TRUE>>=
-
-vfplot(example2, type = "tds")
-
-@
-
-And a similar hybrid plot, but with pattern deviation values and probabilities.
-
-<<fig=TRUE>>=
-
-vfplot(example2, type = "pds")
-
-@
-
 \subsection{Global indices}
@@
 
-\subsection{Putting everything together}
-Single field analysis reports can be created as pdfs or as interactive reports with Shiny. To explore them, type in the R Console the following:
-
-<<eval = FALSE>>=
-vfsfa(example2, file = "test.pdf")
-vfsfashiny(example2)
-@
-
-\subsection{Structure function correlations in glaucoma patients}
-In the last years, techniques that provide detailed imaging of the retina and the optic nerve have made incredible advances.
-Therefore, correlation between structure and function is becoming more important than ever.
-
-To aid comparisons between structural insult and functional defect, it is useful to understand and model the anatomy of a typical healthy eye, even if such modeling through is imperfect given the large inter-individual differences in anatomical features [@IMF:Jan2009] [@IMF:Jan2012] [@IMF:Drasdo2007] [@IMF:Montesano2020]
-
-In glaucoma, the damage occurs in the optic nerve at the levels of the nerve fiber bundles, which consist of the axons of the retinal ganglion cells.
-In the retina, these cells form two layers: the ganglion cell layer that contains the cell bodies, and the retinal nerve fiber layer.
-However, the cell bodies of the retinal ganglion cells are displaced from the area where the photoreceptors are located that they receive input from.
-This displacement is relevant mostly in the center of the visual field, so it would greatly affect the structure-function comparisons on grids such as the 10-2.
-The following figure shows the visual field locations and the corresponding ganglion cell soma being stimulated.
-
-<<echo=FALSE, fig=TRUE>>=
-
-locmap <- locmaps$p10d2
-gcmap <- vf2gc(locmap$coord)
-par(mfrow = c(1, 2))
-plot(x = locmap$coord$x, 
-     y = locmap$coord$y, 
-     type = "n", 
-     asp = 1,
-     xlim = c(-10, 12),
-     ylim = c(-10, 10),
-     yaxt = 'n',
-     xaxt = 'n',
-     bty = 'n',
-     xlab = "",
-     ylab = "",
-     main = "10-2 grid of test locations")
-lines(-c(10, 0.5), c(0, 0), lwd = 2, col = "lightgray")
-lines(c(0.5, 10), c(0, 0), lwd = 2, col = "lightgray")
-lines(c(0, 0), -c(10, 0.5), lwd = 2, col = "lightgray")
-lines(c(0, 0), c(0.5, 10), lwd = 2, col = "lightgray")
-lines(x = c(-0.5, 0.5), y = c(0, 0), lwd = 1.5)
-lines(x = c(0, 0), y = c(-0.5, 0.5), lwd = 1.5)
-for(r in c(2.5, 5, 7.5, 10))
-  draw.circle(0, 0, r, lwd = 2, col = NA, border = "lightgray")
-text(x = 11, y = 0, "10\u00b0", adj = c(0, 0.5))
-text(x = 0, y = 11, "10\u00b0", adj = c(0.5, 0))
-points(x = locmap$coord$x, y = locmap$coord$y, bg = "white")
-plot(x = locmap$coord$x, 
-     y = locmap$coord$y, 
-     type = "n", 
-     asp = 1,
-     xlim = c(-10, 12),
-     ylim = c(-10, 10),
-     yaxt = 'n',
-     xaxt = 'n',
-     bty = 'n',
-     xlab = "",
-     ylab = "",
-     main = "Corresponding ganglion cells")
-lines(-c(10, 0.5), c(0, 0), lwd = 2, col = "lightgray")
-lines(c(0.5, 10), c(0, 0), lwd = 2, col = "lightgray")
-lines(c(0, 0), -c(10, 0.5), lwd = 2, col = "lightgray")
-lines(c(0, 0), c(0.5, 10), lwd = 2, col = "lightgray")
-lines(x = c(-0.5, 0.5), y = c(0, 0), lwd = 1.5)
-lines(x = c(0, 0), y = c(-0.5, 0.5), lwd = 1.5)
-for(r in c(2.5, 5, 7.5, 10))
-  draw.circle(0, 0, r, lwd = 2, col = NA, border = "lightgray")
-text(x = 11, y = 0, "10\u00b0", adj = c(0, 0.5))
-text(x = 0, y = 11, "10\u00b0", adj = c(0.5, 0))
-points(x = gcmap$x, y = gcmap$y, pch = 21, bg = "white")
-@
-
-Therefore, the visualFields package provides function to calculate where the ganglion cells that subserve a given location in the visual field are located and to calculate at which angle the axons of these ganglion cells enter the optic nerve head (optic disc), where they pass through the sclera (wall of the eye) to form the optic nerve.
-This function is based on Montesano et al.
-[@IMF:Montesano2020] which revisited the original work by Drasdo et al.
-[@IMF:Drasdo2007].
-
-<<>>=
-
-print(locmaps$p24d2$coord[23,])
-print(vf2gc(locmaps$p24d2$coord[23,]))
-
-@
-
-The Montesano Ganglion-cell displacement function can be in conjunction with the Jansonius map describing average ganglion cell bundles paths reaching different regions of the retina [@IMF:Jan2009] [@IMF:Jan2012].
-First, we can obtain the angle of incidence on the optic never head (ONH)
-<<>>=
-print(loc2psi(vf2gc(locmaps$p24d2$coord[23,])))
-@
-
-And from the angle of incidence, the nerve fiber bundle path can be calculated.
-Note that the function for calculating the trajectory will extrapolate the trajectory beyond the location of the visual field (usually to the raphe).
-
-<<warning=FALSE, fig=TRUE>>=
-
-# use loc2psi to calculate the incidence of the nerve fiber corresponding to a given location at the ONH
-angle_of_incidence_l19 <-loc2psi(locmaps$p24d2$coord[19, ])
-
-# use cart2jpolar for calculating the position of the visual field location in the polar coordinate system
-locations <- rbind(cart2jpolar(locmaps$p24d2$coord[19, ]),
-                   cart2jpolar(locmaps$p24d2$coord[23, ]),
-                   cart2jpolar(locmaps$p24d2$coord[47, ]),
-                   cart2jpolar(data.frame(x = 0, y = 0)))
-locations$labels <- c("l19", "l23", "l47", "fixation")
-
-# use bundePath to create a function for calculating the bundle path
-path_function_l19 <- bundlePath(angle_of_incidence_l19)
-path_function_l23 <- bundlePath(loc2psi(locmaps$p24d2$coord[23, ]))
-path_function_l47 <- bundlePath(loc2psi(locmaps$p24d2$coord[47, ]))
-
-bundle_path <- data.frame(y = 4:90,
-                          x_l19 = path_function_l19(4:90),
-                          x_l23 = path_function_l23(4:90),
-                          x_l47 = path_function_l47(4:90))
-ggplot(bundle_path, aes(y = y)) +
-  geom_line(aes(x = x_l19)) +
-  geom_point(data = locations, aes(x = -psi, y = r, color = labels, shape = labels == "fixation")) +
-  geom_line(aes(x = x_l23), color = "red") +
-  geom_line(aes(x = x_l47), color = "blue") +
-  coord_polar(start = pi / 2, direction = -1) +
-  scale_x_continuous(breaks = c(-180, -90, 0, 90, 180), limits = c(-180, 180)) +
-  scale_y_continuous(limits = c(0, 60)) +
-  scale_color_manual(values = c("l19" = "black", "l23" = "red", "l47" = "blue", "fixation" = "black")) +
-  guides(color = "none", shape = "none")
-
-@
-
-In this plot, the origin corresponds to the center of the ONH, the black triangle corresponds to the fovea, and the $180^\circ$ angle corresponds to the line between the ONH center and the fovea, which is, in reality, slightly tilted downward in most patients.
-I DO NOT SEE THIS, BUT I MAY BE MISSING SOMETHING.
-ALSO THE FOVEA (TRIANGLE IF I GOT IT RIGHT?) SHOULD BE A BIT BELOW THE HORIZONTAL MERIDIAN.
-THE POSITION OF THE CENTER OF THE BLIND SPOT IS (15,+2) IN THE MODEL (PLUS AND NOT MINUS BECAUSE THE MODEL IS SPECIFIED FOR THE ANATOMICAL RETINA, AND IT NEEDS TO BE TURNED UPSIDE DOWN TO MATCH THE VISUAL FIELD)
-
-[IVAN: I THOUGHT THAT PSI WAS THE ANGLE OF INCIDENCE IN THE OPTIC NERVE HEAD.
-THEREFORE, I CHOSE A POLAR PLOT AND NOT AN X-Y-PLOT.
-ALSO, I THOUGHT THAT THE 0° ANGLE OF INCIDENCE WAS DEFINED AS THE LINE BETWEEN FOVEA AND ONH.
-I DID NOT TURN THE POLAR COORDINATED ACCORDINGLY EVEN THOUGHT I MENTIONED THAT THE FOVEA WAS "TILTED DOWNWARD".
-BUT I DO NOT HAVE MUCH EXPERIENCE WITH THE MODEL, SO PLEASE CORRECT ME IF I'M WRONG.
-HOW WOULD I CALCULATE XY-COORDINATES?]
-
-For convenience, visualFields also contains a function to map the angle of incidence of bundles on the ONH to the corresponding angle of the circumpapillary retinal nerve fiber layer (cpRNFL) scan.
-This allows to map a visual field location to its corresponding ganglion cell location, to its angle of incidence on the ONH and finally to the angle in the cpRNFL scan.
-
-<<>>=
-
-l23 <- locmaps$p24d2$coord[23,]
-
-# location on the visual field to corresponding ganglion cell
-gc23 <- vf2gc(l23)
-print(gc23)
-
-# ganglion cell location to angle of incidence on the ONH
-psi23 <- loc2psi(gc23)
-print(psi23)
-
-# angle of incidence on the ONH to OCT cpRNFL scan angle
-ang23 <- psi2oct(psi23)
-print(ang23 - 360)
-
-@
+%% -- Results ------------------------------------------------------------------
+
+\section{Results} \label{sec:results}
+
+\subsection{Plotting the visual field}
+There are different ways to visualize the perimetric results. These visualizations have to be interpreted like maps in that they represent a simplified version of the hill of vision.
+
+The standard already contains some additional information with a border presenting a color-coded visualization of the deviation and a graphical representation of the blind spot.
+Note that the vfplot with the option "td" needs information on differential light-sensitivities in a normal reference population and on the probability of given deviations from these normal values.
+These data are included for the typical Humphrey Field Analyzer fields, but you need to collect this information if you use other perimeters.
+
+By convention, when both eyes are presented side by side, the left eye is shown on the left and the right eye on the right.
+This corresponds to the visual field of the observer, so that the physiological blindspots are directed temporally (toward the temple, away from the nose) for both eyes.
+This is in contrast to presentation of structural data, which is usually shown like it would look in a patient sitting in front of you, where his right eye is on the observers left side.
+
+Note that you have to specify what type of data is contained in the visual field tables. The default is set to \textbf{total deviation}, `td`.
+
+<<fig=TRUE>>=
+
+vfplot(example2)
+
+@
+
+In this graph, yellow represents probabilities < 0.05, orange probabilities < 0.02, red < 0.01, and dark red < 0.005.
+
+A gallery with all available visual field plots is available online at \url{
+https://rpubs.com/huchzi/1307927}.
 
@@
 \subsection{Plotting progression}
-The sparkline plot only shows the raw data: the series of visual field data (by default total deviation values) over time at each location of the visual field.
-The lines are plotted in red if they are very "noisy", i.e., if the median absolut deviation of the residuals from simple linear regression are greater than a certain threshold `thr` (by default 2 dB).
-Here is an example with the threshold set at 4 dB.
-
-<<fig=TRUE>>=
-
-vfsparklines(example3, thr = 4)
-
-@
-
-The legoplot is a visualization device designed to compare the first visual fields against the last ones on a series in a sort of "wait-and-see" fashion.
-By default it compares the first 3 visual fields against the last 3 of a series.
-And by default it compares total deviation values (and probabilities of damage, which relies on population based statistics).
-This example, however, shows sensitivities.
-
-<<fig=TRUE>>=
-
-vflegoplot(example3, type = "s")
-
-@
-
-The outer polygon shows the grayscale for the average sensitivity over the first 3 visual fields, whereas the inner circles show the grayscale for the average of the last 3.
-The numbers inside the inner circles are the difference in sensitivity between the last 3 and first 3 visual fields.
+
+The plot used for showing progression are shown in the gallery, which can be found at: \url{https://rpubs.com/huchzi/1307927}.
 
@@
 abline(v = prog2$csl, col = "red", lwd = 2)
+
+@
+
+\subsection{Putting everything together}
+Single field analysis reports can be created as pdfs or as interactive reports with Shiny. To explore them, type in the R Console the following:
+
+<<eval = FALSE>>=
+vfsfa(example2, file = "test.pdf")
+vfsfashiny(example2)
+@
+
+\subsection{Structure function correlations in glaucoma patients}
+In the last years, techniques that provide detailed imaging of the retina and the optic nerve have made incredible advances.
+Therefore, correlation between structure and function is becoming more important than ever.
+
+To aid comparisons between structural insult and functional defect, it is useful to understand and model the anatomy of a typical healthy eye, even if such modeling through is imperfect given the large inter-individual differences in anatomical features [@IMF:Jan2009] [@IMF:Jan2012] [@IMF:Drasdo2007] [@IMF:Montesano2020]
+
+In glaucoma, the damage occurs in the optic nerve at the levels of the nerve fiber bundles, which consist of the axons of the retinal ganglion cells.
+In the retina, these cells form two layers: the ganglion cell layer that contains the cell bodies, and the retinal nerve fiber layer.
+However, the cell bodies of the retinal ganglion cells are displaced from the area where the photoreceptors are located that they receive input from.
+This displacement is relevant mostly in the center of the visual field, so it would greatly affect the structure-function comparisons on grids such as the 10-2.
+The following figure shows the visual field locations and the corresponding ganglion cell soma being stimulated.
+
+<<>>=
+
+locmap <- locmaps$p10d2
+gcmap <- vf2gc(locmap$coord)
+
+@
+
+<<echo=TRUE, fig=TRUE>>=
+
+plot_grid <- function() {
+  lines(-c(10, 0.5), c(0, 0), lwd = 2, col = "lightgray")
+  lines(c(0.5, 10), c(0, 0), lwd = 2, col = "lightgray")
+  lines(c(0, 0), -c(10, 0.5), lwd = 2, col = "lightgray")
+  lines(c(0, 0), c(0.5, 10), lwd = 2, col = "lightgray")
+  lines(x = c(-0.5, 0.5), y = c(0, 0), lwd = 1.5)
+  lines(x = c(0, 0), y = c(-0.5, 0.5), lwd = 1.5)
+  for(r in c(2.5, 5, 7.5, 10))
+    draw.circle(0, 0, r, lwd = 2, col = NA, border = "lightgray")
+  text(x = 11, y = 0, "10\u00b0", adj = c(0, 0.5))
+  text(x = 0, y = 11, "10\u00b0", adj = c(0.5, 0))
+}
+
+par(mfrow = c(1, 2))
+plot(x = locmap$coord$x, 
+     y = locmap$coord$y, 
+     type = "n", 
+     asp = 1,
+     xlim = c(-10, 12),
+     ylim = c(-10, 10),
+     yaxt = 'n',
+     xaxt = 'n',
+     bty = 'n',
+     xlab = "",
+     ylab = "",
+     main = "10-2 grid of test locations")
+plot_grid()
+points(x = locmap$coord$x, y = locmap$coord$y, bg = "white")
+
+plot(x = locmap$coord$x, 
+     y = locmap$coord$y, 
+     type = "n", 
+     asp = 1,
+     xlim = c(-10, 12),
+     ylim = c(-10, 10),
+     yaxt = 'n',
+     xaxt = 'n',
+     bty = 'n',
+     xlab = "",
+     ylab = "",
+     main = "Corresponding ganglion cells")
+plot_grid()
+points(x = gcmap$x, y = gcmap$y, pch = 21, bg = "white")
+@
+
+Therefore, the visualFields package provides function to calculate where the ganglion cells that subserve a given location in the visual field are located and to calculate at which angle the axons of these ganglion cells enter the optic nerve head (optic disc), where they pass through the sclera (wall of the eye) to form the optic nerve.
+This function is based on Montesano et al.
+[@IMF:Montesano2020] which revisited the original work by Drasdo et al.
+[@IMF:Drasdo2007].
+
+<<>>=
+
+print(locmaps$p24d2$coord[23,])
+print(vf2gc(locmaps$p24d2$coord[23,]))
+
+@
+
+The Montesano Ganglion-cell displacement function can be in conjunction with the Jansonius map describing average ganglion cell bundles paths reaching different regions of the retina [@IMF:Jan2009] [@IMF:Jan2012].
+First, we can obtain the angle of incidence on the optic never head (ONH)
+
+<<>>=
+
+print(loc2psi(vf2gc(locmaps$p24d2$coord[23,])))
+
+@
+
+And from the angle of incidence, the nerve fiber bundle path can be calculated.
+Note that the function for calculating the trajectory will extrapolate the trajectory beyond the location of the visual field (usually to the raphe).
+
+<<warning=FALSE, fig=TRUE>>=
+
+# use loc2psi to calculate the incidence of the nerve fiber corresponding to a given location at the ONH
+angle_of_incidence_l19 <-loc2psi(locmaps$p24d2$coord[19, ])
+
+# use cart2jpolar for calculating the position of the visual field location in the polar coordinate system
+locations <- rbind(cart2jpolar(locmaps$p24d2$coord[19, ]),
+                   cart2jpolar(locmaps$p24d2$coord[23, ]),
+                   cart2jpolar(locmaps$p24d2$coord[47, ]),
+                   cart2jpolar(data.frame(x = 0, y = 0)))
+locations$labels <- c("l19", "l23", "l47", "fixation")
+
+# use bundePath to create a function for calculating the bundle path
+path_function_l19 <- bundlePath(angle_of_incidence_l19)
+path_function_l23 <- bundlePath(loc2psi(locmaps$p24d2$coord[23, ]))
+path_function_l47 <- bundlePath(loc2psi(locmaps$p24d2$coord[47, ]))
+
+bundle_path <- data.frame(y = 4:90,
+                          x_l19 = path_function_l19(4:90),
+                          x_l23 = path_function_l23(4:90),
+                          x_l47 = path_function_l47(4:90))
+ggplot(bundle_path, aes(y = y)) +
+  geom_line(aes(x = x_l19)) +
+  geom_point(data = locations, aes(x = -psi, y = r, color = labels, shape = labels == "fixation")) +
+  geom_line(aes(x = x_l23), color = "red") +
+  geom_line(aes(x = x_l47), color = "blue") +
+  coord_polar(start = pi / 2, direction = -1) +
+  scale_x_continuous(breaks = c(-180, -90, 0, 90, 180), limits = c(-180, 180)) +
+  scale_y_continuous(limits = c(0, 60)) +
+  scale_color_manual(values = c("l19" = "black", "l23" = "red", "l47" = "blue", "fixation" = "black")) +
+  guides(color = "none", shape = "none")
+
+@
+
+In this plot, the origin corresponds to the center of the ONH, the black triangle corresponds to the fovea, and the $180^\circ$ angle corresponds to the line between the ONH center and the fovea, which is, in reality, slightly tilted downward in most patients.
+I DO NOT SEE THIS, BUT I MAY BE MISSING SOMETHING.
+ALSO THE FOVEA (TRIANGLE IF I GOT IT RIGHT?) SHOULD BE A BIT BELOW THE HORIZONTAL MERIDIAN.
+THE POSITION OF THE CENTER OF THE BLIND SPOT IS (15,+2) IN THE MODEL (PLUS AND NOT MINUS BECAUSE THE MODEL IS SPECIFIED FOR THE ANATOMICAL RETINA, AND IT NEEDS TO BE TURNED UPSIDE DOWN TO MATCH THE VISUAL FIELD)
+
+[IVAN: I THOUGHT THAT PSI WAS THE ANGLE OF INCIDENCE IN THE OPTIC NERVE HEAD.
+THEREFORE, I CHOSE A POLAR PLOT AND NOT AN X-Y-PLOT.
+ALSO, I THOUGHT THAT THE 0° ANGLE OF INCIDENCE WAS DEFINED AS THE LINE BETWEEN FOVEA AND ONH.
+I DID NOT TURN THE POLAR COORDINATED ACCORDINGLY EVEN THOUGHT I MENTIONED THAT THE FOVEA WAS "TILTED DOWNWARD".
+BUT I DO NOT HAVE MUCH EXPERIENCE WITH THE MODEL, SO PLEASE CORRECT ME IF I'M WRONG.
+HOW WOULD I CALCULATE XY-COORDINATES?]
+
+For convenience, visualFields also contains a function to map the angle of incidence of bundles on the ONH to the corresponding angle of the circumpapillary retinal nerve fiber layer (cpRNFL) scan.
+This allows to map a visual field location to its corresponding ganglion cell location, to its angle of incidence on the ONH and finally to the angle in the cpRNFL scan.
+
+<<>>=
+
+l23 <- locmaps$p24d2$coord[23,]
+
+# location on the visual field to corresponding ganglion cell
+gc23 <- vf2gc(l23)
+print(gc23)
+
+# ganglion cell location to angle of incidence on the ONH
+psi23 <- loc2psi(gc23)
+print(psi23)
+
+# angle of incidence on the ONH to OCT cpRNFL scan angle
+ang23 <- psi2oct(psi23)
+print(ang23 - 360)
 
@@
 
-\subsection{Acknowledgments}
+%% -- Summary and discussion ---------------------------------------------------
+
+\section{Summary and discussion} \label{sec:summary}
+
+The \pkg{visualFields} package provides a comprehensive set of tools for analyzing visual field data.
+It is part of the Open Perimetry Initiative and supports a wide range of analyses, from basic visualizations to advanced statistical modeling.
+Future work could include extending the package to support additional perimeters and integrating new methods for visual field analysis.
+
+%% -- Acknowledgements ---------------------------------------------------
+
+\section{Acknowledgments}
 The visualField's package would not have existed if it wasn't for the important contribution of many contributors who partially funded the development of the package, and/or contributed datasets, code, and above all insight on visual fields analysis and about perimetry and visual sciences at large.
@@
   - Mitchell W Dul
-
-%% -- Methods ------------------------------------------------------------------
-
-\section{Methods} \label{sec:methods}
-
-\begin{leftbar}
-This section can include details about the methods implemented in the \pkg{visualFields} package, such as statistical models, algorithms, or visualization techniques.
-For now, this section is left as a placeholder.
-\end{leftbar}
-
-%% -- Results ------------------------------------------------------------------
-
-\section{Results} \label{sec:results}
-
-\begin{leftbar}
-This section can include results obtained using the \pkg{visualFields} package, such as example analyses or visualizations. For now, this section is left as a placeholder.
-\end{leftbar}
-
-%% -- Summary and discussion ---------------------------------------------------
-
-\section{Summary and discussion} \label{sec:summary}
-
-The \pkg{visualFields} package provides a comprehensive set of tools for analyzing visual field data.
-It is part of the Open Perimetry Initiative and supports a wide range of analyses, from basic visualizations to advanced statistical modeling.
-Future work could include extending the package to support additional perimeters and integrating new methods for visual field analysis.
 

### 2025-05-11 jss-article-rnw/article.Rnw
Started shortening Introduction.

@@
 The \pkg{visualFields} package, part of the OPI, is a collection of tools for analyzing the field of vision.
-It provides a framework for the development and use of innovative methods for visualization, statistical analysis, and clinical interpretation of visual-field loss and its change over time.
+It provides a framework for the development and use of innovative methods for statistical analysis, visualization, and clinical interpretation of visual-field loss and its change over time.
 
@@
 
-The visual field of an eye is the extent of the world that can be seen.
-It refers to both the central and the peripheral vision.
+The visual field is the extent of the world that can be seen and refers to both the central and the peripheral vision.
 It is measured in degrees of visual angle, with the center of vision being $0^\circ$, points on the temporal and superior regions of the visual field being positive angles, and those in the inferior and nasal being negative.
-In healthy eyes, the sensitivity to light contrast decreases as the visual angle increases (or as it is commonly said in perimetry, eccentricity).
+
+In healthy eyes, the sensitivity to light contrast decreases with eccentricity (i.e.
+increasing visual angle.
 The difference in light sensitivity over the visual field as a function of eccentricity is known as Traquair's \emph{hill of vision} \citep{IMF:Traquair1938}.
-The figure below shows the hill of vision as rendered by Aulhorn \& Harms \citep{IMF:Aul1967}.
+\ref{hill} below shows the hill of vision as rendered by Aulhorn \& Harms \citep{IMF:Aul1967}.
 
@@
 
-\subsection{What is visual field testing?}
-
-Visual field testing measures a patient's visual sensitivity at different locations of the visual field.
-Its purpose is to reconstruct as best as possible within the confines and limitations of a clinical setting the \textbf{hill of vision}.
+\subsection{What is perimetry?}
+
+Perimetry (visual field testing) measures a patient's sensitivity to light differences at different locations of the visual field.
+Thus, it is a psychometric way of quantifying the visual field.
+Psychophysical tests systematically measure the relationship between physical stimulus strength and the subjective perception.
+This is performed by varying the stimulus and asking the observer to communicate his perception with a limited number of allowed responses.
+In visual field testing, the observer usually presses a button when a light spot is perceived.
+The relationship between stimulus and perception is analyzed using mathematical model.
+Commonly, these models result in an estimate of a \textbf{threshold}, which denotes either the absolute stimulus strength that can just barely be perceived or the smallest difference that can be perceived as different.
+
+Its purpose is to estimate the \textbf{hill of vision} as best as possible within the confines and limitations of a clinical setting.
 The most common type of visual field testing is \textbf{static automated perimetry (SAP)}, which is performed with in a dome-shaped device, so that the distance between the stimulus and the eye equal for all locations within the visual field.
 In contrast, \textbf{campimetry} is performed with a flat screen.
-[CAMPIMETRY MAY BE USED MORE OFTEN IN THE FUTURE BECAUSE COMPUTER SCREENS/ VR DEVICES MAY BE USED FOR TELEMEDICINE; LEAVE IT?]
-
-Usually, perimetry tests the ability to detect increments in stimulus luminance with respect to the background with an achromatic (not colored) stimulus.
-This is a form of perimetry known as \textbf{white-on-white perimetry}.
-However, theoretically, other aspects of vision can be tested at different locations (ability to discriminate dots that are very close to each other, ability to discriminate similar colors, contrast sensitivity).
+Campimetry may be used more often in the future, because computer screens and virtual reality devices may be used for telemedicine.
+
+The visual fields package was created to analyze data from standard automated perimetry\textbf{SAP}, where the light sensitivity is measured at fixed, discrete locations in the visual field.
+Kinetic perimetry, where locations of isosensitivity to fixed, discrete luminance increments, is beyond the scope of the visualFields package.
+
+In SAP, the goal is to quantify sensitivity by increasing and decreasing the light differential (or contrast) between stimulus and background and asking the observer to press the button when they detect the presence of the visual stimulus.
+The contrast threshold is determined at each location of the visual field as the one for which the observer is able to detect the presence of the stimulus 50\% of the times.
+Such thresholds are estimated at a fixed number of locations forming a regular grid as the one in the figure above.
+
+For the typical SAP test, an observer (usually a patient in an ophthalmology or optometry clinic) is comfortably seated in front of a perimeter and puts his head on a chin rest looking into a white Ganzfeld dome (or hemisphere).
+The perimeter is located in a dimly lit, calm room and the observer has been instructed by an experienced technician before the examination.
+During the test, the observer looks at a fixation target (a dot, a cross, a diamond, or other) in the center of the dome.
+The observer presses a button in his hand any time a stimulus appears anywhere in the dome.
+These stimuli are presented only for a short time span (typically 100-200ms), so that the observer has no time to shift his gaze to the stimulus.
+A press of the button will only be linked to a presentation when it occurs within a given time window after the presentation.
+A computer will present stimuli several times at different locations in a grid pattern and in a spatially random order.
+For each location, the stimulus contrast will be decreased if the observer responded to the previous response and increased if not.
+
 The figure below shows the most common grid of locations tested during {SAP}.
@@
 
-The visual fields package was created to analyze data from \textbf{SAP}.
-In SAP, the goal is to quantify sensitivity by increasing and decreasing the light differential (or contrast) between stimulus and background and asking the observer to press the button when they detect the presence of the visual stimulus.
-The contrast threshold is determined at each location of the visual field as the one for which the observer is able to detect the presence of the stimulus 50\% of the times.
-Such thresholds are estimated at a fixed number of locations forming a regular grid as the one in the figure above.
-
-There is another common form of perimetry, \textbf{kinetic perimetry}, where a light spot is moved from the periphery towards the center along a number of meridians.
-In this form of perimetry the location where the spot is first detected by the subject is noted and those points at different meridians are connected to find isopters of equal sensitivity on the retina.
-The statistical analysis for this form of perimetry is very different than for static perimetry and beyond the scope of this visualFields package.
-
-\subsection{Performing visual field tests}
-For the typical visual field test, an observer (usually a patient in an ophthalmology or optometry clinic) is comfortably seated in front of a perimeter and puts his head on a chin rest looking into a white Ganzfeld dome (or hemisphere).
-The perimeter is located in a dimly lit, calm room and the observer has been explained what to do by an experienced technician before the examination.
-In his hand, the observer has access to a clicker.
-
-Observers now look at a fixation target (a dot, a cross, a diamond, or other) right in front of them in the center of the dome and are presented visual stimuli over their visual field.
-Observer are to press the clicker any time a stimulus appears anywhere in the dome These lights are presented only for a short time span (typically 100-200ms), so that the observer has no time to shift his gaze to the stimulus.
-A press of the clicker will be linked to a presentation when it occurs within a given time window after the presentation.
-A computer will present stimuli several times at different locations in a grid pattern and in a spatially random order.
-For each location, the stimulus contrast will be decreased if the observer responded to the previous response and increased if not.
-
 At the end of the test, the inverse of the contrast threshold, $\Delta L_{\rm threshold}$ at each location is recorded in a decibel scale (dB).
@@
 These are the data expected by the visualFields package for statistical analysis and generation of visualization devices.
-
-\subsection{Psychophysical testing}
-Psychophysical tests systematically measure the relationship between physical stimulus strength and the subjective perception.
-This is performed by varying the stimulus and asking the observer to communicate his perception with a limited number of allowed responses.
-In visual field testing, the observer usually presses a button when a light spot is perceived.
-
-The relationship between stimulus and perception is analyzed using mathematical model.
-Commonly, these models result in an estimate of a \textbf{threshold}, which denotes either the absolute stimulus strength that can just barely be perceived or the smallest difference that can be perceived as different.
-
-Real world scenarios where perception is close to threshold include going through a dark forest on a cloudy night or seeing the car in front through a dense fog.
-Stimuli are not clearly seen when close to threshold.
-For this reason, patients sometimes feel that lengthy visual field testing is tedious and that it is as an unreliable examination.
-The latter is not clearly correct when fluctuations are taken into account and data are analyzed with appropriate statistical methods.
 

### 2025-05-11 gallery/createThumbnails.Rmd

--- 2025-05-11 vignette.Rmd
@@
 author:
-  - Cord Huchzermeyer, INSTITUTION
+  - Cord Huchzermeyer, Department of Ophthalmology, University Hospital Erlangen
   - Iván Marín-Franch, Computational Optometry
 date: "`r Sys.Date()`"
-output: html_document
-bibliography: references.bib
-biblio-style: "apalike"
-link-citations: true
+output:   
+  distill::distill_article:
+    theme: styles.css
 ---
@@
 knitr::opts_chunk$set(echo = TRUE)
-```
 
-# The Open Perimetry Initiative (OPI)
-The OPI [@IMF:Tur2012] [@IMF:Marin2022a] is a standard for interfacing with visual field testing machines (perimeters).
-It specifies basic functions that allow many visual field tests to be constructed.
-As of October 2017 it is fully implemented on the Octopus 900 and partially on the Heidelberg Edge Perimeter, the Kowa AP 7000, the CrewT IMO and the Centervue Compass.
-
-The visualFields package, part of the OPI, is a collection of tools for analyzing the field of vision.
-It provides a framework for the development and use of innovative methods for visualization, statistical analysis, and clinical interpretation of visual-field loss and its change over time.
-
-To install the released version of the package, run 
-```{r install from CRAN, eval=FALSE}
-install.packages("visualFields")
-```
-
-To install the newest version of the package, run:
-```{r install from github, eval=FALSE}
-install.packages("devtools")
-library(devtools)
-install_github("imarinfr/vf1/source")
-```
-
-To load the package run
-```{r load package}
 library(visualFields)
-```
-
-```{r echo=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
 library(ggplot2)
 library(plotrix)
-```
-
-# What is the visual field?
-The visual field of an eye is the extent of the world that can be seen.
-It refers to both the central and the peripheral vision.
-It is measured in degrees of visual angle, with the center of vision being $0^\circ$, points on the temporal and superior regions of the visual field being positive angles and those in the inferior and nasal being negative.
-In healthy eyes, the sensitivity to light contrast decreases as the visual angle increases (or as it is commonly said in perimetry, eccentricity).
-The difference in light sensitivity over the visual field as a function of eccentricity is known as the Traquair's *hill of vision* [@IMF:Traquair1938].
-The figure below shows the hill of vision as rendered by Aulhorn & Harms [@IMF:Aul1967].
-Notice the discontinuity in the hill of vision in the lower right panel.
-Sensitivity in this region, the *blind spot*, centered at approximately $(15^\circ, -3^\circ)$ in the visual field corresponds to the optic nerve head [THE BLIND SPOT IS NOT AN ANATOMICAL TERM, BUT A PERIMETRIC ONE], where the optical fiber exits from the eye towards the visual cortex.
-
-<figure>
-  <center><img src="./AulhornHillOfVision.png" alt="Hill of Vision" width="75%"></center>
-  <center><figcaption>Aulhorn and Harms' rendering of Traquair's hill of vision.</figcaption></center>
-</figure>
-
-# What is visual field testing?
-Visual field testing measures a patient's visual sensitivity at different locations of the visual field.
-Its purpose is to reconstruct as best as possible within the confines and limitations of a clinical setting the *hill of vision*.
-The most common type of visual field testing is *static automated perimetry (SAP)*, which is performed with in a dome-shaped device, so that the distance between the stimulus and the eye equal for all locations within the visual field.
-In contrast, *campimetry* is performed with a flat screen.
-[CAMPIMETRY MAY BE USED MORE OFTEN IN THE FUTURE BECAUSE COMPUTER SCREENS/ VR DEVICES MAY BE USED FOR TELEMEDICINE; LEAVE IT?]
-
-Usually, perimetry tests the ability to detect increments in stimulus luminance with respect to the background with an achromatic (not colored) stimulus.
-This is a form of perimetry known as *white-on-white perimetry*.
-However, theoretically, other aspects of vision can be tested at different locations (ability to discriminate dots that are very close to each other, ability to discriminate similar colors, contrast sensitivity).
-The figure below shows the most common grid of locations tested during *SAP*.
-The gray ellipse corresponds to the blind spot of the hill of vision and, therefore, the red dots are expected to have no sensitivity.
-
-```{r echo=FALSE}
-
-default_locmap <- getlocmap()
-
-plot(x = default_locmap$coord$x, 
-     y = default_locmap$coord$y, 
-     type = "n", 
-     asp = 1,
-     xlim = c(-30, 30),
-     ylim = c(-34, 34),
-     yaxt = 'n',
-     xaxt = 'n',
-     bty = 'n',
-     xlab = "",
-     ylab = "",
-     main = "24-2 grid of test locations")
-lines(-c(30, 2.5), c(0, 0), lwd = 2, col = "lightgray")
-lines(c(2.5, 30), c(0, 0), lwd = 2, col = "lightgray")
-lines(c(0, 0), -c(30, 2.5), lwd = 2, col = "lightgray")
-lines(c(0, 0), c(2.5, 30), lwd = 2, col = "lightgray")
-lines(x = c(-2.5, 2.5), y = c(0, 0), lwd = 1.5)
-lines(x = c(0, 0), y = c(-2.5, 2.5), lwd = 1.5)
-for(r in c(10, 20, 30))
-  draw.circle(0, 0, r, lwd = 2, col = NA, border = "lightgray")
-draw.ellipse(15, -1.5, 2.75, 3.75, col = "gray", border = NA)
-text(x = 0, y = -34, "horizontal eccentricity [degrees]")
-text(x = -34, y = -0, "vertical eccentricity [degrees]", srt = 90)
-text(x = 31, y = 0, "30\u00b0", adj = c(0, 0.5))
-text(x = 0, y = 31, "30\u00b0", adj = c(0.5, 0))
-points(x = default_locmap$coord$x[-default_locmap$bs],
-       y = default_locmap$coord$y[-default_locmap$bs])
-points(x = default_locmap$coord$x[default_locmap$bs],
-       y = default_locmap$coord$y[default_locmap$bs],
-       col = "red", pch = 19)
-
-```
-
-The visual fields package was created to analyze data from *SAP*.
-In SAP, the goal is to quantify sensitivity by increasing and decreasing the light differential (or contrast) between stimulus and background and asking the observer to press the button when they detect the presence of the visual stimulus.
-The contrast threshold is determined at each location of the visual field as the one for which the observer is able to detect the presence of the stimulus 50% of the times.
-Such thresholds are estimated at a fixed number of locations forming a regular grid as the one in the figure above.
-
-There is another common form of perimetry, *kinetic perimetry*, where a light spot is moved from the periphery towards the center along a number of meridians.
-In this form of perimetry the location where the spot is first detected by the subject is noted and those points at different meridians are connected to find isopters of equal sensitivity on the retina.
-The statistical analysis for this form of perimetry is very different than for static perimetry and beyond the scope of this visualFields package.
-
-# Performing visual field tests
-For the typical visual field test, an observer (usually a patient in an ophthalmology or optometry clinic) is comfortably seated in front of a perimeter and puts his head on a chin rest looking into a white Ganzfeld dome (or hemisphere).
-The perimeter is located in a dimly lit, calm room and the observer has been explained what to do by an experienced technician before the examination.
-In his hand, the observer has access to a clicker.
-
-Observers now look at a fixation target (a dot, a cross, a diamond, or other) right in front of them in the center of the dome and are presented visual stimuli over their visual field.
-Observer are to press the clicker any time a stimulus appears anywhere in the dome These lights are presented only for a short time span (typically 100-200ms), so that the observer has no time to shift his gaze to the stimulus.
-A press of the clicker will be linked to a presentation when it occurs within a given time window after the presentation.
-A computer will present stimuli several times at different locations in a grid pattern and in a spatially random order.
-For each location, the stimulus contrast will be decreased if the observer responded to the previous response and increased if not.
-
-At the end of the test, the inverse of the contrast threshold, $\Delta L_{\rm threshold}$ at each location is recorded in a decibel scale (dB).
-This dB scale is normally defined as attenuation from the maximum stimulus possible $\Delta L_{\rm max}$ as:
-\begin{equation}
-dB = -10\ \rm log \left (\frac{\Delta L_{\rm thr}}{\Delta L_{\rm max}} \right)\ .
-\end{equation}
-For the most common devices, the Humprhery Field Analyzer and the Octopus 900, the maximum luminance differential $\Delta L_{\rm max}$ is $10000\ (\rm apostilb) / \pi = 3183.1\ cd\ m^{-2}$.
-
-These are the data expected by the visualFields package for statistical analysis and generation of visualization devices.
-
-## Psychophysical testing
-Psychophysical tests systematically measure the relationship between physical stimulus strength and the subjective perception.
-This is performed by varying the stimulus and asking the observer to communicate his perception with a limited number of allowed responses.
-In visual field testing, the observer usually presses a button when a light spot is perceived.
-
-The relationship between stimulus and perception is analyzed using mathematical model.
-Commonly, these models result in an estimate of a *threshold*, which denotes either the absolute stimulus strength that can just barely be perceived or the smallest difference that can be perceived as different.
-
-Real world scenarios where perception is close to threshold include going through a dark forest on a cloudy night or seeing the car in front through a dense fog.
-Stimuli are not clearly seen when close to threshold.
-For this reason, patients sometimes feel that lengthy visual field testing is tedious and that it is as an unreliable examination.
-The latter is not clearly correct when fluctuations are taken into account and data are analyzed with appropriate statistical methods.
-
-## Grid patterns
-The spatial position of the points $l_1$ to $l_k$ in the visual field is found in the location map (locmap) and depends on the pattern used by the perimeter.
-When analyzing data, you need to specify the locmap.
-
-By default, the 24-2 pattern (shown in the previous figure) is used, which consists of 54 locations that are equally spaced across the field.
-
-```{r}
-str(default_locmap)
-```
-
-Beside a name and a description, a locmap dataset contains a coordinate table with the x- and y-coordinates of the locations measured in degrees, and the id of the points that cover the physiological blind spot (the id is the number x in $l_x$ and corresponds to the row in the coord table).
-
-Other grid patterns are also defined in the package. For instance, the 10-2 grid:
-
-```{r}
-ten_two_locmap <- locmaps$p10d2
-str(ten_two_locmap)
-```
-
-And many others:
-
-```{r}
-names(locmaps)
-```
-
-Locations of the locmap that cover the blind spot are reported in the locmaps lists under the variable bs.
-If the blind spot is outside the pattern (for example in the 10° visual fields) or is spared by the grid as in the Octopus G1 pattern, the list item bs is set to `r numeric(0)`.
-
-The grid, the stimulus size, presentation time and the algorithm responsible for varying intensities remain constant throughout the examination and are called a perimetry program.
-Not that examinations performed with different programs cannot be directly compared.
-
-## Measurements
-Although usually not seen in visual field data, the most elementary unit of a static visual field test is a subject's response.
-It consists of the constant stimulus parameters (size, presentation time, background luminance), the stimulus location, the stimulus intensity and whether the subject pressed the button within a predefined time window to indicate that he saw the stimulus.
-
-Below a certain *threshold* luminance, a stimulus is usually not perceivable, above the threshold, the stimulus is usually perceived.
-However, close to threshold, a stimulus is not always reliably perceived.
-By definition, the threshold is the luminance of the stimulus where the subject will see the stimulus with a probability of 50%.
-
-The most straightforward to measure such a threshold would be to present stimuli several time at a number different, fixed stimulus strengths.
-This method is called the method of *constant stimuli* and results in the *psychometric function*, which is usually modeled with a cumulative distribution function (CDF), for example a Gaussian CDF.
-However, this is not the most effective way to estimate a threshold, because stimulus strength far from the threshold will have probabilities to be seen that are close to either 0% or 100%.
-Testing these strengths several times does not yield much information.
-
-Therefore, advanced algorithms were developed for reducing the number of stimulus presentations with limited trade-off concerning the accuracy of the measurements.
-A staircase algorithm where the intensity is decreased by a certain step size when the stimulus was seen, increased when not seen, and the step size is halved upon each revearsal (seen to not seen or vice versa) is a more efficient algorithms that is frequently used.
-More efficient algorithms usually make a trade-off between precision of the threshold estimate and test time, and often incorporate information on responses in neighboring locations into the algorithm.
-Examples include the Swedish Interactive Threshold Algorithm (SITA) of the Humphery Perimeters.
-Note that long test times negatively affect attention and therefore both accuracy and precision of the threshold estimates.
-
-## False negatives and false positives
-
-Residual deviation from the psychometric function may occur due to inattention of the observer but also due to short-term fluctuation of the threshold.
-For example, it has been shown that the threshold underlies short-term fluctuates in the areas where light sensitivity was negatively affected by glaucoma, a disease of the optic nerve.
-
-When the observer does not see a stimulus that is clearly above a level that has been seen before, this is called a false negative response (fnr).
-In contrast, when the subject presses the button although no stimulus is presented, this is a false positive response (fpr).
-
-The cooperation of the subject is also tested in some perimeters by presenting deliberately presenting a stimulus at the location of the blind spot.
-If the subjects sees that stimulus, it must be assumed that he did not fixate correctly (fixation loss) and that the reliability of the field is decreased.
-
-# Clinical applications of perimetry
-In clinical practice, visual field testing is used to identify decreases in visual function caused by disease.
-Such *visual field defects* or *scotomas* can be caused by diseases that affect tissue anywhere along the optic pathway.
-This ranges from the optic media (cornea / lens / vitreous) which transmit light, to the retina, which detects lights and preprocesses visual information, to the optic nerve, which transmits information, and up to the visual cortex.
-
-Often, clinicians can draw reliable conclusions about the localisation of the disease by the pattern of the visual field defects.
-For example, an infarction of the right visual cortex leads complete blindness in the left hemispheres of both eyes with a very distinct border along the vertical meridian.
-Therefore, the patient does not perceive anything to the right of the point he looks at.
-On the other side of the vertical meridian the visual field is normal.
-In contrast, patients with glaucoma start with areas of slightly decreased light sensitivity often in the upper or lower hemifield with a bow-shaped (arcuate) pattern cause by the localisation of damage in the optic nerve head.
-
-Visual field tests can be used to diagnose disease, to appraise the functional consequences of disease (for example: can the patient still drive a vehicle?), and to monitor progression.
-
-[THIS SEEMS TO BE A BIT OUF OF PLACE, BUT I PREFER NOT TO EDIT THIS PART MYSELF]
-The most advanced statistical methods have been created for monitoring glaucoma progression.
-The open-angle glaucomas are a group of slowly progressing, chronic neurodegenerative diseases of the optic nerve, often caused by increased eye pressure.
-Patients need consequent monitoring over many years and slight changes in the visual field must be detected reliably.
-Therefore, many of the functions of the visualField package focus on glaucoma research.
-
-# Getting started with the package
-## Loading data
-The visualField package contains some example data. This dataset contains the results of 42 visual field tests of one patient (id: sample1). The patient has primary open-angle glaucoma (type: pwg).
-
-```{r}
-
-str(vfpwgSunyiu24d2[, 1:10])
-
-```
-
-The first columns characterize the patient, the time and duration of the examination, the diagnosis, and the reliability parameters 'false positive responses', 'false negative responses' and 'fixation losses'.
-
-Each row corresponds to one visual field test.
-Visual field testing is performed monocularly (one eye at a time), so that two examinations exist for each date.
-The abbreviations for the laterality ('eye') is derived from latin (OD: right eye = oculus dexter, and OS: left eye = oculus sinister), when both eyes are examined, the abbreviation is OU.
-
-The environment variable locini contains the number of the first column that contains the actual visual field date.
-Each column $l_{x}$ corresponds to one of {k} locations in the visual field test.
-A concise way to select the columns $l_1$ to $l_{k}$ is to use the *getvfcols* function.
-
-```{r}
-
-getlocini()
-getvfcols()
-
-vfpwgSunyiu24d2[1, getvfcols()]
-
-```
-
-There mustn't be any columns after these columns.
-Use the *vffilter* function to select specific fields and the *vfisvalid* function to check whether a table fulfills the criteria.
-The latter will also tell you what the problem is when it doesn't.
-
-```{r}
 
@@
 
-example1$newVariable <- "not valid"
-
-vfisvalid(example1)
-
-```
-
-## Plotting the visual field
-The most straightforward plots show the sensitivities at the corresponding locations.
-These representations are very exact, but may be difficult to interpret.
-For learning purposes, here, we manually plot some visual field data.
-
-```{r}
-
-setdefaults()
-
 example2 <- vffilter(vfpwgRetest24d2, 
@@
 
-vfisvalid(example2)
-
-example2_dataframe <- data.frame(x = getlocmap()$coord$x, 
-                                 y = getlocmap()$coord$y, 
-                                 td = as.vector(unlist(example2[,getvfcols()])))
-
-ggplot(example2_dataframe, aes(x = x, y = y)) +
-  geom_text(aes(label = td)) +
-  geom_hline(yintercept = 0) +
-  geom_vline(xintercept = 0) +
-  coord_fixed() +
-  theme_void()
+example3 <- vffilter(vfpwgSunyiu24d2, 
+                     id == "sample1" & 
+                       eye == "OS")
 
@@
 
-Note that left eyes have to be treated differently if you plot the points manually, because the location labels are mirrored along the y axis.
-In other words, to plot the data manually, for a left eye x should be `-getlocmap()$coord$x`.
+# Plotting the visual field
 
-By convention, when both eyes are presented side by side, the left eye is shown on the left and the right eye on the right.
-This corresponds to the visual field of the observer, so that the physiological blindspots are directed temporally (toward the temple, away from the nose) for both eyes.
-This is in contrast to presentation of structural data, which is usually shown like it would look in a patient sitting in front of you, where his right eye is on the observers left side.
+[![Projekt 1](thumbnails/projekt1.png)](projekt1.html)
 
@@
 
-OD_dataframe <- data.frame(x = getlocmap()$coord$x, 
-                             y = getlocmap()$coord$y, 
-                             loc_label = paste0("l", 1:54),
-                             is_blindspot = FALSE,
-                             eye = "OD")
-
-# Note the minus in the x variable for the left eye
-OS_dataframe <- data.frame(x = -getlocmap()$coord$x, 
-                             y = getlocmap()$coord$y, 
-                             loc_label = paste0("l", 1:54),
-                             is_blindspot = FALSE,
-                             eye = "OS")
-
-OD_dataframe[getlocmap()$bs, "is_blindspot"] <- TRUE
-OS_dataframe[getlocmap()$bs, "is_blindspot"] <- TRUE
-
-ODOS_dataframe <- rbind(OS_dataframe, OD_dataframe)
-
-# Use factor in order to show left eye on the left side
-ODOS_dataframe$eye <- factor(ODOS_dataframe$eye, 
-                             levels = c("OS", "OD"), 
-                             ordered = TRUE)
-
-ggplot(ODOS_dataframe, aes(x = x, y = y)) +
-  coord_fixed() +
-  xlim(-30, 30) +
-  ylim(-24, 24) +
-  geom_label(aes(label = loc_label, color = is_blindspot), size = 2.75) +
-  scale_color_manual(values = c("black", "red")) +
-  facet_wrap(~ eye)
-
+png("thumbnails/sensitivity.png", width = 200, height = 200, units = "px")
+vfplot(example2, type = "s")
+dev.off()
 ```
-
-The visualField package provides functions for more sophisticated plotting, which will be explained later.
-*These functions automatically take care that right eyes and left eyes are plotted correctly.*
-
-# Mathematical derivatives of contrast thresholds for statisitcal analysis
-
-## Identifying defective locations on the visual field
-
-As mentioned earlier, the sensitivity in dB at each location is calculated from the contrast threshold relative to the maximal contrast.
-
-The sensitivity to light fluctuates considerably between measurements in the same observer (test-retest variability ---and more as the disease progresses) and between observers (inter-individual differences).
-They also decrease at different rates depending on the eccentricity with age.
-
-With these predictable and unpredictable differences in the pointwise sensitivity measurements, how can we provide tools and reports to aid the clinical assessment of the visual field?
-Or more to the point, how can we discriminate between healthy locations and locations that are damaged or suspected of being damaged?
-Clinically, it is more useful to compute and report the deviations from an age-corrected mean normal visual field, since the identification of spatial patterns of damage becomes easier, thus facilitating the clinical assessment of the visual field.
-
-But this analysis requires the use of a *normative dataset* from which to obtain models describing the age effect on sensitivity at each location.
-These functions are usually linear and those *age linear models* are used to obtain the *age-corrected mean normal visual field*.
-
-The resulting *deviations* from age-corrected mean normal visual field are also measured in dB and they are interpreted as follows.
-By convention, in the Humphrey Field Analyzer, a negative deviation value means that the sensitivity at that location is below mean normal.
-In contrast, on Octopus perimeters, negative is above and positive below mean normal (therefore, values are called *defects*).
-In this package, those differences are called *total deviation* and their interpretation are as for the Humphrey perimeter.
-In case you use visualFields with data from a different perimeter it is important to know what is the definition of *deviation* and change the sign if necessary.
-
-In eye diseases like glaucoma, general defects that affect the whole visual field are often less relevant than focal scotomas.
-And often ocular opacities caused by cataract can mask such focal defects.
-For this reason, a statistical post-hoc correction to *total deviation* values is further applied to obtain *pattern deviation* values.
-The correction factor is often called the *general height* and for the 24-2 Humphrey is obtained by selecting the 7th greatest total deviation value.
-The *pattern deviation* values equal the *total deviation* values minus the *general height*.
-
-## Normative dataset and normative references
-The visualFields package contains three datasets of healthy subjects that can be used to obtain normative age linear models and references.
-
-The SUNY-IU dataset [@IMF:Mar2013b] for the 24-2 SITA Standard and the Goldmann Size III:
-```{r SUNYIU datasets}
-str(vfctrSunyiu24d2[,1:10])
-```
-
-The Iowa dataset [@IMF:Marin2018c] for a bimodal ZEST algorithm and the Goldmann Size V:
-```{r Iowa datasets}
-str(vfctrIowaPC26[,1:10])
-str(vfctrIowaPeri[,1:10])
-```
-
-From these datasets, or any other dataset of healthy eyes, normative references can be computed as follows:
-
-```{r nv}
-nv <- nvgenerate(vfctrSunyiu24d2)
-```
-
-For future use, we set `nv` as the normative values to be used in visualFields for the calculation of total and pattern deviation values and other statistical analyses.
-```{r setnv}
-setnv(nv)
-```
-
-### Age linear models and mean normal visual fields
-The normative values and references stored in `nv` contains the coefficients of the age linear models:
-
-```{r agelm}
-print(nv$agem$coeff)
-```
-
-And a function from where the age-corrected mean normal visual fields can be computed:
-```{r agemeannormal}
-age <- 50
-print(nv$agem$model(age))
-```
-
-Notice that the locations corresponding to the anatomical region of the blind spot all have `NA` as values, as they are excluded from the statistical analysis.
-
-## Total deviation
-Total deviation is calculated from the sensitivities and from age-related normal values. These are saved in the environment variable *nv* (normal values).
-
-```{r sensitivities}
-
-print(example2[, getvfcols()[1:10]])
-
-```
-
-```{r total deviation}
-
-age <- example2$age[1]
-
-print(age)
-
-normal_values <- getnv()$agem$model(age)[1:10]
-
-print(normal_values)
-
-total_defects <- example2[, getvfcols()[1:10]] - normal_values
-
-print(total_defects)
-
-```
-
-These values can be automatically calculated with a dedicated function.
 
@@
 
-td <- gettd(example2)
-print(td[, getvfcols()[1:10]])
-
-```
-
-## Pattern deviation
-Pattern deviation can be calculated after obtaining the general height.
-```{r gh}
-gh <- getgh(td)
-print(gh)
-pattern_defects <- td[, getvfcols()[1:10]] - gh
-print(pattern_defects)
-```
-
-These values can be automatically calculated with a dedicated function.
-
-```{r pattern deviation}
-
-pd <- getpd(td)
-print(pd[, getvfcols()[1:10]])
-
-```
-
-## Total and Pattern deviation probabilities
-As mentioned psychophysical sensitivity values vary considerably from test to test and between subjects Therefore, even in perfectly average eyes deviations will be different from zero most of the time.
-Given a visual field from a patient, the clinician has to know if a given deviation is likely to be caused by disease as opposed to normal retest and inter-individual variability.
-
-The function `nvgenerate` computes for the normative dataset, the total deviation and pattern deviation values for healthy eyes and records pointwise empirical probability distribution functions [@IMF:Hei1987b] that can be used to compute the probability that a specific total deviation (or pattern deviation value) value is from a healthy location [@IMF:Hei1987a] [@IMF:Heijl1991].
-Thus, a low probability implies a greater likelihood of the location being damaged.
-Typically (and in our example), the probabilities are categorized as
-```{r probability categories}
-print(getnv()$lut$probs)
-```
-
-where `0.005` means probability $< 0.005$, where `0.01` means probability $< 0.01$, etc. A probability greater than 0.05 is considered within the normal range.
-
-The probability values for the previous values are as follows.
-
-```{r probability maps}
-print(gettdp(td)[, getvfcols()[1:10]])
-print(getpdp(pd)[, getvfcols()[1:10]])
-```
-
-# Plotting the visual field
-There are different ways to visualize the perimetric results. These visualizations have to be interpreted like maps in that they represent a simplified version of the hill of vision.
-
-[WE PROBABLY NEED SOME INTRO TO VFPLOT BEFORE TALKING MAKING NOTES, ETC?]
-The standard already contains some additional information with a border presenting a color-coded visualization of the deviation and a graphical representation of the blind spot.
-Note that the vfplot with the option "td" needs information on differential light-sensitivities in a normal reference population and on the probability of given deviations from these normal values.
-These data are included for the typical Humphrey Field Analyzer fields, but you need to collect this information if you use other perimeters
-
-Note that you have to specify what type of data is contained in the visual field tables. The default is set to *total deviation*, `td`.
-
-```{r}
-
+png("thumbnails/total_deviation.png", width = 200, height = 200, units = "px")
 vfplot(example2)
+dev.off()
 
@@
 
-It is possible to print sensitivities (with darker representing locations with lower sensitivities)
-
 ```{r}
 
-vfplot(example2, type = "s")
+png("thumbnails/pattern_deviation.png", width = 200, height = 200, units = "px")
+vfplot(example2, type = "pd")
+dev.off()
 
@@
 
-```{r}
-
-vfplot(example2, type = "pd")
-
-```
-
 And an hybrid plot showing total deviation values and probabilities along with a grayscale representing the sensitivities.
@@
 
+png("thumbnails/td_sens.png", width = 200, height = 200, units = "px")
 vfplot(example2, type = "tds")
+dev.off()
 
@@
 
-And a similar hybrid plotbut with pattern deviation values and probabilities.
+And a similar hybrid plot but with pattern deviation values and probabilities.
 
@@
 
+png("thumbnails/pd_sens.png", width = 200, height = 200, units = "px")
 vfplot(example2, type = "pds")
-
-```
-
-# Global indices
-Global indices summarize the visual field in a few parameters and their values and probability level can be calculated with the visualField package.
-The resulting table contains the visual field header data (patient id, eye, etc.) and 8 additional columns (msens, ssens, tmd, tsd, pmd, psd, gh, and vfi).
-
-```{r}
-
-print(getgl(example2)[,-(1:getlocini()-1)])
-
-```
-
-The column `msens` is the mean sensitivity (MS); the arithmetic mean of all sensitivity values. The column `ssens` is the standard deviation of sensitivity values.
-
-The column `tmd` is the mean deviation (MD); the arithmetic mean of all total deviation values. The column `tsd` is the standard deviation of total deviation values.
-
-The column `pmd` is the mean of pattern deviation values. The column `psd` is the standard pattern deviation (PSD) or the standard deviation of the pattern deviation values.
-
-The last column, `vfi`, is the visual field index (VFI) [@IMF:Ben2008], and the second to last is the general height (GH).
-
-The probability of the observed values given that the visual field is normal can also be calculated.
-
-```{r}
-
-print(getglp(getgl(example2))[, -(1:getlocini()-1)])
-
-```
-
-# Putting everything together.
-Single field analysis reports can be created as pdfs or as interactive reports with Shiny. To explore them, type in the R Console the following:
-
-```{r eval = FALSE, include = TRUE}
-vfsfa(example2, file = "test.pdf")
-vfsfashiny(example2)
-```
-
-# Structure function correlations in glaucoma patients
-In the last years, techniques that provide detailed imaging of the retina and the optic nerve have made incredible advances.
-Therefore, correlation between structure and function is becoming more important than ever.
-
-To aid comparisons between structural insult and functional defect, it is useful to understand and model the anatomy of a typical healthy eye, even if such modeling through is imperfect given the large inter-individual differences in anatomical features [@IMF:Jan2009] [@IMF:Jan2012] [@IMF:Drasdo2007] [@IMF:Montesano2020]
-
-In glaucoma, the damage occurs in the optic nerve at the levels of the nerve fiber bundles, which consist of the axons of the retinal ganglion cells.
-In the retina, these cells form two layers: the ganglion cell layer that contains the cell bodies, and the retinal nerve fiber layer.
-However, the cell bodies of the retinal ganglion cells are displaced from the area where the photoreceptors are located that they receive input from.
-This displacement is relevant mostly in the center of the visual field, so it would greatly affect the structure-function comparisons on grids such as the 10-2.
-The following figure shows the visual field locations and the corresponding ganglion cell soma being stimulated.
-
-```{r echo=FALSE}
-
-locmap <- locmaps$p10d2
-gcmap <- vf2gc(locmap$coord)
-par(mfrow = c(1, 2))
-plot(x = locmap$coord$x, 
-     y = locmap$coord$y, 
-     type = "n", 
-     asp = 1,
-     xlim = c(-10, 12),
-     ylim = c(-10, 10),
-     yaxt = 'n',
-     xaxt = 'n',
-     bty = 'n',
-     xlab = "",
-     ylab = "",
-     main = "10-2 grid of test locations")
-lines(-c(10, 0.5), c(0, 0), lwd = 2, col = "lightgray")
-lines(c(0.5, 10), c(0, 0), lwd = 2, col = "lightgray")
-lines(c(0, 0), -c(10, 0.5), lwd = 2, col = "lightgray")
-lines(c(0, 0), c(0.5, 10), lwd = 2, col = "lightgray")
-lines(x = c(-0.5, 0.5), y = c(0, 0), lwd = 1.5)
-lines(x = c(0, 0), y = c(-0.5, 0.5), lwd = 1.5)
-for(r in c(2.5, 5, 7.5, 10))
-  draw.circle(0, 0, r, lwd = 2, col = NA, border = "lightgray")
-text(x = 11, y = 0, "10\u00b0", adj = c(0, 0.5))
-text(x = 0, y = 11, "10\u00b0", adj = c(0.5, 0))
-points(x = locmap$coord$x, y = locmap$coord$y, bg = "white")
-plot(x = locmap$coord$x, 
-     y = locmap$coord$y, 
-     type = "n", 
-     asp = 1,
-     xlim = c(-10, 12),
-     ylim = c(-10, 10),
-     yaxt = 'n',
-     xaxt = 'n',
-     bty = 'n',
-     xlab = "",
-     ylab = "",
-     main = "Corresponding ganglion cells")
-lines(-c(10, 0.5), c(0, 0), lwd = 2, col = "lightgray")
-lines(c(0.5, 10), c(0, 0), lwd = 2, col = "lightgray")
-lines(c(0, 0), -c(10, 0.5), lwd = 2, col = "lightgray")
-lines(c(0, 0), c(0.5, 10), lwd = 2, col = "lightgray")
-lines(x = c(-0.5, 0.5), y = c(0, 0), lwd = 1.5)
-lines(x = c(0, 0), y = c(-0.5, 0.5), lwd = 1.5)
-for(r in c(2.5, 5, 7.5, 10))
-  draw.circle(0, 0, r, lwd = 2, col = NA, border = "lightgray")
-text(x = 11, y = 0, "10\u00b0", adj = c(0, 0.5))
-text(x = 0, y = 11, "10\u00b0", adj = c(0.5, 0))
-points(x = gcmap$x, y = gcmap$y, pch = 21, bg = "white")
-```
-
-Therefore, the visualFields package provides function to calculate where the ganglion cells that subserve a given location in the visual field are located and to calculate at which angle the axons of these ganglion cells enter the optic nerve head (optic disc), where they pass through the sclera (wall of the eye) to form the optic nerve.
-This function is based on Montesano et al.
-[@IMF:Montesano2020] which revisited the original work by Drasdo et al.
-[@IMF:Drasdo2007].
-
-```{r}
-
-print(locmaps$p24d2$coord[23,])
-print(vf2gc(locmaps$p24d2$coord[23,]))
-
-```
-
-The Montesano Ganglion-cell displacement function can be in conjunction with the Jansonius map describing average ganglion cell bundles paths reaching different regions of the retina [@IMF:Jan2009] [@IMF:Jan2012].
-First, we can obtain the angle of incidence on the optic never head (ONH)
-```{r}
-print(loc2psi(vf2gc(locmaps$p24d2$coord[23,])))
-```
-
-And from the angle of incidence, the nerve fiber bundle path can be calculated.
-Note that the function for calculating the trajectory will extrapolate the trajectory beyond the location of the visual field (usually to the raphe).
-
-```{r, warning=FALSE}
-
-# use loc2psi to calculate the incidence of the nerve fiber corresponding to a given location at the ONH
-angle_of_incidence_l19 <-loc2psi(locmaps$p24d2$coord[19, ])
-
-# use cart2jpolar for calculating the position of the visual field location in the polar coordinate system
-locations <- rbind(cart2jpolar(locmaps$p24d2$coord[19, ]),
-                   cart2jpolar(locmaps$p24d2$coord[23, ]),
-                   cart2jpolar(locmaps$p24d2$coord[47, ]),
-                   cart2jpolar(data.frame(x = 0, y = 0)))
-locations$labels <- c("l19", "l23", "l47", "fixation")
-
-# use bundePath to create a function for calculating the bundle path
-path_function_l19 <- bundlePath(angle_of_incidence_l19)
-path_function_l23 <- bundlePath(loc2psi(locmaps$p24d2$coord[23, ]))
-path_function_l47 <- bundlePath(loc2psi(locmaps$p24d2$coord[47, ]))
-
-bundle_path <- data.frame(y = 4:90,
-                          x_l19 = path_function_l19(4:90),
-                          x_l23 = path_function_l23(4:90),
-                          x_l47 = path_function_l47(4:90))
-ggplot(bundle_path, aes(y = y)) +
-  geom_line(aes(x = x_l19)) +
-  geom_point(data = locations, aes(x = -psi, y = r, color = labels, shape = labels == "fixation")) +
-  geom_line(aes(x = x_l23), color = "red") +
-  geom_line(aes(x = x_l47), color = "blue") +
-  coord_polar(start = pi / 2, direction = -1) +
-  scale_x_continuous(breaks = c(-180, -90, 0, 90, 180), limits = c(-180, 180)) +
-  scale_y_continuous(limits = c(0, 60)) +
-  scale_color_manual(values = c("l19" = "black", "l23" = "red", "l47" = "blue", "fixation" = "black")) +
-  guides(color = "none", shape = "none")
-
-```
-
-In this plot, the origin corresponds to the center of the ONH, the black triangle corresponds to the fovea, and the $180^\circ$ angle corresponds to the line between the ONH center and the fovea, which is, in reality, slightly tilted downward in most patients.
-
-For convenience, visualFields also contains a function to map the angle of incidence of bundles on the ONH to the corresponding angle of the circumpapillary retinal nerve fiber layer (cpRNFL) scan.
-This allows to map a visual field location to its corresponding ganglion cell location, to its angle of incidence on the ONH and finally to the angle in the cpRNFL scan.
-
-```{r}
-
-l23 <- locmaps$p24d2$coord[23,]
-
-# location on the visual field to corresponding ganglion cell
-gc23 <- vf2gc(l23)
-print(gc23)
-
-# ganglion cell location to angle of incidence on the ONH
-psi23 <- loc2psi(gc23)
-print(psi23)
-
-# angle of incidence on the ONH to OCT cpRNFL scan angle
-ang23 <- psi2oct(psi23)
-print(ang23 - 360)
+dev.off()
 
@@
 # Longitudinal analyses
-Monitoring visual function over time in glaucoma patients is an essential clinical application of visual fields in ophthalmology and optometry practices.
-
-For several reasons, measurements will fluctuate over time.
-Fluctuation between measurements will be greater in eyes with glaucoma, especially in areas of glaucoma-related visual field defects.
-Therefore, it is not possible to compare two field tests and claim that the disease has progressed only because the second test yielded lower sensitivities.
-
-Statistical methods and a series with a minimum number of tests are necessary in order to demonstrate disease progression in the visual fields.
-Statistical methods for progression in glaucoma that can be subdivided into *event-based* and *trend-based* methods.
-Furthermore, methods can be subdivided into those that require datasets which describe the "normal fluctuation" in a given population, those that rely on a mathematical model of visual field progression, and those that do not rely on either.
-
-Event-based methods are not implemented in the visualFields package.
-Briefly, they look for progression between the most recent field and a baseline.
-Progression is characterized by a decrease in sensitivity with a magnitude that is unlikely (< 5%) in the population under examination (population based data is necessary) and needs to be confirmed in further tests (usually at least two).
-
-Trend-based methods assume that progression takes place more slowly, so that several field tests are carried out during this time span.
-This concept is unsettling for the clinician who wants to stop progression the first time that he sees the patient, but regularly takes place even in the case of optimal treatment.
-The advantage is that all measurements are taken into account, not only the baseline test and the most recent tests.
-
-There are several tools to analyze progression with visualFields. To review them, we will select a series of 27 visual fields that were obtained over several years.
 
 ```{r}
-example3 <- vffilter(vfpwgSunyiu24d2, 
-                     id == "sample1" & 
-                       eye == "OS")
+
+png("thumbnails/spark.png", width = 200, height = 200, units = "px")
+vfsparklines(example3, thr = 4)
+dev.off()
+
+```
+
+```{r}
+
+png("thumbnails/lego.png", width = 200, height = 200, units = "px")
+vflegoplot(example3, type = "s")
+dev.off()
+
+```
+
+```{r}
+
+png("thumbnails/plr.png", width = 200, height = 200, units = "px")
+vfplotplr(example3)
+dev.off()
+
 ```
@@
 
-```{r}
-
-vfsparklines(example3, thr = 4)
-
-```
-
 The legoplot is a visualization device designed to compare the first visual fields against the last ones on a series in a sort of "wait-and-see" fashion.
@@
 This example, however, shows sensitivities.
-
-```{r}
-
-vflegoplot(example3, type = "s")
-
-```
 
@@
 
-```{r}
-
-prog1 <- plr(example3)
-str(prog1, max.level = 1)
-
-# show the total number of siginficantly progressions locations
-# (out of the 52 pointwise linear regressions)
-print(sum(prog1$pval[-getlocmap()$bs] < 0.05))
-
-```
-
 The results of `plr` can be displayed using the following:
@@
 ```{r}
-vfplotplr(example3)
-```
-In this graphical display, yellow means that the $p$-value for this location is < 0.05, orange that it is < 0.02, red < 0.01, and dark red < 0.005.
 
-Criteria such as the one suggested earlier based on clusters of locations significantly progressing are useful clinically but do not control for specificity.
-An alternative method that ensure that specificity of progression is mantained at 95% or any other desired level is Permutation of Pointwise Linear Regression or PoPLR [@IMF:Ole2012] [@IMF:Marin2021].
-For this method, all 52 $p$-values are combined into a single $S$-statistic, originally introduced by Fisher [@IMF:Fisher1925].
-In order to avoid reliance on population statistics or parametric statistics.
-(Fisher's original work showed that such statistic followed a chi-square distribution but only if the 52 significance tests were independent of each other and the data was Gaussian, two assumptions that are far from reality.) Thus, random permutations of the series of visual fields are generated and, for each, the $S$-statistic computed.
-If enough random permutations are performed, the resulting set of $S$-values represent an empirical distribution for the null hypothesis that there is no progression.
-A combined $p$-value can then be obtained to test this hypothesis against deterioration of the visual field.
-The interpretation of the $p$-value is as with any significance test.
-
-The *poplr* function yields a large list with many intermediate computations.
-The main result is the $S$-value and its $p$-value.
-Note that there is a left tailed (deterioration), a right tailed (improvement) and a two-tailed (visual field not stable) versions in the current implementation.
-
-```{r}
-
-prog2 <- poplr(example3)
-str(prog2, max.level = 1)
+# hist(prog2$cstats$cslall, xlab = "S-values", xlim = c(0, 200))
+# abline(v = prog2$csl, col = "red", lwd = 2)
 
 ```
-
-An improvement is highly unlikely in this series.
-
-```{r}
-
-print(prog2$csr)
-print(prog2$csrp)
-
-```
-
-In contrast, a deterioration is very likely (p < 0.001).
-
-```{r}
-
-prog2$csl
-prog2$cslp
-
-```
-
-In the histogram representing the null distribution, you can see that a $S$-value of `r round(prog2$csl)` was only very rarely found in the random permutations of the visual field series.
-
-```{r}
-
-hist(prog2$cstats$cslall, xlab = "S-values", xlim = c(0, 200))
-abline(v = prog2$csl, col = "red", lwd = 2)
-
-```
-
-## Creating pretty reports
-Progression reports can be created as pdfs or as interactive reports with shiny. To explore them, type in the R Console the following:
-
-```{r eval = FALSE, include = TRUE}
-
-vfspa(example3, file = "test.pdf")
-vfspashiny(example3)
-
-```
-
-# DICOM standard of visual fields
-DICOM (Digital Imaging and Communications in Medicine) standards for ophthalmic static perimetry are specified in Supplement 146 [CAN YOU PROVIDE A REFERENCE OR SPECIFY SUPPLEMENT IN WHAT?].
-However, there are two functions for loading data generated in the DICOMM format created by the Humphrey Field Analyzer (HFA) by Zeiss.
-
-Currently, the data representation in the visualFields package differs from these standards and data cannot be exported from the package to DICOMM.
-
-# Using a different perimeter
-Different manufacturers have different conventions in perimetry.
-For example, clinicians working with Humphrey Field Analyzers (Carl-Zeiss-Meditec) are often used to report loss of function as negative values (e.g., MD: mean *deviation* of $-10$ dB), while those working with Octopus perimeters (Haag-Streit) often report loss as a positive value (e.g., MD: mean *defect* of 10 dB).
-Furthermore, clinicians working with Octopus perimeters often use different grids, even though the 24-2 and 10-2 grids are also available.
-Usually, fields cannot be directly compared because other setting ($L_{max}$) are also different and because different algorithms are used.
-
-The visualField package provided normative data for the Humphrey Field Analyzer 24-2 and Goldmann size III and for Iowa's full field perimetry [@IMF:Marin2018c], which comprises two different grids, Central $26^\circ$ and Periphery (from $30^\circ$ to more than $60^\circ$), with the Goldmann Size V stimulus on the Octopus 900.
-The `loadoctopus` function also imports some normative data from the Octopus perimeter by Haag-Streit, but, as of now, this does not allow full use of the functionality of the visualFields package.
-Currently, there are no other normative references for any other perimeters or grids.
-Research labs can, however, use their suitable normative datasets to generate normative values for analysis with visual fields.
-
-# Advanced topics
-## Setting up the environment
-A given visual field program has certain properties that remain unchanged during the test.
-In the visualFields packages these are saved in an *environment*.
-After setting this environment, you can only analyse data that is consistent with this environment.
-
-The core elements of the environment that need to be setup are
-
--   The normative reference values (`nv`)
--   The test grid (`locmap`)
--   Graphical parameters used for the different visualization devices (`gpar`)
--   The column where the visual field measurements start in the tables (`locini`)
-
-The `nv` environment variable also contains functions for calculating defects, probabilities and global parameters that can be accessed via the functions stated above.
-
-visualFields provides predefined test grids `locmaps`, normative values `normvals`, and parameters for graphical display (`gpars`)
-
-```{r}
-
-print(names(locmaps))
-print(names(normvals))
-print(names(gpars))
-
-```
-
-They should be consistent.
-For instance, `locmaps$p24d2` can only be used with one of the following `normvals`: `sunyiu_24d2_pw`, `sunyiu_24d2`, `sunyiu_24d2_pw_cps` or `sunyiu_24d2_cps`, but not with, e.g., `iowa_PC26_pw`.
-
-Likewise, `locmaps$p24d2` and `normvals$sunyiu_24d2_pw` or `normvals$sunyiu_24d2` can only be used with `gpars$p24d2`, but not the "_csp" version `gpars$p24d2_cps`
-
-```{r getters}
-
-# current normal values
-print(getnv()$info)
-str(getnv(), 1)
-
-# current locmap
-str(getlocmap(), 1)
-# current parameters for plotting
-str(getgpar(), 1)
-# position of the column l1 in the dataframe
-print(getlocini())
-
-```
-
-The default environment assumes a static white-on-white perimetry with a Humphrey perimeter, using a 24-2 grid, a SITA standard algorithm, a stimulus size of Goldman Size III.
-The stimulus duration and the $\Delta L_{max}$ are not explicitly stated in the info, because they are generally known and it does not impact the statistical analysis of the visual field.
-
-Note that for switching to another perimetry program, you need to set all environment variables. visualFields has setter functions for this purpose.
-
-```{r}
-
-# Set locations for Iowa Central 26 grid
-setlocmap(locmaps$pPC26)
-print(getlocmap()$name)
-# Set normative values for Iowa Central 26 grid
-setnv(normvals$iowa_PC26_pw)
-print(getnv()$info$name)
-# Set graphical parameters for Iowa Central 26 grid
-setgpar(gpars$pPC26)
-
-```
-
-If the structure of the dataframe containing visual field data has changed, and the sensitivity locations start at a differen column, this can also be changed with another setter function.
-
-```{r}
-
-# Set first column with sensitivity values
-setlocini(24)
-print(getlocini())
-
-```
-
-Here is the figure with the test locations of the Iowa central $26^\circ$. Notice that this map does not have any locations falling on the anatomical blind spot.
-
-```{r}
-
-new_locmap <- getlocmap()
-
-plot(x = new_locmap$coord$x, 
-     y = new_locmap$coord$y, 
-     type = "n", 
-     asp = 1,
-     xlim = c(-30, 30),
-     ylim = c(-34, 34),
-     yaxt = 'n',
-     xaxt = 'n',
-     bty = 'n',
-     xlab = "",
-     ylab = "",
-     main = "Iowa Central 26 grid of test locations")
-lines(-c(30, 2.5), c(0, 0), lwd = 2, col = "lightgray")
-lines(c(2.5, 30), c(0, 0), lwd = 2, col = "lightgray")
-lines(c(0, 0), -c(30, 2.5), lwd = 2, col = "lightgray")
-lines(c(0, 0), c(2.5, 30), lwd = 2, col = "lightgray")
-lines(x = c(-2.5, 2.5), y = c(0, 0), lwd = 1.5)
-lines(x = c(0, 0), y = c(-2.5, 2.5), lwd = 1.5)
-for(r in c(10, 20, 30))
-  draw.circle(0, 0, r, lwd = 2, col = NA, border = "lightgray")
-draw.ellipse(15, -1.5, 2.75, 3.75, col = "gray", border = NA)
-text(x = 0, y = -34, "horizontal eccentricity [degrees]")
-text(x = -34, y = -0, "vertical eccentricity [degrees]", srt = 90)
-text(x = 31, y = 0, "30\u00b0", adj = c(0, 0.5))
-text(x = 0, y = 31, "30\u00b0", adj = c(0.5, 0))
-points(x = new_locmap$coord$x, y = new_locmap$coord$y)
-
-```
-
-### Defaults
-The environment can be reset to its default values, the 24-2 pattern measured with the SITA algorithm on a Humphrey Field Analyzer, and with a data frame structure where the sensitivity values start at column 11.
-
-```{r}
-
-setdefaults()
-
-print(getlocini())
-print(getlocmap()$name)
-print(getnv()$info$name)
-print(str(getnv(), 1))
-
-```
-
-## Customizing visualFields
-To make full use of the visualField package with other perimeters, grids, or testing algorithms (staircase, full threshold, different implementations of ZEST, etc.), it is necessary to configure visualFields to the custom test.
-
-The first decision to make is the structure of visual field result data.
-In a data frame (and a loadable csv file) the first 4 columns must have the uniquely identifiable key (`id`, `eye`, `date`, `time`).
-To compute *deviations* from age-corrected mean normal, the age must also be included.
-And it is recommended to add the information about the reliability indices of the exam, such as false positive rate (`fpr`), false negative rate (`fnr`), and fixation losses (`fl`), although the last two have recently been found not to be "reliable".
-To design your structure, looking at the datasets included in the visualFields package may be useful.
-
-The first 10 columns of `vfpwgRetest24d2` contains the key, age, and other information.
-
-```{r}
-str(vfpwgRetest24d2[,1:10])
-```
-
-After the information regarding the test, etc, must come the columns representing the visual field locations, each row containing the sensitivities for a specific exam of the eye of a patient.
-
-```{r}
-str(vfpwgRetest24d2[,11:21])
-```
-
-The name of the columns with sensitivity values of an exam must be prefixed by "l" and followed by the corresponding conecutive location number.
-So for a patient using the 24-2 grid, the full record is:
-
-```{r}
-print(vfpwgRetest24d2[1,])
-```
-
-Depending on the structure the visualField environment variable `locini` must be adjusted so that its value is the column number of the first sensitivity value, `l1`.
-This may be done with the setter `setlocini()`.
-
-For a custom grid of location, it is possible to create and set the location map and its graphical parameters.
-visualFields has a function `vfgpar()` to generate graphical parameters for a given grid.
-Here is an example with the Octopus "G1" map, that uses `vfread` to read visualField's ready visual field data.
-
-[I WANTED TO SHOW SOME OF THE FUNCTIONALITY OF THE LOADOCTOPUS FUNCTION, BUT I COULD DOCUMENT THIS ELSEWHERE]
-
-[THE PROBLEM HERE IS THAT NOW PEOPLE WORKING WITH THE OCTOPUS HAVE TO FIND THE COORDINATES ELSEWHERE, AGAIN, BECAUSE THEY DO NOT HAVE ACCESS TO g1.csv. COULD IT BE PROVIDED IN THE PACKAGE?]
-
-```{r}
-# read the Octopus G1 coordinates and create the locmap
-g1 <- list(coord = read.csv("./g1.csv"), name = "G1", desc = "The G1 location map used by Octopus perimeters", bs = numeric())
-str(g1)
-# set the G1 locmap
-setlocmap(g1)
-# create and set the graphical parameters
-setgpar(vfgpar(g1$coord))
-# read visual fields and plot the sensitivities
-vfplot(vfread("./vfg1.csv"), type = "s")
-```
-
-The function `vfgpar()` that generates the graphical parameters has many input argument to provide as much flexibility as possible regarding the color schemes to use for probability scales, grayscale, etc.
-
-We cannot compute or visualize total deviations, pattern deviations, or their corresponding probability maps, because we do not have normative reference values for this particular test.
-If we had a visualField's ready csv file with a suitable normative dataset, we could read it with `vfread()`, use the previously used function `nvgenerate()`, and set the generated normative values with `setnv()`.
-
-Finally, if we want to go to the default settings, we ran again:
-
-```{r}
-setdefaults()
-```
-
-# Acknowledgments
-The visualField's package would not have existed if it wasn't for the important contribution of many contributors who partially funded the development of the package, and/or contributed datasets, code, and above all insight on visual fields analysis and about perimetry and visual sciences at large.
-Huge thanks to:
-  - William H Swanson
-  - Michael Wall
-  - Andrew Turpin
-  - Paul H Artes
-  - Giovanni Montesano
-  - Mitchell W Dul
-
-# References

### 2025-05-11 gallery/gallery.Rmd

@@
 ---
-title: visualFields 1.x
+title: visualFields 1.0.7
 author:
@@
 
-example1 <- vffilter(vfpwgSunyiu24d2, 
-                     id == "sample1" & 
-                       eye == "OD")[1, ]
-
 example2 <- vffilter(vfpwgRetest24d2, 
@@
 
+*Please klick on the thumbnails to see more detail.*
+
 [![sensitivity-plot](thumbnails/sensitivity.png)](#sensitivity-plot)
@@
 
+*Scroll down or click for ["plotting progression"](#plotting-progression).*
+
 # Plotting progression
+
+*Please klick on the thumbnails to see more detail.*
 
@@
 [![Projekt 1](thumbnails/plr.png)](#pointwise-linear-regression-plot)
+
+*Scroll up or click or ["plotting single fields"](#plotting-single-fields).*
+
+---
 
@@
 
-[Back](#plotting-single-fields)
+[**Back to single fields**](#plotting-single-fields)
+<br>
+[**Back to progression**](#plotting-progression)
 
@@
 
-[Back](#plotting-single-fields)
+[**Back to single fields**](#plotting-single-fields)
+<br>
+[**Back to progression**](#plotting-progression)

### 2025-05-15 jss-article-rnw/article.Rnw
Copyediting of the Introduction.

@@
 
-The Open Perimetry Initiative (OPI) \citep{IMF:Tur2012, IMF:Marin2022a} is a standard for interfacing with visual field testing machines (perimeters).
-It specifies basic functions that allow many visual field tests to be constructed.
-As of October 2017, it is fully implemented on the Octopus 900 and partially on the Heidelberg Edge Perimeter, the Kowa AP 7000, the CrewT IMO, and the Centervue Compass.
-
-The \pkg{visualFields} package, part of the OPI, is a collection of tools for analyzing the field of vision.
-It provides a framework for the development and use of innovative methods for statistical analysis, visualization, and clinical interpretation of visual-field loss and its change over time.
-
+The Open Perimetry Initiative (OPI) \citep{IMF:Tur2012, IMF:Marin2022a} is an open standard for interfacing with visual field testing devices (perimeters).
+It defines a set of core functions that enable the construction of a wide range of visual field tests.
+As of October 2017, OPI is fully implemented on the Octopus 900 and partially supported on the Heidelberg Edge Perimeter, Kowa AP 7000, CrewT IMO, and Centervue Compass.
+
+The \pkg{visualFields} package, developed as part of the OPI ecosystem, provides a suite of tools for the statistical analysis and visualization of visual field data.
+It offers a flexible framework to support the development and application of innovative methods for analyzing visual field loss and its progression over time, with applications in both research and clinical settings.
 \subsection{What is the visual field?}
 
-The visual field is the extent of the world that can be seen and refers to both the central and the peripheral vision.
-It is measured in degrees of visual angle, with the center of vision being $0^\circ$, points on the temporal and superior regions of the visual field being positive angles, and those in the inferior and nasal being negative.
-
-In healthy eyes, the sensitivity to light contrast decreases with eccentricity (i.e.
-increasing visual angle.
-The difference in light sensitivity over the visual field as a function of eccentricity is known as Traquair's \emph{hill of vision} \citep{IMF:Traquair1938}.
-\ref{hill} below shows the hill of vision as rendered by Aulhorn \& Harms \citep{IMF:Aul1967}.
+The visual field refers to the entire area that can be seen at a given moment, encompassing both central and peripheral vision.
+It is typically measured in degrees of visual angle, with the fixation point defined as the origin at $x = 0^\circ$ and $y = 0^\circ$.
+Locations temporal and superior to fixation are represented by positive angles, while those nasal and inferior are denoted by negative angles.
+
+In healthy vision, most visual functions, particularly light sensitivity, decline with increasing eccentricity—i.e., distance from the center of fixation.
+This decline in differential light sensitivity forms a three-dimensional structure known as the hill of vision, also referred to as Traquair’s island of vision \citep{IMF:Traquair1938}.
+Figure~\ref{hill} below shows a classic rendering of the hill of vision by Aulhorn and Harms \citep{IMF:Aul1967}.
 
@@
 
-Perimetry (visual field testing) measures a patient's sensitivity to light differences at different locations of the visual field.
-Thus, it is a psychometric way of quantifying the visual field.
-Psychophysical tests systematically measure the relationship between physical stimulus strength and the subjective perception.
-This is performed by varying the stimulus and asking the observer to communicate his perception with a limited number of allowed responses.
-In visual field testing, the observer usually presses a button when a light spot is perceived.
-The relationship between stimulus and perception is analyzed using mathematical model.
-Commonly, these models result in an estimate of a \textbf{threshold}, which denotes either the absolute stimulus strength that can just barely be perceived or the smallest difference that can be perceived as different.
-
-Its purpose is to estimate the \textbf{hill of vision} as best as possible within the confines and limitations of a clinical setting.
-The most common type of visual field testing is \textbf{static automated perimetry (SAP)}, which is performed with in a dome-shaped device, so that the distance between the stimulus and the eye equal for all locations within the visual field.
-In contrast, \textbf{campimetry} is performed with a flat screen.
-Campimetry may be used more often in the future, because computer screens and virtual reality devices may be used for telemedicine.
-
-The visual fields package was created to analyze data from standard automated perimetry\textbf{SAP}, where the light sensitivity is measured at fixed, discrete locations in the visual field.
-Kinetic perimetry, where locations of isosensitivity to fixed, discrete luminance increments, is beyond the scope of the visualFields package.
-
-In SAP, the goal is to quantify sensitivity by increasing and decreasing the light differential (or contrast) between stimulus and background and asking the observer to press the button when they detect the presence of the visual stimulus.
-The contrast threshold is determined at each location of the visual field as the one for which the observer is able to detect the presence of the stimulus 50\% of the times.
-Such thresholds are estimated at a fixed number of locations forming a regular grid as the one in the figure above.
-
-For the typical SAP test, an observer (usually a patient in an ophthalmology or optometry clinic) is comfortably seated in front of a perimeter and puts his head on a chin rest looking into a white Ganzfeld dome (or hemisphere).
-The perimeter is located in a dimly lit, calm room and the observer has been instructed by an experienced technician before the examination.
-During the test, the observer looks at a fixation target (a dot, a cross, a diamond, or other) in the center of the dome.
-The observer presses a button in his hand any time a stimulus appears anywhere in the dome.
-These stimuli are presented only for a short time span (typically 100-200ms), so that the observer has no time to shift his gaze to the stimulus.
-A press of the button will only be linked to a presentation when it occurs within a given time window after the presentation.
-A computer will present stimuli several times at different locations in a grid pattern and in a spatially random order.
-For each location, the stimulus contrast will be decreased if the observer responded to the previous response and increased if not.
-
-The figure below shows the most common grid of locations tested during {SAP}.
-The gray ellipse corresponds to the blind spot of the hill of vision and, therefore, the red dots are expected to have no sensitivity.
+\emph{Perimetry} (visual field testing) uses psychophysical methods to quantify the visual field by measuring a patient's sensitivity to light differences at various locations within the field.
+Psychophysical tests allow for the scientific measurement of the relationship between physical stimulus strength and subjective perception.
+This is achieved by systematically varying the stimulus strength across discrete trials based on the observer's responses.
+The observer is typically restricted to a limited set of responses, and the stimuli are varied according to a predefined algorithm.
+
+In visual field testing, the observer either presses or does not press a button within a fixed time window following the presentation of a light increment in the visual field (a yes/no response).
+The relationship between stimulus strength and response probability is modeled by a psychometric function, which characterizes the likelihood of a "yes" response as a function of stimulus strength.
+These models yield an estimate of the threshold, defined as the stimulus strength at which the observer detects the stimulus with a probability of 50\%.
+
+The most straightforward method for estimating this threshold is the method of constant stimuli, in which stimuli are presented multiple times at several discrete, fixed intensities around the anticipated threshold.
+This approach yields an accurate estimate of the psychometric function, which is typically modeled using a cumulative distribution function (CDF), such as the Gaussian CDF.
+However, this method is inefficient and requires a large number of trials, since stimulus levels far from the threshold provide little information.
+Long test durations negatively affect attention, thereby reducing both the accuracy and precision of the threshold estimates.
+
+To improve efficiency, adaptive algorithms have been developed to reduce the number of stimulus presentations with minimal compromise in accuracy.
+A commonly used example is the staircase algorithm, in which the stimulus intensity decreases after a "seen" response and increases after a "not seen" response.
+The step size is halved after each reversal (i.e., a change in response direction).
+More sophisticated algorithms balance precision and test duration, often incorporating information from adjacent locations in the visual field.
+A notable example is the Swedish Interactive Threshold Algorithm (SITA) used in the Humphrey Field Analyzer.
+
+At the conclusion of the test, the inverse of the contrast threshold, $\Delta L_{\text{threshold}}$, is recorded at each location using a logarithmic decibel (dB) scale.
+This scale is defined as the attenuation from the maximum possible stimulus contrast, $\Delta L_{\text{max}}$:
+
+\begin{equation}
+dB = -10\ \rm log \left (\frac{\Delta L_{\rm thr}}{\Delta L_{\rm max}} \right)\ .
+\end{equation}
+
+For common devices such as the Humphrey Field Analyzer and the Octopus 900, the maximum luminance differential is $\Delta L_{\text{max}} = \frac{10{,}000\ \text{apostilbs}}{\pi} \approx 3183.1\ \text{cd},\text{m}^{-2}$.
+
+The visualFields package was developed to analyze data from static perimetry, where thresholds are measured at fixed, discrete locations in the visual field.
+Kinetic perimetry, in which a stimulus of constant luminance is moved across the field to identify isosensitivity contours, is outside the scope of this package.
+
+Static perimetry aims to estimate the hill of vision as accurately as possible within the constraints of clinical practice.
+The most common form is standard automated perimetry (SAP), which is performed inside a dome-shaped device (a Ganzfeld perimeter) to ensure a uniform distance between the eye and all stimulus locations.
+In contrast, campimetry uses a flat screen and may become more common with the increasing availability of computer screens and virtual reality devices for telemedicine.
+
+\subsection{How is perimetry performed?}
+
+In a typical SAP test, the patient is seated comfortably in front of a perimeter and places their head on a chin rest while looking into a white Ganzfeld dome.
+The room is dimly lit and quiet, and the patient is briefed by an experienced technician before the examination.
+During the test, the patient fixates on a central target (e.g., a dot, cross, or diamond), and presses a button whenever a light stimulus is perceived.
+
+Stimuli are presented briefly (typically 100–200 ms), minimizing the opportunity for eye movement.
+A response is only considered valid if it occurs within a defined time window following the stimulus.
+A computer presents stimuli at discrete locations in a grid, in randomized order.
+The stimulus intensity at each location is adjusted adaptively based on the patient' responses.
+Because of randomization, the patient cannot anticipate where the next stimulus will appear.
+
+The figure below shows the commonly used 24-2 grid of test locations.
+The gray ellipse represents the anatomical blind spot; the red points fall within this region and are expected to show no sensitivity.
+
+The grid, stimulus size, presentation duration, and intensity algorithm remain constant throughout the test and together define the perimetry program.
+It is important to note that results from different perimetry programs are not directly comparable.
+
+\subsection{False negatives and false positives}
+
+Deviations from the expected psychometric function may arise due to observer inattention or short-term threshold fluctuations.
+In diseases like glaucoma, short-term variability is often elevated in damaged regions of the visual field.
+Response quality is typically assessed by:
+
+\begin{enumeration}
+  \item [False negatives (FNR)] Missed responses to stimuli that were previously seen.
+  \item [False positives (FPR)] Responses when no stimulus is presented
+  \item [Fixation losses] Responses to stimuli presented within the blind spot
+\end{enumeration}
+
+These metrics help assess the reliability of a test.
+
+\subsection{Clinical applications of perimetry}
+
+In clinical settings, perimetry is used to detect and monitor visual field defects or scotomas, which can arise from damage anywhere along the visual pathway—from the ocular media (cornea, lens, vitreous), to the retina, optic nerve, or visual cortex.
+
+The spatial pattern of visual field loss often provides diagnostic clues.
+For instance, a right occipital lobe infarct can result in left homonymous hemianopia—complete loss of the left visual field in both eyes, with a sharp border along the vertical meridian.
+Conversely, glaucoma typically presents as arcuate defects, often beginning in the upper or lower hemifield, reflecting characteristic damage to the optic nerve head.
+
+Perimetry supports not only diagnosis but also evaluation of functional impact (e.g., fitness to drive) and disease progression monitoring.
+
+%% -- Methods ------------------------------------------------------------------
+
+\section{Methods} \label{sec:methods}
+
+\subsection{Installation}
+
+The package is available on CRAN under the Apache 2.0 license and can be installed with the following command:
+
+<<eval=FALSE>>=
+
+install.packages("visualFields")
+
+@
+
+To install the newest version of the package, run:
+
+<<eval=FALSE>>=
+
+library(devtools)
+install_github("imarinfr/vf1/source")
+
+@
+
+<<echo=FALSE>>=
+
+library(ggplot2)
+library(plotrix)
+
+@
 
@@
 
-At the end of the test, the inverse of the contrast threshold, $\Delta L_{\rm threshold}$ at each location is recorded in a decibel scale (dB).
-This dB scale is normally defined as attenuation from the maximum stimulus possible $\Delta L_{\rm max}$ as:
-\begin{equation}
-dB = -10\ \rm log \left (\frac{\Delta L_{\rm thr}}{\Delta L_{\rm max}} \right)\ .
-\end{equation}
-For the most common devices, the Humphrey Field Analyzer and the Octopus 900, the maximum luminance differential $\Delta L_{\rm max}$ is $10000\ (\rm apostilb) / \pi = 3183.1\ cd\ m^{-2}$.
-
-These are the data expected by the visualFields package for statistical analysis and generation of visualization devices.
-
 \subsection{Grid patterns}
@@
 If the blind spot is outside the pattern (for example in the 10° visual fields) or is spared by the grid as in the Octopus G1 pattern, the list item bs is set to `r numeric(0)`.
-
-The grid, the stimulus size, presentation time and the algorithm responsible for varying intensities remain constant throughout the examination and are called a perimetry program.
-Not that examinations performed with different programs cannot be directly compared.
-
-\subsection{Measurements}
-Although usually not seen in visual field data, the most elementary unit of a static visual field test is a subject's response.
-It consists of the constant stimulus parameters (size, presentation time, background luminance), the stimulus location, the stimulus intensity and whether the subject pressed the button within a predefined time window to indicate that he saw the stimulus.
-
-Below a certain \textbf{threshold} luminance, a stimulus is usually not perceivable, above the threshold, the stimulus is usually perceived.
-However, close to threshold, a stimulus is not always reliably perceived.
-By definition, the threshold is the luminance of the stimulus where the subject will see the stimulus with a probability of 50\%.
-
-The most straightforward to measure such a threshold would be to present stimuli several time at a number different, fixed stimulus strengths.
-This method is called the method of \textbf{constant stimuli} and results in the \textbf{psychometric function}, which is usually modeled with a cumulative distribution function (CDF), for example a Gaussian CDF.
-However, this is not the most effective way to estimate a threshold, because stimulus strength far from the threshold will have probabilities to be seen that are close to either 0\% or 100\%.
-Testing these strengths several times does not yield much information.
-
-Therefore, advanced algorithms were developed for reducing the number of stimulus presentations with limited trade-off concerning the accuracy of the measurements.
-A staircase algorithm where the intensity is decreased by a certain step size when the stimulus was seen, increased when not seen, and the step size is halved upon each revearsal (seen to not seen or vice versa) is a more efficient algorithms that is frequently used.
-More efficient algorithms usually make a trade-off between precision of the threshold estimate and test time, and often incorporate information on responses in neighboring locations into the algorithm.
-Examples include the Swedish Interactive Threshold Algorithm (SITA) of the Humphery Perimeters.
-Note that long test times negatively affect attention and therefore both accuracy and precision of the threshold estimates.
-
-\subsection{False negatives and false positives}
-
-Residual deviation from the psychometric function may occur due to inattention of the observer but also due to short-term fluctuation of the threshold.
-For example, it has been shown that the threshold underlies short-term fluctuates in the areas where light sensitivity was negatively affected by glaucoma, a disease of the optic nerve.
-
-When the observer does not see a stimulus that is clearly above a level that has been seen before, this is called a false negative response (fnr).
-In contrast, when the subject presses the button although no stimulus is presented, this is a false positive response (fpr).
-
-The cooperation of the subject is also tested in some perimeters by presenting deliberately presenting a stimulus at the location of the blind spot.
-If the subjects sees that stimulus, it must be assumed that he did not fixate correctly (fixation loss) and that the reliability of the field is decreased.
-
-\subsection{Clinical applications of perimetry}
-In clinical practice, visual field testing is used to identify decreases in visual function caused by disease.
-Such \textbf{visual field defects} or \textbf{scotomas} can be caused by diseases that affect tissue anywhere along the optic pathway.
-This ranges from the optic media (cornea / lens / vitreous) which transmit light, to the retina, which detects lights and preprocesses visual information, to the optic nerve, which transmits information, and up to the visual cortex.
-
-Often, clinicians can draw reliable conclusions about the localisation of the disease by the pattern of the visual field defects.
-For example, an infarction of the right visual cortex leads complete blindness in the left hemispheres of both eyes with a very distinct border along the vertical meridian.
-Therefore, the patient does not perceive anything to the right of the point he looks at.
-On the other side of the vertical meridian the visual field is normal.
-In contrast, patients with glaucoma start with areas of slightly decreased light sensitivity often in the upper or lower hemifield with a bow-shaped (arcuate) pattern cause by the localisation of damage in the optic nerve head.
-
-Visual field tests can be used to diagnose disease, to appraise the functional consequences of disease (for example: can the patient still drive a vehicle?), and to monitor progression.
-
-[THIS SEEMS TO BE A BIT OUF OF PLACE, BUT I PREFER NOT TO EDIT THIS PART MYSELF]
-The most advanced statistical methods have been created for monitoring glaucoma progression.
-The open-angle glaucomas are a group of slowly progressing, chronic neurodegenerative diseases of the optic nerve, often caused by increased eye pressure.
-Patients need consequent monitoring over many years and slight changes in the visual field must be detected reliably.
-Therefore, many of the functions of the visualField package focus on glaucoma research.
-
-%% -- Methods ------------------------------------------------------------------
-
-\section{Methods} \label{sec:methods}
-
-\subsection{Installation}
-
-The package is available on CRAN under the Apache 2.0 license and can be installed with the following command:
-
-<<eval=FALSE>>=
-
-install.packages("visualFields")
-
-@
-
-To install the newest version of the package, run:
-
-<<eval=FALSE>>=
-
-library(devtools)
-install_github("imarinfr/vf1/source")
-
-@
-
-<<echo=FALSE>>=
-
-library(ggplot2)
-library(plotrix)
-
-@
 

### 2025-05-15 jss-article-rnw/article.Rnw
Copyedited rest of the article.

@@
 
-The package is available on CRAN under the Apache 2.0 license and can be installed with the following command:
+The \pkg{visualFields} package is available from CRAN under the Apache 2.0 license and can be installed with:
 
 <<eval=FALSE>>=
-
 install.packages("visualFields")
-
 @
 
-To install the newest version of the package, run:
+The latest development version can be installed from GitHub using:
 
 <<eval=FALSE>>=
-
 library(devtools)
 install_github("imarinfr/vf1/source")
-
-@
-
-<<echo=FALSE>>=
-
-library(ggplot2)
-library(plotrix)
-
-@
-
-<<echo=FALSE, fig=TRUE>>=
-
-library(visualFields)
-library(plotrix)
-
-default_locmap <- getlocmap()
-
-plot(x = default_locmap$coord$x, 
-     y = default_locmap$coord$y, 
-     type = "n", 
-     asp = 1,
-     xlim = c(-30, 30),
-     ylim = c(-34, 34),
-     yaxt = 'n',
-     xaxt = 'n',
-     bty = 'n',
-     xlab = "",
-     ylab = "",
-     main = "24-2 grid of test locations")
-lines(-c(30, 2.5), c(0, 0), lwd = 2, col = "lightgray")
-lines(c(2.5, 30), c(0, 0), lwd = 2, col = "lightgray")
-lines(c(0, 0), -c(30, 2.5), lwd = 2, col = "lightgray")
-lines(c(0, 0), c(2.5, 30), lwd = 2, col = "lightgray")
-lines(x = c(-2.5, 2.5), y = c(0, 0), lwd = 1.5)
-lines(x = c(0, 0), y = c(-2.5, 2.5), lwd = 1.5)
-for(r in c(10, 20, 30))
-  draw.circle(0, 0, r, lwd = 2, col = NA, border = "lightgray")
-draw.ellipse(15, -1.5, 2.75, 3.75, col = "gray", border = NA)
-text(x = 0, y = -34, "horizontal eccentricity [degrees]")
-text(x = -34, y = -0, "vertical eccentricity [degrees]", srt = 90)
-text(x = 31, y = 0, "30\u00b0", adj = c(0, 0.5))
-text(x = 0, y = 31, "30\u00b0", adj = c(0.5, 0))
-points(x = default_locmap$coord$x[-default_locmap$bs],
-       y = default_locmap$coord$y[-default_locmap$bs])
-points(x = default_locmap$coord$x[default_locmap$bs],
-       y = default_locmap$coord$y[default_locmap$bs],
-       col = "red", pch = 19)
-
 @
@@
 \subsection{Grid patterns}
-The spatial position of the points $l_1$ to $l_k$ in the visual field is found in the location map (locmap) and depends on the pattern used by the perimeter.
-When analyzing data, you need to specify the locmap.
 
-By default, the 24-2 pattern (shown in the previous figure) is used, which consists of 54 locations that are equally spaced across the field.
+Test locations are defined by location maps (\texttt{locmap}), which specify the spatial coordinates of stimulus presentation points in degrees of visual angle.
+These maps reflect the layout used by visual field testing devices.
+When analyzing data, the correct \texttt{locmap} must be specified.
+
+By default, the \texttt{24-2} pattern is used, consisting of 54 regularly spaced locations across the central visual field.
 
 <<>>=
+library(visualFields)
+default_locmap <- getlocmap()
 str(default_locmap)
@@
 
-Beside a name and a description, a locmap dataset contains a coordinate table with the x- and y-coordinates of the locations measured in degrees, and the id of the points that cover the physiological blind spot (the id is the number x in $l_x$ and corresponds to the row in the coord table).
+Each \texttt{locmap} contains:
+\begin{itemize}
+  \item A coordinate table with $x$- and $y$-positions (in degrees),
+  \item A vector \texttt{bs} indicating locations corresponding to the physiological blind spot.
+\end{itemize}
 
-Other grid patterns are also defined in the package. For instance, the 10-2 grid:
+Other patterns are available in the package, including the \texttt{10-2} grid:
 
@@
 
-And many others:
+All available patterns can be listed with:
 
@@
 
-Locations of the locmap that cover the blind spot are reported in the locmaps lists under the variable bs.
-If the blind spot is outside the pattern (for example in the 10° visual fields) or is spared by the grid as in the Octopus G1 pattern, the list item bs is set to `r numeric(0)`.
+If the blind spot is outside the tested region (e.g., in the 10$^\circ$ pattern) or is skipped (e.g., in the Octopus G1), the \texttt{bs} field is set to \texttt{numeric(0)}.
 
@@
 
-The first columns characterize the patient, the time and duration of the examination, the diagnosis, and the reliability parameters 'false positive responses', 'false negative responses' and 'fixation losses'.
+\subsection{Loading data}
 
-Each row corresponds to one visual field test.
-Visual field testing is performed monocularly (one eye at a time), so that two examinations exist for each date.
-The abbreviations for the laterality ('eye') is derived from latin (OD: right eye = oculus dexter, and OS: left eye = oculus sinister), when both eyes are examined, the abbreviation is OU.
-
-The environment variable locini contains the number of the first column that contains the actual visual field date.
-Each column $l_{x}$ corresponds to one of {k} locations in the visual field test.
-A concise way to select the columns $l_1$ to $l_{k}$ is to use the \textbf{getvfcols} function.
+The package includes example data from a glaucoma patient (\texttt{vfpwgSunyiu24d2}), consisting of 42 visual field tests.
 
 <<>>=
+str(vfpwgSunyiu24d2[, 1:10])
+@
 
+Each row represents a monocular test.
+The eye is labeled using Latin abbreviations: OD (right eye), OS (left eye), and OU (both eyes).
+Initial columns store patient information and reliability metrics.
+Actual test values begin at the column number stored in the \texttt{locini} variable:
+
+<<>>=
 getlocini()
 getvfcols()
-
 vfpwgSunyiu24d2[1, getvfcols()]
-
 @
 
-There mustn't be any columns after these columns.
-Use the \textbf{vffilter} function to select specific fields and the \textbf{vfisvalid} function to check whether a table fulfills the criteria.
-The latter will also tell you what the problem is when it doesn't.
+No columns should appear after these tsest values. Use \texttt{vffilter} to select specific tests and \texttt{vfisvalid} to verify data structure:
 
 <<>>=
-
-example1 <- vffilter(vfpwgSunyiu24d2, 
-                     id == "sample1" & 
-                       eye == "OD")[1, ]
-
-example1$newVariable <- "not valid"
-
+example1 <- vffilter(vfpwgSunyiu24d2, id == "sample1" & eye == "OD")[1, ]
+example1$newVariable <- "invalid"
 vfisvalid(example1)
-
 @
 
-\subsection{Mathematical derivatives of contrast thresholds for statisitcal analysis}
+\subsection{Statistical analysis of contrast thresholds}
 
-\subsubsection{Identifying defective locations on the visual field}
+\subsubsection{Defining defective locations}
 
-As mentioned earlier, the sensitivity in dB at each location is calculated from the contrast threshold relative to the maximal contrast.
+Sensitivity is measured in decibels (dB), derived from contrast thresholds. Sensitivity varies:
+\begin{itemize}
+  \item Intra-individually (test-retest variability),
+  \item Inter-individually (e.g., with age),
+  \item By eccentricity.
+\end{itemize}
 
-The sensitivity to light fluctuates considerably between measurements in the same observer (test-retest variability ---and more as the disease progresses) and between observers (inter-individual differences).
-They also decrease at different rates depending on the eccentricity with age.
+To detect abnormal points, we compare observed sensitivities with an age-corrected mean normal field.
+These comparisons require a \textbf{normative dataset} and location-wise \textbf{age-linear models}.
 
-With these predictable and unpredictable differences in the pointwise sensitivity measurements, how can we provide tools and reports to aid the clinical assessment of the visual field?
-Or more to the point, how can we discriminate between healthy locations and locations that are damaged or suspected of being damaged?
-Clinically, it is more useful to compute and report the deviations from an age-corrected mean normal visual field, since the identification of spatial patterns of damage becomes easier, thus facilitating the clinical assessment of the visual field.
+The difference between observed and normal values is the \textbf{total deviation (TD)}.
+In this package, negative values indicate lower-than-normal sensitivity, following the Humphrey Field Analyzer convention.
+For perimeters that reverse this definition (e.g., Octopus), users must invert the sign.
 
-But this analysis requires the use of a \textbf{normative dataset} from which to obtain models describing the age effect on sensitivity at each location.
-These functions are usually linear and those \textbf{age linear models} are used to obtain the \textbf{age-corrected mean normal visual field}.
+To reveal focal defects masked by generalized depression (e.g., due to cataract), a \textbf{pattern deviation (PD)} is computed by subtracting a general height (GH) estimate from the TD.
+For 24-2, GH is defined as the 7th highest TD value.
 
-The resulting \textbf{deviations} from age-corrected mean normal visual field are also measured in dB and they are interpreted as follows.
-By convention, in the Humphrey Field Analyzer, a negative deviation value means that the sensitivity at that location is below mean normal.
-In contrast, on Octopus perimeters, negative is above and positive below mean normal (therefore, values are called \textbf{defects}).
-In this package, those differences are called \textbf{total deviation} and their interpretation are as for the Humphrey perimeter.
-In case you use visualFields with data from a different perimeter it is important to know what is the definition of \textbf{deviation} and change the sign if necessary.
+\subsubsection{Normative datasets}
 
-In eye diseases like glaucoma, general defects that affect the whole visual field are often less relevant than focal scotomas.
-And often ocular opacities caused by cataract can mask such focal defects.
-For this reason, a statistical post-hoc correction to \textbf{total deviation} values is further applied to obtain \textbf{pattern deviation} values.
-The correction factor is often called the \textbf{general height} and for the 24-2 Humphrey is obtained by selecting the 7th greatest total deviation value.
-The \textbf{pattern deviation} values equal the \textbf{total deviation} values minus the \textbf{general height}.
+The package includes three normative datasets:
 
-\subsubsection{Normative dataset and normative references}
-The visualFields package contains three datasets of healthy subjects that can be used to obtain normative age linear models and references.
+\begin{itemize}
+  \item SUNY-IU dataset for 24-2 with Goldmann size III \citep{IMF:Mar2013b}
+  \item Iowa PC26 and Peri datasets for Goldmann size V \citep{IMF:Marin2018c}
+\end{itemize}
 
-The SUNY-IU dataset [@IMF:Mar2013b] for the 24-2 SITA Standard and the Goldmann Size III:
-<<SUNYIU datasets>>=
-str(vfctrSunyiu24d2[,1:10])
+<<>>=
+str(vfctrSunyiu24d2[, 1:10])
+str(vfctrIowaPC26[, 1:10])
+str(vfctrIowaPeri[, 1:10])
 @
 
-The Iowa dataset [@IMF:Marin2018c] for a bimodal ZEST algorithm and the Goldmann Size V:
-<<Iowa datasets>>=
-str(vfctrIowaPC26[,1:10])
-str(vfctrIowaPeri[,1:10])
-@
+To build the normative model:
 
-From these datasets, or any other dataset of healthy eyes, normative references can be computed as follows:
-
-<<nv>>=
+<<>>=
 nv <- nvgenerate(vfctrSunyiu24d2)
-@
-
-For future use, we set `nv` as the normative values to be used in visualFields for the calculation of total and pattern deviation values and other statistical analyses.
-<<setnv>>=
 setnv(nv)
@@
 
-\subsubsubsection{Age linear models and mean normal visual fields}
-The normative values and references stored in `nv` contains the coefficients of the age linear models:
+The normative object \texttt{nv} includes coefficients of age-linear models:
 
-<<agelm>>=
+<<>>=
 print(nv$agem$coeff)
@@
 
-And a function from where the age-corrected mean normal visual fields can be computed:
-<<agemeannormal>>=
+The expected field at age 50 is:
+
+<<>>=
 age <- 50
@@
 
-Notice that the locations corresponding to the anatomical region of the blind spot all have `NA` as values, as they are excluded from the statistical analysis.
+Blind spot locations are returned as \texttt{NA}.
 
 \subsubsection{Total deviation}
-Total deviation is calculated from the sensitivities and from age-related normal values. These are saved in the environment variable \textbf{nv} (normal values).
 
-<<sensitivities>>=
+Total deviation is the difference between observed and expected sensitivity:
 
-example2 <- vffilter(vfpwgRetest24d2, 
-                     id == "1" & 
-                       eye == "OD")[1,]
-
-print(example2[, getvfcols()[1:10]])
-
+<<>>=
+example2 <- vffilter(vfpwgRetest24d2, id == "1", eye == "OD")[1, ]
+age <- example2$age[1]
+normal_values <- getnv()$agem$model(age)[1:10]
+example2[, getvfcols()[1:10]] - normal_values
 @
 
-<<total deviation>>=
-
-age <- example2$age[1]
-
-print(age)
-
-normal_values <- getnv()$agem$model(age)[1:10]
-
-print(normal_values)
-
-total_defects <- example2[, getvfcols()[1:10]] - normal_values
-
-print(total_defects)
-
-@
-
-These values can be automatically calculated with a dedicated function.
+Or, use the automated function:
 
 <<>>=
-
 td <- gettd(example2)
 print(td[, getvfcols()[1:10]])
-
 @
@@
 \subsubsection{Pattern deviation}
-Pattern deviation can be calculated after obtaining the general height.
-<<gh>>=
+
+Pattern deviation is obtained by subtracting the general height:
+
+<<>>=
 gh <- getgh(td)
-print(gh)
 pattern_defects <- td[, getvfcols()[1:10]] - gh
-print(pattern_defects)
 @
 
-These values can be automatically calculated with a dedicated function.
+Or automatically:
 
-<<pattern deviation>>=
-
+<<>>=
 pd <- getpd(td)
 print(pd[, getvfcols()[1:10]])
-
 @
 
-\subsubsection{Total and Pattern deviation probabilities}
-As mentioned psychophysical sensitivity values vary considerably from test to test and between subjects Therefore, even in perfectly average eyes deviations will be different from zero most of the time.
-Given a visual field from a patient, the clinician has to know if a given deviation is likely to be caused by disease as opposed to normal retest and inter-individual variability.
+\subsubsection{Probability values}
 
-The function `nvgenerate` computes for the normative dataset, the total deviation and pattern deviation values for healthy eyes and records pointwise empirical probability distribution functions [@IMF:Hei1987b] that can be used to compute the probability that a specific total deviation (or pattern deviation value) value is from a healthy location [@IMF:Hei1987a] [@IMF:Heijl1991].
-Thus, a low probability implies a greater likelihood of the location being damaged.
-Typically (and in our example), the probabilities are categorized as
-<<probability categories>>=
+Even healthy eyes may show small deviations.
+To interpret these values, we calculate the probability that a deviation reflects pathology using empirical distributions from the normative data \citep{IMF:Hei1987a, IMF:Heijl1991}.
+
+<<>>=
 print(getnv()$lut$probs)
-@
-
-where `0.005` means probability $< 0.005$, where `0.01` means probability $< 0.01$, etc. A probability greater than 0.05 is considered within the normal range.
-
-The probability values for the previous values are as follows.
-
-<<probability maps>>=
 print(gettdp(td)[, getvfcols()[1:10]])
@@
 
+Probabilities less than 0.05 suggest the location is likely abnormal.
+
 \subsection{Global indices}
-Global indices summarize the visual field in a few parameters and their values and probability level can be calculated with the visualField package.
-The resulting table contains the visual field header data (patient id, eye, etc.) and 8 additional columns (msens, ssens, tmd, tsd, pmd, psd, gh, and vfi).
+
+Global indices summarize the visual field:
 
 <<>>=
-
-print(getgl(example2)[,-(1:getlocini()-1)])
-
+gl <- getgl(example2)
+print(gl[, -(1:getlocini() - 1)])
 @
 
-The column `msens` is the mean sensitivity (MS); the arithmetic mean of all sensitivity values. The column `ssens` is the standard deviation of sensitivity values.
+\begin{itemize}
+  \item \texttt{msens}: Mean sensitivity (MS)
+  \item \texttt{ssens}: SD of sensitivity
+  \item \texttt{tmd}: Mean total deviation (MD)
+  \item \texttt{tsd}: SD of total deviation
+  \item \texttt{pmd}: Mean pattern deviation
+  \item \texttt{psd}: SD of pattern deviation
+  \item \texttt{gh}: General height
+  \item \texttt{vfi}: Visual field index (VFI) \citep{IMF:Ben2008}
+\end{itemize}
 
-The column `tmd` is the mean deviation (MD); the arithmetic mean of all total deviation values. The column `tsd` is the standard deviation of total deviation values.
-
-The column `pmd` is the mean of pattern deviation values. The column `psd` is the standard pattern deviation (PSD) or the standard deviation of the pattern deviation values.
-
-The last column, `vfi`, is the visual field index (VFI) [@IMF:Ben2008], and the second to last is the general height (GH).
-
-The probability of the observed values given that the visual field is normal can also be calculated.
+Associated probabilities:
 
 <<>>=
-
-print(getglp(getgl(example2))[, -(1:getlocini()-1)])
-
+print(getglp(gl)[, -(1:getlocini() - 1)])
 @
@@
 \subsection{Plotting the visual field}
-There are different ways to visualize the perimetric results. These visualizations have to be interpreted like maps in that they represent a simplified version of the hill of vision.
 
-The standard already contains some additional information with a border presenting a color-coded visualization of the deviation and a graphical representation of the blind spot.
-Note that the vfplot with the option "td" needs information on differential light-sensitivities in a normal reference population and on the probability of given deviations from these normal values.
-These data are included for the typical Humphrey Field Analyzer fields, but you need to collect this information if you use other perimeters.
+There are various ways to visualize perimetric data. These visualizations should be interpreted as simplified representations of the hill of vision.
 
-By convention, when both eyes are presented side by side, the left eye is shown on the left and the right eye on the right.
-This corresponds to the visual field of the observer, so that the physiological blindspots are directed temporally (toward the temple, away from the nose) for both eyes.
-This is in contrast to presentation of structural data, which is usually shown like it would look in a patient sitting in front of you, where his right eye is on the observers left side.
+Standard displays include color-coded probability maps, graphical representations of the blind spot, and border enhancements that encode diagnostic information.
+For example, 	exttt{vfplot} with the argument 	exttt{"td"} generates plots that require normative data on differential light sensitivity and the probabilities of deviations from these norms.
+This information is included for the Humphrey Field Analyzer patterns but must be provided separately for other perimeters.
 
-Note that you have to specify what type of data is contained in the visual field tables. The default is set to \textbf{total deviation}, `td`.
+By convention, when displaying both eyes side-by-side, the left eye appears on the left and the right eye on the right.
+This reflects the patient's visual fields, with blind spots positioned temporally.
+This is in contrast to structural imaging, which is typically displayed as viewed by the examiner.
+
+It is important to specify the type of data represented in the visual field tables. The default is set to \textbf{total deviation} (	exttt{td}).
 
@@
 
-In this graph, yellow represents probabilities < 0.05, orange probabilities < 0.02, red < 0.01, and dark red < 0.005.
+In this plot, yellow indicates a probability $<$ 0.05, orange $<$ 0.02, red $<$ 0.01, and dark red $<$ 0.005.
 
-A gallery with all available visual field plots is available online at \url{
-https://rpubs.com/huchzi/1307927}.
+A gallery of all available plot types is available at \url{https://rpubs.com/huchzi/1307927}.
 
 \subsection{Longitudinal analyses}
-Monitoring visual function over time in glaucoma patients is an essential clinical application of visual fields in ophthalmology and optometry practices.
 
-For several reasons, measurements will fluctuate over time.
-Fluctuation between measurements will be greater in eyes with glaucoma, especially in areas of glaucoma-related visual field defects.
-Therefore, it is not possible to compare two field tests and claim that the disease has progressed only because the second test yielded lower sensitivities.
+Monitoring visual function over time is a key application of perimetry in managing glaucoma.
 
-Statistical methods and a series with a minimum number of tests are necessary in order to demonstrate disease progression in the visual fields.
-Statistical methods for progression in glaucoma that can be subdivided into \textbf{event-based} and \textbf{trend-based} methods.
-Furthermore, methods can be subdivided into those that require datasets which describe the "normal fluctuation" in a given population, those that rely on a mathematical model of visual field progression, and those that do not rely on either.
+Sensitivity measurements naturally fluctuate between tests.
+This variability increases in eyes with glaucoma, especially at affected locations.
+Hence, two tests showing differences cannot alone indicate progression.
+Instead, statistical analysis across a series of tests is required to detect significant change.
 
-Event-based methods are not implemented in the visualFields package.
-Briefly, they look for progression between the most recent field and a baseline.
-Progression is characterized by a decrease in sensitivity with a magnitude that is unlikely (< 5\%) in the population under examination (population based data is necessary) and needs to be confirmed in further tests (usually at least two).
+Progression analysis methods fall into two broad categories: \textbf{event-based} and \textbf{trend-based} approaches.
+These may further be classified by their reliance on population-based normative data, mathematical models of progression, or model-free methods.
 
-Trend-based methods assume that progression takes place more slowly, so that several field tests are carried out during this time span.
-This concept is unsettling for the clinician who wants to stop progression the first time that he sees the patient, but regularly takes place even in the case of optimal treatment.
-The advantage is that all measurements are taken into account, not only the baseline test and the most recent tests.
+Event-based methods (not implemented in \pkg{visualFields}) compare a test to a baseline and flag change if the sensitivity reduction is unlikely ($p < 0.05$) based on normative variability, and if confirmed in subsequent tests.
 
-There are several tools to analyze progression with visualFields. To review them, we will select a series of 27 visual fields that were obtained over several years.
+Trend-based methods assume slower change and analyze all tests in a series. Though less intuitive for clinicians, these methods are statistically robust and reflect real-world progression patterns.
+
+To illustrate, we select a time series of 27 visual fields from one eye:
 
 <<>>=
-example3 <- vffilter(vfpwgSunyiu24d2, 
-                     id == "sample1" & 
-                       eye == "OS")
+example3 <- vffilter(vfpwgSunyiu24d2, id == "sample1" & eye == "OS")
 @
@@
 
-The plot used for showing progression are shown in the gallery, which can be found at: \url{https://rpubs.com/huchzi/1307927}.
+Example plots used for monitoring progression are available in the gallery: \url{https://rpubs.com/huchzi/1307927}.
 
 \subsection{Statistical analysis}
-Statistical analysis can be performed with linear regression of global indices or with pointwise linear regression (one regression line for each location), which results in as many slopes and $p$-values as there are visual field locations (52 for the 24-2 grid after removing the two that fall on the anatomical region of the blind spot).
 
-The \textbf{glr} function performs linear regression on global parameters.
-In the following example, the patient loses `r round(abs(glr(getgl(example3))$sl), 2)` dB each year and this change is statistically significant, with a $p$-value < `r format(ceiling(1000 * glr(getgl(example3))$pval) / 1000, nsmall = 3)`.
+Statistical progression analysis may be performed via:
+\begin{itemize}
+\item \textbf{Global linear regression} (e.g., on MD)
+\item \textbf{Pointwise linear regression} (PLR)
+\end{itemize}
+
+The \texttt{glr()} function estimates a regression model for global indices.
+For the example series, the patient loses \texttt{r round(abs(glr(getgl(example3))$sl), 2)} dB/year, with $p < \texttt{r format(ceiling(1000 * glr(getgl(example3))$pval) / 1000, nsmall = 3)}$.
+
+Pointwise linear regression is performed using \texttt{plr()}, resulting in one slope and $p$-value per location:
 
 <<>>=
-
-# md_regression <- data.frame(time = glr(getgl(example3))$years,
-#            md = glr(getgl(example3))$data)
-# 
-# globallm <- glr(getgl(example3))
-# slope <- globallm$sl
-# intercept <- globallm$int
-# pval <- globallm$pval
-# 
-# ggplot(md_regression, aes(x = time, y = md)) +
-#   geom_point() +
-#   geom_abline(slope = slope, intercept = intercept)
-# 
-# print(paste("intercept:", intercept))
-# print(paste("slope:", slope))
-# print(paste("p-value:", pval))
-
+prog1 <- plr(example3)
+str(prog1, max.level = 1)
+print(sum(prog1$pval[-getlocmap()$bs] < 0.05))
 @
 
-The `plr` function performs pointwise linear regression (i.e.
-linear regression for each location in the field separately).
-Specific parameters have to be chosen, in order to detect disease progression.
-For example, progression might be defined as at least 3 locations having a slope less than 0 dB / year and a $p$-value < 0.05.
+To visualize these results:
+
+<<fig=TRUE>>=
+vfplotplr(example3)
+@
+
+In this display, yellow indicates $p < 0.05$, orange $< 0.02$, red $< 0.01$, and dark red $< 0.005$.
+
+PLR-based criteria (e.g., 3 or more progressing points) are clinically useful but lack explicit control for specificity.
+\textbf{Permutation of Pointwise Linear Regression (PoPLR)} addresses this by combining the 52 $p$-values into a single $S$-statistic (per Fisher's method) and comparing it to a null distribution from permuted series [@IMF:Ole2012; @IMF:Marin2021].
 
 <<>>=
-
-prog1 <- plr(example3)
-str(prog1, max.level = 1)
-
-# show the total number of siginficantly progressions locations
-# (out of the 52 pointwise linear regressions)
-print(sum(prog1$pval[-getlocmap()$bs] < 0.05))
-
+prog2 <- poplr(example3)
+str(prog2, max.level = 1)
 @
 
-The results of `plr` can be displayed using the following:
-
-<<fig=TRUE>>=
-
-vfplotplr(example3)
-
-@
-In this graphical display, yellow means that the $p$-value for this location is < 0.05, orange that it is < 0.02, red < 0.01, and dark red < 0.005.
-
-Criteria such as the one suggested earlier based on clusters of locations significantly progressing are useful clinically but do not control for specificity.
-An alternative method that ensure that specificity of progression is mantained at 95\% or any other desired level is Permutation of Pointwise Linear Regression or PoPLR [@IMF:Ole2012] [@IMF:Marin2021].
-For this method, all 52 $p$-values are combined into a single $S$-statistic, originally introduced by Fisher [@IMF:Fisher1925].
-In order to avoid reliance on population statistics or parametric statistics.
-(Fisher's original work showed that such statistic followed a chi-square distribution but only if the 52 significance tests were independent of each other and the data was Gaussian, two assumptions that are far from reality.) Thus, random permutations of the series of visual fields are generated and, for each, the $S$-statistic computed.
-If enough random permutations are performed, the resulting set of $S$-values represent an empirical distribution for the null hypothesis that there is no progression.
-A combined $p$-value can then be obtained to test this hypothesis against deterioration of the visual field.
-The interpretation of the $p$-value is as with any significance test.
-
-The \textbf{poplr} function yields a large list with many intermediate computations.
-The main result is the $S$-value and its $p$-value.
-Note that there is a left tailed (deterioration), a right tailed (improvement) and a two-tailed (visual field not stable) versions in the current implementation.
+The null distribution and observed statistic:
 
 <<>>=
-
-prog2 <- poplr(example3)
-str(prog2, max.level = 1)
-
+print(prog2$csr)
+print(prog2$csrp)
+print(prog2$csl)
+print(prog2$cslp)
 @
 
-An improvement is highly unlikely in this series.
+Visualization:
 
 <<>>=
-
-print(prog2$csr)
-print(prog2$csrp)
-
+hist(prog2$cstats$cslall, xlab = "S-values", xlim = c(0, 200))
+abline(v = prog2$csl, col = "red", lwd = 2)
 @
 
-In contrast, a deterioration is very likely (p < 0.001).
+\subsection{Single field reports}
 
-<<>>=
+Static reports and Shiny applications can be generated:
 
-prog2$csl
-prog2$cslp
-
-@
-
-In the histogram representing the null distribution, you can see that a $S$-value of `r round(prog2$csl)` was only very rarely found in the random permutations of the visual field series.
-
-<<>>=
-
-hist(prog2$cstats$cslall, xlab = "S-values", xlim = c(0, 200))
-abline(v = prog2$csl, col = "red", lwd = 2)
-
-@
-
-\subsection{Putting everything together}
-Single field analysis reports can be created as pdfs or as interactive reports with Shiny. To explore them, type in the R Console the following:
-
-<<eval = FALSE>>=
+<<eval=FALSE>>=
 vfsfa(example2, file = "test.pdf")
@@
 
-\subsection{Structure function correlations in glaucoma patients}
-In the last years, techniques that provide detailed imaging of the retina and the optic nerve have made incredible advances.
-Therefore, correlation between structure and function is becoming more important than ever.
+\subsection{Structure-function correlation}
 
-To aid comparisons between structural insult and functional defect, it is useful to understand and model the anatomy of a typical healthy eye, even if such modeling through is imperfect given the large inter-individual differences in anatomical features [@IMF:Jan2009] [@IMF:Jan2012] [@IMF:Drasdo2007] [@IMF:Montesano2020]
+Anatomical modeling aids comparison between structural imaging and functional testing in glaucoma.
 
-In glaucoma, the damage occurs in the optic nerve at the levels of the nerve fiber bundles, which consist of the axons of the retinal ganglion cells.
-In the retina, these cells form two layers: the ganglion cell layer that contains the cell bodies, and the retinal nerve fiber layer.
-However, the cell bodies of the retinal ganglion cells are displaced from the area where the photoreceptors are located that they receive input from.
-This displacement is relevant mostly in the center of the visual field, so it would greatly affect the structure-function comparisons on grids such as the 10-2.
-The following figure shows the visual field locations and the corresponding ganglion cell soma being stimulated.
+The ganglion cell bodies stimulated by a visual field location are displaced from the photoreceptor input region, especially in the central field.
+This displacement is modeled in \pkg{visualFields} based on Montesano et al.
+[@IMF:Montesano2020].
 
 <<>>=
-
 locmap <- locmaps$p10d2
 gcmap <- vf2gc(locmap$coord)
+@
+
+A plot of the 10-2 grid and corresponding ganglion cell bodies:
+
+<<echo=TRUE, fig=TRUE>>=
+
+plotting code from original omitted for brevity
 
@@
 
-<<echo=TRUE, fig=TRUE>>=
+Nerve fiber paths from each location to the optic disc can also be modeled, using Jansonius' framework [@IMF:Jan2009; @IMF:Jan2012].
 
-plot_grid <- function() {
-  lines(-c(10, 0.5), c(0, 0), lwd = 2, col = "lightgray")
-  lines(c(0.5, 10), c(0, 0), lwd = 2, col = "lightgray")
-  lines(c(0, 0), -c(10, 0.5), lwd = 2, col = "lightgray")
-  lines(c(0, 0), c(0.5, 10), lwd = 2, col = "lightgray")
-  lines(x = c(-0.5, 0.5), y = c(0, 0), lwd = 1.5)
-  lines(x = c(0, 0), y = c(-0.5, 0.5), lwd = 1.5)
-  for(r in c(2.5, 5, 7.5, 10))
-    draw.circle(0, 0, r, lwd = 2, col = NA, border = "lightgray")
-  text(x = 11, y = 0, "10\u00b0", adj = c(0, 0.5))
-  text(x = 0, y = 11, "10\u00b0", adj = c(0.5, 0))
-}
-
-par(mfrow = c(1, 2))
-plot(x = locmap$coord$x, 
-     y = locmap$coord$y, 
-     type = "n", 
-     asp = 1,
-     xlim = c(-10, 12),
-     ylim = c(-10, 10),
-     yaxt = 'n',
-     xaxt = 'n',
-     bty = 'n',
-     xlab = "",
-     ylab = "",
-     main = "10-2 grid of test locations")
-plot_grid()
-points(x = locmap$coord$x, y = locmap$coord$y, bg = "white")
-
-plot(x = locmap$coord$x, 
-     y = locmap$coord$y, 
-     type = "n", 
-     asp = 1,
-     xlim = c(-10, 12),
-     ylim = c(-10, 10),
-     yaxt = 'n',
-     xaxt = 'n',
-     bty = 'n',
-     xlab = "",
-     ylab = "",
-     main = "Corresponding ganglion cells")
-plot_grid()
-points(x = gcmap$x, y = gcmap$y, pch = 21, bg = "white")
+<<>>=
+psi23 <- loc2psi(vf2gc(locmaps$p24d2$coord[23,]))
+ang23 <- psi2oct(psi23)
+print(ang23 - 360)
 @
 
-Therefore, the visualFields package provides function to calculate where the ganglion cells that subserve a given location in the visual field are located and to calculate at which angle the axons of these ganglion cells enter the optic nerve head (optic disc), where they pass through the sclera (wall of the eye) to form the optic nerve.
-This function is based on Montesano et al.
-[@IMF:Montesano2020] which revisited the original work by Drasdo et al.
-[@IMF:Drasdo2007].
+\subsection{Progression reports}
+
+Automated reports can also be generated for progression series:
+
+<<eval = FALSE>>=
+vfspa(example3, file = "test.pdf")
+vfspashiny(example3)
+@
+
+\subsection{DICOM support}
+
+DICOM standards for static perimetry are defined in Supplement 146. The package includes preliminary functions for reading HFA-generated DICOM files, but exporting to DICOM is not yet supported.
+
+\subsection{Support for other perimeters}
+
+Perimetry conventions vary across manufacturers.
+The package includes normative data for Humphrey 24-2 (Goldmann III) and Iowa full-field perimetry with Goldmann V [@IMF:Marin2018c].
+The \texttt{loadoctopus} function provides partial support for Octopus G1 fields, but normative support is incomplete.
+
+Users may provide custom normative datasets using \texttt{nvgenerate()} and \texttt{setnv()}.
+
+\subsection{Setting the environment}
+
+Each analysis requires environment variables:
+\begin{itemize}
+\item \texttt{nv}: normative values
+\item \texttt{locmap}: test grid
+\item \texttt{gpar}: graphical parameters
+\item \texttt{locini}: first sensitivity column
+\end{itemize}
 
 <<>>=
-
-print(locmaps$p24d2$coord[23,])
-print(vf2gc(locmaps$p24d2$coord[23,]))
-
-@
-
-The Montesano Ganglion-cell displacement function can be in conjunction with the Jansonius map describing average ganglion cell bundles paths reaching different regions of the retina [@IMF:Jan2009] [@IMF:Jan2012].
-First, we can obtain the angle of incidence on the optic never head (ONH)
-
-<<>>=
-
-print(loc2psi(vf2gc(locmaps$p24d2$coord[23,])))
-
-@
-
-And from the angle of incidence, the nerve fiber bundle path can be calculated.
-Note that the function for calculating the trajectory will extrapolate the trajectory beyond the location of the visual field (usually to the raphe).
-
-<<warning=FALSE, fig=TRUE>>=
-
-# use loc2psi to calculate the incidence of the nerve fiber corresponding to a given location at the ONH
-angle_of_incidence_l19 <-loc2psi(locmaps$p24d2$coord[19, ])
-
-# use cart2jpolar for calculating the position of the visual field location in the polar coordinate system
-locations <- rbind(cart2jpolar(locmaps$p24d2$coord[19, ]),
-                   cart2jpolar(locmaps$p24d2$coord[23, ]),
-                   cart2jpolar(locmaps$p24d2$coord[47, ]),
-                   cart2jpolar(data.frame(x = 0, y = 0)))
-locations$labels <- c("l19", "l23", "l47", "fixation")
-
-# use bundePath to create a function for calculating the bundle path
-path_function_l19 <- bundlePath(angle_of_incidence_l19)
-path_function_l23 <- bundlePath(loc2psi(locmaps$p24d2$coord[23, ]))
-path_function_l47 <- bundlePath(loc2psi(locmaps$p24d2$coord[47, ]))
-
-bundle_path <- data.frame(y = 4:90,
-                          x_l19 = path_function_l19(4:90),
-                          x_l23 = path_function_l23(4:90),
-                          x_l47 = path_function_l47(4:90))
-ggplot(bundle_path, aes(y = y)) +
-  geom_line(aes(x = x_l19)) +
-  geom_point(data = locations, aes(x = -psi, y = r, color = labels, shape = labels == "fixation")) +
-  geom_line(aes(x = x_l23), color = "red") +
-  geom_line(aes(x = x_l47), color = "blue") +
-  coord_polar(start = pi / 2, direction = -1) +
-  scale_x_continuous(breaks = c(-180, -90, 0, 90, 180), limits = c(-180, 180)) +
-  scale_y_continuous(limits = c(0, 60)) +
-  scale_color_manual(values = c("l19" = "black", "l23" = "red", "l47" = "blue", "fixation" = "black")) +
-  guides(color = "none", shape = "none")
-
-@
-
-In this plot, the origin corresponds to the center of the ONH, the black triangle corresponds to the fovea, and the $180^\circ$ angle corresponds to the line between the ONH center and the fovea, which is, in reality, slightly tilted downward in most patients.
-I DO NOT SEE THIS, BUT I MAY BE MISSING SOMETHING.
-ALSO THE FOVEA (TRIANGLE IF I GOT IT RIGHT?) SHOULD BE A BIT BELOW THE HORIZONTAL MERIDIAN.
-THE POSITION OF THE CENTER OF THE BLIND SPOT IS (15,+2) IN THE MODEL (PLUS AND NOT MINUS BECAUSE THE MODEL IS SPECIFIED FOR THE ANATOMICAL RETINA, AND IT NEEDS TO BE TURNED UPSIDE DOWN TO MATCH THE VISUAL FIELD)
-
-[IVAN: I THOUGHT THAT PSI WAS THE ANGLE OF INCIDENCE IN THE OPTIC NERVE HEAD.
-THEREFORE, I CHOSE A POLAR PLOT AND NOT AN X-Y-PLOT.
-ALSO, I THOUGHT THAT THE 0° ANGLE OF INCIDENCE WAS DEFINED AS THE LINE BETWEEN FOVEA AND ONH.
-I DID NOT TURN THE POLAR COORDINATED ACCORDINGLY EVEN THOUGHT I MENTIONED THAT THE FOVEA WAS "TILTED DOWNWARD".
-BUT I DO NOT HAVE MUCH EXPERIENCE WITH THE MODEL, SO PLEASE CORRECT ME IF I'M WRONG.
-HOW WOULD I CALCULATE XY-COORDINATES?]
-
-For convenience, visualFields also contains a function to map the angle of incidence of bundles on the ONH to the corresponding angle of the circumpapillary retinal nerve fiber layer (cpRNFL) scan.
-This allows to map a visual field location to its corresponding ganglion cell location, to its angle of incidence on the ONH and finally to the angle in the cpRNFL scan.
-
-<<>>=
-
-l23 <- locmaps$p24d2$coord[23,]
-
-# location on the visual field to corresponding ganglion cell
-gc23 <- vf2gc(l23)
-print(gc23)
-
-# ganglion cell location to angle of incidence on the ONH
-psi23 <- loc2psi(gc23)
-print(psi23)
-
-# angle of incidence on the ONH to OCT cpRNFL scan angle
-ang23 <- psi2oct(psi23)
-print(ang23 - 360)
-
-@
-
-\subsection{Creating pretty reports}
-Progression reports can be created as pdfs or as interactive reports with shiny. To explore them, type in the R Console the following:
-
-<<eval = FALSE>>=
-
-vfspa(example3, file = "test.pdf")
-vfspashiny(example3)
-
-@
-
-\subsection{DICOM standard of visual fields}
-DICOM (Digital Imaging and Communications in Medicine) standards for ophthalmic static perimetry are specified in Supplement 146 [CAN YOU PROVIDE A REFERENCE OR SPECIFY SUPPLEMENT IN WHAT?].
-However, there are two functions for loading data generated in the DICOMM format created by the Humphrey Field Analyzer (HFA) by Zeiss.
-
-Currently, the data representation in the visualFields package differs from these standards and data cannot be exported from the package to DICOMM.
-
-\subsection{Using a different perimeter}
-Different manufacturers have different conventions in perimetry.
-For example, clinicians working with Humphrey Field Analyzers (Carl-Zeiss-Meditec) are often used to report loss of function as negative values (e.g., MD: mean \textbf{deviation} of $-10$ dB), while those working with Octopus perimeters (Haag-Streit) often report loss as a positive value (e.g., MD: mean \textbf{defect} of 10 dB).
-Furthermore, clinicians working with Octopus perimeters often use different grids, even though the 24-2 and 10-2 grids are also available.
-Usually, fields cannot be directly compared because other setting ($L_{max}$) are also different and because different algorithms are used.
-
-The visualField package provided normative data for the Humphrey Field Analyzer 24-2 and Goldmann size III and for Iowa's full field perimetry [@IMF:Marin2018c], which comprises two different grids, Central $26^\circ$ and Periphery (from $30^\circ$ to more than $60^\circ$), with the Goldmann Size V stimulus on the Octopus 900.
-The `loadoctopus` function also imports some normative data from the Octopus perimeter by Haag-Streit, but, as of now, this does not allow full use of the functionality of the visualFields package.
-Currently, there are no other normative references for any other perimeters or grids.
-Research labs can, however, use their suitable normative datasets to generate normative values for analysis with visual fields.
-
-\subsection{Advanced topics}
-\subsection{Setting up the environment}
-A given visual field program has certain properties that remain unchanged during the test.
-In the visualFields packages these are saved in an \textbf{environment}.
-After setting this environment, you can only analyse data that is consistent with this environment.
-
-The core elements of the environment that need to be setup are
-
--   The normative reference values (`nv`)
--   The test grid (`locmap`)
--   Graphical parameters used for the different visualization devices (`gpar`)
--   The column where the visual field measurements start in the tables (`locini`)
-
-The `nv` environment variable also contains functions for calculating defects, probabilities and global parameters that can be accessed via the functions stated above.
-
-visualFields provides predefined test grids `locmaps`, normative values `normvals`, and parameters for graphical display (`gpars`)
-
-<<>>=
-
 print(names(locmaps))
@@
 print(names(gpars))
-
 @
 
-They should be consistent.
-For instance, `locmaps$p24d2` can only be used with one of the following `normvals`: `sunyiu_24d2_pw`, `sunyiu_24d2`, `sunyiu_24d2_pw_cps` or `sunyiu_24d2_cps`, but not with, e.g., `iowa_PC26_pw`.
+These must be consistent. For example:
 
-Likewise, `locmaps$p24d2` and `normvals$sunyiu_24d2_pw` or `normvals$sunyiu_24d2` can only be used with `gpars$p24d2`, but not the "_csp" version `gpars$p24d2_cps`
-
-<<getters>>=
-
-# current normal values
-print(getnv()$info)
-str(getnv(), 1)
-
-# current locmap
-str(getlocmap(), 1)
-# current parameters for plotting
-str(getgpar(), 1)
-# position of the column l1 in the dataframe
-print(getlocini())
-
+<<>>=
+setlocmap(locmaps$pPC26)
+setnv(normvals$iowa_PC26_pw)
+setgpar(gpars$pPC26)
+setlocini(24)
 @
 
-The default environment assumes a static white-on-white perimetry with a Humphrey perimeter, using a 24-2 grid, a SITA standard algorithm, a stimulus size of Goldman Size III.
-The stimulus duration and the $\Delta L_{max}$ are not explicitly stated in the info, because they are generally known and it does not impact the statistical analysis of the visual field.
-
-Note that for switching to another perimetry program, you need to set all environment variables. visualFields has setter functions for this purpose.
-
-<<>>=
-
-# Set locations for Iowa Central 26 grid
-setlocmap(locmaps$pPC26)
-print(getlocmap()$name)
-# Set normative values for Iowa Central 26 grid
-setnv(normvals$iowa_PC26_pw)
-print(getnv()$info$name)
-# Set graphical parameters for Iowa Central 26 grid
-setgpar(gpars$pPC26)
-
-@
-
-If the structure of the dataframe containing visual field data has changed, and the sensitivity locations start at a differen column, this can also be changed with another setter function.
-
-<<>>=
-
-# Set first column with sensitivity values
-setlocini(24)
-print(getlocini())
-
-@
-
-Here is the figure with the test locations of the Iowa central $26^\circ$. Notice that this map does not have any locations falling on the anatomical blind spot.
-
-<<fig=TRUE>>=
-
-new_locmap <- getlocmap()
-
-plot(x = new_locmap$coord$x, 
-     y = new_locmap$coord$y, 
-     type = "n", 
-     asp = 1,
-     xlim = c(-30, 30),
-     ylim = c(-34, 34),
-     yaxt = 'n',
-     xaxt = 'n',
-     bty = 'n',
-     xlab = "",
-     ylab = "",
-     main = "Iowa Central 26 grid of test locations")
-lines(-c(30, 2.5), c(0, 0), lwd = 2, col = "lightgray")
-lines(c(2.5, 30), c(0, 0), lwd = 2, col = "lightgray")
-lines(c(0, 0), -c(30, 2.5), lwd = 2, col = "lightgray")
-lines(c(0, 0), c(2.5, 30), lwd = 2, col = "lightgray")
-lines(x = c(-2.5, 2.5), y = c(0, 0), lwd = 1.5)
-lines(x = c(0, 0), y = c(-2.5, 2.5), lwd = 1.5)
-for(r in c(10, 20, 30))
-  draw.circle(0, 0, r, lwd = 2, col = NA, border = "lightgray")
-draw.ellipse(15, -1.5, 2.75, 3.75, col = "gray", border = NA)
-text(x = 0, y = -34, "horizontal eccentricity [degrees]")
-text(x = -34, y = -0, "vertical eccentricity [degrees]", srt = 90)
-text(x = 31, y = 0, "30\u00b0", adj = c(0, 0.5))
-text(x = 0, y = 31, "30\u00b0", adj = c(0.5, 0))
-points(x = new_locmap$coord$x, y = new_locmap$coord$y)
-
-@
-
-\subsubsection{Defaults}
-The environment can be reset to its default values, the 24-2 pattern measured with the SITA algorithm on a Humphrey Field Analyzer, and with a data frame structure where the sensitivity values start at column 11.
-
-<<>>=
-
-setdefaults()
-
-print(getlocini())
-print(getlocmap()$name)
-print(getnv()$info$name)
-print(str(getnv(), 1))
-
-@
-
-\subsection{Customizing visualFields}
-To make full use of the visualField package with other perimeters, grids, or testing algorithms (staircase, full threshold, different implementations of ZEST, etc.), it is necessary to configure visualFields to the custom test.
-
-The first decision to make is the structure of visual field result data.
-In a data frame (and a loadable csv file) the first 4 columns must have the uniquely identifiable key (`id`, `eye`, `date`, `time`).
-To compute \textbf{deviations} from age-corrected mean normal, the age must also be included.
-And it is recommended to add the information about the reliability indices of the exam, such as false positive rate (`fpr`), false negative rate (`fnr`), and fixation losses (`fl`), although the last two have recently been found not to be "reliable".
-To design your structure, looking at the datasets included in the visualFields package may be useful.
-
-The first 10 columns of `vfpwgRetest24d2` contains the key, age, and other information.
-
-<<>>=
-str(vfpwgRetest24d2[,1:10])
-@
-
-After the information regarding the test, etc, must come the columns representing the visual field locations, each row containing the sensitivities for a specific exam of the eye of a patient.
-
-<<>>=
-str(vfpwgRetest24d2[,11:21])
-@
-
-The name of the columns with sensitivity values of an exam must be prefixed by "l" and followed by the corresponding conecutive location number.
-So for a patient using the 24-2 grid, the full record is:
-
-<<>>=
-print(vfpwgRetest24d2[1,])
-@
-
-Depending on the structure the visualField environment variable `locini` must be adjusted so that its value is the column number of the first sensitivity value, `l1`.
-This may be done with the setter `setlocini()`.
-
-For a custom grid of location, it is possible to create and set the location map and its graphical parameters.
-visualFields has a function `vfgpar()` to generate graphical parameters for a given grid.
-Here is an example with the Octopus "G1" map, that uses `vfread` to read visualField's ready visual field data.
-
-[I WANTED TO SHOW SOME OF THE FUNCTIONALITY OF THE LOADOCTOPUS FUNCTION, BUT I COULD DOCUMENT THIS ELSEWHERE]
-
-[THE PROBLEM HERE IS THAT NOW PEOPLE WORKING WITH THE OCTOPUS HAVE TO FIND THE COORDINATES ELSEWHERE, AGAIN, BECAUSE THEY DO NOT HAVE ACCESS TO g1.csv. COULD IT BE PROVIDED IN THE PACKAGE?]
-
-<<eval=FALSE>>=
-# read the Octopus G1 coordinates and create the locmap
-g1 <- list(coord = read.csv("./g1.csv"), name = "G1", desc = "The G1 location map used by Octopus perimeters", bs = numeric())
-str(g1)
-# set the G1 locmap
-setlocmap(g1)
-# create and set the graphical parameters
-setgpar(vfgpar(g1$coord))
-# read visual fields and plot the sensitivities
-vfplot(vfread("./vfg1.csv"), type = "s")
-@
-
-The function `vfgpar()` that generates the graphical parameters has many input argument to provide as much flexibility as possible regarding the color schemes to use for probability scales, grayscale, etc.
-
-We cannot compute or visualize total deviations, pattern deviations, or their corresponding probability maps, because we do not have normative reference values for this particular test.
-If we had a visualField's ready csv file with a suitable normative dataset, we could read it with `vfread()`, use the previously used function `nvgenerate()`, and set the generated normative values with `setnv()`.
-
-Finally, if we want to go to the default settings, we ran again:
+To reset to default (Humphrey 24-2):
 
@@
 
+\subsection{Custom configurations}
+
+Custom test structures must:
+\begin{itemize}
+\item Start with columns: \texttt{id}, \texttt{eye}, \texttt{date}, \texttt{time}
+\item Include \texttt{age}, optionally \texttt{fpr}, \texttt{fnr}, \texttt{fl}
+\item Have sensitivity columns named \texttt{l1}--\texttt{lk}
+\end{itemize}
+
+To import custom locations:
+
+<<eval=FALSE>>=
+g1 <- list(coord = read.csv("./g1.csv"), name = "G1", desc = "G1 map", bs = numeric())
+setlocmap(g1)
+setgpar(vfgpar(g1$coord))
+vfplot(vfread("./vfg1.csv"), type = "s")
+@
+
+If normative data exist, use \texttt{vfread()}, \texttt{nvgenerate()}, and \texttt{setnv()} to enable deviation computations.
+
 %% -- Summary and discussion ---------------------------------------------------
@@
 
-The \pkg{visualFields} package provides a comprehensive set of tools for analyzing visual field data.
-It is part of the Open Perimetry Initiative and supports a wide range of analyses, from basic visualizations to advanced statistical modeling.
-Future work could include extending the package to support additional perimeters and integrating new methods for visual field analysis.
+The \pkg{visualFields} package provides a robust suite of tools for statistical analysis and visualization of visual field data.
+It supports single-field reports, progression analysis, anatomical modeling, and customization for non-standard perimeters.
+As part of the Open Perimetry Initiative, it aims to foster reproducibility and transparency in perimetry research.
+Future development could focus on expanding normative datasets, full DICOM integration, and extended support for additional testing strategies and devices.
 

### 2025-05-17 jss-article-rnw/article.Rnw
Copyedited introduction.

--- 2025-05-15 jss-article-rnw/article.Rnw
@@
 
-\author{Cord Huchzermeyer~\orcidlink{0000-0002-0408-8981}\\Department of Ophthalmology, University Hospital Erlangen
+\author{Cord Huchzermeyer~\orcidlink{0000-0002-0408-8981}\\Department of Ophthalmology\\University Hospital Erlangen
    \And Iván Marín-Franch\\Computational Optometry}
@@
 
-\section[Introduction: Visual Fields in R]{Introduction: Visual Fields in \proglang{R}} \label{sec:intro}
-
 The Open Perimetry Initiative (OPI) \citep{IMF:Tur2012, IMF:Marin2022a} is an open standard for interfacing with visual field testing devices (perimeters).
@@
 library(plotrix)
-
-@
-
-<<echo=FALSE, fig=TRUE>>=
-
 library(visualFields)
-library(plotrix)
-
-default_locmap <- getlocmap()
-
-plot(x = default_locmap$coord$x, 
-     y = default_locmap$coord$y, 
-     type = "n", 
-     asp = 1,
-     xlim = c(-30, 30),
-     ylim = c(-34, 34),
-     yaxt = 'n',
-     xaxt = 'n',
-     bty = 'n',
-     xlab = "",
-     ylab = "",
-     main = "24-2 grid of test locations")
-lines(-c(30, 2.5), c(0, 0), lwd = 2, col = "lightgray")
-lines(c(2.5, 30), c(0, 0), lwd = 2, col = "lightgray")
-lines(c(0, 0), -c(30, 2.5), lwd = 2, col = "lightgray")
-lines(c(0, 0), c(2.5, 30), lwd = 2, col = "lightgray")
-lines(x = c(-2.5, 2.5), y = c(0, 0), lwd = 1.5)
-lines(x = c(0, 0), y = c(-2.5, 2.5), lwd = 1.5)
-for(r in c(10, 20, 30))
-  draw.circle(0, 0, r, lwd = 2, col = NA, border = "lightgray")
-draw.ellipse(15, -1.5, 2.75, 3.75, col = "gray", border = NA)
-text(x = 0, y = -34, "horizontal eccentricity [degrees]")
-text(x = -34, y = -0, "vertical eccentricity [degrees]", srt = 90)
-text(x = 31, y = 0, "30\u00b0", adj = c(0, 0.5))
-text(x = 0, y = 31, "30\u00b0", adj = c(0.5, 0))
-points(x = default_locmap$coord$x[-default_locmap$bs],
-       y = default_locmap$coord$y[-default_locmap$bs])
-points(x = default_locmap$coord$x[default_locmap$bs],
-       y = default_locmap$coord$y[default_locmap$bs],
-       col = "red", pch = 19)
-
-@
-
-\subsection{Grid patterns}
-The spatial position of the points $l_1$ to $l_k$ in the visual field is found in the location map (locmap) and depends on the pattern used by the perimeter.
-When analyzing data, you need to specify the locmap.
-
-By default, the 24-2 pattern (shown in the previous figure) is used, which consists of 54 locations that are equally spaced across the field.
-
-<<>>=
-str(default_locmap)
-@
-
-Beside a name and a description, a locmap dataset contains a coordinate table with the x- and y-coordinates of the locations measured in degrees, and the id of the points that cover the physiological blind spot (the id is the number x in $l_x$ and corresponds to the row in the coord table).
-
-Other grid patterns are also defined in the package. For instance, the 10-2 grid:
-
-<<>>=
-ten_two_locmap <- locmaps$p10d2
-str(ten_two_locmap)
-@
-
-And many others:
-
-<<>>=
-names(locmaps)
-@
-
-Locations of the locmap that cover the blind spot are reported in the locmaps lists under the variable bs.
-If the blind spot is outside the pattern (for example in the 10° visual fields) or is spared by the grid as in the Octopus G1 pattern, the list item bs is set to `r numeric(0)`.
+
+@
 

### 2025-05-17 jss-article-rnw/article.Rnw
Copyedited rest of manuscript. Compiles now.

--- 2025-05-15 jss-article-rnw/article.Rnw
@@
 library("MASS")
+library("plotrix")
+library("ggplot2")
 @
@@
 %% -- Introduction -------------------------------------------------------------
-
-\section[Introduction: Visual Fields in R]{Introduction: Visual Fields in \proglang{R}} \label{sec:intro}
 
@@
 If the blind spot is outside the tested region (e.g., in the 10$^\circ$ pattern) or is skipped (e.g., in the Octopus G1), the \texttt{bs} field is set to \texttt{numeric(0)}.
-
-\subsection{Loading data}
-The visualField package contains some example data. This dataset contains the results of 42 visual field tests of one patient (id: sample1). The patient has primary open-angle glaucoma (type: pwg).
-
-<<>>=
-
-str(vfpwgSunyiu24d2[, 1:10])
-
-@
 
@@
 
-The \texttt{glr()} function estimates a regression model for global indices.
-For the example series, the patient loses \texttt{r round(abs(glr(getgl(example3))$sl), 2)} dB/year, with $p < \texttt{r format(ceiling(1000 * glr(getgl(example3))$pval) / 1000, nsmall = 3)}$.
+The \texttt{glr()} function estimates a regression model for global indices. For the example series, the patient loses \texttt{getgl(example3)} dB/year, with $p < \texttt{getgl(example3)}$.
+
+%\texttt{r round(abs(glr(getgl(example3))$sl), 2)}
+%\texttt{r format(ceiling(1000 * glr(getgl(example3))$pval) / 1000, nsmall = 3)}
 
@@
 
-plotting code from original omitted for brevity
+plot_grid <- function() {
+  lines(-c(10, 0.5), c(0, 0), lwd = 2, col = "lightgray")
+  lines(c(0.5, 10), c(0, 0), lwd = 2, col = "lightgray")
+  lines(c(0, 0), -c(10, 0.5), lwd = 2, col = "lightgray")
+  lines(c(0, 0), c(0.5, 10), lwd = 2, col = "lightgray")
+  lines(x = c(-0.5, 0.5), y = c(0, 0), lwd = 1.5)
+  lines(x = c(0, 0), y = c(-0.5, 0.5), lwd = 1.5)
+  for(r in c(2.5, 5, 7.5, 10))
+    draw.circle(0, 0, r, lwd = 2, col = NA, border = "lightgray")
+  text(x = 11, y = 0, "10\u00b0", adj = c(0, 0.5))
+  text(x = 0, y = 11, "10\u00b0", adj = c(0.5, 0))
+}
 
+par(mfrow = c(1, 2))
+plot(x = locmap$coord$x, 
+     y = locmap$coord$y, 
+     type = "n", 
+     asp = 1,
+     xlim = c(-10, 12),
+     ylim = c(-10, 10),
+     yaxt = 'n',
+     xaxt = 'n',
+     bty = 'n',
+     xlab = "",
+     ylab = "",
+     main = "10-2 grid of test locations")
+plot_grid()
+points(x = locmap$coord$x, y = locmap$coord$y, bg = "white")
+
+plot(x = locmap$coord$x, 
+     y = locmap$coord$y, 
+     type = "n", 
+     asp = 1,
+     xlim = c(-10, 12),
+     ylim = c(-10, 10),
+     yaxt = 'n',
+     xaxt = 'n',
+     bty = 'n',
+     xlab = "",
+     ylab = "",
+     main = "Corresponding ganglion cells")
+plot_grid()
+points(x = gcmap$x, y = gcmap$y, pch = 21, bg = "white")
 @

### 2025-05-17 jss-article-rnw/article.Rnw
Fixed problem with merge: ganglion cell displacement plot added.
Cleaned up the manuscript.

--- 2025-05-17 jss-article-rnw/article.Rnw
@@
 \usepackage{framed}
+\usepackage{listings}
 
@@
 <<preliminaries, echo=FALSE, results=hide>>=
-options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
+options(prompt = "R> ", continue = "+  ", width = 60, useFancyQuotes = FALSE)
 library("MASS")
@@
 library("ggplot2")
+library("utils")
+library("visualFields")
 @
@@
 %%\SweaveOpts{concordance=TRUE}
+\lstset{breaklines=true, breakatwhitespace=true}
 
 %% -- Introduction -------------------------------------------------------------
+
+\section[Introduction: Visual Fields in R]{Introduction: Visual Fields} \label{sec:intro}
 
@@
 This decline in differential light sensitivity forms a three-dimensional structure known as the hill of vision, also referred to as Traquair’s island of vision \citep{IMF:Traquair1938}.
-Figure~\ref{hill} below shows a classic rendering of the hill of vision by Aulhorn and Harms \citep{IMF:Aul1967}.
+Figure~\ref{fig:hill} below shows a classic rendering of the hill of vision by Aulhorn and Harms \citep{IMF:Aul1967}.
 
@@
 
-At the conclusion of the test, the inverse of the contrast threshold, $\Delta L_{\text{threshold}}$, is recorded at each location using a logarithmic decibel (dB) scale.
-This scale is defined as the attenuation from the maximum possible stimulus contrast, $\Delta L_{\text{max}}$:
+At the conclusion of the test, the inverse of the contrast threshold, $\Delta L_{threshold}$, is recorded at each location using a logarithmic decibel (dB) scale.
+This scale is defined as the attenuation from the maximum possible stimulus contrast, $\Delta L_{max}$:
 
@@
 
-For common devices such as the Humphrey Field Analyzer and the Octopus 900, the maximum luminance differential is $\Delta L_{\text{max}} = \frac{10{,}000\ \text{apostilbs}}{\pi} \approx 3183.1\ \text{cd},\text{m}^{-2}$.
+For common devices such as the Humphrey Field Analyzer and the Octopus 900, the maximum luminance differential is $\Delta L_{max} = \frac{10{,}000\ apostilbs}{\pi} \approx 3183.1\ cd,{m}^{-2}$.
 
@@
 
-\begin{enumeration}
+\begin{description}
   \item [False negatives (FNR)] Missed responses to stimuli that were previously seen.
@@
   \item [Fixation losses] Responses to stimuli presented within the blind spot
-\end{enumeration}
+\end{description}
 
@@
 
-Test locations are defined by location maps (\texttt{locmap}), which specify the spatial coordinates of stimulus presentation points in degrees of visual angle.
+Test locations are defined by location maps (\code{locmaps}), which specify the spatial coordinates of stimulus presentation points in degrees of visual angle.
 These maps reflect the layout used by visual field testing devices.
-When analyzing data, the correct \texttt{locmap} must be specified.
-
-By default, the \texttt{24-2} pattern is used, consisting of 54 regularly spaced locations across the central visual field.
-
-<<>>=
-library(visualFields)
-default_locmap <- getlocmap()
-str(default_locmap)
-@
-
-Each \texttt{locmap} contains:
-\begin{itemize}
-  \item A coordinate table with $x$- and $y$-positions (in degrees),
-  \item A vector \texttt{bs} indicating locations corresponding to the physiological blind spot.
-\end{itemize}
-
-Other patterns are available in the package, including the \texttt{10-2} grid:
-
-<<>>=
-ten_two_locmap <- locmaps$p10d2
-str(ten_two_locmap)
-@
-
-All available patterns can be listed with:
+When analyzing data, the correct \code{locmap} must be specified.
 
@@
 
-If the blind spot is outside the tested region (e.g., in the 10$^\circ$ pattern) or is skipped (e.g., in the Octopus G1), the \texttt{bs} field is set to \texttt{numeric(0)}.
-
-\subsection{Loading data}
-
-The package includes example data from a glaucoma patient (\texttt{vfpwgSunyiu24d2}), consisting of 42 visual field tests.
+By default, the \code{24-2} pattern is used, consisting of 54 regularly spaced locations across the central visual field.
 
 <<>>=
-str(vfpwgSunyiu24d2[, 1:10])
+default_locmap <- getlocmap()
+str(default_locmap, strict.width = "wrap")
 @
 
-Each row represents a monocular test.
-The eye is labeled using Latin abbreviations: OD (right eye), OS (left eye), and OU (both eyes).
+Each \code{locmap} contains:
+\begin{itemize}
+  \item A coordinate table with $x$- and $y$-positions (in degrees),
+  \item A vector \code{bs} indicating locations corresponding to the physiological blind spot.
+\end{itemize}
+
+Other patterns are available in the package, including the \code{10-2} grid.
+
+<<>>=
+ten_two_locmap <- locmaps$p10d2
+str(ten_two_locmap, strict.width = "wrap")
+@
+
+If the blind spot is outside the tested region (e.g., in the 10-2 pattern) or is skipped (e.g., in the Octopus G1), \code{bs} is set to \code{numeric(0)}.
+
+\subsection{Working with visual field data}
+
+The package includes example data from a glaucoma patient (\code{vfpwgSunyiu24d2}), consisting of 42 visual field tests.
+
+<<>>=
+str(vfpwgSunyiu24d2[, 1:10], strict.width = "wrap")
+@
+
+Each row represents a monocular test. The eye is labeled using Latin abbreviations:
+
+\begin{description}
+\item [OD] oculus dexter -> right eye
+\item [OS] oculus sinister -> left eye
+\item [OU] oculus uterque -> both eyes
+\end{description}
+
 Initial columns store patient information and reliability metrics.
-Actual test values begin at the column number stored in the \texttt{locini} variable:
+Actual test values (sensitivities) begin at the column number stored in the \code{locini} variable.
+The function \code{getvfcols()} returns the numbers of the columns that contain the actual test values at the locations \code{l1} to \code{ln}, where 1..n corresponds to the rows of \code{coords} in the \code{locmaps}.
 
@@
 getlocini()
-getvfcols()
 vfpwgSunyiu24d2[1, getvfcols()]
@@
 
-No columns should appear after these tsest values. Use \texttt{vffilter} to select specific tests and \texttt{vfisvalid} to verify data structure:
+No columns should appear after these test values. Use \code{vffilter()} to select specific tests and \code{vfisvalid()} to verify data structure:
 
@@
 
-\subsection{Statistical analysis of contrast thresholds}
+The \code{vfread()} and \code{vfread()} functions can be used to read and write data in this format from a CSV file.
+However, these functions cannot be used to import native data from the perimeters.
+For this, the package includes functions to import data from the Octopus 900 and the Humphrey Field Analyzer.
 
-\subsubsection{Defining defective locations}
+\begin{itemize}
+  \item \code{loadoctopus()} for Octopus 900 (Haag-Streit, Switzerland)
+  \item \code{loadhfaxml()}, \code{loadhfadicom()}, \code{loadhfaxmlbatch()}, \code{loadhfadicombatch()} for the Humphrey Field Analyzer (Zeiss)
+\end{itemize}
+
+Detailed examples for exporting data from the Octopus 900 and using \code{loadoctopus()} can be found here: \url{https://huchzi.github.io/vf1/loadOctopus.html}.
+
+\subsection{Normative data}
 
@@
 
-To detect abnormal points, we compare observed sensitivities with an age-corrected mean normal field.
-These comparisons require a \textbf{normative dataset} and location-wise \textbf{age-linear models}.
+To detect abnormal points, we compare observed sensitivities with an age-corrected mean normal field. These comparisons require a \emph{normative dataset} and location-wise \emph{age-linear models}.
 
-The difference between observed and normal values is the \textbf{total deviation (TD)}.
+The difference between observed and normal values is the \emph{total deviation (TD)}.
 In this package, negative values indicate lower-than-normal sensitivity, following the Humphrey Field Analyzer convention.
@@
 
-To reveal focal defects masked by generalized depression (e.g., due to cataract), a \textbf{pattern deviation (PD)} is computed by subtracting a general height (GH) estimate from the TD.
+To reveal focal defects masked by generalized depression (e.g., due to cataract), a \emph{pattern deviation (PD)} is computed by subtracting a general height (GH) estimate from the TD.
 For 24-2, GH is defined as the 7th highest TD value.
 
-\subsubsection{Normative datasets}
-
-The package includes three normative datasets:
+The package includes three datasets with visual fields from normal observers:
 
 \begin{itemize}
-  \item SUNY-IU dataset for 24-2 with Goldmann size III \citep{IMF:Mar2013b}
-  \item Iowa PC26 and Peri datasets for Goldmann size V \citep{IMF:Marin2018c}
+  \item SUNY-IU dataset (\code{vfctrsunyiu_24d2}) for 24-2 with Goldmann size III \citep{IMF:Mar2013b}
+  \item Iowa PC26 (\code{vfctriowa_PC26}) and Peri (\code{vfctrsunyiu_10d2}) datasets for Goldmann size V \citep{IMF:Marin2018c}
 \end{itemize}
@@
 <<>>=
-str(vfctrSunyiu24d2[, 1:10])
-str(vfctrIowaPC26[, 1:10])
-str(vfctrIowaPeri[, 1:10])
+str(
+  vfctrSunyiu24d2, 
+  strict.width = "wrap")
 @
 
-To build the normative model:
+The normal values derived from these data can be found in \code{normvals}.
+
+<<>>=
+str(
+  normvals$sunyiu_24d2, 
+  max.level = 1, strict.width = "wrap")
+@
+
+To build the normative model for a custom dataset, use \code{nvgenerate()}.
 
@@
 
-The normative object \texttt{nv} includes coefficients of age-linear models:
+The normative object \code{nv} includes coefficients of age-linear models (\code{nv$agem$coeff}).:
 
 <<>>=
-print(nv$agem$coeff)
+head(nv$agem$coeff, 5)
 @
 
-The expected field at age 50 is:
+It also contains a function that calculates normal sensitivity for a given age (\code{nv$agem$model()}), Thus, the expected field at age 50 is:
 
@@
 age <- 50
-print(nv$agem$model(age))
+nv$agem$model(age)
 @
 
-Blind spot locations are returned as \texttt{NA}.
+Blind spot locations are returned as \code{NA}.
+
+%% -- Results ------------------------------------------------------------------
+
+\section{Results} \label{sec:results}
+
+\subsection{Metric for individual locations}
 
@@
 
-Total deviation is the difference between observed and expected sensitivity:
+Total deviation is the difference between observed and expected sensitivity. Here, it is shown how it is calculated.
 
@@
 
-Or, use the automated function:
+To calculate the total deviation, use the function from the \code{nv} object:
 
@@
 
-Global indices summarize the visual field:
+Global indices summarize the visual field by applying summary statistics on sensitivity, total deviation or pattern deviation.
+These indices are calculated using \code{getgl()}.
+The resulting \code{data.frame} also contains the patient's age, eye, and date of the test.
 
@@
 gl <- getgl(example2)
-print(gl[, -(1:getlocini() - 1)])
+str(gl, max.level = 1, strict.width = "wrap")
+gl[, -(1:getlocini() - 1)]
 @
 
-\begin{itemize}
-  \item \texttt{msens}: Mean sensitivity (MS)
-  \item \texttt{ssens}: SD of sensitivity
-  \item \texttt{tmd}: Mean total deviation (MD)
-  \item \texttt{tsd}: SD of total deviation
-  \item \texttt{pmd}: Mean pattern deviation
-  \item \texttt{psd}: SD of pattern deviation
-  \item \texttt{gh}: General height
-  \item \texttt{vfi}: Visual field index (VFI) \citep{IMF:Ben2008}
-\end{itemize}
+\begin{description}
+  \item [msens]: Mean sensitivity (MS)
+  \item [ssens]: SD of sensitivity
+  \item [tmd]: Mean total deviation (MD)
+  \item [tsd]: SD of total deviation
+  \item [pmd]: Mean pattern deviation
+  \item [psd]: SD of pattern deviation
+  \item [gh]: General height
+  \item [vfi]: Visual field index (VFI) \citep{IMF:Ben2008}
+\end{description}
 
-Associated probabilities:
+Associated probabilities are calculated with \code{getglp()}:
 
 <<>>=
-print(getglp(gl)[, -(1:getlocini() - 1)])
+getglp(gl)[, -(1:getlocini() - 1)]
 @
-
-%% -- Results ------------------------------------------------------------------
-
-\section{Results} \label{sec:results}
 
@@
 Standard displays include color-coded probability maps, graphical representations of the blind spot, and border enhancements that encode diagnostic information.
-For example, 	exttt{vfplot} with the argument 	exttt{"td"} generates plots that require normative data on differential light sensitivity and the probabilities of deviations from these norms.
+For example, \code{vfplot} with the argument 	\code{"td"} generates plots that require normative data on differential light sensitivity and the probabilities of deviations from these norms.
 This information is included for the Humphrey Field Analyzer patterns but must be provided separately for other perimeters.
@@
 
-It is important to specify the type of data represented in the visual field tables. The default is set to \textbf{total deviation} (	exttt{td}).
+It is important to specify the type of data represented in the visual field tables. The default is set to \emph{total deviation} (	exttt{td}).
 
@@
 
-Progression analysis methods fall into two broad categories: \textbf{event-based} and \textbf{trend-based} approaches.
+Progression analysis methods fall into two broad categories: \emph{event-based} and \emph{trend-based} approaches.
 These may further be classified by their reliance on population-based normative data, mathematical models of progression, or model-free methods.
@@
 
-\subsection{Plotting progression}
+\subsubsection{Plotting progression}
 
@@
 
-\subsection{Statistical analysis}
+\subsubsection{Statistical analysis}
 
@@
 \begin{itemize}
-\item \textbf{Global linear regression} (e.g., on MD)
-\item \textbf{Pointwise linear regression} (PLR)
+\item \emph{Global linear regression} (e.g., on MD)
+\item \emph{Pointwise linear regression} (PLR)
 \end{itemize}
 
-The \texttt{glr()} function estimates a regression model for global indices. For the example series, the patient loses \texttt{getgl(example3)} dB/year, with $p < \texttt{getgl(example3)}$.
+The \code{glr()} function estimates a regression model for global indices. For the example series, the patient loses \code{getgl(example3)} dB/year, with $p < getgl(example3)$.
 
-%\texttt{r round(abs(glr(getgl(example3))$sl), 2)}
-%\texttt{r format(ceiling(1000 * glr(getgl(example3))$pval) / 1000, nsmall = 3)}
+%\code{r round(abs(glr(getgl(example3))$sl), 2)}
+%\code{r format(ceiling(1000 * glr(getgl(example3))$pval) / 1000, nsmall = 3)}
 
-Pointwise linear regression is performed using \texttt{plr()}, resulting in one slope and $p$-value per location:
+Pointwise linear regression is performed using \code{plr()}, resulting in one slope and $p$-value per location:
 
@@
 prog1 <- plr(example3)
-str(prog1, max.level = 1)
+str(prog1, max.level = 1, strict.width = "wrap")
 print(sum(prog1$pval[-getlocmap()$bs] < 0.05))
@@
 PLR-based criteria (e.g., 3 or more progressing points) are clinically useful but lack explicit control for specificity.
-\textbf{Permutation of Pointwise Linear Regression (PoPLR)} addresses this by combining the 52 $p$-values into a single $S$-statistic (per Fisher's method) and comparing it to a null distribution from permuted series [@IMF:Ole2012; @IMF:Marin2021].
+\emph{Permutation of Pointwise Linear Regression (PoPLR)} addresses this by combining the 52 $p$-values into a single $S$-statistic (per Fisher's method) and comparing it to a null distribution from permuted series \citep{IMF:Ole2012,IMF:Marin2021}.
 
@@
 prog2 <- poplr(example3)
-str(prog2, max.level = 1)
+str(prog2, max.level = 1, strict.width = "wrap")
 @
@@
 
-\subsection{Single field reports}
+\subsection{Automated reports}
 
@@
 
+Automated reports can also be generated for progression series:
+
+<<eval = FALSE>>=
+vfspa(example3, file = "test.pdf")
+vfspashiny(example3)
+@
+
 \subsection{Structure-function correlation}
@@
 This displacement is modeled in \pkg{visualFields} based on Montesano et al.
-[@IMF:Montesano2020].
+\citep{IMF:Montesano2020}.
 
@@
 
-Nerve fiber paths from each location to the optic disc can also be modeled, using Jansonius' framework [@IMF:Jan2009; @IMF:Jan2012].
+Nerve fiber paths from each location to the optic disc can also be modeled, using Jansonius' framework \citep{IMF:Jan2009, IMF:Jan2012}.
 
@@
 ang23 <- psi2oct(psi23)
-print(ang23 - 360)
-@
 
-\subsection{Progression reports}
-
-Automated reports can also be generated for progression series:
-
-<<eval = FALSE>>=
-vfspa(example3, file = "test.pdf")
-vfspashiny(example3)
+ang23 - 360
 @
@@
 Perimetry conventions vary across manufacturers.
-The package includes normative data for Humphrey 24-2 (Goldmann III) and Iowa full-field perimetry with Goldmann V [@IMF:Marin2018c].
-The \texttt{loadoctopus} function provides partial support for Octopus G1 fields, but normative support is incomplete.
+The package includes normative data for Humphrey 24-2 (Goldmann III) and Iowa full-field perimetry with Goldmann V \citep{IMF:Marin2018c}.
+The \code{loadoctopus} function provides partial support for Octopus G1 fields, but normative support is incomplete.
 
-Users may provide custom normative datasets using \texttt{nvgenerate()} and \texttt{setnv()}.
+Users may provide custom normative datasets using \code{nvgenerate()} and \code{setnv()}.
 
@@
 \begin{itemize}
-\item \texttt{nv}: normative values
-\item \texttt{locmap}: test grid
-\item \texttt{gpar}: graphical parameters
-\item \texttt{locini}: first sensitivity column
+\item \code{nv}: normative values
+\item \code{locmap}: test grid
+\item \code{gpar}: graphical parameters
+\item \code{locini}: first sensitivity column
 \end{itemize}
@@
 <<>>=
-print(names(locmaps))
-print(names(normvals))
-print(names(gpars))
+names(locmaps)
+names(normvals)
+names(gpars)
 @
@@
 \begin{itemize}
-\item Start with columns: \texttt{id}, \texttt{eye}, \texttt{date}, \texttt{time}
-\item Include \texttt{age}, optionally \texttt{fpr}, \texttt{fnr}, \texttt{fl}
-\item Have sensitivity columns named \texttt{l1}--\texttt{lk}
+\item Start with columns: \code{id}, \code{eye}, \code{date}, \code{time}
+\item Include \code{age}, optionally \code{fpr}, \code{fnr}, \code{fl}
+\item Have sensitivity columns named \code{l1}--\code{lk}
 \end{itemize}
@@
 <<eval=FALSE>>=
-g1 <- list(coord = read.csv("./g1.csv"), name = "G1", desc = "G1 map", bs = numeric())
+g1 <- list(coord = read.csv("./g1.csv"), 
+           name = "G1", 
+           desc = "G1 map", 
+           bs = numeric())
 setlocmap(g1)
@@
 
-If normative data exist, use \texttt{vfread()}, \texttt{nvgenerate()}, and \texttt{setnv()} to enable deviation computations.
+If normative data exist, use \code{vfread()}, \code{nvgenerate()}, and \code{setnv()} to enable deviation computations.
 

### 2025-05-18 jss-article-rnw/article.Rnw
Incremental improvement of manuscript.

@@
 \usepackage{framed}
-\usepackage{listings}
 
@@
 %%\SweaveOpts{concordance=TRUE}
-\lstset{breaklines=true, breakatwhitespace=true}
+%\lstset{breaklines=true, breakatwhitespace=true}
 
@@
 
-\subsection{How is perimetry performed?}
+For analyzing visual field data, the package provides the following functionality:
+\begin{itemize}
+  \item A format for representing visual field data
+  \item Functions for importing data from perimeters
+  \item Functions for reading and writing data in CSV format, and for handlings data
+  \item A format for representing the underlying spatial pattern ({data(locmaps)})
+  \item A format for representing age-corrected normal values ({data(normvals)}), including probabilities
+  \item Functions for calculating total deviation (TD), pattern deviation (PD), and global indices
+  \item Functions for calculating probabilities of TD and PD
+  \item A format for representing parameters for plotting the visual fields ({data(gpars)})
+  \item Functions for plotting visual fields, including longitudinal analyses of visual field series
+  \item Functions for calculating structural parameters pertinent to visual field testing (Jansonius maps, ganglion cell displacement)
+\end{itemize}
 
-In a typical SAP test, the patient is seated comfortably in front of a perimeter and places their head on a chin rest while looking into a white Ganzfeld dome.
-The room is dimly lit and quiet, and the patient is briefed by an experienced technician before the examination.
-During the test, the patient fixates on a central target (e.g., a dot, cross, or diamond), and presses a button whenever a light stimulus is perceived.
-
-Stimuli are presented briefly (typically 100–200 ms), minimizing the opportunity for eye movement.
-A response is only considered valid if it occurs within a defined time window following the stimulus.
-A computer presents stimuli at discrete locations in a grid, in randomized order.
-The stimulus intensity at each location is adjusted adaptively based on the patient' responses.
-Because of randomization, the patient cannot anticipate where the next stimulus will appear.
-
-The figure below shows the commonly used 24-2 grid of test locations.
-The gray ellipse represents the anatomical blind spot; the red points fall within this region and are expected to show no sensitivity.
-
-The grid, stimulus size, presentation duration, and intensity algorithm remain constant throughout the test and together define the perimetry program.
-It is important to note that results from different perimetry programs are not directly comparable.
-
-\subsection{False negatives and false positives}
-
-Deviations from the expected psychometric function may arise due to observer inattention or short-term threshold fluctuations.
-In diseases like glaucoma, short-term variability is often elevated in damaged regions of the visual field.
-Response quality is typically assessed by:
-
-\begin{description}
-  \item [False negatives (FNR)] Missed responses to stimuli that were previously seen.
-  \item [False positives (FPR)] Responses when no stimulus is presented
-  \item [Fixation losses] Responses to stimuli presented within the blind spot
-\end{description}
-
-These metrics help assess the reliability of a test.
-
-\subsection{Clinical applications of perimetry}
-
-In clinical settings, perimetry is used to detect and monitor visual field defects or scotomas, which can arise from damage anywhere along the visual pathway—from the ocular media (cornea, lens, vitreous), to the retina, optic nerve, or visual cortex.
-
-The spatial pattern of visual field loss often provides diagnostic clues.
-For instance, a right occipital lobe infarct can result in left homonymous hemianopia—complete loss of the left visual field in both eyes, with a sharp border along the vertical meridian.
-Conversely, glaucoma typically presents as arcuate defects, often beginning in the upper or lower hemifield, reflecting characteristic damage to the optic nerve head.
-
-Perimetry supports not only diagnosis but also evaluation of functional impact (e.g., fitness to drive) and disease progression monitoring.
+By default, the package is set up to handle Humphrey Field Analyzer 24-2 visual fields.
+It is recommended to use such data for exploring the package and data is provided by the package.
+However, other patterns and devices can be used.
+The package contains all necessary data for the HFA 10-2 pattern.
+The package is designed to be extensible, allowing users to add new patterns, devices, and normative datasets.
 
@@
 @
+
+\subsection{Visual field data}
+
+The package includes example data from a glaucoma patient (\code{vfpwgSunyiu24d2}), consisting of 42 visual field tests.
+
+The first ten columns contain metadata for each test, including reliability indices (false negatives, false positives, and fixation losses).
+The remaining columns contain the actual test values (sensitivities) at the locations defined in the \code{locmap}.
+
+<<>>=
+str(vfpwgSunyiu24d2[, 1:10], strict.width = "wrap")
+@
+
+The \code{date} variable is stored as a \code{POSIXct} object, which can be converted to a date format using the \code{as.Date()} function. Each row represents a monocular test.
+
+The eye is labeled using Latin abbreviations:
+
+\begin{description}
+\item [OD] oculus dexter -> right eye
+\item [OS] oculus sinister -> left eye
+\item [OU] oculus uterque -> both eyes
+\end{description}
+
+The \code{type} variable allows assignment of tests to different groups.
+
+\subsubsection{Reliability indices}
+
+Deviations from the expected psychometric function may arise due to observer inattention or short-term threshold fluctuations.
+In diseases like glaucoma, short-term variability is often elevated in damaged regions of the visual field.
+Response quality is typically assessed by:
+
+\begin{description}
+  \item [False negatives (FNR)] Missed responses to stimuli that were previously seen.
+  \item [False positives (FPR)] Responses when no stimulus is presented
+  \item [Fixation losses] Responses to stimuli presented within the blind spot
+\end{description}
+
+These metrics help assess the reliability of a test.
+
+\subsubsection{Psychophysical data: sensitivities}
+
+Initial columns store patient information and reliability metrics.
+Actual test values (sensitivities) begin at the column number stored in the \code{locini} variable.
+The function \code{getvfcols()} returns the numbers of the columns that contain the actual test values at the locations \code{l1} to \code{ln}, where 1..n corresponds to the rows of \code{coords} in the \code{locmaps}.
+
+<<>>=
+getlocini()
+vfpwgSunyiu24d2[1, getvfcols()]
+@
+
+No columns should appear after these test values. Use \code{vffilter()} to select specific tests and \code{vfisvalid()} to verify data structure:
+
+<<>>=
+example1 <- vffilter(vfpwgSunyiu24d2, id == "sample1" & eye == "OD")[1, ]
+example1$newVariable <- "invalid"
+vfisvalid(example1)
+@
+
+The \code{vfread()} and \code{vfread()} functions can be used to read and write data in this format from a CSV file.
+However, these functions cannot be used to import native data from the perimeters.
+For this, the package includes functions to import data from the Octopus 900 and the Humphrey Field Analyzer.
+
+\begin{itemize}
+  \item \code{loadoctopus()} for Octopus 900 (Haag-Streit, Switzerland)
+  \item \code{loadhfaxml()}, \code{loadhfadicom()}, \code{loadhfaxmlbatch()}, \code{loadhfadicombatch()} for the Humphrey Field Analyzer (Zeiss)
+\end{itemize}
+
+Detailed examples for exporting data from the Octopus 900 and using \code{loadoctopus()} can be found here: \url{https://huchzi.github.io/vf1/loadOctopus.html}.
 
@@
 If the blind spot is outside the tested region (e.g., in the 10-2 pattern) or is skipped (e.g., in the Octopus G1), \code{bs} is set to \code{numeric(0)}.
-
-\subsection{Working with visual field data}
-
-The package includes example data from a glaucoma patient (\code{vfpwgSunyiu24d2}), consisting of 42 visual field tests.
-
-<<>>=
-str(vfpwgSunyiu24d2[, 1:10], strict.width = "wrap")
-@
-
-Each row represents a monocular test. The eye is labeled using Latin abbreviations:
-
-\begin{description}
-\item [OD] oculus dexter -> right eye
-\item [OS] oculus sinister -> left eye
-\item [OU] oculus uterque -> both eyes
-\end{description}
-
-Initial columns store patient information and reliability metrics.
-Actual test values (sensitivities) begin at the column number stored in the \code{locini} variable.
-The function \code{getvfcols()} returns the numbers of the columns that contain the actual test values at the locations \code{l1} to \code{ln}, where 1..n corresponds to the rows of \code{coords} in the \code{locmaps}.
-
-<<>>=
-getlocini()
-vfpwgSunyiu24d2[1, getvfcols()]
-@
-
-No columns should appear after these test values. Use \code{vffilter()} to select specific tests and \code{vfisvalid()} to verify data structure:
-
-<<>>=
-example1 <- vffilter(vfpwgSunyiu24d2, id == "sample1" & eye == "OD")[1, ]
-example1$newVariable <- "invalid"
-vfisvalid(example1)
-@
-
-The \code{vfread()} and \code{vfread()} functions can be used to read and write data in this format from a CSV file.
-However, these functions cannot be used to import native data from the perimeters.
-For this, the package includes functions to import data from the Octopus 900 and the Humphrey Field Analyzer.
-
-\begin{itemize}
-  \item \code{loadoctopus()} for Octopus 900 (Haag-Streit, Switzerland)
-  \item \code{loadhfaxml()}, \code{loadhfadicom()}, \code{loadhfaxmlbatch()}, \code{loadhfadicombatch()} for the Humphrey Field Analyzer (Zeiss)
-\end{itemize}
-
-Detailed examples for exporting data from the Octopus 900 and using \code{loadoctopus()} can be found here: \url{https://huchzi.github.io/vf1/loadOctopus.html}.
 
@@
 
+\subsection{Graphical parameters}
+
+<<>>=
+data(gpars)
+names(gpars)
+str(gpars$p24d2, 
+    max.level = 1, strict.width = "wrap")
+@
+
+The \code{gpars} object contains graphical parameters for plotting visual fields.
+The default parameters are set for the 24-2 pattern, but can be customized for other patterns.
+It contains a list of the coords, which must be identical to the \code{locmap} used for the test.
+The \code{gpar} object also contains information on how to tesselate the borders in the plots, and color palettes for both the fields and their borders in the visual field plots.
+
 %% -- Results ------------------------------------------------------------------
@@
 
-\subsection{Metric for individual locations}
+\subsection{Metrics for individual locations}
 
@@
 td <- gettd(example2)
-print(td[, getvfcols()[1:10]])
+td[, getvfcols()[1:10]]
 @
@@
 pd <- getpd(td)
-print(pd[, getvfcols()[1:10]])
+pd[, getvfcols()[1:10]]
 @
@@
 <<>>=
-print(getnv()$lut$probs)
-print(gettdp(td)[, getvfcols()[1:10]])
-print(getpdp(pd)[, getvfcols()[1:10]])
+getnv()$lut$probs
+gettdp(td)[, getvfcols()[1:10]]
+getpdp(pd)[, getvfcols()[1:10]]
 @
@@
 str(prog1, max.level = 1, strict.width = "wrap")
-print(sum(prog1$pval[-getlocmap()$bs] < 0.05))
+sum(prog1$pval[-getlocmap()$bs] < 0.05)
 @
@@
 <<>>=
-print(prog2$csr)
-print(prog2$csrp)
-print(prog2$csl)
-print(prog2$cslp)
+summary(prog2)
+prog2[c("csr", "csrp", "csl", "cslp")]
 @
@@
 
-<<>>=
-hist(prog2$cstats$cslall, xlab = "S-values", xlim = c(0, 200))
-abline(v = prog2$csl, col = "red", lwd = 2)
+<<fig=TRUE>>=
+ggplot(data.frame(x = prog2$cstats$cslall), aes(x = x)) + 
+         geom_histogram() +
+         geom_vline(xintercept = prog2$csl, color = "red", size = 1) +
+         scale_x_continuous("S-values", breaks = seq(0, 200, 20))
 @
@@
 
-\section{More technical details} \label{app:technical}
+\section{Clinical aspects} \label{app:clinical}
 
-\begin{leftbar}
-Appendices can be included after the bibliography (with a page break). Each
-section within the appendix should have a proper section title (rather than
-just \emph{Appendix}).
+\subsection{How is perimetry performed?}
 
-For more technical style details, please check out JSS's style FAQ at
-\url{https://www.jstatsoft.org/pages/view/style#frequently-asked-questions}
-which includes the following topics:
-\begin{itemize}
-  \item Title vs.\ sentence case.
-  \item Graphics formatting.
-  \item Naming conventions.
-  \item Turning JSS manuscripts into \proglang{R} package vignettes.
-  \item Trouble shooting.
-  \item Many other potentially helpful details\dots
-\end{itemize}
-\end{leftbar}
+In a typical SAP test, the patient is seated comfortably in front of a perimeter and places their head on a chin rest while looking into a white Ganzfeld dome.
+The room is dimly lit and quiet, and the patient is briefed by an experienced technician before the examination.
+During the test, the patient fixates on a central target (e.g., a dot, cross, or diamond), and presses a button whenever a light stimulus is perceived.
+
+Stimuli are presented briefly (typically 100–200 ms), minimizing the opportunity for eye movement.
+A response is only considered valid if it occurs within a defined time window following the stimulus.
+A computer presents stimuli at discrete locations in a grid, in randomized order.
+The stimulus intensity at each location is adjusted adaptively based on the patient' responses.
+Because of randomization, the patient cannot anticipate where the next stimulus will appear.
+
+The figure below shows the commonly used 24-2 grid of test locations.
+The gray ellipse represents the anatomical blind spot; the red points fall within this region and are expected to show no sensitivity.
+
+The grid, stimulus size, presentation duration, and intensity algorithm remain constant throughout the test and together define the perimetry program.
+It is important to note that results from different perimetry programs are not directly comparable.
+
+\subsection{Clinical applications of perimetry}
+
+In clinical settings, perimetry is used to detect and monitor visual field defects or scotomas, which can arise from damage anywhere along the visual pathway—from the ocular media (cornea, lens, vitreous), to the retina, optic nerve, or visual cortex.
+
+The spatial pattern of visual field loss often provides diagnostic clues.
+For instance, a right occipital lobe infarct can result in left homonymous hemianopia—complete loss of the left visual field in both eyes, with a sharp border along the vertical meridian.
+Conversely, glaucoma typically presents as arcuate defects, often beginning in the upper or lower hemifield, reflecting characteristic damage to the optic nerve head.
+
+Perimetry supports not only diagnosis but also evaluation of functional impact (e.g., fitness to drive) and disease progression monitoring.
 

### 2025-05-19 jss-article-rnw/article.Rnw
Modified sensitivity scale.

--- 2025-05-17 jss-article-rnw/article.Rnw
@@
 
-At the conclusion of the test, the inverse of the contrast threshold, $\Delta L_{threshold}$, is recorded at each location using a logarithmic decibel (dB) scale.
-This scale is defined as the attenuation from the maximum possible stimulus contrast, $\Delta L_{max}$:
+At the conclusion of the test, the perimetric sensitivity is recorded in a logarithmic decibel (dB) scale based on the luminance of the stimulus at threshold ($L_{threshold}$).
+In perimetry, luminance is usually reported in apostilbs rather than $cd/m^2$, where $1 asb = \frac{1}{\pi} cd/m^2$.
+This scale is defined as the attenuation from the maximum possible stimulus contrast, $L_{max}$:
 
 \begin{equation}
-dB = -10\ \rm log \left (\frac{\Delta L_{\rm thr}}{\Delta L_{\rm max}} \right)\ .
+dB = 10\ \rm log \left (\frac{L_{\rm max}}{L_{\rm threshold}} \right)\ .
 \end{equation}
 
-For common devices such as the Humphrey Field Analyzer and the Octopus 900, the maximum luminance differential is $\Delta L_{max} = \frac{10{,}000\ apostilbs}{\pi} \approx 3183.1\ cd,{m}^{-2}$.
+$L_{max}$ may differ between parameter types, and the Humphrey Field Analyzer uses 10,000 asb, whereas recent Octopus 900 perimeters commonly use 4000 asb.
 

### 2025-05-19 jss-article-rnw/article.Rnw
Added references.

@@
 In visual field testing, the observer either presses or does not press a button within a fixed time window following the presentation of a light increment in the visual field (a yes/no response).
-The relationship between stimulus strength and response probability is modeled by a psychometric function, which characterizes the likelihood of a "yes" response as a function of stimulus strength.
+The relationship between stimulus strength and response probability is modeled by a \emph{psychometric function}, which characterizes the likelihood of a "yes" response as a function of stimulus strength \citep{CH:knoblauch_modeling_2012}.
 These models yield an estimate of the threshold, defined as the stimulus strength at which the observer detects the stimulus with a probability of 50\%.
@@
 More sophisticated algorithms balance precision and test duration, often incorporating information from adjacent locations in the visual field.
-A notable example is the Swedish Interactive Threshold Algorithm (SITA) used in the Humphrey Field Analyzer.
+A notable example is the \emph{Swedish Interactive Threshold Algorithm (SITA)} used in the Humphrey Field Analyzer.
 
@@
 
-$L_{max}$ may differ between parameter types, and the Humphrey Field Analyzer uses 10,000 asb, whereas recent Octopus 900 perimeters commonly use 4000 asb.
+$L_{max}$ may differ between parameter types, and the Humphrey Field Analyzer uses 10{,}000\,asb, whereas recent Octopus 900 perimeters commonly use 4000\,asb \citep{CH:racette_visual_nodate}.
 

### 2025-06-16 jss-article-rnw/article.Rnw
Added captions and did some copyediting.

@@
 %% need no \usepackage{Sweave}
-\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE}
+\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE, width=7, height=5, out.width='0.1\\textwidth'}
 
@@
 
-\title{visualFields 1.0.7: Tools for Analyzing the Field of Vision in \proglang{R}}
-\Plaintitle{visualFields 1.0.7: Tools for Analyzing the Field of Vision in R}
+\title{visualFields 1.0.7: Tools for Analyzing Visual Fields in \proglang{R}}
+\Plaintitle{visualFields 1.0.7: Tools for Analyzing Visual Fields in R}
 \Shorttitle{visualFields 1.0.7 in \proglang{R}}
@@
   including visualization, statistical analysis, and clinical interpretation of 
-  visual-field loss and its change over time.
+  visual-field loss and its progression over time.
 }
@@
 
-The Open Perimetry Initiative (OPI) \citep{IMF:Tur2012, IMF:Marin2022a} is an open standard for interfacing with visual field testing devices (perimeters).
+The Open Perimetry Initiative (OPI) \citep{IMF:Tur2012, IMF:Marin2022a} provides an open standard for interfacing with visual field testing devices (perimeters).
 It defines a set of core functions that enable the construction of a wide range of visual field tests.
-As of October 2017, OPI is fully implemented on the Octopus 900 and partially supported on the Heidelberg Edge Perimeter, Kowa AP 7000, CrewT IMO, and Centervue Compass.
+As of October 2017, OPI has been fully implemented on the Octopus 900 and partially supports the Heidelberg Edge Perimeter, Kowa AP 7000, CrewT IMO, and Centervue Compass.
 
@@
 It is typically measured in degrees of visual angle, with the fixation point defined as the origin at $x = 0^\circ$ and $y = 0^\circ$.
-Locations temporal and superior to fixation are represented by positive angles, while those nasal and inferior are denoted by negative angles.
+Locations temporal and superior to fixation are assigned positive angles, whereas nasal and inferior locations are assigned negative angles.
 
 In healthy vision, most visual functions, particularly light sensitivity, decline with increasing eccentricity—i.e., distance from the center of fixation.
-This decline in differential light sensitivity forms a three-dimensional structure known as the hill of vision, also referred to as Traquair’s island of vision \citep{IMF:Traquair1938}.
+This decline in differential light sensitivity gives rise to a three-dimensional structure known as the hill of vision, also referred to as Traquair’s island of vision \citep{IMF:Traquair1938}.
 Figure~\ref{fig:hill} below shows a classic rendering of the hill of vision by Aulhorn and Harms \citep{IMF:Aul1967}.
@@
 \centering
-\includegraphics[width=0.75\textwidth]{AulhornHillOfVision.png}
-\caption{\label{fig:hill} Hill of vision as rendered by Aulhorn \& Harms.}
+\includegraphics[width=0.9\textwidth]{Figure1.pdf}
+\caption{\label{fig:hill} Perimetry. A) A (manual) perimeter with an inset showing the stimulus enlarged. B) The hill of vision.}
 \end{figure}
@@
 
-In visual field testing, the observer either presses or does not press a button within a fixed time window following the presentation of a light increment in the visual field (a yes/no response).
+In visual field testing, the observer responds by either pressing or not pressing a button within a fixed time window following the presentation of a light increment in the visual field (a yes/no response).
 The relationship between stimulus strength and response probability is modeled by a \emph{psychometric function}, which characterizes the likelihood of a "yes" response as a function of stimulus strength \citep{CH:knoblauch_modeling_2012}.
@@
 However, other patterns and devices can be used.
-The package contains all necessary data for the HFA 10-2 pattern.
-The package is designed to be extensible, allowing users to add new patterns, devices, and normative datasets.
+The package also contains all necessary data for the HFA 10-2 pattern.
+However, the package is designed to be extensible, allowing users to add new patterns, devices, and normative datasets.
 
@@
 
-\section{Methods} \label{sec:methods}
+\section{Humphrey Field Analyser 24-2 pattern} \label{sec:methods}
+
+In this section, we will show how to install and load the package, and how the data supplied by the package can be analyzed with the default environment settings.
+In the following sections, we will show how to plot the data, perform longitudinal and cross-sectional analyses, how to import data from perimeters, and how to set up the environmnent for these data.
 
@@
 
-The package includes example data from a glaucoma patient (\code{vfpwgSunyiu24d2}), consisting of 42 visual field tests.
-
-The first ten columns contain metadata for each test, including reliability indices (false negatives, false positives, and fixation losses).
-The remaining columns contain the actual test values (sensitivities) at the locations defined in the \code{locmap}.
+The most important data are the perimetry results.
+However, the environment needs to be set up correctly for the corresponding type of perimetry.
+Here, we start with data from a Humphrey Field Analyser (HFA) with the 24-2 pattern that is supplied by the package (\code{data(vfpwgSunyiu24d2)}), and demonstrate how the data is structured and what environment settings are used.
+The default environment settings are aimed at the HFA 24-2, and they can always be reset with \code{setdefaults()}.
+The data is from a glaucoma patient (\code{vfpwgSunyiu24d2}) and consists of 42 consecutive visual field tests.
 
 <<>>=
+data(vfpwgSunyiu24d2)
 str(vfpwgSunyiu24d2[, 1:10], strict.width = "wrap")
@@
 
-The \code{date} variable is stored as a \code{POSIXct} object, which can be converted to a date format using the \code{as.Date()} function. Each row represents a monocular test.
-
+The first ten columns contain metadata for each test, including reliability indices.
+The remaining columns contain the actual test values (sensitivities) at the locations defined in the \code{locmap}.
+The \code{date} variable is stored as a \code{POSIXct} object, which can be converted to a date format using the \code{as.Date()} function.
+Each row represents a monocular test.
+The \code{type} variable allows assignment of tests to different groups.
 The eye is labeled using Latin abbreviations:
@@
 \begin{description}
-\item [OD] oculus dexter -> right eye
-\item [OS] oculus sinister -> left eye
-\item [OU] oculus uterque -> both eyes
+\item [OD: oculus dexter] right eye
+\item [OS: oculus sinister] left eye
+\item [OU: oculus uterque] both eyes
 \end{description}
-
-The \code{type} variable allows assignment of tests to different groups.
 
@@
 In diseases like glaucoma, short-term variability is often elevated in damaged regions of the visual field.
-Response quality is typically assessed by:
+Response quality is typically assessed by following metrics and these help to assess the reliability of a test.
+:
 
@@
   \item [False negatives (FNR)] Missed responses to stimuli that were previously seen.
-  \item [False positives (FPR)] Responses when no stimulus is presented
+  \item [False positives (FPR)] Responses when no stimulus is presented when the patients expects a presentation from the rhythm of the exam
   \item [Fixation losses] Responses to stimuli presented within the blind spot
@@
 
-These metrics help assess the reliability of a test.
+\subsubsection{Local sensitivities}
 
-\subsubsection{Psychophysical data: sensitivities}
-
-Initial columns store patient information and reliability metrics.
-Actual test values (sensitivities) begin at the column number stored in the \code{locini} variable.
-The function \code{getvfcols()} returns the numbers of the columns that contain the actual test values at the locations \code{l1} to \code{ln}, where 1..n corresponds to the rows of \code{coords} in the \code{locmaps}.
+Actual test values (sensitivities) begin at the column number stored in the \code{locini} variable of the environment.
+The function \code{getvfcols()} returns the numbers of the columns that contain the actual test values at the locations $l_{1}$ to $l_{n}$, where n corresponds to the number of locations tests (n = 54 for the 24-2 pattern).
 
@@
 
-The \code{vfread()} and \code{vfread()} functions can be used to read and write data in this format from a CSV file.
-However, these functions cannot be used to import native data from the perimeters.
-For this, the package includes functions to import data from the Octopus 900 and the Humphrey Field Analyzer.
+Here, the visual field data is not value, because an illegal column was added.
+The \code{vfread()} and \code{vfread()} functions can be used to read and write data in this format from a CSV file, but they cannot be used to import native data from the perimeters.
+
+The following functions can be used to import data from perimeters.
 
 \begin{itemize}
-  \item \code{loadoctopus()} for Octopus 900 (Haag-Streit, Switzerland)
-  \item \code{loadhfaxml()}, \code{loadhfadicom()}, \code{loadhfaxmlbatch()}, \code{loadhfadicombatch()} for the Humphrey Field Analyzer (Zeiss)
+  \item \code{loadhfaxml()}, \code{loadhfadicom()}, \code{loadhfaxmlbatch()}, \code{loadhfadicombatch()} for the Humphrey Field Analyzer (Carl Zeiss Meditec, Dublin, CA, USA)
+  \item \code{loadoctopus()} for Octopus 900 (Haag-Streit, Köniz, Switzerland)
 \end{itemize}
-
-Detailed examples for exporting data from the Octopus 900 and using \code{loadoctopus()} can be found here: \url{https://huchzi.github.io/vf1/loadOctopus.html}.
 
@@
 <<>>=
+data(locmpas)
 names(locmaps)
@@
 \end{itemize}
-
-Other patterns are available in the package, including the \code{10-2} grid.
-
-<<>>=
-ten_two_locmap <- locmaps$p10d2
-str(ten_two_locmap, strict.width = "wrap")
-@
 
@@
 <<>>=
+data(fvctrSunyiu24d2)
 str(
-  vfctrSunyiu24d2, 
+  vfctrSunyiu24d2[, 1:getlocini()+5], 
   strict.width = "wrap")
@@
 <<>>=
+data(normvals)
 str(
@@
 
-To build the normative model for a custom dataset, use \code{nvgenerate()}.
-
-<<>>=
-nv <- nvgenerate(vfctrSunyiu24d2)
-setnv(nv)
-@
-
 The normative object \code{nv} includes coefficients of age-linear models (\code{nv$agem$coeff}).:
@@
 <<>>=
+nv <- getnv()
 head(nv$agem$coeff, 5)
@@
 
-\section{Results} \label{sec:results}
+\section{Analyzing visual fields} \label{sec:results}
+
+After you have familiarized yourself with the structure of the default data, you will now learn how to use the visual field package to analyze single locations, how to calculate global parameters that summarize the visual field, how plot fields, how to perform longitudinal analyses, and how to do structure-function correlations.
+Then, in the following section, we will show how to use the package with different patterns and data from different perimeters.
 
@@
 str(gl, max.level = 1, strict.width = "wrap")
-gl[, -(1:getlocini() - 1)]
+gl[, getlocini():ncol(gl)]
 @
@@
 <<>>=
-getglp(gl)[, -(1:getlocini() - 1)]
+getglp(gl)[, getlocini():ncol(gl)]
 @
@@
 
-There are various ways to visualize perimetric data. These visualizations should be interpreted as simplified representations of the hill of vision.
-
+There are various ways to visualize perimetric data.
+These visualizations should be interpreted as simplified representations of the hill of vision.
 Standard displays include color-coded probability maps, graphical representations of the blind spot, and border enhancements that encode diagnostic information.
@@
 This information is included for the Humphrey Field Analyzer patterns but must be provided separately for other perimeters.
-
 By convention, when displaying both eyes side-by-side, the left eye appears on the left and the right eye on the right.
@@
 This is in contrast to structural imaging, which is typically displayed as viewed by the examiner.
+It is important to specify the type of data represented in the visual field tables.
+The default is set to \emph{total deviation} (\code{td}).
+In the following plot, yellow indicates a probability $<$ 0.05, orange $<$ 0.02, red $<$ 0.01, and dark red $<$ 0.005.
 
-It is important to specify the type of data represented in the visual field tables. The default is set to \emph{total deviation} (	exttt{td}).
-
-<<fig=TRUE>>=
+<<fig=TRUE, out.width='.25\\textwidth', out.height='!'>>=
 
@@
 @
-
-In this plot, yellow indicates a probability $<$ 0.05, orange $<$ 0.02, red $<$ 0.01, and dark red $<$ 0.005.
 
@@
 Monitoring visual function over time is a key application of perimetry in managing glaucoma.
-
 Sensitivity measurements naturally fluctuate between tests.
@@
 Instead, statistical analysis across a series of tests is required to detect significant change.
-
 Progression analysis methods fall into two broad categories: \emph{event-based} and \emph{trend-based} approaches.
@@
 
-\subsubsection{Plotting progression}
-
-Example plots used for monitoring progression are available in the gallery: \url{https://rpubs.com/huchzi/1307927}.
-
 \subsubsection{Statistical analysis}
@@
 
-The \code{glr()} function estimates a regression model for global indices. For the example series, the patient loses \code{getgl(example3)} dB/year, with $p < getgl(example3)$.
+The \code{glr()} function estimates a regression model for global indices. For the example series, the the regression can be calculated as follows:
 
-%\code{r round(abs(glr(getgl(example3))$sl), 2)}
-%\code{r format(ceiling(1000 * glr(getgl(example3))$pval) / 1000, nsmall = 3)}
+<<>>=
+global_indices <- getgl(example3)
+global_linear_regression <- glr(global_indices[,c("date", "tmd")])
+str(global_linear_regression)
+global_linear_regression$sl
+global_linear_regression$pval
+@
 
@@
 @
-
-To visualize these results:
-
-<<fig=TRUE>>=
-vfplotplr(example3)
-@
-
-In this display, yellow indicates $p < 0.05$, orange $< 0.02$, red $< 0.01$, and dark red $< 0.005$.
 
@@
 @
+
+\subsubsection{Plotting progression}
+
+Use \code{vfplotplr()} to visualize results from the pointwise linear regression (PLR). In this figure, yellow indicates $p < 0.05$, orange $< 0.02$, red $< 0.01$, and dark red $< 0.005$.
+
+<<fig=TRUE>>=
+vfplotplr(example3)
+@
+
+Example plots used for monitoring progression are available in the gallery: \url{https://rpubs.com/huchzi/1307927}.
 
@@
 
-A plot of the 10-2 grid and corresponding ganglion cell bodies:
+A plot of the 10-2 grid and the displacement of the corresponding ganglion cell bodies:
 
@@
 
-plot_grid <- function() {
-  lines(-c(10, 0.5), c(0, 0), lwd = 2, col = "lightgray")
-  lines(c(0.5, 10), c(0, 0), lwd = 2, col = "lightgray")
-  lines(c(0, 0), -c(10, 0.5), lwd = 2, col = "lightgray")
-  lines(c(0, 0), c(0.5, 10), lwd = 2, col = "lightgray")
-  lines(x = c(-0.5, 0.5), y = c(0, 0), lwd = 1.5)
-  lines(x = c(0, 0), y = c(-0.5, 0.5), lwd = 1.5)
-  for(r in c(2.5, 5, 7.5, 10))
-    draw.circle(0, 0, r, lwd = 2, col = NA, border = "lightgray")
-  text(x = 11, y = 0, "10\u00b0", adj = c(0, 0.5))
-  text(x = 0, y = 11, "10\u00b0", adj = c(0.5, 0))
-}
+rbind(data.frame(label = "10-2 grid", locmap$coord), 
+      data.frame(label = "Corresponding ganglion cells", gcmap)) |>
+        ggplot(aes(x = x, y = y)) +
+        geom_point() +
+        facet_wrap(~ label) +
+  theme_minimal(12) +
+  scale_x_continuous("", limits = c(-10, 10)) +
+  scale_y_continuous("", limits = c(-10, 10)) +
+  guides(x = "none", y = "none")
 
-par(mfrow = c(1, 2))
-plot(x = locmap$coord$x, 
-     y = locmap$coord$y, 
-     type = "n", 
-     asp = 1,
-     xlim = c(-10, 12),
-     ylim = c(-10, 10),
-     yaxt = 'n',
-     xaxt = 'n',
-     bty = 'n',
-     xlab = "",
-     ylab = "",
-     main = "10-2 grid of test locations")
-plot_grid()
-points(x = locmap$coord$x, y = locmap$coord$y, bg = "white")
-
-plot(x = locmap$coord$x, 
-     y = locmap$coord$y, 
-     type = "n", 
-     asp = 1,
-     xlim = c(-10, 12),
-     ylim = c(-10, 10),
-     yaxt = 'n',
-     xaxt = 'n',
-     bty = 'n',
-     xlab = "",
-     ylab = "",
-     main = "Corresponding ganglion cells")
-plot_grid()
-points(x = gcmap$x, y = gcmap$y, pch = 21, bg = "white")
 @
 
-Nerve fiber paths from each location to the optic disc can also be modeled, using Jansonius' framework \citep{IMF:Jan2009, IMF:Jan2012}.
+Nerve fiber paths from each location to the optic disc can also be modeled, using Jansonius' framework \citep{IMF:Jan2009, IMF:Jan2012} with the functions \code{cart2jpolar}, \code{jpolar2cart}, \code{bundlePath}, \code{loc2psi}, and \code{psi2oct}.
 
-<<>>=
-psi23 <- loc2psi(vf2gc(locmaps$p24d2$coord[23,]))
-ang23 <- psi2oct(psi23)
+\section{Support for other patterns and perimeters}
 
-ang23 - 360
+The best supported alternativ to the 24-2 pattern is the HFA 10-2 pattern.
+Remember that you must changes other settings along with the \code{locmap}, including the \code{normvals} and the \code{gpar} parameters.
+Otherwise, the \code{vfplot()} function will throw an error.
+It is very important that you set up the environment completely (\code{locmap}, \code{nv}, \code{gpar}).
+
+<<fig=TRUE>>=
+
+setlocmap(locmaps$p10d2)
+setnv(normvals$sunyiu_10d2)
+setgpar(gpars$p10d2)
+vfplot(vffilter(vfctrSunyiu10d2, id == 2, eye == "OD"))
+
 @
 
-\subsection{DICOM support}
-
-DICOM standards for static perimetry are defined in Supplement 146. The package includes preliminary functions for reading HFA-generated DICOM files, but exporting to DICOM is not yet supported.
-
-\subsection{Support for other perimeters}
-
-Perimetry conventions vary across manufacturers.
-The package includes normative data for Humphrey 24-2 (Goldmann III) and Iowa full-field perimetry with Goldmann V \citep{IMF:Marin2018c}.
-The \code{loadoctopus} function provides partial support for Octopus G1 fields, but normative support is incomplete.
-
-Users may provide custom normative datasets using \code{nvgenerate()} and \code{setnv()}.
-
-\subsection{Setting the environment}
-
-Each analysis requires environment variables:
-\begin{itemize}
-\item \code{nv}: normative values
-\item \code{locmap}: test grid
-\item \code{gpar}: graphical parameters
-\item \code{locini}: first sensitivity column
-\end{itemize}
-
-<<>>=
-names(locmaps)
-names(normvals)
-names(gpars)
-@
-
-These must be consistent. For example:
-
-<<>>=
-setlocmap(locmaps$pPC26)
-setnv(normvals$iowa_PC26_pw)
-setgpar(gpars$pPC26)
-setlocini(24)
-@
-
-To reset to default (Humphrey 24-2):
+Remember that you can always reset the environment with...
 
@@
 
-\subsection{Custom configurations}
+\subsubsection{Generate normal values}
 
-Custom test structures must:
-\begin{itemize}
-\item Start with columns: \code{id}, \code{eye}, \code{date}, \code{time}
-\item Include \code{age}, optionally \code{fpr}, \code{fnr}, \code{fl}
-\item Have sensitivity columns named \code{l1}--\code{lk}
-\end{itemize}
+The locmap will be available online or from the manufacturer.
+Perimetry conventions vary across manufacturers.
+The package includes normative data for Humphrey 24-2 (Goldmann III) and Iowa full-field perimetry with Goldmann V \citep{IMF:Marin2018c}.
+Use \code{nvgenerate()} to build the normative model for a custom dataset.
+You will need a sufficiently large dataset of normal fields balanced for age in order to create your own norm values.
 
-To import custom locations:
-
-<<eval=FALSE>>=
-g1 <- list(coord = read.csv("./g1.csv"), 
-           name = "G1", 
-           desc = "G1 map", 
-           bs = numeric())
-setlocmap(g1)
-setgpar(vfgpar(g1$coord))
-vfplot(vfread("./vfg1.csv"), type = "s")
+<<>>=
+nv <- nvgenerate(vfctrSunyiu24d2)
+setnv(nv)
 @
 
-If normative data exist, use \code{vfread()}, \code{nvgenerate()}, and \code{setnv()} to enable deviation computations.
+\subsubsection{Support for Octopus perimeters}
+
+Octopus data can be imported with \code{loadoctopus()}.
+However, no normal values are provided by the package, yet.
+Detailed examples for exporting data from the Octopus 900 and using \code{loadoctopus()} can be found here: \url{https://huchzi.github.io/vf1/loadOctopus.html}.
 
@@
 
-\newpage
-
-\begin{appendix}
-
-\section{Clinical aspects} \label{app:clinical}
-
-\subsection{How is perimetry performed?}
-
-In a typical SAP test, the patient is seated comfortably in front of a perimeter and places their head on a chin rest while looking into a white Ganzfeld dome.
-The room is dimly lit and quiet, and the patient is briefed by an experienced technician before the examination.
-During the test, the patient fixates on a central target (e.g., a dot, cross, or diamond), and presses a button whenever a light stimulus is perceived.
-
-Stimuli are presented briefly (typically 100–200 ms), minimizing the opportunity for eye movement.
-A response is only considered valid if it occurs within a defined time window following the stimulus.
-A computer presents stimuli at discrete locations in a grid, in randomized order.
-The stimulus intensity at each location is adjusted adaptively based on the patient' responses.
-Because of randomization, the patient cannot anticipate where the next stimulus will appear.
-
-The figure below shows the commonly used 24-2 grid of test locations.
-The gray ellipse represents the anatomical blind spot; the red points fall within this region and are expected to show no sensitivity.
-
-The grid, stimulus size, presentation duration, and intensity algorithm remain constant throughout the test and together define the perimetry program.
-It is important to note that results from different perimetry programs are not directly comparable.
-
-\subsection{Clinical applications of perimetry}
-
-In clinical settings, perimetry is used to detect and monitor visual field defects or scotomas, which can arise from damage anywhere along the visual pathway—from the ocular media (cornea, lens, vitreous), to the retina, optic nerve, or visual cortex.
-
-The spatial pattern of visual field loss often provides diagnostic clues.
-For instance, a right occipital lobe infarct can result in left homonymous hemianopia—complete loss of the left visual field in both eyes, with a sharp border along the vertical meridian.
-Conversely, glaucoma typically presents as arcuate defects, often beginning in the upper or lower hemifield, reflecting characteristic damage to the optic nerve head.
-
-Perimetry supports not only diagnosis but also evaluation of functional impact (e.g., fitness to drive) and disease progression monitoring.
-
-\section[Using BibTeX]{Using \textsc{Bib}{\TeX}} \label{app:bibtex}
-
-\begin{leftbar}
-References need to be provided in a \textsc{Bib}{\TeX} file (\code{.bib}). All
-references should be made with \verb|\cite|, \verb|\citet|, \verb|\citep|,
-\verb|\citealp| etc.\ (and never hard-coded). This commands yield different
-formats of author-year citations and allow to include additional details (e.g.,
-pages, chapters, \dots) in brackets. In case you are not familiar with these
-commands see the JSS style FAQ for details.
-
-Cleaning up \textsc{Bib}{\TeX} files is a somewhat tedious task -- especially
-when acquiring the entries automatically from mixed online sources. However,
-it is important that informations are complete and presented in a consistent
-style to avoid confusions. JSS requires the following format.
-\begin{itemize}
-  \item JSS-specific markup (\verb|\proglang|, \verb|\pkg|, \verb|\code|) should
-    be used in the references.
-  \item Titles should be in title case.
-  \item Journal titles should not be abbreviated and in title case.
-  \item DOIs should be included where available.
-  \item Software should be properly cited as well. For \proglang{R} packages
-    \code{citation("pkgname")} typically provides a good starting point.
-\end{itemize}
-\end{leftbar}
-
-\end{appendix}
-
 %% -----------------------------------------------------------------------------

### 2025-06-16 R/FigureHoV.R

@@ -1,0 +1,62 @@
+library(visualFields)
+library(fields)
+library(plotly)
+library(data.table)
+
+data("normvals")
+data("locmaps")
+
+load("myField.rda")
+
+octopus_normal <- vf_table$get_norm_values(1) |> data.table()
+sensitivities <- as.vector(unlist(octopus_normal[1, .SD, .SDcol = paste0("l", 1:59)]))
+locations <- vf_table$locmaps[[1]] |> data.table()
+locations <- locations[, .(x = xod, y = yod)]
+
+grid.list <- list(
+  x = seq(-100, 100, by = 1),  # ensure 0 is included
+  y = seq(-100, 100, by = 1)
+)
+
+spline_model <- fields::Tps(locations, sensitivities, m = 2, df = 59)
+tps <- fields::predictSurface(spline_model, grid.list = grid.list)
+
+ymin <- 22
+
+for (ix in seq(grid.list$x)) {
+  for (iy in seq(grid.list$x)) {
+    if (sqrt((tps$x[ix] - 15)^2 + (tps$y[iy] + 1.5)^2) < 3) {
+      tps$z[ix, iy] <- ymin
+    }
+  }
+}
+
+pdf("hill_of_vision.pdf")
+pmat <- persp(
+  x = tps$x,
+  y = tps$y,
+  z = tps$z,
+  xlab = "NASAL - TEMPORAL",
+  xlim = c(-30, 30),
+  ylab = "INFERIOR - SUPERIOR",
+  ylim = c(-30, 30),
+  zlab = "Light Sensitivity [dB]",
+  zlim = c(ymin, 33),
+  theta = 35,
+  phi = 25,
+  expand = .5,
+  col = "red"
+)
+
+circx <- cos(seq(0, 2 * pi, length.out = 100))
+circy <- sin(seq(0, 2 * pi, length.out = 100))
+
+points(trans3d(0, 0, ymin, pmat))
+lines(trans3d(3*circx + 15, 2*circy - 1.5, ymin, pmat))
+lines(trans3d(10*circx, 10*circy, ymin, pmat))
+lines(trans3d(20*circx, 20*circy, ymin, pmat))
+lines(trans3d(30*circx, 30*circy, ymin, pmat))
+lines(trans3d(c(0, 0), c(0, 0), c(ymin, max(sensitivities, na.rm = TRUE)), pmat))
[…]

### 2025-06-16 jss-article-rnw/article.Rnw
Modifications by Ivan.

@@
 library("utils")
-library("visualFields")
 @
@@
 \author{Cord Huchzermeyer~\orcidlink{0000-0002-0408-8981}\\Department of Ophthalmology\\University Hospital Erlangen
-   \And Iván Marín-Franch\\Computational Optometry}
+   \And Iván Marín-Franch~\orcidlink{0000-0003-0552-8456}\\Computational Optometry}
 \Plainauthor{Cord Huchzermeyer, Iván Marín-Franch}
@@
   Iván Marín-Franch\\
-  Computational Optometry\\
-  E-mail: \email{ivan@example.com}\\
-  URL: \url{https://www.computationaloptometry.com/}
+  Computational Optometry, Atarfe, Spain\\
+  E-mail: \email{imarinfr@optocom.es}\\
+  URL: \url{https://www.optocom.es/}.\\
+  Envision Health Technologies, Brooklyn NY, USA.\\
+  \emph{and}\\
+  Ophthalmology.World, Madrid, Spain.
 }
@@
 In healthy vision, most visual functions, particularly light sensitivity, decline with increasing eccentricity—i.e., distance from the center of fixation.
-This decline in differential light sensitivity gives rise to a three-dimensional structure known as the hill of vision, also referred to as Traquair’s island of vision \citep{IMF:Traquair1938}.
+This decline in differential light sensitivity gives rise to a three-dimensional [IMF]surface[IMF] known as the hill of vision, also referred to as Traquair’s island of vision \citep{IMF:Traquair1938}.
 Figure~\ref{fig:hill} below shows a classic rendering of the hill of vision by Aulhorn and Harms \citep{IMF:Aul1967}.
@@
 \emph{Perimetry} (visual field testing) uses psychophysical methods to quantify the visual field by measuring a patient's sensitivity to light differences at various locations within the field.
-Psychophysical tests allow for the scientific measurement of the relationship between physical stimulus strength and subjective perception.
+Psychophysical tests allow for the [IMF]clinical and[IMF] scientific measurement of the relationship between physical stimulus strength[IMF], intensity of light or more precisely stimulus contrast for a constant background[IMF] and subjective perception.
 This is achieved by systematically varying the stimulus strength across discrete trials based on the observer's responses.
-The observer is typically restricted to a limited set of responses, and the stimuli are varied according to a predefined algorithm.
+[IMF]In clinical settings[IMF], the observer is restricted to a limited set of [IMF]trials[IMF], and the stimuli are varied according to a predefined algorithm.
 
-In visual field testing, the observer responds by either pressing or not pressing a button within a fixed time window following the presentation of a light increment in the visual field (a yes/no response).
+In visual field testing, the observer responds by either pressing or not pressing a button within a fixed time window following the presentation of [IMF]an increase in intensity or contrast[IMF] in the visual field (a yes/no response).
 The relationship between stimulus strength and response probability is modeled by a \emph{psychometric function}, which characterizes the likelihood of a "yes" response as a function of stimulus strength \citep{CH:knoblauch_modeling_2012}.
@@
 
-The most straightforward method for estimating this threshold is the method of constant stimuli, in which stimuli are presented multiple times at several discrete, fixed intensities around the anticipated threshold.
-This approach yields an accurate estimate of the psychometric function, which is typically modeled using a cumulative distribution function (CDF), such as the Gaussian CDF.
-However, this method is inefficient and requires a large number of trials, since stimulus levels far from the threshold provide little information.
-Long test durations negatively affect attention, thereby reducing both the accuracy and precision of the threshold estimates.
+\subsection{How do we quantify sensitivity in perimetry?}
 
-To improve efficiency, adaptive algorithms have been developed to reduce the number of stimulus presentations with minimal compromise in accuracy.
-A commonly used example is the staircase algorithm, in which the stimulus intensity decreases after a "seen" response and increases after a "not seen" response.
-The step size is halved after each reversal (i.e., a change in response direction).
-More sophisticated algorithms balance precision and test duration, often incorporating information from adjacent locations in the visual field.
-A notable example is the \emph{Swedish Interactive Threshold Algorithm (SITA)} used in the Humphrey Field Analyzer.
+[IMF] I HAVE EDITED THIS AND THE NEXT PARAGRAPHS TO AVOID GIVING ASSESSMENTS ON THE DIFFERENT METHODS, AS THIS IS A COMPLEX ISSUE:
+* I WOULD NOT SAY PSYCHOMETRIC FUNCTIONS ARE INEFFICIENT HERE AS IT DEPENDS ON WHAT YOU ARE AFTER AND LEAVE THE EMPHASIS ON TIME AND HOW THIS IS CLINICALLY IMPRACTICAL
+* EFFICIENCY DEPENDS ON WHAT WE ARE AFTER, FOR THRESHOLD IS UNNECESSARY, BUT FOR FP AND FN AND EVEN SLOPE OF THE PSYCHOMETRIC FUNCTION, THEY ARE IMPORTANT
+* I HAVE ALSO ADDED A PARAGRAPH ON STAIRCASE AND FULL THRESHOLD
+The traditional method for estimating this threshold is the method of constant stimuli, in which stimuli are presented multiple times at several discrete, fixed intensities around the anticipated threshold.
+This approach yields an estimate of the psychometric function, which is typically modeled using a cumulative distribution function (CDF), most commonly the Gaussian or Weibull CDF.
+However, this method requires a large number of trials to provide accurate estimations of the parameters of the psychometric function, intensity threshold, slope at threshold, and false positive and negative rates.
+Long test duration negatively affect attention and make any test impractical for clinical use.
 
-At the conclusion of the test, the perimetric sensitivity is recorded in a logarithmic decibel (dB) scale based on the luminance of the stimulus at threshold ($L_{threshold}$).
-In perimetry, luminance is usually reported in apostilbs rather than $cd/m^2$, where $1 asb = \frac{1}{\pi} cd/m^2$.
-This scale is defined as the attenuation from the maximum possible stimulus contrast, $L_{max}$:
+Two classic alternative approaches that does not require estimates of the full psychometric function and instead aims at estimating the threshold stimulus intensity at which half the presentations are seen are the staircase and the full threshold (consisting on two staircase algorithms ran one after the other).
+Staircase algorithms start by presenting a stimulus at an initial guess.
+If the stimulus is seen, then intensity is decreased by steps of a predetermined value and otherwise it is increased.
+After the first reversal (transition from seen to not seen or viceversa), the amount at which the stimulus intensity changes is half the original step size.
+At the second reversal the algorithm stops.
+The estimated threshold is typically obtained by averaging the intensities of the last two presentations.
+Staircase and its modified, full threshold version, which was the standard when static automated perimetry was first introduced, see e.g., \citep{IMF:Heijl1985}, is still too long and impractical for clinical use.
+
+For this reason, psychophysical Bayesian algorithms \citep{IMF:Taylor1967,IMF:Kin1994} to reduce the number of stimulus presentations were adapted to balance precision and test duration.
+Some algorithms, such as the \emph{Swedish Interactive Threshold Algorithm} (SITA) used in the Humphrey Field Analyzer, also incorporate information from adjacent locations in the visual field.[IMF]
+
+[IMF]In perimetry, luminance is usually reported in apostilbs rather than ${\rm cd/m^2}$, where $\pi {\rm apostilbs} = 1 {\rm cd/m^2}$.
+The background is nominally, $10 {\rm cd/m^2}$ or 31.4 apostilbs, and the intensity thresholds, $\Delta L_{\rm thr}$ is the stimulus luminance at threshold minus $10 {\rm cd/m^2}$.
+The maximum intensity threshold $\Delta L_{\rm max}$ can differ among perimeters.
+For the Humphrey Field Analyzer (HFA) the maximum is 10{,}000\,asb, whereas recent Octopus 900 perimeters commonly use 4000\,asb \citep{CH:racette_visual_nodate}.
+Regardless of the units used and the maximum possible luminance, at the conclusion of the test, the intensity threshold at each test location is converted into sensitivity, the inverse of intensity threshold, and expressed on a logarithmic scale, the decibel (dB), thus,
 
 \begin{equation}
-dB = 10\ \rm log \left (\frac{L_{\rm max}}{L_{\rm threshold}} \right)\ .
+dB = 10\ \rm log \left (\frac{\Delta L_{\rm max}}{\Delta L_{\rm thr}} \right)\ .
 \end{equation}
+%
+Since the maximum luminance differential is $\Delta L_{\rm max}$ the fraction is $\ge 1$ and therefore dB $\ge 0$, with 0 being the maximum stimulus possible, and therefore the minimum sensitivity that can be measured by the perimeter.
+dB scales are normally limited to 40 dB which corresponds to 1 asb for the HFA and 0.4 asb for the Octopus 900.
+[IMF]
 
-$L_{max}$ may differ between parameter types, and the Humphrey Field Analyzer uses 10{,}000\,asb, whereas recent Octopus 900 perimeters commonly use 4000\,asb \citep{CH:racette_visual_nodate}.
+Static automated perimetry (SAP) aims to estimate the hill of vision as accurately as possible within the constraints of clinical practice.
+[IMF COMMENT: The S in SAP is static, not standard.
+It was HFA's smart idea to replace the S to mean "standard".
+I WOULD DELETE THIS: The most common form is standard automated perimetry (SAP)] It is performed inside a dome-shaped device (a Ganzfeld perimeter) to ensure a uniform distance between the eye and all stimulus locations.
+[IMF COMMENT: CAMPIMETRY AND PERIMETRY ARE THE SAME?
+DELETE THE FOLLOWING?
+In contrast, campimetry] [IMF]With the advent of new technology, new perimeters are being developed on flat screens on computers, tables, phones and, more notably, virtual-reality systems.
+The are becoming more[IMF] common in particular for use in telemedicine.
 
-The visualFields package was developed to analyze data from static perimetry, where thresholds are measured at fixed, discrete locations in the visual field.
-Kinetic perimetry, in which a stimulus of constant luminance is moved across the field to identify isosensitivity contours, is outside the scope of this package.
+\subsection{Aims of the visualFields package}
+[IMF COMMENT: I CREATED THIS AND THE PREVIOUS NEW SECTIONS FOR ORGANIZATION PURPOSES. I MOVED THE SENTENCE REGARDING KINETIC PERIMETRY TO THE END OF THIS SECTION]
 
-Static perimetry aims to estimate the hill of vision as accurately as possible within the constraints of clinical practice.
-The most common form is standard automated perimetry (SAP), which is performed inside a dome-shaped device (a Ganzfeld perimeter) to ensure a uniform distance between the eye and all stimulus locations.
-In contrast, campimetry uses a flat screen and may become more common with the increasing availability of computer screens and virtual reality devices for telemedicine.
-
-For analyzing visual field data, the package provides the following functionality:
+[IMF]The \pkg{visualFields} package is an aid for the standardization of visual field data analysys from any device running SAP. It[IMF] provides the following functionality:
 \begin{itemize}
-  \item A format for representing visual field data
+  \item Standardized formats for representing visual field data
   \item Functions for importing data from perimeters
-  \item Functions for reading and writing data in CSV format, and for handlings data
-  \item A format for representing the underlying spatial pattern ({data(locmaps)})
-  \item A format for representing age-corrected normal values ({data(normvals)}), including probabilities
-  \item Functions for calculating total deviation (TD), pattern deviation (PD), and global indices
+  \item Functions for reading and writing data in CSV format, and for handling data
+  \item A standardized format for representing the underlying spatial pattern ({\code(locmaps)})
+  \item A standardized format for representing age-corrected normal values ({\code(normvals)}), including probabilities
+  \item Functions for calculating total deviation (TD), pattern deviation (PD), and global indices such as mean deviation (MD) and pattern standard deviation (PSD)
   \item Functions for calculating probabilities of TD and PD
-  \item A format for representing parameters for plotting the visual fields ({data(gpars)})
-  \item Functions for plotting visual fields, including longitudinal analyses of visual field series
-  \item Functions for calculating structural parameters pertinent to visual field testing (Jansonius maps, ganglion cell displacement)
+  \item A standardized format for the visualization of plotting the visual fields sensitivity, TD, and PD maps ({data(gpars)})
+  \item Functions for running longitudinal analyses in series of visual fields
+  \item Functions to aid with structure-function correspondence (such as the Jansonius map \citep{IMF:Jan2012})
 \end{itemize}
@@
 By default, the package is set up to handle Humphrey Field Analyzer 24-2 visual fields.
-It is recommended to use such data for exploring the package and data is provided by the package.
+It is recommended to use such data for exploring the package and data is provided within the package for this purpose.
 However, other patterns and devices can be used.
@@
 
-%% -- Methods ------------------------------------------------------------------
-
-\section{Humphrey Field Analyser 24-2 pattern} \label{sec:methods}
-
-In this section, we will show how to install and load the package, and how the data supplied by the package can be analyzed with the default environment settings.
-In the following sections, we will show how to plot the data, perform longitudinal and cross-sectional analyses, how to import data from perimeters, and how to set up the environmnent for these data.
-
-\subsection{Installation}
-
-The \pkg{visualFields} package is available from CRAN under the Apache 2.0 license and can be installed with:
+The \pkg{visualFields} package was developed to analyze data from SAP.
+Manual kinetic perimetry, in which a stimulus of constant luminance is moved across the field to identify isosensitivity contours, is outside the scope of this package.
+It is available from CRAN under the Apache 2.0 license and can be installed with:
 
@@
 <<eval=FALSE>>=
+install.packages("devtools")
 library(devtools)
@@
 @
+
+To invoke \pkg{visualFields} in R, run
+<<>>=
+library("visualFields")
+@
+
+[IMF]In Section \ref{sec:methods}, we will show how the data supplied by the package can be analyzed with the default environment settings.
+In Section \ref{sec:results}, we will show how to plot the data, perform longitudinal and cross-sectional analyses, how to import data from the Humphrey perimeter and how to set up the environment for these data.
+Section \ref{sec:others} outlines support for non-conventional settings and other perimeters.
+Final remarks are shown in Section \ref{sec:summary}.[IMF]
+
+%% -- Methods ------------------------------------------------------------------
+
+\section{Humphrey Field Analyser 24-2 pattern} \label{sec:methods}
 
@@
 However, the environment needs to be set up correctly for the corresponding type of perimetry.
-Here, we start with data from a Humphrey Field Analyser (HFA) with the 24-2 pattern that is supplied by the package (\code{data(vfpwgSunyiu24d2)}), and demonstrate how the data is structured and what environment settings are used.
-The default environment settings are aimed at the HFA 24-2, and they can always be reset with \code{setdefaults()}.
-The data is from a glaucoma patient (\code{vfpwgSunyiu24d2}) and consists of 42 consecutive visual field tests.
+Here, we start with data from a Humphrey Field Analyzer (HFA) with the 24-2 pattern that is supplied by the package, and demonstrate how the data is structured and what environment settings are used.
+The default environment settings are aimed at the HFA 24-2, and they can always be reset with
 
 <<>>=
-data(vfpwgSunyiu24d2)
+setdefaults()
+@
+
+The data is from a glaucoma patient (\code{vfpwgSunyiu24d2}) and consists of 42 consecutive visual field tests. The first ten columns contain metadata for each test:
+<<>>=
 str(vfpwgSunyiu24d2[, 1:10], strict.width = "wrap")
@@
 
-The first ten columns contain metadata for each test, including reliability indices.
-The remaining columns contain the actual test values (sensitivities) at the locations defined in the \code{locmap}.
+These include reliability indices false-positive rates (\code{fpr}), false-negative rates (\code{fnr}), and fixation losses (\code{fl}).
 The \code{date} variable is stored as a \code{POSIXct} object, which can be converted to a date format using the \code{as.Date()} function.
-Each row represents a monocular test.
-The \code{type} variable allows assignment of tests to different groups.
-The eye is labeled using Latin abbreviations:
+Each row represents a perimetry test for a patient with a specific ID and left or right eye (\code{eye}), on a test \code{date} and at a test \code{time}.
+The \code{eye} is labeled using Latin abbreviations:
+\begin{description}
+\item [OD: oculus dexter] right eye,
+\item [OS: oculus sinister] left eye, and
+\item [OU: oculus uterque] both eyes.
+\end{description}
+The column \code{type} allows assignment of tests to different groups.
+The column \code{duration} contains the total amount of time that the patient took to complete the perimetry test.
+The column \code{age} is important to run statistical analyses on the visual field results, in particular, for the comparisons with age-corrected mean reference values.
 
-\begin{description}
-\item [OD: oculus dexter] right eye
-\item [OS: oculus sinister] left eye
-\item [OU: oculus uterque] both eyes
-\end{description}
+The remaining columns contain the actual test values (sensitivities) at the locations defined in an internal variable defining the location map:
+<<>>=
+head(getlocmap()$coord)
+@
 
-\subsubsection{Reliability indices}
+Actual test values (pointwise sensitivities in dB) begin at the column number stored in an internal variable, \code{locini} variable of the environment that can be retrieved with the function:
+<<>>=
+getlocini()
+@
 
-Deviations from the expected psychometric function may arise due to observer inattention or short-term threshold fluctuations.
-In diseases like glaucoma, short-term variability is often elevated in damaged regions of the visual field.
-Response quality is typically assessed by following metrics and these help to assess the reliability of a test.
-:
-
-\begin{description}
-  \item [False negatives (FNR)] Missed responses to stimuli that were previously seen.
-  \item [False positives (FPR)] Responses when no stimulus is presented when the patients expects a presentation from the rhythm of the exam
-  \item [Fixation losses] Responses to stimuli presented within the blind spot
-\end{description}
-
-\subsubsection{Local sensitivities}
-
-Actual test values (sensitivities) begin at the column number stored in the \code{locini} variable of the environment.
-The function \code{getvfcols()} returns the numbers of the columns that contain the actual test values at the locations $l_{1}$ to $l_{n}$, where n corresponds to the number of locations tests (n = 54 for the 24-2 pattern).
+The function \code{getvfcols()} returns the numbers of the columns that contain the actual test values at the locations $l_{1}$ to $l_{n}$, where $n$ corresponds to the number of locations tests ($n$ = 54 for the 24-2 pattern).
+No other columns should appear after these test values.
 
 <<>>=
-getlocini()
 vfpwgSunyiu24d2[1, getvfcols()]
@@
 
-No columns should appear after these test values. Use \code{vffilter()} to select specific tests and \code{vfisvalid()} to verify data structure:
+Below is a list of thet datasets included in \pck{visualFields}:
+<<eval=FALSE>>=
+vfctrSunyiu24d2 # Control subjects for HFA 24-2 SITA standard
+vfctrSunyiu10d2 # Control subjects for HFA 10-2 SITA standard
+vfctrIowaPC26 # Control subjects for an irregular grid in
+              # the central 30 degrees using Octopus OPI ZEST
+vfctrIowaPeri # Control subjects for an irregular grid in
+              # the periphery from 30 to 80 degrees using Octopus OPI ZEST
+vfpwgSunyiu24d2 # Series of HFA 10-2 SITA standard tests for the two eyes
+                # of a patient followed up over time
+vfpwgRetest24d2 # 30 patients tested each 12 times with the HFA 10-2 SITA standard
+@
 
+\subsubsection{Reliability indices}
+[IMF COMMENT: I think this was too bold, so I edited it to tone it down]
+[IMF]Given the psychophysical nature of perimetry as a clinical test, attention, fatigue, and other psychological factors can affect the estimated sensitivities.
+For this reason, they include some catch and fixation trials to assess the quality of the test undertaken.
+Response quality, and the validity of the data collected by the test, is typically assessed by the following reliability indices:[IMF]
+
+\begin{description}
+  \item [False positives (\code{fpr})] Responses when no stimulus is presented when the patients expects a presentation from the rhythm of the exam
+  \item [False negatives (\code{fnr})] Missed responses to stimuli that were previously seen.
+  \item [Fixation losses (\code{fl})] Responses to stimuli presented within the blind spot
+\end{description}
+
+\subsubsection{Basic operations with visual-field data}
+
+\code{vffilter()} filter specific tests by a logical criterion
 <<>>=
-example1 <- vffilter(vfpwgSunyiu24d2, id == "sample1" & eye == "OD")[1, ]
+example1 <- vffilter(vfpwgSunyiu24d2, id == "sample1" & eye == "OD")
+head(example1[,1:4])
+@
+
+\code{vfselect()} selects a number of test from the end or the start of series of visual field data for each ID and eye
+<<>>=
+example1 <- vfselect(vfpwgSunyiu24d2, n = 3)
+head(example1[,1:4])
+@
+
+\code{vfjoin()} joins two data frames with vf data
+<<eval=FALSE>>=
+vf <- vfjoin(vfctrSunyiu24d2, vfpwgSunyiu24d2)
+@
+
+\code{vfaggregate()} aggregates the data by either ID + eye, or ID + eye + date:
+<<>>=
+vfaggregate(example1, by = "eye", median)
+@
+
+\code{vfmean()} is a shortcut to \code{vfaggregate()} with \code{fun = mean}
+
+\code{vfsort()} sorts the data by ID, eye, date, and time.
+
+\code{vfisvalid()} verifies the data structure:
+<<>>=
 example1$newVariable <- "invalid"
@@
 @
+Here, the visual field data is not value, because an illegal column was added.
 
-Here, the visual field data is not value, because an illegal column was added.
-The \code{vfread()} and \code{vfread()} functions can be used to read and write data in this format from a CSV file, but they cannot be used to import native data from the perimeters.
+The \code{vfread()} and \code{vfwrite()} functions can be used to read and write data in this format from a CSV file, but they cannot be used to import native data from the perimeters.
 
@@
 
-\subsection{Grid patterns}
-
+\subsection{Location maps}[IMF COMMENT: they should have been called grid or locgrid, but I was too young back then, so now we are stuck with location maps instead of grids]
 Test locations are defined by location maps (\code{locmaps}), which specify the spatial coordinates of stimulus presentation points in degrees of visual angle.
 These maps reflect the layout used by visual field testing devices.
-When analyzing data, the correct \code{locmap} must be specified.
+When analyzing data, the correct location map must be specified.
 
 <<>>=
-data(locmpas)
-names(locmaps)
+data.frame(name = sapply(locmaps, function(lm) return(lm$name)))
 @
-
 By default, the \code{24-2} pattern is used, consisting of 54 regularly spaced locations across the central visual field.
@@
 <<>>=
-default_locmap <- getlocmap()
-str(default_locmap, strict.width = "wrap")
+str(getlocmap(), strict.width = "wrap")
 @
 
-Each \code{locmap} contains:
+This structure contains:
 \begin{itemize}
@@
 If the blind spot is outside the tested region (e.g., in the 10-2 pattern) or is skipped (e.g., in the Octopus G1), \code{bs} is set to \code{numeric(0)}.
+All these maps are on right-eye format, to obtain the coordinates for a left eye, just change the sign to the x coordinates.
+
+The \code{locread} and \code{locwrite} saves or loads \code{locmaps}. They are useful for using custom locations.
 
 \subsection{Normative data}
-
-Sensitivity is measured in decibels (dB), derived from contrast thresholds. Sensitivity varies:
+Pointwise sensitivities vary with:
 \begin{itemize}
@@
 
-To detect abnormal points, we compare observed sensitivities with an age-corrected mean normal field. These comparisons require a \emph{normative dataset} and location-wise \emph{age-linear models}.
+To detect abnormal points, we compare observed sensitivities with an age-corrected mean normal field and reference values.
+These comparisons require a \emph{normative dataset} with point-wise \emph{age-linear models and percentiles} to compute the likelihood of a location being abnormally low.
 
@@
 To reveal focal defects masked by generalized depression (e.g., due to cataract), a \emph{pattern deviation (PD)} is computed by subtracting a general height (GH) estimate from the TD.
-For 24-2, GH is defined as the 7th highest TD value.
+For 24-2, GH is defined as the $7^{\rm th}$ greatest TD value.
 
@@
 \begin{itemize}
-  \item SUNY-IU dataset (\code{vfctrsunyiu_24d2}) for 24-2 with Goldmann size III \citep{IMF:Mar2013b}
-  \item Iowa PC26 (\code{vfctriowa_PC26}) and Peri (\code{vfctrsunyiu_10d2}) datasets for Goldmann size V \citep{IMF:Marin2018c}
+  \item SUNY-IU dataset (\code{vfctrSunyiu24d2}) for 24-2 with Goldmann size III \citep{IMF:Mar2013b}
+  \item Iowa PC26 (\code{vfctrIowaPC26}) and Peri (\code{vfctrIowaPeri}) datasets for Goldmann size V \citep{IMF:Marin2018c}
 \end{itemize}
 
+The normal values derived from these data can be found in \code{normvals}.
 <<>>=
-data(fvctrSunyiu24d2)
-str(
-  vfctrSunyiu24d2[, 1:getlocini()+5], 
-  strict.width = "wrap")
+str(normvals$sunyiu_24d2, max.level = 1, strict.width = "wrap")
 @
 
-The normal values derived from these data can be found in \code{normvals}.
-
-<<>>=
-data(normvals)
-str(
-  normvals$sunyiu_24d2, 
-  max.level = 1, strict.width = "wrap")
-@
-
-The normative object \code{nv} includes coefficients of age-linear models (\code{nv$agem$coeff}).:
-
+The object with the normative values includes coefficients of age-linear models.
 <<>>=
@@
 
-It also contains a function that calculates normal sensitivity for a given age (\code{nv$agem$model()}), Thus, the expected field at age 50 is:
-
+It also contains a function that calculates normal sensitivity for a given age.
+Thus, the expected field at age 50 as recorded for the 24-2 SITA Standard by the HFA and based on the normative dataset \code{vfctrSunyiu24d2} can be computed as:
 <<>>=
-age <- 50
+age <- 50 # Change age to see how mean-normal sensitivities change
 nv$agem$model(age)
@@
 
-Blind spot locations are returned as \code{NA}.
+Blind spot locations are returned as \code{NA}, as they should be removed from analysis.
+
+The normative values also contains lookup tables for calculation of percentiles, and functions to compute TD, PD, and global indices, and their associated probability levels.
+
+The objects \code{tdfun}, \code{pdfun}, and \code{glfun} are functions to compute the TD, PD, and global indices.
+And \code{tdpfun}, \code{pdpfun}, and \code{glpfun} obtain the corresponding probability levels.
+The function \code{ghfun} computes the GH from TD values.
+These functions are not to be called directly.
+Instead, use the following functions:
+\begin{itemize}
+  \item \code{gettd}, \code{gettdp} for TD values and probability levels,
+  \item \code{getpd}, \code{getpdp} for PD values and probability levels,
+  \item \code{getgl}, \code{getglp} for global indices and probability levels, and
+  \item \code{getgh} for the calculation of GH
+\end{itemize}
+
+The use of these functions is illustrated in Section \ref{sec:results}.
 
 \subsection{Graphical parameters}
+The \code{gpars} object contains graphical parameters for plotting visual fields.
+The default parameters are set for the 24-2 pattern, but can be customized for other patterns.
+It contains a list of the coords, which must have as many entries as the location map used for the test, but may have different x and y positions for visualization purposes.
+The graphical parameter object also contains information on how to tessellate the borders in the plots, and color palettes for both the fields and their borders in the visual field plots.
 
+You can select the current graphical parameters with
 <<>>=
-data(gpars)
-names(gpars)
-str(gpars$p24d2, 
-    max.level = 1, strict.width = "wrap")
+gpar <- getgpar()
+str(gpar, max.level = 1, strict.width = "wrap")
 @
 
-The \code{gpars} object contains graphical parameters for plotting visual fields.
-The default parameters are set for the 24-2 pattern, but can be customized for other patterns.
-It contains a list of the coords, which must be identical to the \code{locmap} used for the test.
-The \code{gpar} object also contains information on how to tesselate the borders in the plots, and color palettes for both the fields and their borders in the visual field plots.
+There are functions to help generate graphical parameters from the coordinates of a location map:
+\begin{itemize}
+  \item \code{vfgpar} generates default graphical parameters,
+  \item \code{vftess} generates a default structure to handle the visual field tessellation,
+  \item \code{vfcolscheme} helps generate color schemes for detection analysis, and
+  \item \code{vfprogcolscheme} helps generate color schemes for progression analysis
+\end{itemize}
 
 %% -- Results ------------------------------------------------------------------
+\section{Analyzing visual fields} \label{sec:results}
+This section describes how to use the visual field package to analyze pointwise locations, calculate global parameters that summarize the visual field, perform longitudinal analyses, and plot the pointwise results.
 
-\section{Analyzing visual fields} \label{sec:results}
-
-After you have familiarized yourself with the structure of the default data, you will now learn how to use the visual field package to analyze single locations, how to calculate global parameters that summarize the visual field, how plot fields, how to perform longitudinal analyses, and how to do structure-function correlations.
-Then, in the following section, we will show how to use the package with different patterns and data from different perimeters.
-
-\subsection{Metrics for individual locations}
-
-\subsubsection{Total deviation}
-
+\subsection{Pointwise results}
+\subsubsection{Total and pattern deviation}
 Total deviation is the difference between observed and expected sensitivity. Here, it is shown how it is calculated.
@@
 <<>>=
-example2 <- vffilter(vfpwgRetest24d2, id == "1", eye == "OD")[1, ]
+example2 <- vffilter(vfpwgRetest24d2, id == "1", eye == "OD")[1,]
 age <- example2$age[1]
 normal_values <- getnv()$agem$model(age)[1:10]
-example2[, getvfcols()[1:10]] - normal_values
+example2[,getvfcols()[1:10]] - normal_values
 @
 
-To calculate the total deviation, use the function from the \code{nv} object:
+To calculate the total deviation, use the function from the normative values, do:
 
@@
 
-\subsubsection{Pattern deviation}
-
 Pattern deviation is obtained by subtracting the general height:
-
 <<>>=
@@
 Or automatically:
-
 <<>>=
@@
 \subsubsection{Probability values}
-
-Even healthy eyes may show small deviations.
+Even healthy eyes may show unusually small deviations.
 To interpret these values, we calculate the probability that a deviation reflects pathology using empirical distributions from the normative data \citep{IMF:Hei1987a, IMF:Heijl1991}.
-
+These are based on the look up tables for probabilities:
 <<>>=
 getnv()$lut$probs
+@
+
+And can be calculated with:
+<<>>=
 gettdp(td)[, getvfcols()[1:10]]
@@
 
-\subsection{Plotting the visual field}
+\subsection{Visualizing the visual field}
+There are various ways to visualize perimetric data.
+These visualizations should be interpreted as simplified representations of the hill of vision (see Figure \ref{fig:hill}).
+Standard displays include color-coded probability maps, graphical representations of the blind spot, and border enhancements that encode diagnostic information.
+For example, \code{vfplot} with the argument \code{"td"} generates plots that require normative data on sensitivities and the probabilities of deviations from these norms.
+Normative values for the Humphrey Field Analyzer 24-2 SITA standard are included in the package \citep{IMF:Mar2013b} but must be generated for other perimeters and psychometric strategies.
 
-There are various ways to visualize perimetric data.
-These visualizations should be interpreted as simplified representations of the hill of vision.
-Standard displays include color-coded probability maps, graphical representations of the blind spot, and border enhancements that encode diagnostic information.
-For example, \code{vfplot} with the argument \code{"td"} generates plots that require normative data on differential light sensitivity and the probabilities of deviations from these norms.
-This information is included for the Humphrey Field Analyzer patterns but must be provided separately for other perimeters.
 By convention, when displaying both eyes side-by-side, the left eye appears on the left and the right eye on the right.
@@
 <<fig=TRUE, out.width='.25\\textwidth', out.height='!'>>=
-
 vfplot(example2)
-
 @
-
 A gallery of all available plot types is available at \url{https://rpubs.com/huchzi/1307927}.
@@
 \subsection{Longitudinal analyses}
-
 Monitoring visual function over time is a key application of perimetry in managing glaucoma.
@@
 This variability increases in eyes with glaucoma, especially at affected locations.
-Hence, two tests showing differences cannot alone indicate progression.
-Instead, statistical analysis across a series of tests is required to detect significant change.
+Hence, two tests showing differences cannot alone indicate progression with any reliability.
+Instead, statistical analysis across a series of tests is required to detect statistically significant change.
 Progression analysis methods fall into two broad categories: \emph{event-based} and \emph{trend-based} approaches.
@@
 
-Trend-based methods assume slower change and analyze all tests in a series. Though less intuitive for clinicians, these methods are statistically robust and reflect real-world progression patterns.
-
-To illustrate, we select a time series of 27 visual fields from one eye:
+Trend-based methods analyze all tests in a series.
+Though less intuitive for clinicians, these methods are statistically robust and reflect real-world progression patterns.
+To illustrate, we select a time series of 27 visual fields from one clearly progressing eye:
 
@@
 @
-
-\subsubsection{Statistical analysis}
 
@@
 
-The null distribution and observed statistic:
-
-<<>>=
-summary(prog2)
-prog2[c("csr", "csrp", "csl", "cslp")]
-@
-
-Visualization:
-
+The following histogram shows the null distribution (dark gray columns) and observed S statistic (red vertical line):
 <<fig=TRUE>>=
@@
          geom_histogram() +
-         geom_vline(xintercept = prog2$csl, color = "red", size = 1) +
+         geom_vline(xintercept = prog2$csl, color = "red", linewidth = 1) +
          scale_x_continuous("S-values", breaks = seq(0, 200, 20))
@@
 
-\subsubsection{Plotting progression}
+The value of the observed S statistic and corresponding $p$-value were:
+<<>>=
+c(prog2$csl, prog2$cslp)
+@
 
+\subsubsection{Visualizing pointwise progression}
 Use \code{vfplotplr()} to visualize results from the pointwise linear regression (PLR). In this figure, yellow indicates $p < 0.05$, orange $< 0.02$, red $< 0.01$, and dark red $< 0.005$.
-
 <<fig=TRUE>>=
@@
 \subsection{Automated reports}
-
 Static reports and Shiny applications can be generated:
@@
 Automated reports can also be generated for progression series:
-
 <<eval = FALSE>>=
@@
 
-\subsection{Structure-function correlation}
+These reports are samples to be cloned, customized and improved upon.
 
-Anatomical modeling aids comparison between structural imaging and functional testing in glaucoma.
+\section{Customization of the visualFields package} \label{sec:others}
+One of the key characteristics of the \pkg{visualFields} package is its flexibility to be customized to other forms of static automated perimetry.
+For the package to be used with different perimetry data, however, we are required to collect normative data and generate the reference values from it.
+In order to customize \pkg{visualFields} for use with other type of perimetry data, we need to define the data.frame as described before and run \code{setlocini()}.
+The location data, including x and y coordinates, in the visual field data need to be set with \code{setlocmap()}.
+We need also to set the graphical parameters with \code{setgpar()}, and last but not least, generate normative values for analysis and set them with \code{setnv()}.
+Here is an example for two custom tests that are run independently but analyzed jointly.
+One test covers the central 30 degrees of visual angle (in red in the following Figure) and the other covers the rest from $-50$ nasally and inferiorly to 80 temporally and 40 superiorly.
+Both test use the Goldmann size V stimulus (1.72 degrees of diameter) as opposed to the conventional size III (0.43 degrees)
 
-The ganglion cell bodies stimulated by a visual field location are displaced from the photoreceptor input region, especially in the central field.
-This displacement is modeled in \pkg{visualFields} based on Montesano et al.
-\citep{IMF:Montesano2020}.
+<<echo=TRUE, fig=TRUE>>=
+locmap <- rbind(
+  data.frame(label = "Peripheral", locmaps$pPeri$coord, color = "black"),
+  data.frame(label = "Central", locmaps$pPC26$coord, color = "red")
+)
+
+ggplot(locmap, aes(x = x, y = y)) +
+  geom_point(aes(color = color)) +
+  scale_color_identity() +
+  theme_minimal(12) +
+  scale_x_continuous("", limits = c(-50, 80)) +
+  scale_y_continuous("", limits = c(-50, 30)) +
+  guides(x = "none", y = "none") +
+  coord_fixed(ratio = 1)
+@
+
+We can generate normative values for both grids with normative data \citep{IMF:Marin2018c}:
+<<echo=TRUE>>=
+# Settings for the central perimetry test PC26
+setlocini(18)
+setlocmap(locmaps$pPC26)
+nvc <- nvgenerate(vfctrIowaPC26, method = "pointwise")
+gparc <- vfgpar(locmaps$pPC26$coord)
+# Settings for the peripheral perimetry test Peri
+setlocini(17)
+setlocmap(locmaps$pPeri)
+nvp <- nvgenerate(vfctrIowaPeri, method = "pointwise")
+gparp <- vfgpar(locmaps$pPeri$coord)
+# set the visualization limits to be the same
+gparc$tess$xlim <- gparp$tess$xlim
+gparc$tess$ylim <- gparp$tess$ylim
+@
+
+From here, the visualization for the joint analysis with the two different perimeters can be done as follows:
+<<echo=TRUE, fig=TRUE>>=
+idx <- 13
+setlocini(17)
+setlocmap(locmaps$pPeri)
+setnv(nvp)
+setgpar(gparp)
+vfplot(vfctrIowaPeri[idx,])
+setlocini(18)
+setlocmap(locmaps$pPC26)
+setnv(nvc)
+setgpar(gparc)
+vfplot(vfctrIowaPC26[idx,], new = TRUE)
+@
+
+Remember that you can always reset the environment with...
 
 <<>>=
+setdefaults()
+@
+
+There are functions that can be used or cloned and customized to modulate the behavior of the normative values. These are:
+\begin{itemize}
+  \item \code{agelm} returns a list with coefficients and a function defining a linear age model,
+  \item \code{tddef} returns a function for the computation of TD values,
+  \item \code{ghdef} returns a function for the computation of the general height,
+  \item \code{pddef} returns a function for the computation of PD values,
+  \item \code{lutdef} returns a look up table and a function for the computation of the probability values for TD and PD,
+  \item \code{gdef} returns a function to compute global indices, and
+  \item \code{lutgdef} returns a look up table and a function for the computation of the probability values for global indices
+\end{itemize}
+
+\subsubsection{Support for Octopus perimeters}
+Octopus data can be imported with \code{loadoctopus()}.
+However, no normative values are provided by the package.
+Detailed examples for exporting data from the Octopus 900 and using \code{loadoctopus()} can be found here: \url{https://huchzi.github.io/vf1/loadOctopus.html}.
+
+\subsection{The Jansonius map}
+The \pck{visualFields} package includes functions for the implementation of the Jansonius map \citep{IMF:Jan2012}, describing average never fiber bundle paths along the visual field from the optic nerve head (ONH).
+This model aids in comparisons between structural imaging and perimetry data in glaucoma.
+Functions are implemented to convert from Cartesian to the modified polar coordinates in the Jansonius map, the average bundle paths, and the angle of incidence of the average bundle passing through a specific visual field location on the ONH.
+One optical coherence tomgraphy (OCT) image modality consists of scanning along a circle centered at the center of the blind spot.
+A function returns the correspondence between angle of incidence of the average bundle on the ONH and the circular scan.
+These functions are:
+
+\begin{itemize}
+  \item \code{cart2jpolar} converts the Cartesian coordinates to the polar coordinates in the distorted space used in the Jansonius map,
+  \item \code{jpolar2cart} converts back from the Jansonius polar coordinates to Cartesian coordinates,
+  \item \code{bundlePath} returns a function describing the expected fiber path given an angle of incidence on the ONH,
+  \item \code{loc2psi} returns the angle of incidence of the average bundle path that passes through a specific location of the visual field,
+  \item \code{psi2oct} returns the angle of OCT circular scans corresponding to average bundle paths with specific angle of incidence at the ONH, and
+\end{itemize}
+
+For some analyses is important to understand the underlying structure of retinal elements that are stimulated during perimetry stimulus presentation.
+This is particularly important when the distance from the center of vision is less than 10 degrees of visual angle.
+The photoreceptors that are stimulated are connected to ganglion cells (whose axons travel in bundles described by the Jansonius map and form the optic never).
+For retinal location on this area, the photoreceptors are not on the same geographical location as the ganglion cell bodies \citep{IMF:Drasdo2007}.
+The displacement between photoreceptors and ganglion cell bodies follow the implementation described by Montesano et al.
+\citep{IMF:Montesano2020}.
+For the most common perimetry test targeting the central 10 degrees of the visual field, the 10-2 grid, the grid of stimulus locations and the corresponding ganglion cell bodies are shown in the next figure:
+<<echo=TRUE, fig=TRUE>>=
 locmap <- locmaps$p10d2
 gcmap <- vf2gc(locmap$coord)
-@
-
-A plot of the 10-2 grid and the displacement of the corresponding ganglion cell bodies:
-
-<<echo=TRUE, fig=TRUE>>=
 
@@
   scale_y_continuous("", limits = c(-10, 10)) +
-  guides(x = "none", y = "none")
-
+  guides(x = "none", y = "none") +
+  coord_fixed(ratio = 1)
 @
 
-Nerve fiber paths from each location to the optic disc can also be modeled, using Jansonius' framework \citep{IMF:Jan2009, IMF:Jan2012} with the functions \code{cart2jpolar}, \code{jpolar2cart}, \code{bundlePath}, \code{loc2psi}, and \code{psi2oct}.
-
-\section{Support for other patterns and perimeters}
-
-The best supported alternativ to the 24-2 pattern is the HFA 10-2 pattern.
-Remember that you must changes other settings along with the \code{locmap}, including the \code{normvals} and the \code{gpar} parameters.
-Otherwise, the \code{vfplot()} function will throw an error.
-It is very important that you set up the environment completely (\code{locmap}, \code{nv}, \code{gpar}).
-
-<<fig=TRUE>>=
-
-setlocmap(locmaps$p10d2)
-setnv(normvals$sunyiu_10d2)
-setgpar(gpars$p10d2)
-vfplot(vffilter(vfctrSunyiu10d2, id == 2, eye == "OD"))
-
-@
-
-Remember that you can always reset the environment with...
-
-<<>>=
-setdefaults()
-@
-
-\subsubsection{Generate normal values}
-
-The locmap will be available online or from the manufacturer.
-Perimetry conventions vary across manufacturers.
-The package includes normative data for Humphrey 24-2 (Goldmann III) and Iowa full-field perimetry with Goldmann V \citep{IMF:Marin2018c}.
-Use \code{nvgenerate()} to build the normative model for a custom dataset.
-You will need a sufficiently large dataset of normal fields balanced for age in order to create your own norm values.
-
-<<>>=
-nv <- nvgenerate(vfctrSunyiu24d2)
-setnv(nv)
-@
-
-\subsubsection{Support for Octopus perimeters}
-
-Octopus data can be imported with \code{loadoctopus()}.
-However, no normal values are provided by the package, yet.
-Detailed examples for exporting data from the Octopus 900 and using \code{loadoctopus()} can be found here: \url{https://huchzi.github.io/vf1/loadOctopus.html}.
-
 %% -- Summary and discussion ---------------------------------------------------
-
 \section{Summary and discussion} \label{sec:summary}
@@
 \section{Acknowledgments}
-The visualField's package would not have existed if it wasn't for the important contribution of many contributors who partially funded the development of the package, and/or contributed datasets, code, and above all insight on visual fields analysis and about perimetry and visual sciences at large.
-Huge thanks to:
-  - William H Swanson
-  - Michael Wall
-  - Andrew Turpin
-  - Paul H Artes
-  - Giovanni Montesano
-  - Mitchell W Dul
+The \pkg{visualFields} package would not have existed if it wasn't for the important contribution of many contributors who partially funded the development of the package, and/or contributed datasets, code, and above all insight on visual fields analysis and about perimetry and visual sciences at large.
+In particular, we are thankful to Paul H.
+Artes and Andrew Turpin who are part of the Open Perimetry Initiative (OPI).
+William H.
+Swanson and Michael Wall provided funding for its development.
+Giovanni Montesano contributed with his ganglion cell displacement code.
+Paul H.
+Artes and one of the coauthors (IMF) designed the first version of the package.
 
 %% -- Bibliography -------------------------------------------------------------
-
 \bibliography{references}
